<!-- DEPRECATED: This RMD pipeline is kept for documentation purposes only.
     The actively maintained pipeline is MARMOT_Pipeline.qmd (Quarto). -->
---
title: "{{PIPELINE_NAME}}"
author: ''
date: "`r format(Sys.time(), '%Y-%m-%d_%H.%M.%S')`"
output:
  html_document:
    self_contained: true
    highlight: zenburn
    theme: cosmo
    code_folding: hide
    toc: true
    toc_depth: 6
    dev: ragg_png
---

```{=html}
<style>
    body .main-container {
        max-width: 1600px;
    }
</style>
```

```{r knitr_settings}

# Global chunk defaults
knitr::opts_chunk$set(
  comment = "",
  fig.width = 8,
  fig.height = 8,
  warning = FALSE,
  error = FALSE,
  message = FALSE,
  cache = FALSE,
  cache.lazy = FALSE,
  echo = FALSE
)

# Timestamp tag for output folder
timeRun <- format(Sys.time(), "%Y-%m-%d_%H.%M.%S")

# Render options
options(
  # bitmapType = "cairo",
  kableExtra_view_html = FALSE
)
```

```{r input_settings}

## 🔧 File paths
fp <- NULL      # Folder with all FCS files
md_fp <- NULL      # Full path to metadata Excel file
condaDir <- NULL      # Where conda/mamba lives (for PARC/PaCMAP)
parcScriptDir <- NULL # Folder with f_pacmap.py and f_parc.py

## 🧪 Preprocessing
downsampleTo <- NULL  # e.g. 1000 to downsample each FCS file, NULL = no downsampling

## 🔍 Clustering
clusteringMethodToUse <- NULL   # One of: "Rphenograph", "FastPG", "PARC", "FlowSOM"
markersToClusterBy <- NULL       # One of: "all", "type", "state"
kValuesIWant <- NULL
knn <- NULL          # Must be in kValuesIWant

## 📉 Dimension Reduction
dimRedMethodToUse <- NULL   # One of: "TSNE", "UMAP", "pacmap"
markersToDimRedBy <- NULL   # One of: "all", "type", "state"

## 🧼 QC
runQC <- NULL      # One of: "FlowAI", "PeacoQC", or "None" to skip
useQC <- NULL            # Whether to keep only QC-passed cells

## 💾 Output
RDataFolder <- NULL # Use existing results (skip re-analysis)
excludeTheseSamples <- NULL           # e.g. c("Sample_001", "Sample_026"), or NULL
gimmePDFs <- NULL
greyscalePlots <- FALSE
quantileNormaliseAll <- NULL

## 🧠 Parallelism
runInParallel <- NULL
nCores <- NULL
ramPerCore <- NULL

## 🎨 Aesthetics
themeToUse <- NULL    # ggplot2 theme
viridisColour <- NULL     # Palette: "magma", "inferno", "plasma", etc.

## WIP
runScGate <- FALSE

## ✅ Done editing? Knit away!
```

```{r check_variables}

required_vars <- c(
  "fp", "md_fp", "condaDir", "parcScriptDir", "downsampleTo",
  "clusteringMethodToUse", "markersToClusterBy", "kValuesIWant", "knn",
  "dimRedMethodToUse", "markersToDimRedBy",
  "runQC", "useQC", "RDataFolder", "excludeTheseSamples",
  "greyscalePlots", "gimmePDFs", "quantileNormaliseAll",
  "runInParallel", "nCores", "ramPerCore", "themeToUse", "viridisColour"
)

missingVariables <- names(unlist(setNames(lapply(required_vars, function(x) {
  if (!exists(x)) {
    x
  }
}), required_vars)))

if(!is.null(missingVariables)) {
  stop(paste0("❌ Error: ", missingVariables, " is missing. Please check in the input settings chunk and make sure the variable exists/isn't commented out."))
}
```

```{r load_packages}

# Load packages
pacman::p_load(
  BiocGenerics, S4Vectors, SummarizedExperiment, SingleCellExperiment, future.apply,
  flowCore, FlowSOM, tidyverse, CATALYST, readxl, cowplot, diffcyt, ggpubr, flowSpecs,
  ggprism, gridExtra, rstatix, Rphenograph, plotly, slingshot, grDevices,
  RColorBrewer, reshape2, grid, ComplexHeatmap, circlize, clustree, kableExtra,
  DT, flowAI, parallel, qs, ggrepel, gtools, reticulate, ggbeeswarm, Rcpp, Seurat,
  PeacoQC, MARMOT, future, SCpubr, scGate
)

# Define class union for ExpData
if (!methods::isClass("ExpData")) {
  try({
    methods::setClassUnion("ExpData", c("matrix", "SummarizedExperiment"), where = .GlobalEnv)
  }, silent = TRUE)
}

# Detect available cores
max_cores <- parallel::detectCores(logical = FALSE)
nCores <- min(nCores, max_cores - 1)  # leave one core free

# Set future plan based on OS and core count
if (nCores <= 1) {
  # Single-core mode — no parallelism
  plan(sequential)
} else if (.Platform$OS.type == "windows") {
  # Windows: use multisession
  options(future.fork.enable = FALSE)
  plan(multisession, workers = nCores)
} else if (Sys.info()["sysname"] == "Darwin") {
  # macOS (Intel or Apple Silicon)
  arch_info <- system("uname -m", intern = TRUE)
  r_arch <- R.version$arch

  if (r_arch == "x86_64" && arch_info == "arm64") {
    # Running x86_64 R under Rosetta on ARM Mac
    options(future.fork.enable = FALSE)
    plan(multisession, workers = min(nCores, 4))
  } else {
    # Native macOS (ARM or Intel)
    options(future.fork.enable = TRUE)
    plan(multicore, workers = nCores)
  }
} else {
  # Linux or other Unix-like systems
  options(future.fork.enable = TRUE)
  plan(multicore, workers = nCores)
}

# Set future options
options(future.globals.maxSize = (ramPerCore * nCores) * 1024^3)
options(future.wait.timeout = 120)
options(future.rng.onMisuse = "ignore")

# Reproducibility
future.seed <- TRUE
set.seed(2)
```

```{r load_rdata}

if (!is.null(RDataFolder)) {
  message(glue::glue("🔁 Loading results from: {RDataFolder}"))

  qs_files <- list.files(RDataFolder, pattern = "\\.(qs)$", full.names = TRUE, recursive = TRUE)
  if (length(qs_files) == 0) stop("No .qs files found in the specified RDataFolder.")

  for (file in qs_files) {
    obj_name <- tools::file_path_sans_ext(basename(file))
    message(glue::glue("→ Loading {obj_name}"))
    assign(obj_name, qs::qread(file, nthreads = nCores), envir = .GlobalEnv)
  }
}

# Pick the merge method depending on clustering backend
mergeBy <- dplyr::case_when(
  clusteringMethodToUse %in% c("Rphenograph", "FastPG") ~ "k",
  clusteringMethodToUse == "PARC" ~ "p",
  clusteringMethodToUse == "FlowSOM" ~ "meta",
  TRUE ~ NA_character_  # temp placeholder
)
if (is.na(mergeBy)) stop("Unsupported clustering method: ", clusteringMethodToUse)

# Heatmap colour palette
hmPalToUse <- if (greyscalePlots) {
  RColorBrewer::brewer.pal(9, "Greys")
} else {
  rev(RColorBrewer::brewer.pal(11, "RdBu"))
}
```

```{r import_metadata}

# Load metadata
md  <- readxl::read_xlsx(md_fp, sheet = "File Data")
smd <- readxl::read_xlsx(md_fp, sheet = "Study Data")

# Paths
fcsDir <- fp

# Order sample IDs naturally
md <- md[gtools::mixedorder(md$sample_id), ]
md$sample_id <- factor(md$sample_id, levels = md$sample_id)

# Apply condition ordering if available
conditionOrder <- smd$`Conditions Order` |> na.omit()
if (length(conditionOrder) > 0) {
  md$condition <- factor(md$condition, levels = conditionOrder)
} else {
  md$condition <- as.factor(md$condition)
}

# Extract marker names and types 
markers_to_use_and_their_type <- smd$`Marker Type`[!is.na(smd$`Marker Type`)]
names(markers_to_use_and_their_type) <- smd$`Markers to include`[!is.na(smd$`Marker Type`)]
# knitr::knit_exit()

# Define feature type from marker type
featureType <- if ("Marker Type" %in% names(smd)) {
  types <- na.omit(smd$`Marker Type`)
  if ("type" %in% types) "type" else NULL
} else {
  stop("Metadata file must contain 'Marker Type' column.")
}

# Markers to exclude
removeMarkers <- na.omit(smd$`Markers to exclude completely`)

# Cofactors
cfToUse <- na.omit(smd$`Cofactors for markers to use`)

# Contrasts
contrastsToTest <- na.omit(smd$`Conditions To Test`)
contrastsToTestS <- paste(contrastsToTest, collapse = ",")

# Subsampling settings
raw_subsamp <- na.omit(smd$`Cells per condition in UMAPs etc.`)
drCellPerCondition <- setNames(
  lapply(raw_subsamp, function(expr) eval(parse(text = expr))),
  conditionOrder
)

# Identify all condition columns
conditions <- c("condition", setdiff(names(md), c("file_name", "sample_id", "condition")))
conditions <- gsub("-", ".", conditions)

# Exclude samples if specified
if (!is.null(excludeTheseSamples)) {
  md <- md[!md$sample_id %in% excludeTheseSamples, ]
}

# Sort again by condition
md <- md[order(md$condition), ]
md$patient_id <- factor(md$sample_id, levels = md$sample_id)
md$sample_id  <- factor(md$sample_id, levels = md$sample_id)
```

```{r cluster_colours}

# Define base palette used in CATALYST-inspired plots
catalystCols <- c(
  "#DC050C", "#FB8072", "#1965B0", "#7BAFDE", "#882E72", "#B17BA6",
  "#FF7F00", "#FDB462", "#E7298A", "#E78AC3", "#33A02C", "#B2DF8A",
  "#55A1B1", "#8DD3C7", "#A6761D", "#E6AB02", "#7570B3", "#BEAED4",
  "#666666", "#999999", "#AA8282", "#D4B7B7", "#8600BF", "#BA5CE3",
  "#808000", "#AEAE5C", "#1E90FF", "#00BFFF", "#56FF0D", "#CFBF11"
)

# Darken the palette and append (to allow >30 colours with contrast)
darkenedCols <- colorspace::darken(catalystCols, 0.4)
catalystCols <- paste0(c(catalystCols, darkenedCols), "FF")  # Add full opacity
```

```{r sample_colours}

if (is.null(RDataFolder)) {
  # Collect all unique factor levels from the specified metadata columns
  allConditionLevels <- unlist(lapply(conditions, \(x) levels(factor(md[[x]]))), use.names = FALSE)
  
  # Assign colours to each unique level using a palette combo (expand as needed)
  palettePool <- c(
    RColorBrewer::brewer.pal(9, "Set1"),
    RColorBrewer::brewer.pal(8, "Dark2"),
    RColorBrewer::brewer.pal(12, "Paired"),
    RColorBrewer::brewer.pal(8, "Set3")
  )
  
  allConditionLevelColours <- setNames(palettePool[seq_along(allConditionLevels)], allConditionLevels)
  
  # Create named list of colours per condition
  coloursList <- setNames(
    lapply(conditions, \(x) {
      levels <- levels(factor(md[[x]]))
      allConditionLevelColours[levels]
    }),
    conditions
  )
  
  # Assign sample-level colours
  sampleCols <- setNames(
    colorRampPalette(brewer.pal(8, "Dark2"))(length(levels(md$sample_id))),
    levels(md$sample_id)
  )
  
  # Assign condition-level colours
  conditionCols <- setNames(
    colorRampPalette(brewer.pal(8, "Set1"))(length(levels(md$condition))),
    levels(md$condition)
  )
}
```

```{r import_fcs_files}

if (is.null(RDataFolder)) {
  message("📥 Importing FCS files...")
  
  # List all .fcs files
  fcs_files <- list.files(fcsDir, pattern = "\\.fcs$", full.names = TRUE)
  matches <- unlist(lapply(md$file_name, function(x) grep(x, fcs_files, fixed = TRUE)))
  fcs_files <- fcs_files[unique(matches)]
  fcs_names <- tools::file_path_sans_ext(basename(fcs_files))
  
  # Read each FCS file into a named list
  frames <- setNames(
    lapply(fcs_files, function(f) {
      flowCore::read.FCS(f, transformation = FALSE, truncate_max_range = FALSE)
    }),
    fcs_names
  )
  
  # Exclude samples, if specified
  if (!is.null(excludeTheseSamples)) {
    keep_samples <- md$file_name[!md$sample_id %in% excludeTheseSamples]
    frames <- frames[keep_samples]
  }
  
  framesList <- list(
    "Untransformed" = frames
  )
}
```

```{r create_panel}

if (is.null(RDataFolder)) {
  # Extract parameter metadata
  param_df <- pData(parameters(framesList$Untransformed[[1]])) %>%
    as.data.frame() %>%
    mutate(desc = ifelse(is.na(desc), name, desc))
  
  # Build panel
  panel <- param_df %>%
    select(fcs_colname = name, marker_name = desc) %>%
    mutate(
      antigen       = marker_name,
      marker_class  = ifelse(marker_name %in% names(markers_to_use_and_their_type), markers_to_use_and_their_type[marker_name], "state"),
      use_channel   = marker_name %in% names(markers_to_use_and_their_type)
    ) %>%
    dplyr::filter(!marker_name %in% removeMarkers) %>%
    dplyr::filter(marker_name %in% names(markers_to_use_and_their_type)) %>%
    arrange(match(marker_name, names(markers_to_use_and_their_type)))
  
  # Sanity check
  missing <- setdiff(names(markers_to_use_and_their_type), panel$marker_name)
  if (length(missing)) {
    stop("❌ Markers in metadata not found in FCS: ", paste(missing, collapse = ", "))
  }
  
  message("⚙️ Preparing cofactors for arcsinh transformation...")
  
  # Base cofactor vector, matched to FCS column names
  cf <- setNames(cfToUse, panel$fcs_colname[match(names(markers_to_use_and_their_type), panel$marker_name)])
  
  # Fill in default (5) for other channels
  all_channels <- panel$fcs_colname
  extra_channels <- setdiff(all_channels, names(cf))
  cf[extra_channels] <- 2500 # Default cofactor for unused channels
  
  # Ensure correct order
  cf <- cf[match(all_channels, names(cf))]
  
  # Alternative version cf2, named by marker name
  cf2 <- setNames(cf[match(panel$fcs_colname, names(cf))], panel$marker_name)
  cf2 <- cf2[panel$use_channel]
  
  # Assume `fs` is your flowSet and `cofactors` is a named numeric vector
  asinh_transforms <- transformList(
    names(cf),
    lapply(cf, function(cof) arcsinhTransform(a = 0, b = 1 / cof, c = 0))
  )
  # Apply to the flowSet
  framesList$Transformed <- future_lapply(seq_along(framesList$Untransformed), function(i) {
    flowCore::transform(framesList$Untransformed[[i]], asinh_transforms)
  }, future.seed = TRUE) %>% setNames(names(framesList$Untransformed))
}
```

```{r run_flow_auto_qc}

if (is.null(RDataFolder)) {
  if (runQC == "FlowAI") {
  
    message("🔍 Starting automated quality control...")
  
    metricsQC <- "FR_FS"
  
    qc_dir <- file.path(fp, "resultsQC_flowAI")
    expected_qc_files <- gsub(".fcs$", "_QC.fcs", md$file_name)
    qc_present <- dir.exists(qc_dir) &&
      all(expected_qc_files %in% list.files(qc_dir, pattern = "_QC.fcs"))
  
    if (qc_present) {
      message("📁 QC files found — loading from disk.")
    } else {
      message("⚙️  QC files not found — running `flow_auto_qc()`.")
      flow_auto_qc(
        fcsfiles      = as(framesList$Untransformed, "flowSet"),
        folder_results = qc_dir,
        remove_from   = metricsQC,
        fcs_highQ     = "_QC_highQ",
        fcs_lowQ      = "_QC_lowQ"
      )
    }
  
    # Load the QC FCS files
    fcsFilesQC <- list.files(qc_dir, full.names = TRUE, pattern = "_QC_highQ.fcs")
    fcsFilesQC <- fcsFilesQC[grep(paste((md$file_name %>% gsub("\\.fcs", "", .)), collapse = "|"), fcsFilesQC)]
    fcsNames <- sub("_QC_highQ.fcs$", "", basename(fcsFilesQC))
  
    message("📥 Loading ", length(fcsFilesQC), " QC FCS files into memory...")
    framesList[["FlowAI"]] <- setNames(
      lapply(fcsFilesQC, \(f) read.FCS(f, truncate_max_range = FALSE, transformation = FALSE)),
      fcsNames
    )
  
    # Force evaluation of ranges to avoid downstream surprises
    invisible(lapply(framesList[["QC"]], range, type = "data"))
  
    # Load and annotate QC summary table
    QCmini <- read_tsv(file.path(qc_dir, "QCmini.txt"), show_col_types = FALSE) |>
      distinct(`Name file`, .keep_all = TRUE)
    QCmini <- QCmini[grep(paste0(fcsNames, collapse = "|"), QCmini$`Name file`), ]
  
    message("🧾 Processing QC summary metadata...")
  
    qc_names_clean <- sub("_QC_highQ|\\.fcs", "", QCmini$`Name file`)
    md_names_clean <- sub("\\.fcs$", "", md$file_name)
  
    QCmini$`Sample Name` <- md$sample_id[match(qc_names_clean, md_names_clean)]
    QCmini <- QCmini[gtools::mixedorder(QCmini$`Name file`), ]
  
    # Use QC data as default if requested
    if (useQC) {
      
      framesList <- framesList[c("QC", setdiff(names(framesList), "QC"))]
    } else {
      message("📦 QC-filtered data loaded but not used in primary workflow.")
    }
  
    message("🎉 QC step complete.")
  }
}
```

```{r run_peacoqc}

if (is.null(RDataFolder)) {
  if (runQC == "PeacoQC") {
  
    message("🔍 Starting automated quality control with PeacoQC...🦚")
  
    qc_dir <- file.path(fp, "resultsQC_peacoQC")
    qc_present <- dir.exists(qc_dir) && {
      expected_qc_files <- gsub(".fcs$", "_QC.fcs", md$file_name)
      actual_qc_paths <- list.files(qc_dir, pattern = "_QC\\.fcs$", recursive = TRUE, full.names = TRUE)
      all(expected_qc_files %in% basename(actual_qc_paths))
    }
  
    if (qc_present) {
      message("📁 QC files found — loading from disk.")
    } else {
      message("⚙️  QC files not found — running PeacoQC...")
  
      dir.create(qc_dir, showWarnings = FALSE, recursive = TRUE)
  
      fs <- as(framesList$Transformed, "flowSet")
      all_channels <- colnames(fs[[1]])
      fluor_channels <- all_channels[!grepl("^(FSC|SSC|Time)", all_channels, ignore.case = TRUE)]
      # Also exclude common technical channels if present
      fluor_channels <- fluor_channels[!grepl("(Width|Height|Trigger)", fluor_channels, ignore.case = TRUE)]
      message("🔬 Using channels for QC: ", paste(fluor_channels, collapse = ", "))
      message("🚫 Excluded channels: ", paste(setdiff(all_channels, fluor_channels), collapse = ", "))
      
      fs_qc <- lapply(seq_along(fs), function(i) {
        fcs_name <- sampleNames(fs)[i]
        ff <- fs[[i]]
        qc_result <- PeacoQC::PeacoQC(
          ff = ff, 
          channels = fluor_channels, 
          plot = TRUE, 
          save_fcs = TRUE, 
          output_directory = qc_dir, 
          name_directory = fcs_name, 
          report = TRUE, 
          determine_good_cells = "all", 
          time_channel_parameter = "Time"
        )
        keep <- qc_result$GoodCells
        ff[keep, ]
      })
      fs_qc <- flowSet(fs_qc)
      
      # Generate a QC summary table
      qc_files <- list.files(path = qc_dir, pattern = "report.txt", recursive = TRUE, full.names = TRUE)
      QCmini <- read_tsv(qc_files[[1]], show_col_types = FALSE)
      for (i in 2:length(qc_files)) {
        xx <- read_tsv(qc_files[[i]], show_col_types = FALSE)
        QCmini[i,] <- xx[1,]
      }
      write_tsv(QCmini, file.path(qc_dir, "QCmini.txt"))
      message("✅ PeacoQC processing complete. QC files saved to: ", qc_dir)
    }
    
    # Load the QC FCS files
    fcsFilesQC <- list.files(qc_dir, full.names = TRUE, pattern = "_QC.fcs$", recursive = T)
    fcsFilesQC <- fcsFilesQC[grep(paste(md$file_name, collapse = "|"), fcsFilesQC)]
    fcsNames <- sub("_QC.fcs$", "", basename(fcsFilesQC))
  
    message("📥 Loading ", length(fcsFilesQC), " QC FCS files into memory...")
    framesList[["PeacoQC"]] <- setNames(
      lapply(fcsFilesQC, \(f) read.FCS(f, truncate_max_range = FALSE, transformation = FALSE)),
      fcsNames
    )
  
    # Force evaluation of ranges
    invisible(lapply(framesList[["QC"]], range, type = "data"))
  
    # Load QC summary table
    QCmini <- read_tsv(file.path(qc_dir, "QCmini.txt"), show_col_types = FALSE)
    QCmini <- QCmini[grep(paste0(fcsNames, collapse = "|"), QCmini$Filename), ]
    message("🧾 Processing QC summary metadata...")
  
    message("🎉 QC step complete.")
  }
}
```

```{r arrange_frames_list}

if (is.null(RDataFolder)) {
  message("📦 Converting frame lists into flowSets...")
  
  # Convert to flowSet and set $CYT keyword to "FACS"
  fsList <- purrr::imap(framesList, function(sample_list, name) {
    fs <- as(sample_list, "flowSet")
    for (j in seq_along(fs)) {
      keyword(fs[[j]])[["$CYT"]] <- "FACS"
    }
    fs
  })
}
```

```{r plot_cofactor_histograms}

message("📉 Generating cofactor histograms for visualising transformations...")

cofactorsToTry <- c(5, 50, 500, 1000, 2500, 5000, 10000, 15000)

# Aggressively downsample
framesList[["Marker Histos"]] <- lapply(framesList$Untransformed, function(fcs) {
  if (nrow(fcs) > 1000) {
    set.seed(2)
    fcs[sample(nrow(fcs), 1000), ]
  } else {
    fcs
  }
})

# Function to strip data from ggplot object
minify_plot_data <- function(p, colname, group_col = "condition") {
  # Keep only the columns needed for plotting (x and fill/colour)
  vars_needed <- unique(c(colname, group_col))
  p$data <- p$data[, vars_needed, drop = FALSE]
  p
}

# Plot generation without persistent cofactorDFList
cofactorPlotList <- lapply(names(markers_to_use_and_their_type), function(marker) {
  colname <- panel$fcs_colname[panel$marker_name == marker]
  marker_expr <- purrr::map(framesList[["Marker Histos"]], ~ exprs(.x)[, colname])
  
  # Flat vector of expression values
  expr_vec <- unlist(marker_expr)
  sample_vec <- rep(names(marker_expr), times = purrr::map_int(marker_expr, length))
  cond_vec <- md$condition[match(sample_vec, (md$file_name %>% gsub("\\.fcs", "", .)))]

  # Build only what's needed: wide format DF with just transformed values + condition
  df <- tibble(
    condition = cond_vec,
    pretransform = expr_vec
  )

  # Transform in-place
  transformed_cols <- purrr::map_dfc(cofactorsToTry, function(cof) {
    tibble(!!paste0("Cofactor_", cof) := asinh(expr_vec / cof))
  })

  df <- bind_cols(df, transformed_cols)

  # Generate plots and strip ggplot data
  plot_names <- c("pretransform", paste0("Cofactor_", cofactorsToTry))
  plot_set <- purrr::map(plot_names, function(colname) {
    p <- ggplot(df, aes(x = .data[[colname]], fill = condition, colour = condition)) +
      geom_density(alpha = 0.1) +
      ggtitle(colname) +
      theme_prism() +
      theme(text = element_text(size = 15), legend.position = "none") +
      scale_colour_manual(values = coloursList$condition) +
      scale_fill_manual(values = coloursList$condition) +
      geom_vline(xintercept = mean(df[[colname]], na.rm = TRUE), linetype = "dashed", colour = "red")
    minify_plot_data(p, colname)
  })

  setNames(plot_set, plot_names)
})

names(cofactorPlotList) <- names(markers_to_use_and_their_type)
```

```{r create_sce}

if (is.null(RDataFolder)) {
  if (useQC) {
    message("✅ Using QC-filtered data for downstream analysis.")
    fsToUse <- fsList[[runQC]]
  } else {
    fsToUse <- fsList[["Transformed"]]
  }
  
  if (!is.null(downsampleTo)) {
    message("🔽 Downsampling to ", downsampleTo, " cells per sample.")
    fsToUse <- fsApply(fsToUse, function(x) {
      set.seed(2)
      n_events <- nrow(x)
      sample_size <- min(downsampleTo, n_events)
      sampled_indices <- sample(n_events, sample_size)
      x[sampled_indices, ]
    })
  }
  
  message("📦 Creating SingleCellExperiment objects with CATALYST::prepData...")
  
  sce <- CATALYST::prepData(
    x = fsToUse,
    panel = panel,
    md = md,
    features = panel$fcs_colname,
    transform = FALSE,
    md_cols = list(factors = conditions)
  )
  assay(sce, "exprs") <- assay(sce, "counts")
  assay(sce, "exprsTransformed") <- assay(sce, "exprs")
  
  # 99% quantile normalisation
  message("📐 Applying quantile normalisation...")
  
  exprs_mat <- assay(sce, "exprs")
  qs <- matrixStats::rowQuantiles(exprs_mat, probs = c(0.01, 0.99))
  norm_exprs <- sweep(exprs_mat, 1, qs[, 1], "-")
  norm_exprs <- sweep(norm_exprs, 1, qs[, 2] - qs[, 1], "/")
  norm_exprs[norm_exprs < 0 | is.na(norm_exprs)] <- 0
  norm_exprs[norm_exprs > 1] <- 1
  
  assay(sce, "exprsQuantNorm") <- norm_exprs
  
  # Also include a scaled version
  scale.coef <- matrixStats::rowQuantiles(exprs_mat, probs = c(0.01, 0.99))
  assay(sce, "norm") <- t(scale(t(exprs_mat), center = scale.coef[, 1], scale = scale.coef[, 2] - scale.coef[, 1]))
}
```

```{r select_assay_and_markers}

# Define which markers to use for clustering
if (!is.null(featureType) && markersToClusterBy == "type") {
  featuresToCluster <- panel$marker_name[panel$marker_class == "type"]
} else if (!is.null(featureType) && markersToClusterBy == "state") {
  featuresToCluster <- panel$marker_name[panel$marker_class == "state"]
} else if (markersToClusterBy %in% c("type", "state", "all")) {
  featuresToCluster <- panel$marker_name
} else {
  warning("⚠️  Unknown value for `markersToClusterBy`, falling back to all markers.")
  featuresToCluster <- panel$marker_name
}

if(quantileNormaliseAll) {
  assay(sce, "exprs") <- assay(sce, "exprsQuantNorm")
}
set.seed(2)
if (quantileNormaliseAll) {
  exprsToUse = "exprsQuantNorm"
} else {
  exprsToUse = "exprsTransformed"
}
```

```{r clustering_graph}

if (is.null(RDataFolder)) {
    
  if (clusteringMethodToUse %in% c("Rphenograph", "FastPG")) {
    message("🧠 Running clustering using ", clusteringMethodToUse, "...")
  
    # Expression matrix for clustering
    es <- t(assay(sce, exprsToUse)[featuresToCluster, ])
  
    # Define cluster function to use
    cluster_fn <- function(k) {
      set.seed(2)
      if (clusteringMethodToUse == "FastPG") {
        clusters <- FastPG::fastCluster(data = es, k = k, num_threads = max(1, floor(nCores / length(kValuesIWant))))
        ids <- factor(clusters$communities + 1)
      } else if (clusteringMethodToUse == "Rphenograph") {
        clusters <- Rphenograph::Rphenograph(data = es, k = k)
        ids <- factor(membership(clusters[[2]]))
      }
      # Add "k" prefix and apply mixed sort
      ids <- paste0("k", ids)
      ids <- factor(ids, levels = gtools::mixedsort(unique(ids)))
      list(kids = ids, clusters = clusters)
    }
  
    # Run clustering (parallel if requested)
    kList <- if (runInParallel && length(kValuesIWant) > 1) {
      tryCatch({
        setNames(future_lapply(kValuesIWant, cluster_fn, future.seed = TRUE), paste0("k", kValuesIWant))
      }, error = function(e) {
        message("Parallel failed, using sequential: ", e$message)
        setNames(lapply(kValuesIWant, cluster_fn), paste0("k", kValuesIWant))
      })
    } else {
      setNames(lapply(kValuesIWant, cluster_fn), paste0("k", kValuesIWant))
    }
  
    # Store all clusterings in SCE object
    for (k in names(kList)) {
      sce[[k]] <- kList[[k]]$kids
    }
  
    # Select default cluster to use
    selected_k <- paste0("k", knn)
    sce$cluster_id <- sce[[selected_k]]
    
    metadata(sce)$cluster_codes <- eval(parse(text = paste0(
      "data.frame(", clusteringMethodToUse , " = factor(levels(sce[[selected_k]])))"
    )))
    kids <- unlist(sce[[paste0("k", knn)]])
    numbersOfClusters <- unique(kids)
  }
}
```

```{r parc_clustering, include = F}

if (clusteringMethodToUse == "PARC") {
  message("🐍 Running PARC clustering via reticulate...")
  
  # Check and setup Python environment
  if (is.null(reticulate::conda_binary())) {
    stop("❌ Conda not found. Please install conda or miniconda first.")
  }
  
  envs <- reticulate::conda_list()
  if ("p4r" %in% envs$name) {
    # Use bundled environment and scripts
    message("✅ Using bundled p4r environment")
    reticulate::use_condaenv("p4r")
    parc_script   <- system.file("python", "f_parc.py", package = "MARMOT", mustWork = TRUE)
    reticulate::source_python(parc_script)
  } else {
    # Use user-specified paths
    message("📁 Using user-specified conda and script paths")
    use_condaenv("p4r", conda = condaDir)
    source_python(file.path(parcScriptDir, "f_parc.py"))
  }
  
  set.seed(2)
  esPARC <- t(assay(sce, exprsToUse)[featuresToCluster, ])
  
  parcClustList <- setNames(future_lapply(kValuesIWant, function(k) {
    parc_clust(esPARC, knn = as.integer(k), num_threads = as.integer(nCores)) %>% unlist
  }, future.seed = TRUE), paste0("p", kValuesIWant))
  
  for (k in kValuesIWant) {
    cluster_labels <- factor(parcClustList[[paste0("p", k)]] + 1)
    cluster_labels <- paste0("p", cluster_labels)
    cluster_labels <- factor(cluster_labels, levels = gtools::mixedsort(unique(cluster_labels)))
    sce[[paste0("p", k)]] <- cluster_labels
  }
  
  chosen_k <- paste0("p", knn)
  sce$cluster_id <- sce[[chosen_k]]
  
  # Select default cluster to use
  selected_k <- paste0("p", knn)
  sce$cluster_id <- sce[[selected_k]]
  metadata(sce)$cluster_codes <- eval(parse(text = paste0(
    "data.frame(", clusteringMethodToUse , " = factor(levels(sce[[selected_k]])))"
  )))
  kids <- unlist(sce[[paste0("p", knn)]])
  numbersOfClusters <- levels(kids)
}
```

```{r FlowSOM}

if (is.null(RDataFolder)) {
  if (clusteringMethodToUse == "FlowSOM") {
    message("🌾 Running FlowSOM clustering...")
  
    sceSOM <- cluster(
      sce,
      features = featuresToCluster,
      xdim = 10,
      ydim = 10,
      maxK = max(kValuesIWant),
      seed = 42
    )
  
    # Store all clusterings
    for (k in kValuesIWant) {
      sce[[paste0("meta", k)]] <- as.character(cluster_ids(sceSOM, paste0("meta", k)))
      # Add "m" prefix and apply mixed sort
      sce[[paste0("meta", k)]] <- paste0("m", sce[[paste0("meta", k)]])
      sce[[paste0("meta", k)]] <- factor(sce[[paste0("meta", k)]], levels = gtools::mixedsort(unique(sce[[paste0("meta", k)]])))
    }
  
    # Assign selected clustering
    selected_k <- paste0("meta", knn)
    sce$cluster_id <- sce[[paste0("meta", knn)]]
  
    metadata(sce)$cluster_codes <- data.frame(FlowSOM = factor(levels(sce[[selected_k]]), levels = levels(sce[[selected_k]])))
    kids <- unlist(sce[[paste0("meta", knn)]])
    numbersOfClusters <- unique(kids)
  }
}
```

```{r dim_reduction, include = F}

if (is.null(RDataFolder)) {
  message("📉 Running dimensionality reduction: ", dimRedMethodToUse)

  # Fix seed so DR is identical regardless of which clustering method ran before this
  set.seed(2)

  # Select features for DR
  if (!is.null(featureType) && markersToDimRedBy == "type") {
    featuresToUmap <- panel$marker_name[panel$marker_class == "type"]
  } else if (!is.null(featureType) && markersToDimRedBy == "state") {
    featuresToUmap <- panel$marker_name[panel$marker_class == "state"]
  } else {
    featuresToUmap <- panel$marker_name
  }
  
  # Always run PCA first (needed for pacmap, helpful for others)
  sce <- runDR(x = sce, dr = "PCA", features = featuresToUmap, assay = exprsToUse)
  
  # Run PaCMAP if requested 
  if (dimRedMethodToUse == "pacmap") {
    
    # Check and setup Python environment
    if (is.null(reticulate::conda_binary())) {
      stop("❌ Conda not found. Please install conda or miniconda first.")
    }
  
    envs <- reticulate::conda_list()
    if ("p4r" %in% envs$name) {
      # Use bundled environment and scripts
      message("✅ Using bundled p4r environment")
      reticulate::use_condaenv("p4r")
      pacmap_script <- system.file("python", "f_pacmap.py", package = "MARMOT", mustWork = TRUE)
      reticulate::source_python(pacmap_script)
    } else {
      # Use user-specified paths
      message("📁 Using user-specified conda and script paths")
      use_condaenv("p4r", conda = condaDir)
      source_python(file.path(parcScriptDir, "f_pacmap.py"))
    }
  
    es <- as.data.frame(t(assay(sce, exprsToUse)[featuresToUmap, ]))
    embedding <- reducedDim(sce, dimRedMethodToUse) <- pacmap_fit(es, verbose = TRUE, apply_pca = FALSE)
    colnames(embedding) <- c("x", "y")
    
    if (is.null(embedding)) {
      message("⚠️ PaCMAP unavailable → skipping.")
      dimRedMethodToUse <- "UMAP"
    } else {
      reducedDim(sce, dimRedMethodToUse) <- embedding
    }
  }
  
  # Native DRs
  sce <- runDR(x = sce, dr = "UMAP", features = featuresToUmap, assay = exprsToUse)
  if (dimRedMethodToUse == "TSNE") {
    sce <- runDR(x = sce, dr = "TSNE", features = featuresToUmap, assay = exprsToUse)
  }
}
```

```{r make_seurat_object}

if (is.null(RDataFolder)) {
  message("🖼️Generating Seurat object...")
  # Seurat conversion and scaling
  scData <- suppressWarnings(Seurat::as.Seurat(sce, counts = exprsToUse, data = "exprsQuantNorm"))
  scData <- Seurat::ScaleData(scData, assay = "originalexp", verbose = FALSE)
}
```

```{r run_scGate}

if (is.null(RDataFolder) && runScGate) {
  message("🌉 Running scGate on marker pairs...")
  markerPairs <- smd$`Marker Pairs`[!is.na(smd$`Marker Pairs`)]
  markerPairTypes <- markerPairs %>% gsub("\\:.*", "", .) %>% unique
  markerPairs2 <- markerPairs
  names(markerPairs2) <- markerPairTypes
  models <- lapply(markerPairTypes, function(x) {
    markerPairs2[x][[1]] %>% strsplit(" ") %>% .[[1]] %>% .[2:3]
  }) %>% setNames(markerPairTypes)
  for (x in names(models)) {
    if (!models[[x]][[1]] %in% rownames(scData)) {
        models[[x]][[1]] <- agrep(models[[x]][[1]], rownames(scData), value = TRUE)
    }
    if (!models[[x]][[2]] %in% rownames(scData)) {
        models[[x]][[2]] <- agrep(models[[x]][[2]], rownames(scData), value = TRUE)
    }
    model <- gating_model(name = x, signature = c(models[[x]][[1]], models[[x]][[2]]))
    scData <- scGate(data = scData, model = model, pos.thr = 0.65)
    y <- x %>% gsub("\\ |\\-", "", .)
    scData[[paste0("is_", y)]] <- scData$scGate_multi
    scData[[paste0("is_", y)]][!is.na(scData[[paste0("is_", y)]])] <- x
  }
  
  # Get UCell results and determine gated cells
  ucell_cols <- grep("UCell", colnames(scData@meta.data), value = TRUE)
  scData$Gated_Cells <- lapply(1:nrow(scData@meta.data), function(i) {
    (ucell_cols[which.max(scData@meta.data[i, ucell_cols])] %>% gsub("_UCell", "", .))
  }) %>% unlist
  
  # Transfer scGate results back to sce
  sce$Gated_Cells <- scData$Gated_Cells
  
  # Transfer all scGate-related metadata
  scgate_cols <- grep("^is_|UCell", colnames(scData@meta.data), value = TRUE)
  for (col in scgate_cols) {
    sce[[col]] <- scData@meta.data[[col]]
  }
  
  # Add gate colours to the colour list 
  catalystColsGates <- catalystCols[1:length(unique(scData$Gated_Cells))]
  names(catalystColsGates) <- unique(scData$Gated_Cells)
  coloursList[["Gated_Cells"]] <- catalystColsGates
}
```

```{r prepare_dimreds_all}

if (is.null(RDataFolder)) {
  message("🧭 Preparing dimensionality reduction dataframes for all methods...")
  
  # Get available DR methods with 2D coords
  dr_methods <- reducedDimNames(sce)
  valid_methods <- dr_methods[sapply(dr_methods, function(m) ncol(reducedDim(sce, m)) >= 2)]
  
  # Function to build one DR dataframe
  build_dr_df <- function(method) {
    coords <- reducedDim(sce, method)[, 1:2]
    colnames(coords) <- c("x", "y")
    df <- data.frame(
      colData(sce),
      coords,
      DRMethod = method,
      check.names = FALSE
    )
    
    # Add cluster annotations
    if (!is.null(metadata(sce)$cluster_codes)) {
      df <- left_join(df, cluster_codes(sce), by = c("cluster_id" = clusteringMethodToUse))
    }
    df[[clusteringMethodToUse]] <- df$cluster_id
    
    # Add quantile-normalised expression
    qn <- t(assay(sce, "exprsQuantNorm"))
    df <- bind_cols(df, as.data.frame(qn))
    
    return(df)
  }
  
  # Build full DR list
  umapDFList <- lapply(valid_methods, build_dr_df) %>% setNames(valid_methods)
  
  # Optionally downsample per condition
  if (!is.null(drCellPerCondition)) {
    set.seed(2)
    for (method in valid_methods) {
      df <- umapDFList[[method]]
      down_df <- df %>%
        group_by(sample_id) %>%
        group_modify(~ {
          cond <- unique(md$condition[md$sample_id == .y$sample_id])
          target_n <- as.numeric(drCellPerCondition[as.character(cond)])
          if (nrow(.x) > target_n) .x[sample(nrow(.x), target_n), ] else .x
        }) %>%
        ungroup()
      umapDFList[[paste0("Downsampled.", method)]] <- down_df
    }
  }
  
  # Create label vectors for each DR method
  labsList <- lapply(valid_methods, function(method) {
    source_group <- if (!is.null(drCellPerCondition)) paste0("Downsampled.", method) else method
    df <- umapDFList[[source_group]]
    labs <- df %>%
      group_by(condition) %>%
      summarise(n = n(), .groups = "drop") %>%
      mutate(label = paste0(condition, "\n n = ", n)) %>%
      deframe()
    names(labs) <- levels(factor(df$condition))
    return(labs)
  })
  names(labsList) <- valid_methods
  
  umapDFList$All <- umapDFList[[dimRedMethodToUse]]
  umapDFList$Downsampled <- umapDFList[[paste0("Downsampled.", dimRedMethodToUse)]]
}
```

```{r cluster_colours2}

if (is.null(RDataFolder)) {
  message("🎨 Assigning colours to clusters...")
  
  # Base cluster_id colours
  cluster_levels <- levels(factor(umapDFList$All$cluster_id))
  coloursList[["cluster_id"]] <- setNames(catalystCols[seq_along(cluster_levels)], cluster_levels)
  
  # Add colours for each k-based clustering
  for (k in kValuesIWant) {
    colname <- paste0(mergeBy, k)
    if (colname %in% names(umapDFList$All)) {
      cluster_levels_k <- levels(factor(umapDFList$All[[colname]]))
      coloursList[[colname]] <- setNames(catalystCols[seq_along(cluster_levels_k)], cluster_levels_k)
    }
  }
}
```

```{r da_ds_analysis}
message("🚀 Starting differential abundance (DA) and differential state (DS) analysis...")

daPValToUse <- "p_adj"

# ───────────────────────────────────────────────────────
# 🧹 Tidy condition names and prepare design matrix
ei <- metadata(sce)$experiment_info
ei$condition <- gsub(" ", ".", ei$condition)

ei$conditionToTest <- ei$condition %>%
  gsub(" ", ".", .) %>%
  gsub("-", "0", .) %>%
  gsub("\\+", "1", .) %>%
  gsub("/", "", .) %>%
  as.factor()

cTTforEdgeR <- contrastsToTest %>%
  gsub(" ", ".", .) %>%
  gsub("-", "0", .) %>%
  gsub("\\+", "1", .) %>%
  gsub("\\.over\\.", "-", .) %>%
  gsub("\\.vs\\.", "-", .) %>%
  gsub("/", "", .)

cTTForPlots <- contrastsToTest

design <- model.matrix(~0 + conditionToTest, data = ei)
colnames(design) <- gsub("conditionToTest", "", colnames(design))

# ───────────────────────────────────────────────────────
# ✍️ Create contrast matrix
message("🧮 Constructing contrasts for edgeR...")
contrasts <- setNames(lapply(cTTforEdgeR, \(ctr) {
  eval(parse(text = paste0("limma::makeContrasts(", ctr, ", levels = design)")))
}), cTTforEdgeR)

# ───────────────────────────────────────────────────────
# 📊 Differential abundance analysis (DA)
message("📈 Running DA analysis using diffcyt...")
daList <- setNames(lapply(seq_along(contrasts), function(i) {
  res <- diffcyt(
    d_input            = sce,
    design             = design,
    contrast           = contrasts[[i]],
    analysis_type      = "DA",
    method_DA          = "diffcyt-DA-edgeR",
    clustering_to_use  = clusteringMethodToUse,
    min_cells          = 5,
    transform          = FALSE,
    normalize          = TRUE
  )
  as.data.frame(rowData(res$res))
}), cTTForPlots)

# ───────────────────────────────────────────────────────
# 🧬 Differential state analysis (DS)
message("🧪 Running DS analysis using diffcyt...")
if (is.null(featureType)) {
  rowData(sce)$marker_class <- "state"
}
sceDS <- sce
ds_formula1 <- createFormula(ei, cols_fixed = "conditionToTest", cols_random = "sample_id")

dsList <- setNames(vector("list", length(contrasts)), cTTForPlots)

for (i in seq_along(contrasts)) {
  res_DS <- diffcyt(
    d_input           = sceDS,
    clustering_to_use = clusteringMethodToUse,
    analysis_type     = "DS",
    method_DS         = "diffcyt-DS-limma",
    design            = design,
    contrast          = contrasts[[i]],
    transform         = FALSE,
    normalize         = TRUE,
    verbose           = FALSE
  )
  dsList[[i]] <- list(res_DS = res_DS, tbl_DS = rowData(res_DS$res))
}

# ───────────────────────────────────────────────────────
# 🔗 Merge clusters and rerun DS
message("🔁 Merging clusters and rerunning DS on merged data...")
mergedID <- paste0(clusteringMethodToUse, "_merged")
mergedTable <- data.frame(
  old_cluster = levels(as.factor(sce[[paste0(mergeBy, knn)]])),
  new_cluster = "all"
)

sceDS <- mergeClusters(
  x     = sceDS,
  k     = clusteringMethodToUse,
  id    = mergedID,
  table = mergedTable
)

ds_formula1 <- createFormula(ei, cols_fixed = "conditionToTest")

for (i in seq_along(contrasts)) {
  res_DS <- diffcyt(
    d_input           = sceDS,
    formula           = ds_formula1,
    clustering_to_use = mergedID,
    analysis_type     = "DS",
    method_DS         = "diffcyt-DS-limma",
    design            = design,
    contrast          = contrasts[[i]],
    transform         = FALSE,
    normalize         = TRUE,
    verbose           = FALSE
  )
  dsList[[paste(cTTForPlots[i], "Merged")]] <- list(
    res_DS = res_DS,
    tbl_DS = rowData(res_DS$res)
  )
}

# ───────────────────────────────────────────────────────
# 🔍 Identify significant DA clusters
message("🔬 Extracting significant DA clusters...")

# Pre-allocate list for efficiency
selectedClustersList <- vector("list", length(contrasts) * 2)
names_list <- character(length(contrasts) * 2)

for (i in seq_along(contrasts)) {
  contrast_conditions <- strsplit(contrastsToTest[i], " over ")[[1]]
  da_table <- daList[[i]]
  
  # Filter for complete cases and significant results
  significant_rows <- complete.cases(da_table[c("cluster_id", "logFC", daPValToUse)]) &
                     da_table[[daPValToUse]] < 0.05
  
  significant_data <- da_table[significant_rows, ]
  
  # Extract clusters by direction
  up_clusters <- as.character(unique(significant_data$cluster_id[significant_data$logFC >= 0]))
  down_clusters <- as.character(unique(significant_data$cluster_id[significant_data$logFC <= 0]))
  
  # Store results
  idx_up <- 2 * i - 1
  idx_down <- 2 * i
  
  selectedClustersList[[idx_up]] <- up_clusters
  selectedClustersList[[idx_down]] <- down_clusters
  
  names_list[idx_up] <- paste0("Contrast ", i, ": ", contrast_conditions[1], " Up")
  names_list[idx_down] <- paste0("Contrast ", i, ": ", contrast_conditions[2], " Up")
}

names(selectedClustersList) <- names_list

# ───────────────────────────────────────────────────────
# 🎨 Assign colours to selected clusters
message("🎨 Assigning colours to selected clusters...")
for (i in seq_along(selectedClustersList)) {
  label <- names(selectedClustersList)[i]
  clusterIDs <- selectedClustersList[[i]]
  sc <- coloursList[[paste0(mergeBy, knn)]][clusterIDs]
  coloursList[[label]] <- sc
}

# 🗺️ Map clusters to colours
message("🗺️ Generating cluster-to-colour mapping list...")
clustersToMapList <- lapply(names(selectedClustersList), function(label) {
  ids <- selectedClustersList[[label]]
  clrs <- coloursList[[label]]
  names(clrs) <- ids
  clrs
})
names(clustersToMapList) <- names(selectedClustersList)

message("✅ DA and DS analysis complete.")
```

```{r output_directories}

# Create Results root directory if not resuming from saved RData
if (is.null(RDataFolder)) {
  resultsDir <- file.path(fp, paste0("Results_Files_", timeRun))
  dir.create(resultsDir, showWarnings = FALSE, recursive = TRUE)

  qsDir <- file.path(resultsDir, "R_files")
  dir.create(qsDir, showWarnings = FALSE)
}

# PDFs
if (gimmePDFs) {
  pdfDir <- if (is.null(RDataFolder)) {
    file.path(resultsDir, "PDF_figures")
  } else {
    file.path(RDataFolder, "PDF_figures")
  }
  dir.create(pdfDir, showWarnings = FALSE)
}

# Excel outputs
excelDir <- if (is.null(RDataFolder)) {
  file.path(resultsDir, "Excel_Files")
} else {
  file.path(RDataFolder, "Excel_Files")
}
dir.create(excelDir, showWarnings = FALSE)
```

```{r write_results_to_file}

if (is.null(RDataFolder)) {
  message("💾 Saving R objects to disk...")
  vars_to_save <- c(
    "QCmini", "cf2", "dimRedMethodToUse", "markers_to_use_and_their_type", "framesList", "frames", "nrowsCofactors",
    "sce", "sce2", "sceDS", "coloursList", "md", "downsampleTo", "sceList", "lvlsColours",
    "featuresToCluster", "featuresToUmap", "smd", "knn", "clusteringMethodToUse",
    "selectedClustersList", "kids", "numbersOfClusters", "kList", "umapDFList", "kValuesIWant",
    "clustersToMapList", "daList", "dsList", "daPValToUse", "panel", "labs", "scData"
  )
  
  # Use knitr::knit_global() instead of .GlobalEnv
  knit_env <- knitr::knit_global()
  existing_vars <- vars_to_save[sapply(vars_to_save, exists, envir = knit_env)]
  missing_vars <- setdiff(vars_to_save, existing_vars)
  
  message("📊 Found ", length(existing_vars), " out of ", length(vars_to_save), " variables")
  if(length(missing_vars) > 0) {
    message("❌ Missing variables: ", paste(missing_vars, collapse = ", "))
  }
  
  write_tsv(data.frame(variable = existing_vars), file = file.path(qsDir, "existing_vars.txt"))
  
  # Save each variable from the correct environment
  purrr::walk(existing_vars, function(varname) {
    out_path <- file.path(qsDir, paste0(varname, ".qs"))
    qs::qsave(get(varname, envir = knit_env), file = out_path, nthreads = nCores, preset = "archive")
    message("✔️ Saved: ", varname)
  })
}
```

```{r nice_umap_function}
geom_text_repel_quiet <- function(..., max.overlaps = 10) {
  suppressWarnings(ggrepel::geom_text_repel(..., max.overlaps = max.overlaps))
}
geom_label_repel_quiet <- function(..., max.overlaps = 10) {
  suppressWarnings(ggrepel::geom_label_repel(..., max.overlaps = max.overlaps))
}

plotUMAPNice <- function(
    df = umapDF,
    colour_by = "condition",
    show_density = FALSE,
    size = 0.7,
    alpha = 1,
    text_size = 12,
    facet_by = NULL,
    show_grids = FALSE,
    name = "Normalised\nExpression\n",
    show_cluster_labels = FALSE,
    cluster_labels = "cluster_id"
) {
  gp <- ggplot(df, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[[colour_by]]))
  
  # Density overlay
  if (show_density) {
    gp <- gp +
      geom_density_2d(
        data = df[, c("x", "y")],
        aes(x = x, y = y),
        colour = "lightgrey",
        size = 0.5,
        bins = 20,
        alpha = 0.8
      ) +
      coord_cartesian(
        xlim = range(df$x, na.rm = TRUE) + c(-1, 1),
        ylim = range(df$y, na.rm = TRUE) + c(-1, 1)
      )
  }
  
  gp <- gp +
    geom_point(size = size, alpha = alpha, shape = 16) +
    labs(x = paste(dimRedMethodToUse, "dim 1"),
         y = paste(dimRedMethodToUse, "dim 2"),
         title = colour_by) +
    theme_prism() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      legend.text = element_text(size = 10),
      panel.border = element_rect(colour = "black", fill = NA, size = 1.5)
    )
  
  # Guides
  if (colour_by == "condition" | colour_by == "Gated_Cells") {
    gp <- gp + guides(colour = guide_legend(override.aes = list(size = 2)))
  }
  
  # Cluster label overlays
  if (show_cluster_labels) {
    label_coords <- data.frame(
      df %>%
        group_by(.data[[cluster_labels]]) %>%
        summarise(x = median(x), y = median(y)))
    if (colour_by %in% names(markers_to_use_and_their_type)) {
      label_coords[[colour_by]] <- lapply(levels(df[[cluster_labels]]), function(level) {
        mean(df[[colour_by]][df[[cluster_labels]] == level])
      }) %>% unlist()
      gp <- gp + geom_label_repel(data = label_coords, aes(x = x, y = y, label = .data[[cluster_labels]], fill = .data[[colour_by]]), colour = "black", show.legend = F)
    } else {
      gp <- gp + geom_label_repel(data = label_coords, aes(x = x, y = y, label = .data[[cluster_labels]], fill = .data[[cluster_labels]]), colour = "black", show.legend = F)
    }
  }
  
  # Show the grid lines?
  if (!show_grids) {
    gp <- gp +
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  }
  
  # Gradient or manual colour scale
  if (colour_by %in% names(markers_to_use_and_their_type)) {
    gp <- gp + scale_colour_gradientn(name, colours = hcl.colors(64, viridisColour))
    gp <- gp + scale_fill_gradientn(name, colours = hcl.colors(64, viridisColour))
  } else {
    gp <- gp + scale_colour_manual(values = coloursList[[colour_by]])
    gp <- gp + scale_fill_manual(values = coloursList[[colour_by]])
  }
  
  # Faceting
  if (!is.null(facet_by)) {
    gp <- gp + facet_wrap(facets = as.formula(paste("~", facet_by)))
  }
  
  return(gp)
}

```

Started on `r paste(timeRun)`

# Results {.tabset}

## QC {.tabset}

```{r, label = "print QC", results = 'asis'}
if (runQC %in% c("FlowAI", "PeacoQC")) {
  cat("### QC Cell Filtering \n\n")
  if(runQC %in% c("FlowAI", "PeacoQC") & !useQC) {
    cat("\n\n")
    cat(
      "You have selected to run the QC from", useQC, ", but *not* to use it. 
      The data here is presented purely for your information and does not reflect any of the results in this report. 
      If you wish to also use only the cells that passed QC, please change the `useQC` variable to TRUE in the metadata file.")
    cat("\n\n")
  }
  print(kable(QCmini, format = "html", escape = FALSE) %>% kable_styling())
  cat("\n\n")
  openxlsx::write.xlsx(as.data.frame(QCmini), file = file.path(excelDir, "QCtable.xlsx"))
}
```

### Cofactor Histograms {.tabset}

```{r, label = "print cofactor plots", results = 'asis', fig.height=12, fig.width=12}
for (marker in names(markers_to_use_and_their_type)) {
  cat("####", marker, "\n")
  cat(paste(marker, "Cofactor you chose:", cf2[[marker]]))
  cat("\n\n")
  grid.arrange(grobs = cofactorPlotList[[marker]], ncol = 3, top = marker)
  cat("\n\n")
}
nrowsCofactors <- ceiling(length(names(markers_to_use_and_their_type))/4)
```

### Marker Histograms

These are the histograms of all markers using the selected cofactors. Downsampled to 2000 cells per sample.

```{r, label = "plot marker histograms 2", fig.height = (nrowsCofactors*2)+1, fig.width = 15}
y <- assay(sce, "exprs")
df <- data.frame(t(y), colData(sce), check.names = FALSE)
gg_df <- melt(df, value.name = "value", variable.name = "antigen", 
  id.vars = names(colData(sce)))
mh <- ggplot(gg_df, aes(x = value, y = ..ndensity.., fill = condition, color = condition)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~antigen, scales = "free") + 
  ylab("Normalised Density") + 
  xlab("Transformed Value") +
  theme_prism(axis_text_angle = 90) + theme(panel.grid = element_blank(),
    strip.background = element_blank(), strip.text = element_text(face = "bold"),
    axis.text = element_text(color = "black"), axis.title = element_text(color = "black"),
    text = element_text(size = 15)) +
  scale_colour_manual(values = coloursList$condition) +
  scale_fill_manual(values = coloursList$condition)
mh
```

```{r, label = "save marker histogram", results="hide"}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "marker_histograms.pdf"), width = 15, height = (nrowsCofactors*2), bg = "transparent")
  print(mh)
  dev.off()
}
```

### Cell Counts

The number of samples per conditions is:

```{r, label = "print cell counts1"}
table(md$condition)
```

```{r, label = "warn qc counts"}
if (useQC & !is.null(downsampleTo)) {
  cat("\n\n")
  print("You chose to downsample AND use the QC-filtered counts. Therefore, cells were QC'd and then downsampled. The numbers here reflect this two-step approach.")
  cat("\n\n")
}
```

The number of cells per conditions is:

```{r, label = "print cell counts2"}
table(sce$condition)
```

The number of cells per sample is:

```{r, label = "print cell counts3"}
n_cells(sce)
```

```{r, label = "plot cell counts 1", fig.height=5, fig.width=ceiling(length(conditions)*3)}
plotCounts(sce, color_by = "condition") + scale_fill_manual(values = coloursList$condition) + theme_prism(axis_text_angle = 45)
```

```{r, label = "plot cell counts 2", fig.height=4, fig.width=ceiling(length(conditions)*4)}
plotCounts(sce, color_by = "condition", group_by = "sample_id") + scale_fill_manual(values = coloursList$condition) + theme_prism(axis_text_angle = 45)
```

```{r, label = "plot full counts", results = 'asis', fig.height=4, fig.width=ceiling(length(conditions)*4)}
if (!is.null(downsampleTo)) {
  cat("\n\nHowever, because you downsampled, here are those numbers for the full dataset. The black line is what you downsampled to.\n\n")
  cat("\n\n")
  originalCellNumbers <- lapply(names(framesList$Untransformed), function(x) {
    nrow(framesList$Untransformed[[x]])
  }) %>% setNames(names(framesList$Untransformed))
  originalCellNumbers <- originalCellNumbers %>% data.frame(check.names = F) %>% t %>% data.frame
  colnames(originalCellNumbers)[[1]] <- "Cells"
  originalCellNumbers <- rownames_to_column(.data = originalCellNumbers, "file_name")
  originalCellNumbers$file_name <- paste0(originalCellNumbers$file_name, ".fcs")
  originalCellNumbers <- left_join(originalCellNumbers, md, by = "file_name")
  barplotAllCells <- ggplot(originalCellNumbers, aes(x = sample_id, y = Cells, fill = condition)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = coloursList$condition) +
    labs(x = "Sample ID", y = "Number of Cells", title = "All Cells (before downsampling)") +
    theme_prism(axis_text_angle = 45) +
    geom_hline(yintercept = downsampleTo, linetype = "dashed", colour = "black")
  print(barplotAllCells)
  cat("\n\n")
}
```

### Pseudo-bulk MDS Plots {.tabset}

```{r, label = "plot mds by condition", fig.height=6, fig.width=9, results = 'asis'}
mdsList <- setNames(lapply(conditions, function(x) {
  mds1 <- CATALYST::pbMDS(x = sce, color_by = x, size_by = T, features = featureType)
  mds1 <- mds1 + scale_colour_manual(values = coloursList[[x]]) + theme_prism(base_size = 16) + theme(panel.grid = element_blank())
  mds1
}), conditions)
lapply(names(mdsList), function(x) {
  cat("####", x, "\n")
  print(mdsList[[x]])
  cat("\n\n")
})
```

```{r, label = "save mds plots", include = FALSE}
if (gimmePDFs) {
  lapply(names(mdsList), function(x) {
    pdf(file = file.path(pdfDir, paste0(x, "_MDS.pdf")), width = 7, height = 5, bg = "transparent")
    print(mdsList[[x]])
    dev.off()
  })
}
```

### Non-redundancy score plot

```{r, label = "plot nrs", fig.height=4, fig.width=12}
plotNRS(sce, color_by = "condition", features = featureType) + scale_colour_manual(values = coloursList$condition)  + theme_prism(axis_text_angle = 90)
```

## `r paste(clusteringMethodToUse)` Cluster Analysis {.tabset}

### K Tree {.tabset}

The current k selected is `r paste(knn)`

The markers used for clustering were:

```{r, label = "show feature to cluster"}
paste(featuresToCluster)
```

```{r, label = "get cluster numbers", include = F}
numbersOfClustersPerKnn <- setNames(lapply(paste0(mergeBy, kValuesIWant), function(x) {
  gtools::mixedsort(levels(sce[[x]]))
}), paste0(mergeBy, kValuesIWant))
```

```{r, label = "clustree", fig.height = ceiling(length(kValuesIWant)*3), fig.width = ceiling(max(lengths(numbersOfClustersPerKnn))/2.5)}
if(length(kValuesIWant) >= 2) {
  clustree::clustree(sce, prefix = mergeBy, exprs = "exprsTransformed")
}
```

```{r, label = "marker pair scatterplots", results = 'asis', fig.height=5, fig.width=5, warning = F}
runMarkerPairPlots <- ("type" %in% as.character(markers_to_use_and_their_type)) && ("state" %in% as.character(markers_to_use_and_their_type)) && ("Marker Pairs" %in% colnames(smd))
if (runMarkerPairPlots) {
  cat("\n\n")
  cat("### Marker-Pair Scatterplots {.tabset} \n\n")
  es <- as.data.frame(t(assay(sce, "exprsTransformed")))
  es[[mergeBy]] <- sce[[paste0(mergeBy, knn)]]
  set.seed(2)
  if (!is.null(downsampleTo)) {
    if (nrow(es) > downsampleTo) {
      es <- es[sample(rownames(es), downsampleTo),]
    }
  }
  esQN <- as.data.frame(t(assay(sce, "exprsQuantNorm")))
  esQN[[mergeBy]] <- sce[[paste0(mergeBy, knn)]]
  if (!is.null(downsampleTo)) {
    if (nrow(esQN) > downsampleTo) {
      esQN <- esQN[sample(rownames(esQN), downsampleTo),]
    }
  }
  
  markerPairs <- smd$`Marker Pairs`[!is.na(smd$`Marker Pairs`)]
  markerPairTypes <- markerPairs %>% gsub("\\:.*", "", .) %>% unique
  markerPairsList <- setNames(lapply(markerPairTypes, function(x) {
    toPlot <- markerPairs[grep(paste0("^", x), markerPairs)]
    plots <- lapply(toPlot, function(y) {
      markersToPlot <- y %>% strsplit("\\: ") %>% .[[1]] %>% .[[2]] %>% strsplit(., " ") %>% .[[1]]
      if (!markersToPlot[[1]] %in% colnames(es)) {
        markersToPlot[[1]] <- agrep(markersToPlot[[1]], colnames(df), value = TRUE)
      }
      if (!markersToPlot[[2]] %in% colnames(es)) {
        markersToPlot[[2]] <- agrep(markersToPlot[[1]], colnames(df), value = TRUE)
      }
      
      if (!quantileNormaliseAll) {
        median <- data.table::rbindlist(lapply(unique(es[[mergeBy]]), function(x) {
          data.frame(
            "V1" = x,
            "x" = median(es[[markersToPlot[[1]]]][es[[mergeBy]] == x]),
            "y" = median(es[[markersToPlot[[2]]]][es[[mergeBy]] == x])
          )
        }))
        colnames(median)[1] <- mergeBy
        q1 <- ggplot(es, aes(x = .data[[markersToPlot[1]]], y = .data[[markersToPlot[2]]], color = .data[[mergeBy]])) + 
          geom_point(size = 0.001) +
          geom_label_repel_quiet(data = median, aes(label = .data[[mergeBy]], x = x, y = y, fill = .data[[mergeBy]]), show.legend = FALSE, size = 4, colour = "black") +
          scale_color_manual(values = coloursList[[paste0(mergeBy, knn)]]) + 
          scale_fill_manual(values = coloursList[[paste0(mergeBy, knn)]]) + 
          theme_prism() +
          theme(legend.position = "hide") +
          ggtitle(paste0(x, "\nTransformed"))
      
      }
      
      median <- data.table::rbindlist(lapply(unique(esQN[[mergeBy]]), function(x) {
        data.frame(
          "V1" = x,
          "x" = median(esQN[[markersToPlot[[1]]]][esQN[[mergeBy]] == x]),
          "y" = median(esQN[[markersToPlot[[2]]]][esQN[[mergeBy]] == x])
        )
      }))
      colnames(median)[1] <- mergeBy
      p1 <- ggplot(esQN, aes(x = .data[[markersToPlot[1]]], y = .data[[markersToPlot[2]]], color = .data[[mergeBy]])) + 
        geom_point(size = 0.001) +
        geom_label_repel_quiet(data = median, aes(label = .data[[mergeBy]], x = x, y = y, fill = .data[[mergeBy]]), show.legend = FALSE, size = 4, colour = "black") +
        scale_color_manual(values = coloursList[[paste0(mergeBy, knn)]]) + 
        scale_fill_manual(values = coloursList[[paste0(mergeBy, knn)]]) + 
        theme_prism() +
        theme(legend.position = "hide") +
        ggtitle(paste0(x, "\nTransformed & QN"))
      
      return(list(q1, p1))
    }) %>% setNames(toPlot)
  }), markerPairTypes)
  
  lapply(names(markerPairsList), function(x) {
    cat("####", x, " {.tabset} \n\n")
    lapply(names(markerPairsList[[x]]), function(y) {
      cat("#####", y, " {.tabset} \n\n")
      print(markerPairsList[[x]][[y]][[1]])
      if (length(markerPairsList[[x]][[y]]) == 2)
        print(markerPairsList[[x]][[y]][[2]])
      cat("\n\n")
    })
    cat("\n\n")
  })
  cat("\n\n")
}
```

```{r, label = "save marker pair scatterplots", results = "hide", include = F}
if (gimmePDFs) {
  lapply(names(markerPairsList), function(x) {
    lapply(names(markerPairsList[[x]]), function(y) {
      pdf(file = file.path(pdfDir, paste0(y, "_scatter.pdf")), width = 5, height = 5, bg = "transparent")
      print(markerPairsList[[x]][[y]][[1]])
      if (length(markerPairsList[[x]][[y]]) == 2)
        print(markerPairsList[[x]][[y]][[2]])
      dev.off()
    })
  })
}
```

### Heatmap {.tabset}

#### Marker Per Cluster {.tabset}

```{r, label = "type marker heatmap", results = 'asis', fig.height=ceiling(length(numbersOfClusters)/4), fig.width=ceiling(length(names(markers_to_use_and_their_type))/3.5)}
heatmapsList <- list()

cl <- coloursList[[paste0(mergeBy, knn)]]

if("type" %in% as.character(markers_to_use_and_their_type) & "state" %in% as.character(markers_to_use_and_their_type)) {
  cat("##### Type Markers Only \n\n")
  pET <- MARMOT::plotExprHeatmap(
    x = sce, 
    assay = exprsToUse, 
    features = "type", 
    by = "cluster_id",
    k = clusteringMethodToUse, 
    bars = TRUE,
    perc = TRUE, 
    col_clust = FALSE, 
    hm_pal = hmPalToUse,
    k_pal = cl
    )
  pET
  heatmapsList[["Type"]] <- pET
  print(pET)
  pETdf <- pET@matrix[row_order(pET),]
  topLineageTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(pETdf))
  )
  topLineageTable$Top_Lineage_Markers <- NA
  for (i in seq_along(1:nrow(topLineageTable))) {
    topLineageTable$Top_Lineage_Markers[i] <- paste(names(head(sort(pETdf[i,] %>% .[. > 0.4], decreasing = T), n = 2)), collapse = " ")
  }
  topLineageTable <- topLineageTable[mixedorder(topLineageTable$Cluster), ]
  print(
    htmltools::tagList(
      DT::datatable(
        data = topLineageTable, 
        rownames = F, 
        colnames = c("Cluster", "Top 2 Type Markers"), 
        filter = "top", 
        width = 800, 
        caption = "Top Lineage Markers have to have mean expression > 0.4 to be included, so you may only see 1 marker")
      )
    )
  cat("\n\n")
  
  cat("##### State Markers Only \n\n")
  pES <- MARMOT::plotExprHeatmap(
    x = sce, 
    features = "state", 
    by = "cluster_id", 
    k = clusteringMethodToUse,
    bars = TRUE, 
    perc = TRUE, 
    col_clust = FALSE, 
    hm_pal = hmPalToUse,
    k_pal = cl
    )
  heatmapsList[["State"]] <- pES
  print(pES)
  pESdf <- pES@matrix[row_order(pES),]
  topStateTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(pESdf))
  )
  topStateTable$Top_State_Markers <- NA
  for (i in seq_along(1:nrow(topStateTable))) {
    topStateTable$Top_State_Markers[i] <- paste(names(head(sort(pESdf[i,], decreasing = T), n = 3)), collapse = " ")
    topStateTable$Bottom_State_Markers[i] <- paste(names(tail(sort(pESdf[i,], decreasing = T), n = 3)), collapse = " ")
  }
  topStateTable <- topStateTable[mixedorder(topStateTable$Cluster), ]
  print(
    htmltools::tagList(
      DT::datatable(
        data = topStateTable, 
        rownames = F, 
        colnames = c("Cluster", "Top 3 State Markers", "Bottom 3 State Markers"), 
        filter = "top", 
        width = 800
        )
      )
    )
  cat("\n\n")
}
```

```{r, label = "save type/state heatmaps", results = "hide", include = F}
if (gimmePDFs) {
  if (length(heatmapsList) == 2) {
    for (n in names(heatmapsList)) {
      pdf(file = file.path(pdfDir, paste0(n, "_Marker_Heatmap.pdf")), height = ceiling(length(numbersOfClusters)/4.5), width = ceiling(length(names(markers_to_use_and_their_type))/4))
      print(heatmapsList[[n]])
      dev.off()
    }
  }
}
```

##### All Markers

```{r, label = "plot marker expression cluster", fig.height=ceiling(length(numbersOfClusters)/4), fig.width=if((length(names(markers_to_use_and_their_type))/2) < 10){10}else{(ceiling(length(names(markers_to_use_and_their_type))/2.5))}}
exH <- MARMOT::plotExprHeatmap(
  x = sce, 
  features = NULL, 
  by = "cluster_id", 
  k = clusteringMethodToUse, 
  bars = TRUE, 
  perc = TRUE, 
  col_clust = FALSE, 
  hm_pal = hmPalToUse, 
  k_pal = cl
  )
exH
```

```{r, label = "top marker table", include = T}
if("type" %in% as.character(markers_to_use_and_their_type) & "state" %in% as.character(markers_to_use_and_their_type)) { 
  topMarkerTable <- left_join(topLineageTable, topStateTable)
  openxlsx::write.xlsx(as.data.frame(topMarkerTable), file.path(excelDir, "topMarkerTable.xlsx"))
  DT::datatable(
    data = topMarkerTable, 
    rownames = F, 
    filter = "top", 
    width = 800, 
    colnames = c("Cluster ID", "Top 2 Lineage Markers", "Top 3 State Markers", "Bottom 3 State Markers"),
    caption = "Top Lineage Markers have to have mean expression > 0.4 to be included, so you may only see 1 marker")
} else { 
  f <- NULL
  exT <- plotExprHeatmap(x = sce, features = f, by = "cluster_id", k = clusteringMethodToUse, bars = TRUE, perc = TRUE, col_clust = FALSE)
  exTDF <- exT@matrix[row_order(exH),]
  topMarkerTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(exTDF))
  )
  topMarkerTable$Top_Markers <- NA
  for (i in seq_along(1:nrow(topMarkerTable))) {
    topMarkerTable$Top_Markers[i] <- print(paste(names(head(sort(exTDF[i,], decreasing = T), n = 3)), collapse = " "))
  }
  DT::datatable(topMarkerTable, rownames = F, filter = "top", width = 800)
}
```

```{r, label = "save expression cluster", results = "hide"}
if (gimmePDFs) {
  pdf(
    file = file.path(pdfDir, "marker_per_cluster_heatmap.pdf"), 
    width = if((length(names(markers_to_use_and_their_type))/2) < 10){10}else{(ceiling(length(names(markers_to_use_and_their_type))/3))}, 
    height = ceiling(length(numbersOfClusters)/4.5), bg = "transparent"
    )
  draw(
    MARMOT::plotExprHeatmap(
      x = sce, 
      features = NULL, 
      by = "cluster_id", 
      k = clusteringMethodToUse, 
      bars = TRUE, 
      perc = TRUE, 
      col_clust = FALSE, 
      hm_pal = hmPalToUse, 
      k_pal = cl
      ),
    merge_legend = TRUE
    )
  dev.off()
}
```

#### Sample Per Cluster {.tabset}

##### All Clusters

```{r, label = "sample per cluster heatmap", fig.height=ceiling(length(numbersOfClusters)/3.5), fig.width=(ceiling(length(md$sample_id)/1.2))}
pFH <- MARMOT::plotFreqHeatmap(
  x = sce,
  k = clusteringMethodToUse, 
  normalize = TRUE, 
  col_clust = FALSE, 
  bars = TRUE, 
  perc = TRUE, 
  hm_pal = hmPalToUse,
  k_pal = cl
  )
for (con in conditions) {
  pFH@top_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pFH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pFH@top_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pFH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pFH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pFH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pFH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pFH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
}
draw(pFH, merge_legend = TRUE)
```

```{r, label = "save cluster per sample", results = "hide"}
if (gimmePDFs) {
  pdf(
    file = file.path(pdfDir, "sample_per_cluster_heatmap.pdf"), 
    height = ceiling(length(numbersOfClusters)/4),
    width = (ceiling(length(md$sample_id)/1.6)),
    bg = "transparent"
  )
  print(pFH)
  dev.off()
}
```

```{r, label = "function for custom heatmap"}
x <- sce
by <- "cluster_id"
k <- clusteringMethodToUse
assay = "exprs"
fun = "median"
scale = "first"
q = 0.01
features = NULL
hm_pal = rev(brewer.pal(11, "RdYlBu"))
left_anno = TRUE
row_clust = TRUE
row_anno = TRUE 
col_anno = TRUE 
row_clust = TRUE 
col_clust = TRUE
row_dend = TRUE
col_dend = TRUE
linkage = c("average")

x <- x[unique(CATALYST:::.get_features(x, features)), ]
if (by != "sample_id") {
  CATALYST:::.check_k(x, k)
  x$cluster_id <- cluster_ids(x, k)
}
if (by == "both") {
  by <- c("cluster_id", "sample_id")
}
.do_agg <- function() {
  z <- CATALYST:::.agg(x, by, fun, assay)
  if (length(by) == 1) {
    return(z)
  }
  set_rownames(do.call("rbind", z), levels(x$cluster_id))
}
.do_scale <- function() {
  if (scale == "first") {
    z <- assay(x, assay)
    z <- CATALYST:::.scale_exprs(z, 1, q)
    assay(x, assay, FALSE) <- z
    return(x)
  } else {
    CATALYST:::.scale_exprs(z, 1, q)
  }
}
z <- switch(scale,
            first = {
              x <- .do_scale()
              .do_agg()
            },
            last = {
              z <- .do_agg()
              .do_scale()
            },
            never = {
              .do_agg()
            }
)
if (length(by) == 1) {
  z <- t(z)
}
if (scale != "never" && !(assay == "counts" && fun == "sum")) {
  qs <- round(quantile(z, c(0.01, 0.99)) * 5) / 5
  lgd_aes <- list(at = seq(qs[1], qs[2], 0.2))
} else {
  lgd_aes <- list()
}
lgd_aes$title_gp <- gpar(
  fontsize = 10, fontface = "bold",
  lineheight = 0.8
)
```

##### Selected Clusters {.tabset}

```{r, label = "heatmap by clusters", fig.height = 4, fig.width = 8, results = 'asis', message = F}
heatmapClustersList <- list()
for (i in seq_along(selectedClustersList)) {
  if (length(selectedClustersList[[i]]) >= 2) {
    
    cat("######", names(selectedClustersList)[i], "\n\n")
    cl <- selectedClustersList[[i]]
    x1 <- filterSCE(x, cluster_id %in% cl, k = clusteringMethodToUse)
    x1$cluster_id <- droplevels(x1$cluster_id)
    
    ns <- table(x[[by[1]]])
    ns2 <- ns[which(names(ns) %in% cl)]
    fq <- round(ns/sum(ns) * 100, 2)
    txt <- sprintf("%s%%(%s)", fq, names(fq))
    txt <- txt[which((txt %>% gsub(".*\\(|\\).*", "" , .)) %in% names(ns2))]
    foo <- row_anno_text(txt, just = "center", gp = gpar(fontsize = 8), location = unit(0.5, "npc"))
    right_anno <- rowAnnotation(n_cells = row_anno_barplot(x = as.matrix(ns2), width = unit(2, "cm"), gp = gpar(fill = "grey", col = "white"), border = FALSE, axis = TRUE, bar_width = 0.8), foo = foo)
    
    left_anno <- CATALYST:::.anno_clusters(x1, "cluster_id", NULL, coloursList[[names(selectedClustersList)[[i]]]], NULL)
    
    z1 <- z[selectedClustersList[[i]], ]
    h1 <- Heatmap(
      matrix = z1, 
      name = paste0("median scaled\nexpression"),
      col = hmPalToUse,
      cluster_rows = TRUE,
      cluster_columns = FALSE,
      show_row_dend = TRUE,
      show_column_dend = TRUE,
      show_row_names = (is.null(left_anno) || isTRUE(by == "sample_id")) && !perc, row_names_side = ifelse(by[1] == "cluster_id" || isFALSE(row_anno) && !row_dend || isFALSE(row_clust), "left", "right"),
      left_annotation = left_anno,
      right_annotation = right_anno,
      rect_gp = gpar(col = "white"),
      heatmap_legend_param = lgd_aes
    )
    print(h1)
    heatmapClustersList[[i]] <- h1
    cx <- names(selectedClustersList)[[i]] %>% gsub(": .*", "", .) %>% gsub("Contrast ", "Contrast_", .) %>% gsub("\\/", ".", .)
    sx <- names(selectedClustersList)[[i]] %>% gsub(".*\\: ", "", .) %>% gsub(" ", "_", .) %>% gsub("\\/", ".", .)
    if (gimmePDFs) {
      pdf(file = paste0(pdfDir, "/", cx, "_", sx, "_heatmaps.pdf"), width = 10, height = ceiling(length(cl)*0.5)+1)
      print(h1)
      dev.off()
    }
    cat("\n\n")
  } else {
    cat("######", names(selectedClustersList)[i], "\n\n")
    cat("There is only one (or zero) cluster(s) in this group, and the heatmaps currently do not support a single cluster. Sorry!")
    cat("\n\n")
  }
}
```

```{r, label = "Prepare Seurat object for heatmap", eval = F}
#### Per-Cell Heatmap
scData <- suppressWarnings(Seurat::as.Seurat(x = sce, counts = "exprsTransformed", data = "exprsQuantNorm"))
scData <- Seurat::ScaleData(scData, assay = "originalexp", verbose = FALSE)
Seurat::Idents(scData) <- "cluster_id"
Seurat::DoHeatmap(
  object = scData,
  assay = "originalexp",
  slot = "scale.data",
  features = (names(markers_to_use_and_their_type) %>% gsub("_", "-", .)),
  group.by = "cluster_id",
  group.colors = coloursList[[paste0(mergeBy, knn)]],
  size = 4,
  vjust = 0.1,
  group.bar = TRUE
)
```

### Ridge Plot

```{r, label = "ridge plot", fig.height=ceiling(length(names(markers_to_use_and_their_type))/3)*2.5, fig.width=16}
pCE <- plotClusterExprs(sce, k = clusteringMethodToUse, features = NULL)
pCE$facet$params$ncol <- 12
pCE$facet$params$nrow <- ceiling(length(names(markers_to_use_and_their_type))/12)
pCE + theme_prism()
```

```{r, label = "save ridge plot", include = F}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Ridge_Plot.pdf"), width = 16, height = ceiling(length(names(markers_to_use_and_their_type))/3)*2.5)
  print(pCE + theme_prism())
  dev.off()
}
```

### `r paste(dimRedMethodToUse)` {.tabset}

The markers used for `r paste(dimRedMethodToUse)` generation were:

```{r, label = "show features to DR"}
paste(featuresToUmap)
```

#### Full `r paste(dimRedMethodToUse)` {.tabset}

```{r, label = "DR all cells", fig.height=6, fig.width=7, results = 'asis'}
if (!is.null(drCellPerCondition)) {
  cat("##### All Cells", dimRedMethodToUse, "\n\n")
  cat("Since you clustered on all available cells (after initial downsampling), here is a DR plot of all available cells. All subsequent DR plots are subset from this main DR plot.\n\n")
  dr0 <- plotUMAPNice(df = umapDFList$All, colour_by = "cluster_id", show_density = FALSE, size = 0.2, alpha = 0.9, show_cluster_labels = TRUE, show_grids = F, text_size = 14) + 
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
  dr0
}
```

```{r, label = "save DF all cells", include = F}
if (!is.null(drCellPerCondition)) {
  if (gimmePDFs) {
    pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_All_Cells.pdf")), width = (3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, height = 5)
    print(dr0)
    dev.off()
  }
}
```

```{r, label = "DR all cells unlabelled", fig.height=6, fig.width=(3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, results = 'asis'}
if (!is.null(drCellPerCondition)) {
  cat("\n\n")
  dr0a <- plotUMAPNice(
    df = umapDFList$All, 
    colour_by = "cluster_id", 
    show_density = FALSE, 
    size = 0.2, 
    alpha = 0.9, 
    show_cluster_labels = FALSE, 
    show_grids = F, 
    text_size = 14
    ) + 
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
  dr0a
  cat("\n\n")
}
```

```{r, label = "save DF all cells unlabelled", include = F}
if (!is.null(drCellPerCondition)) {
  if (gimmePDFs) {
    pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_All_Cells_unlabelled.pdf")), width = (3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, height = 5)
    print(dr0a)
    dev.off()
  }
}
```

##### All Samples: Selected Clusters

```{r, label = "main cluster DR 1", fig.height=6, fig.width=7}
if (is.null(drCellPerCondition)) {
  umapsDFToPlot <- "All"
} else {
  umapsDFToPlot <- "Downsampled"
}
openxlsx::write.xlsx(x = as.data.frame(umapDFList[[umapsDFToPlot]]), file = file.path(excelDir, paste0(dimRedMethodToUse, "_Coordinates.xlsx")))
dr1a <- plotUMAPNice(
  df = as.data.frame(umapDFList[[umapsDFToPlot]]),
  colour_by = "cluster_id", 
  show_density = FALSE, 
  size = 0.2, 
  alpha = 0.9, 
  show_cluster_labels = TRUE, 
  show_grids = F, 
  text_size = 14
  ) + 
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr1a
```

```{r, label = "main cluster DR 2", fig.height=6, fig.width=7}
dr1b <- plotUMAPNice(
  df = as.data.frame(umapDFList[[umapsDFToPlot]]), 
  colour_by = "cluster_id", 
  show_density = FALSE, 
  size = 0.2, 
  alpha = 0.8, 
  show_cluster_labels = FALSE, 
  show_grids = F, 
  text_size = 14
  ) + 
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr1b
```

```{r, label =  "save full DR", results = "hide"}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_full_labelled.pdf")), width = 7, height = 6, bg = "transparent")
  print(dr1a)
  dev.off()
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_full_unlabelled.pdf")), width = 7, height = 6, bg = "transparent")
  print(dr1b)
  dev.off()
}
```

##### All Samples: Other k values

```{r, label = "plot k DRs", fig.height=ceiling(6*length(kValuesIWant)), fig.width=8}
pDRList <- setNames(lapply(kValuesIWant, function(i) {
  plotUMAPNice(df = umapDFList[[umapsDFToPlot]], colour_by = paste0(mergeBy, i), show_grids = F, size = 0.4)  + 
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
}), paste0(mergeBy, kValuesIWant))
plot_grid(plotlist = pDRList, ncol = 1, align = "hv")
```

#### Condition `r paste(dimRedMethodToUse)`

```{r, label = "make DRs split by condition"}
## Get the number of cells per condition for the UMAP
conds <- as.factor(sce@metadata$experiment_info$condition)
cellN <- n_cells(sce)
names(cellN) <- conds
condLevels <- levels(conds)
labs <- NA
for (i in seq_along(condLevels)) {
  labs[i] <- paste(condLevels[i], "\n n =", as.numeric(nrow(umapDFList[[umapsDFToPlot]][which(umapDFList[[umapsDFToPlot]]$condition == condLevels[i]),])))
}
names(labs) <- condLevels
dr2a <- plotUMAPNice(df = as.data.frame(umapDFList[[umapsDFToPlot]]), colour_by = "cluster_id", show_density = FALSE, size = 0.3, alpha = 0.85, text_size = 14, show_cluster_labels = TRUE, show_grids = F) +
  facet_wrap("condition", labeller = labeller("condition" = labs), ncol = 2) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
```

```{r, label = "plot DRs split by condition", fig.height=ceiling(length(condLevels)/2)*6, fig.width=12}
dr2a
```

```{r, label = "plot DRs split by condition 2", fig.height=ceiling(length(condLevels)/2)*6, fig.width=12}
dr2b <- plotUMAPNice(df = as.data.frame(umapDFList[[umapsDFToPlot]]), colour_by = "cluster_id", show_density = FALSE, size = 0.3, alpha = 0.85, text_size = 14, show_cluster_labels = F, show_grids = F) +
  facet_wrap("condition", labeller = labeller("condition" = labs), ncol = 2) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr2b
```

```{r, label = "save DRs split by condition", results = "hide"}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_condition_labelled.pdf")), width = 12, height = ceiling(length(condLevels)/2)*6, bg = "transparent")
  print(dr2a)
  dev.off()
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_condition_unlabelled.pdf")), width = 12, height = ceiling(length(condLevels)/2)*6, bg = "transparent")
  print(dr2b)
  dev.off()
}
```

```{r, label = "full DR by condition", fig.height=(6+length(conditionOrder)*2.5), fig.width=7}
u1_by_cond <- plotUMAPNice(umapDFList[[umapsDFToPlot]], "condition", size = 0.2)
u2_by_cond <- plotUMAPNice(df = umapDFList[[umapsDFToPlot]], colour_by = "condition", show_density = FALSE, size = 0.3) + 
  facet_wrap("condition", labeller = labeller("condition" = labs))
plot_grid(u1_by_cond, u2_by_cond, ncol = 1, align = "hv", rel_heights = c(1.4, 1))
```

#### `r paste(dimRedMethodToUse)` Per Marker {.tabset}

All values have been normalised to 0 - 1.

```{r, label = "DR per marker 1", results = 'asis', include = F}
UMAPList <- setNames(lapply(c("With", "Without"), function(yn) {
  var <- switch(
    yn,
    "With" = TRUE,
    "Without" = FALSE
  )
  l1 <- setNames(lapply(names(markers_to_use_and_their_type), function(marker) {
    m1 <- plotUMAPNice(
      df = as.data.frame(umapDFList[[umapsDFToPlot]]), 
      colour_by = marker, 
      name = "Normalised\nExpression\n", 
      size = 0.3, 
      show_cluster_labels = var
      )
    m1 <- m1 + theme(plot.title = element_blank(), axis.title = element_blank())
    m1 <- m1 + annotate(geom = "text", x = max(m1$data$x)*0.8, y = max(m1$data$y)*0.95, label = marker, color = "black", size = 6)
    m1
  }), names(markers_to_use_and_their_type))
  l1[["legend"]] <- cowplot::get_legend(l1[[1]])
  for (marker in names(markers_to_use_and_their_type)) {
    l1[[marker]] <- l1[[marker]] + theme(legend.position = "hide")
  }
  return(l1)
}), c("With Labels", "Without Labels"))
```

```{r, label = "DR per per marker 2", fig.width=16, fig.height=ceiling(length(names(markers_to_use_and_their_type))/3)*3, dependson="umaps_per_marker1", results = 'asis'}
lapply(names(UMAPList), function(yn){
  cat("#####", yn, "\n\n")
  print(plot_grid(plotlist = UMAPList[[yn]], ncol = 4, align = "hv"))
  cat("\n\n")
})
```

```{r, label = "save DR per marker", results="hide"}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, paste0("marker_", dimRedMethodToUse)))
  lapply(names(UMAPList), function(m) {
    dir.create(file.path(pdfDir, paste0("marker_", dimRedMethodToUse), m))
    pdf(file = file.path(pdfDir, paste0(m, "_all_markers_", dimRedMethodToUse, ".pdf")), width = 16, height = ceiling(length(names(markers_to_use_and_their_type))/3)*3)
    print(plot_grid(plotlist = UMAPList[[m]], ncol = 4))
    dev.off()
    lapply(names(UMAPList[[m]]), function(n) {
      pdf(file = file.path(pdfDir, paste0("marker_", dimRedMethodToUse), m, paste0(n, "_", dimRedMethodToUse, ".pdf")), width = 4, height = 4, bg = "transparent")
      if (n != "legend") {
        print(UMAPList[[m]][[n]])
      } else if (n == "legend") {
        plot(UMAPList[[m]][[n]])
      }
      dev.off()
    })
  })
}
```

#### `r paste(dimRedMethodToUse)` Per Marker Per Condition {.tabset}

```{r, label = "DR per marker per cond 1", results = 'asis', include = F}
DRperCondList <- setNames(lapply(c("With", "Without"), function(yn) {
  var <- switch(
    yn,
    "With" = TRUE,
    "Without" = FALSE
  )
  l1 <- setNames(lapply(names(markers_to_use_and_their_type), function(marker) {
    m1 <- plotUMAPNice(
      df = as.data.frame(umapDFList[[umapsDFToPlot]]), 
      colour_by = marker, 
      name = "Normalised\nExpression\n", 
      size = 0.3, 
      show_cluster_labels = var
      ) + 
      facet_wrap(~condition, ncol = 2)
    m1 <- m1 + theme(plot.title = element_blank(), axis.title = element_blank(), legend.position = "hide")
    m1
  }), names(markers_to_use_and_their_type))
  return(l1)
}), c("With Labels", "Without Labels"))
```

```{r, label = "DR per marker per cond 2", fig.width=10, fig.height=5*(ceiling(length(condLevels)/2)), results = 'asis'}
lapply(names(DRperCondList), function(yn){
  cat("#####", yn, "{.tabset} \n\n")
  lapply(names(DRperCondList[[yn]]), function(marker) {
    cat("######", marker, "\n\n")
    print(DRperCondList[[yn]][[marker]])
    cat("\n\n")
  })
  cat("\n\n")
})
```

```{r, label = "save DR per marker per cond", results="hide"}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse)))
  lapply(names(DRperCondList), function(m) {
    dir.create(file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse), m))
    lapply(names(DRperCondList[[m]]), function(n) {
      pdf(file = file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse), m, paste0(n, "_", dimRedMethodToUse, ".pdf")), width = 10, height = 5*(ceiling(length(condLevels)/2)), bg = "transparent")
      if (n != "legend") {
        print(DRperCondList[[m]][[n]])
      } else if (n == "legend") {
        plot(DRperCondList[[m]][[n]])
      }
      dev.off()
    })
  })
}
```

#### `r paste(dimRedMethodToUse)` of selected clusters {.tabset}

##### All Cells

```{r, label = "DR of selected clusters", fig.height=5, fig.width=5, results = 'asis', warning = F, message = F}
DRSelectedClusters <- setNames(lapply(seq_along(clustersToMapList), function(i) {
  clustersToMap <- names(clustersToMapList[[i]])
  cTM <- clustersToMapList[[i]]
  umapDF <- umapDFList[[umapsDFToPlot]]
  umapDF2 <- umapDF[umapDF[["cluster_id"]] %in% clustersToMap,]
  umapDF2[["cluster_id"]] <- droplevels(umapDF2[["cluster_id"]])
  umapDF3 <- umapDF[!umapDF[["cluster_id"]] %in% clustersToMap, ]

  ## Get the mean x/y coordinate for each cluster so we can add labels manually 
  cluster_centre <- umapDF2 %>%
  group_by(cluster_id) %>%
  summarize(center_x = mean(x), center_y = mean(y))

  gp <- ggplot(umapDF, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[["cluster_id"]]))
  ## gp <- gp + geom_density_2d(data = umapDF[,c("x","y")], aes(x = x, y = y), colour = "lightgrey", size = 0.5, bins = 20, alpha = 0.8)
  gp <- gp + ylim(min(umapDF$y[!is.na(umapDF$y)])-1, max(umapDF$y[!is.na(umapDF$y)])+1)
  gp <- gp + xlim(min(umapDF$x[!is.na(umapDF$x)])-1, max(umapDF$x[!is.na(umapDF$x)])+1)
  gp <- gp + geom_point(data = umapDF3, size = 0.2, alpha = 0.2, shape = 16)
  gp <- gp + geom_point(data = umapDF2, size = 0.3, alpha = 0.95, shape = 16) 
  gp <- gp + geom_label_repel_quiet(data = cluster_centre, aes(x = center_x, y = center_y, label = cluster_id, fill = cluster_id), size = 4, colour = "black")
  gp <- gp + scale_color_manual(values = c(cTM, na.value = "grey33"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + scale_fill_manual(values = c(cTM, na.value = "grey33"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + labs(x = paste(dimRedMethodToUse, "dim 1"), y = paste(dimRedMethodToUse, "dim 2"))
  gp <- gp + ggtitle(names(clustersToMapList)[i])
  gp <- gp + theme_prism() +
    theme(
      axis.text = element_blank(), 
      axis.ticks = element_blank(), 
      axis.line = element_blank(), 
      legend.text = element_text(size = 10), 
      panel.border = element_rect(colour = "black", fill = NA, size = 1.5))
  gp <- gp + guides(colour = guide_legend(override.aes = list(size = 3)))
  gp <- gp + theme(legend.position = "hide")
  print(gp)
  cx <- names(clustersToMapList)[[i]] %>% gsub(": .*", "", .) %>% gsub("Contrast ", "Contrast_", .) %>% gsub("\\/", "", .)
  sx <- names(clustersToMapList)[[i]] %>% gsub(".*\\: ", "", .) %>% gsub(" ", "_", .) %>% gsub("\\/", "", .)
  return(list("gp" = gp, "cx" = cx, "sx" = sx))
}), names(clustersToMapList))
```

```{r, label = "save DR of selected clusters", include = F}
if (gimmePDFs) {
  lapply(names(DRSelectedClusters), function(x) {
    pdf(file = paste0(pdfDir, "/", DRSelectedClusters[[x]][["cx"]], "_", DRSelectedClusters[[x]][["sx"]], "_", dimRedMethodToUse, "_full.pdf"), width = 5, height = 5)
    print(DRSelectedClusters[[x]][["gp"]])
    dev.off()
  })
}
```

```{r, label = "DR of selected clusters 2", fig.height= 6*(ceiling(length(condLevels)/2)), fig.width=10, results = 'asis', warning = F, message = F}
DRSelectedClusters2 <- setNames(lapply(seq_along(clustersToMapList), function(i) {
  cat("#####" , names(clustersToMapList)[i], "\n\n")
  clustersToMap <- (names(clustersToMapList[[i]]))
  cTM <- clustersToMapList[[i]]
  umapDF <- umapDFList[[umapsDFToPlot]]
  umapDF2 <- umapDF[umapDF[["cluster_id"]] %in% clustersToMap,]
  umapDF2[["cluster_id"]] <- droplevels(umapDF2[["cluster_id"]])
  umapDF3 <- umapDF[!umapDF[["cluster_id"]] %in% clustersToMap, ]

  ## Get the mean x/y coordinate for each cluster so we can add labels manually 
  cluster_centre <- umapDF2 %>%
  group_by(cluster_id) %>%
  summarize(center_x = mean(x), center_y = mean(y))
  
  gp <- ggplot(umapDF, aes(x = .data[["x"]], .data[["y"]], col = .data[["cluster_id"]]))
  ## gp <- gp + geom_density_2d(data = umapDF[,c("x","y")], aes(x = x, y = y), colour = "lightgrey", size = 0.5, bins = 30, alpha = 0.8)
  gp <- gp + geom_point(data = umapDF3, size = 0.5, alpha = 0.25, shape = 16)
  gp <- gp + geom_point(data = umapDF2, size = 0.5, alpha = 0.95, shape = 16)
  gp <- gp + geom_label_repel_quiet(data = cluster_centre, aes(x = center_x, y = center_y, label = cluster_id, fill = cluster_id), size = 4, show.legend = F, colour = "black")
  gp <- gp + scale_color_manual(values = c(cTM, na.value = "black"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + scale_fill_manual(values = c(cTM, na.value = "black"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + ylim(min(umapDF$y[!is.na(umapDF$y)])-1, max(umapDF$y[!is.na(umapDF$y)])+1)
  gp <- gp + xlim(min(umapDF$x[!is.na(umapDF$x)])-1, max(umapDF$x[!is.na(umapDF$x)])+1)
  gp <- gp + labs(x = paste(dimRedMethodToUse, "dim 1"), y = paste(dimRedMethodToUse, "dim 2"))
  gp <- gp + facet_wrap(~ condition, ncol = 2)
  gp <- gp + ggtitle(names(clustersToMapList)[i])
  gp <- gp + theme_prism() +
    theme(
      axis.text = element_blank(), 
      axis.ticks = element_blank(), 
      axis.line = element_blank(), 
      legend.text = element_text(size = 10), 
      panel.border = element_rect(colour = "black", fill = NA, size = 1.5),
      legend.position = "hide")
  gp <- gp + guides(colour = guide_legend(override.aes = list(size = 3)))
  print(gp)
  cat("\n\n")
  return(gp)
}), names(clustersToMapList))
```

```{r, label = "save DR of selected clusters 2", include = F}
if (gimmePDFs) {
  lapply(names(DRSelectedClusters2), function(x) {
    pdf(file = paste0(pdfDir, "/", x, "_", dimRedMethodToUse, "_subset.pdf"), width = 10, height =  6*(ceiling(length(condLevels)/2)))
    print(DRSelectedClusters2[[x]])
    dev.off()
  })
}
```

```{r label = "print gated DRs", fig.height=6, fig.width=18, results = 'asis'}
if (runScGate) {
  cat("#### Labelled Cells {.tabset}\n\n")
  cat("\n\n##### All \n\n")
  
  # Prepare the marker pairs and colours 
  markerPairs <- smd$`Marker Pairs`[!is.na(smd$`Marker Pairs`)]
  markerPairTypes <- markerPairs %>% gsub("\\:.*", "", .) %>% unique
  markerPairs2 <- markerPairs
  names(markerPairs2) <- markerPairTypes
  models <- lapply(markerPairTypes, function(x) {
    markerPairs2[x][[1]] %>% strsplit(" ") %>% .[[1]] %>% .[2:3]
  }) %>% setNames(markerPairTypes)
  
  print(
    plotUMAPNice(
      df = umapDFList$Downsampled.UMAP, 
      colour_by = "Gated_Cells", 
      show_density = FALSE, 
      size = 0.2, 
      alpha = 1, 
      facet_by = "condition", 
      show_cluster_labels = TRUE, 
      cluster_labels = "Gated_Cells"
    )
  )
  
  cat("\n\n")
  lapply(names(models), function(mpt) {
    cat("#####", mpt, "\n\n")
    y <- mpt %>% gsub("\\ |\\-", "", .)
    if (!models[[mpt]][[1]] %in% rownames(scData)) {
      models[[mpt]][[1]] <- agrep(models[[mpt]][[1]], rownames(scData), value = TRUE)
    }
    if (!models[[mpt]][[2]] %in% rownames(scData)) {
      models[[mpt]][[2]] <- agrep(models[[mpt]][[2]], rownames(scData), value = TRUE)
    }
    print(
      cowplot::plot_grid(
        SCpubr::do_DimPlot(
          sample = scData, 
          group.by = paste0("is_", y), 
          reduction = dimRedMethodToUse,
          legend.position = "none", 
          label = T, pt.size = 0.2) + 
          ggtitle(paste0("Are ", mpt, "?")) +
            theme(panel.border = element_rect(color = "black", fill = NA, size = 1)),
        SCpubr::do_FeaturePlot(
          sample = scData, 
          features = paste0(mpt, "_UCell"), 
          reduction = dimRedMethodToUse, 
          legend.position = "none", 
          use_viridis = T, 
          viridis.palette = "mako", 
          viridis.direction = 1, 
          pt.size = 0.2) + 
          ggtitle(paste0(mpt, " UCell Score")) +
          theme(panel.border = element_rect(color = "black", fill = NA, size = 1)),
        SCpubr::do_NebulosaPlot(
          sample = scData,
          reduction = dimRedMethodToUse, 
          features = c(models[[mpt]][[1]], models[[mpt]][[2]]),
          joint = T, 
          return_only_joint = T, 
          use_viridis = T, 
          viridis.palette = "mako",
          viridis.direction = 1, 
          legend.position = "none", 
          pt.size = 0.2) +
          theme(panel.border = element_rect(color = "black", fill = NA, size = 1)),
        align = "hv", ncol = 3
      )
    )
    cat("\n\n")
  })
  cat("\n\n")
}
```


### Barplots {.tabset}

#### Cells per Cluster

```{r, label = "cells per cluster", fig.height=4, fig.width=8}
ns <- table(cluster_id = cluster_ids(sce, clusteringMethodToUse), sample_id = sample_ids(sce))
df <- as.data.frame(ns)
df$condition <- md$condition[match(df$sample_id, md$sample_id)]
df$cluster_id <- factor(df$cluster_id, levels = levels(sce$cluster_id))
cellsPerClusterBarplot <- ggplot(df, aes(x = cluster_id, y = Freq, fill = cluster_id)) + 
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = coloursList[[paste0(mergeBy, knn)]]) + 
  theme_bw() +
  ylab("Number of cells") + 
  xlab("Cluster") + 
  theme_prism(axis_text_angle = 90) +
  theme(legend.position = "hide", panel.grid.major.x = element_blank())
cellsPerClusterBarplot
```

```{r, label = "cells per cluster facet", fig.height = ceiling(length(condLevels)/2)*4, fig.width = 16}
cellsPerClusterBarplot + facet_wrap(~condition, scales = "free") + theme_prism(axis_text_angle = 90) + theme(legend.position = "hide")
```

```{r, label = "cells per cluster facet sample", fig.height=(4*(ceiling(length(unique(df$sample_id)))/3)), fig.width = 18}
cellsPerClusterBarplot + facet_wrap(~sample_id, ncol = 3, scales = "free")
```

```{r, label = "marker frequency per cluster", fig.height=6, include = F}
x1 <- as.data.frame(t(assay(sce, "exprsQuantNorm")))
x1$cluster_id <- cluster_ids(sce, clusteringMethodToUse)

x1mean <- data.frame("cluster_id" = levels(as.factor(x1$cluster_id)))
for (m in names(markers_to_use_and_their_type)) {
  x1mean[[m]] <- NA
  for (cl in levels(as.factor(x1$cluster_id))) {
    x1mean[cl,m] <- mean(x1[[m]][which(x1$cluster_id == cl)])
  }
}
x1rf <- x1mean
for (m in names(markers_to_use_and_their_type)) {
  for (cl in levels(as.factor(x1$cluster_id))) {
    x1rf[cl,m] <- x1mean[cl,m]/sum(x1mean[cl,-1])
  }
}
x1meanmelt <- melt(x1mean)
x1meanmelt$cluster_id <- factor(x1meanmelt$cluster_id, levels = unique(x1meanmelt$cluster_id))
x1rfmelt <- melt(x1rf)
x1rfmelt$cluster_id <- factor(x1rfmelt$cluster_id, levels = unique(x1rfmelt$cluster_id))
ggplot(x1rfmelt, aes(x = cluster_id, y = value, fill = variable)) + 
  geom_bar(stat = "identity") + 
  theme_prism() + 
  scale_fill_manual(values = coloursList[[paste0(mergeBy, knn)]]) +
  xlab("Cluster ID") +
  ylab("Proportion of the Quantile Scaled Mean")
```

```{r, label = "cells per cluster per sample table"}
DT::datatable(spread(df, cluster_id, Freq) %>% dplyr::select(-condition) %>% column_to_rownames("sample_id") %>% t %>% data.frame %>% rownames_to_column("cluster_id"), rownames = F)
openxlsx::write.xlsx(
  x = (spread(df, cluster_id, Freq) %>% dplyr::select(-condition) %>% column_to_rownames("sample_id") %>% t %>% data.frame %>% rownames_to_column("cluster_id")), 
  file = file.path(excelDir, "Cells_Per_Cluster_Per_Sample.xlsx")
)
```

#### Cluster Frequency

```{r, label = "Cluster barplot 1", fig.width = ceiling(length(condLevels)/2)*10, fig.height=5}
plotAbundances(sce, k = clusteringMethodToUse, by = "sample_id", k_pal = coloursList[[paste0(mergeBy, knn)]]) + facet_grid(~condition, space = "free", scales="free") + theme_prism(axis_text_angle = 45)
```

```{r, label = "save cluster barplot", results = "hide"}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "cluster_barplot.pdf"), width = ceiling(length(condLevels)/2)*10, height = 5, bg = "transparent")
  print(plotAbundances(sce, k = clusteringMethodToUse, by = "sample_id", k_pal = coloursList[[paste0(mergeBy, knn)]]) + facet_grid(~condition, space = "free", scales="free") + theme_prism())
  dev.off()
}
```

```{r, label = "Cluster barplot 2"}
FDR_cutoff <- 0.05

ns <- table(cluster_id = cluster_ids(sce, clusteringMethodToUse), sample_id = sample_ids(sce))
fq <- prop.table(ns, 2) * 100
df <- as.data.frame(fq)

m <- match(df$sample_id, sce$sample_id)
for (i in c(NULL, "condition")) df[[i]] <- sce[[i]][m]

dfList <- list()
for (i in seq_along(clustersToMapList)) {
  dfC1 <- df[df$cluster_id %in% names(clustersToMapList[[i]]), ]
  dfList[[names(selectedClustersList)[[i]]]] <- dfC1
}
```

```{r, label = "Cluster barplot 3", fig.width=10, fig.height=5, results = 'asis'}
selectedBarplotList <- setNames(lapply(seq_along(dfList), function(i) {
  cat("\n\n#### DA:", names(dfList)[i], "\n\n")
  if (nrow(dfList[[i]]) >= 1) {
    g <- ggplot(dfList[[i]], aes(x = .data[["sample_id"]], y = .data[["Freq"]], fill = .data[["cluster_id"]]))
    g <- g + geom_bar(stat = "identity")
    g <- g + facet_grid(~condition, scales = "free", space = "free")
    g <- g + scale_fill_manual(values = coloursList[[names(selectedClustersList)[[i]]]])
    g <- g + ggtitle(names(dfList)[i])
    g <- g + scale_x_discrete(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0), labels = seq(0, 100, 25), limits = c(0, 100))
    g <- g + theme_prism()
    g <- g + theme(panel.border = element_blank(), panel.spacing.x = unit(1, "lines"))
    g <- g + ylab("Proportion [%]") + xlab("Sample ID")
    print(g)
    cx <- names(dfList)[[i]] %>% gsub(": .*", "", .) %>% gsub("Contrast ", "Contrast_", .) %>% gsub("\\/", "", .)
    sx <- names(dfList)[[i]] %>% gsub(".*\\: ", "", .) %>% gsub(" ", "_", .) %>% gsub("\\/", "", .)
    return(list("g" = g, "cx" = cx, "sx" = sx))
  } else {
    print("There's no DA clusters in this group.")
  }
  cat("\n\n")
}), names(dfList))
```

```{r, label = "Save Cluster Barplots 3", include = F}
if (gimmePDFs) {
  lapply(names(selectedBarplotList), function(x) {
    pdf(file = paste0(pdfDir, "/", selectedBarplotList[[x]][["cx"]], "_", selectedBarplotList[[x]][["sx"]], "_barplot.pdf"), width = 10, height = 5)
    print(selectedBarplotList[[x]][["g"]])
    dev.off()
  })
}
```

### Cluster Boxplots {.tabset}

```{r, label = "cluster boxplot dimensions", include = F}
nColsToPlot <- 6
nGroupsToPlot <- nlevels(sce$condition)
nClustersToPlot <- nlevels(sce$cluster_id)
nRowsToPlot <- ceiling(nClustersToPlot/nColsToPlot)
heightToPlot <- ceiling(nRowsToPlot*5)
widthToPlot <- ceiling((5*1.2)*nColsToPlot)
```

```{r, label = "cluster boxplot", fig.width = widthToPlot, fig.height = heightToPlot}
## From CATALYST plotFreqHeatmap function
ns <- table(sce$cluster_id, sce$sample_id)
fq <- prop.table(ns, 2)
fq.z <- as.matrix(unclass(fq))
clusterBoxPlotsDF <- CATALYST:::.z_normalize(asin(sqrt(fq.z)))
clusterBoxPlotsDF <- as.data.frame(t(clusterBoxPlotsDF))
clusterBoxPlotsDF <- rownames_to_column(clusterBoxPlotsDF, "sample_id") %>% left_join(md)
clusterBoxPlotsDFMelt <- melt(clusterBoxPlotsDF)

ctsList <- lapply(names(daList), function(da) { da %>% strsplit(" over ") %>% unlist}) %>% setNames(names(daList))

clusterPvalues <- lapply(levels(kids), function(k) {
  pvals <- lapply(names(daList), function(da) { daList[[da]]$p_adj[daList[[da]]$cluster_id == k] }) %>% setNames(names(daList)) %>% unlist
  if(any(is.na(pvals))) {pvals[is.na(pvals)] <- 1}
  y_max <- max(clusterBoxPlotsDFMelt$value)
  spacing <- (y_max - min(clusterBoxPlotsDFMelt$value)) * 0.2
  y_positions <- y_max + (1:length(ctsList)) * spacing
  pval_df <- data.frame(
    group1 = sapply(ctsList, `[`, 1),
    group2 = sapply(ctsList, `[`, 2),
    p = pvals,
    y.position = y_positions
  )
  pval_df <- add_significance(pval_df, p.col = "p", output.col = "p.signif")
  
  pval_df
}) %>% setNames(levels(kids))
maxY <- lapply(seq_along(clusterPvalues), function(i) {clusterPvalues[[i]]$y.position}) %>% unlist %>% max

clusterBoxplotList <- lapply(levels(kids), function(k) {
  ctp <- as.character(k)
  ggplot(clusterBoxPlotsDF, aes(x = .data[["condition"]], y = .data[[ctp]])) + 
    ## geom_vline(xintercept = (seq_along(condLevels)+0.5)[-length(condLevels)], linetype = "dashed", alpha = 0.6) +
    scale_y_continuous(limits = c(min(clusterBoxPlotsDFMelt$value), maxY)) +
    geom_boxplot(aes(fill = condition), alpha = 0.2, show.legend = F, outlier.shape = NA) +
    geom_beeswarm(aes(fill = condition), pch = 21, size = 3, stroke = 0.3, cex = 2, corral = "random", corral.width = 0.9) +
    scale_fill_manual(values = coloursList$condition) +
    stat_pvalue_manual(clusterPvalues[[ctp]], label = "p.signif", tip.length = 0, hide.ns = T, label.size = 9, bracket.size = 0.8, bracket.shorten = 0.1, bracket.nudge.y = -0.1) +
    theme_prism(axis_text_angle = 45) +
    ggtitle(paste("Cluster", ctp)) +
    theme(legend.position = "hide", axis.title = element_blank())
})
ylab <- textGrob("Normalised Frequency", rot = 90, gp = gpar(fontface = "bold", fontsize = 16))
xlab <- textGrob("Condition", gp = gpar(fontface = "bold", fontsize = 16))
grid.arrange(grobs = clusterBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
```

```{r, label = "save cluster boxplot", include = F}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Cluster_Boxplot.pdf"), width = widthToPlot, height = heightToPlot)
  grid.arrange(grobs = clusterBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
  dev.off()
}
```

### Markers Boxplots {.tabset}

#### Markers per Condition

```{r, label = "generate marker tables"}
medianScaledDF <- CATALYST:::.agg(sce, "sample_id", "median", exprsToUse)
markerBoxplotsMeltDF <- melt(medianScaledDF, varnames = c("antigen", "sample_id"[length("sample_id")]))
i <- match(markerBoxplotsMeltDF$sample_id, sce$sample_id)
j <- setdiff(names(colData(sce)), c(names(markerBoxplotsMeltDF), "cluster_id"))
markerBoxplotsMeltDF <- cbind(markerBoxplotsMeltDF, colData(sce)[i, j, drop = FALSE])
medianScaledDF <- medianScaledDF %>%
  t %>% 
  data.frame(check.names = F) %>%
  rownames_to_column("sample_id") %>%
  left_join(md, by = "sample_id")
openxlsx::write.xlsx(as.data.frame(medianScaledDF), file = file.path(excelDir, "Median_Markers_Per_Sample_Expression.xlsx"))
```

```{r, label = "draw marker boxplots", fig.height = heightToPlot*0.7, fig.width = widthToPlot}
markerPvalues <- lapply(names(markers_to_use_and_their_type), function(marker) {
  pvals <- lapply(names(dsList)[grep("Merged", names(dsList))], function(ds) { dsList[[ds]]$tbl_DS$p_adj[dsList[[ds]]$tbl_DS$marker_id == marker] }) %>% setNames(names(dsList)[grep("Merged", names(dsList))]) %>% unlist
  if(length(pvals) == 0) {pvals <- 1}
  y_max <- max(markerBoxplotsMeltDF$value)
  spacing <- (y_max - min(markerBoxplotsMeltDF$value)) * 0.1
  y_positions <- y_max + (1:length(ctsList)) * spacing
  pval_df <- data.frame(
    group1 = sapply(ctsList, `[`, 1),
    group2 = sapply(ctsList, `[`, 2),
    p = pvals,
    y.position = y_positions
  )
  pval_df <- add_significance(pval_df, p.col = "p", output.col = "p.signif")
  pval_df
}) %>% setNames(names(markers_to_use_and_their_type))
maxY <- lapply(seq_along(markerPvalues), function(i) {markerPvalues[[i]]$y.position}) %>% unlist %>% max

markerBoxplotList <- lapply(names(markers_to_use_and_their_type), function(marker) {
  ggplot(medianScaledDF, aes(x = .data[["condition"]], y = .data[[marker]])) + 
    ## geom_vline(xintercept = (seq_along(condLevels)+0.5)[-length(condLevels)], linetype = "dashed", alpha = 0.6) +
    scale_y_continuous(limits = c(min(markerBoxplotsMeltDF$value), maxY)) +
    geom_boxplot(aes(fill = condition), alpha = 0.2, show.legend = F, outlier.shape = NA) +
    geom_beeswarm(aes(fill = condition), pch = 21, size = 3, stroke = 0.3, cex = 2, corral = "random", corral.width = 0.9) +
    scale_fill_manual(values = coloursList$condition) +
    stat_pvalue_manual(markerPvalues[[marker]], label = "p.signif", tip.length = 0, hide.ns = T, label.size = 9, bracket.size = 0.8, bracket.shorten = 0.1, bracket.nudge.y = -0.1) +
    theme_prism(axis_text_angle = 45) +
    ggtitle(marker) +
    theme(legend.position = "hide", axis.title = element_blank())
})
ylabtext <- switch(exprsToUse, "exprsTransformed" = "Transformed", "exprsQuantNorm" = "Quantile Normalised")
ylab <- textGrob(paste0("Median ", ylabtext, " Expression"), rot = 90, gp = gpar(fontface = "bold", fontsize = 16))
xlab <- textGrob("Condition", gp = gpar(fontface = "bold", fontsize = 16))
grid.arrange(grobs = markerBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
```

```{r, label = "save marker boxplots", include = F}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Marker_per_Sample_boxplot.pdf"), width = widthToPlot, height = heightToPlot)
  grid.arrange(grobs = markerBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
  dev.off()
}
```

#### Markers within Clusters {.tabset}

```{r, label = "markers per cluster boxplot", fig.width=18, fig.height=5, results = 'asis'}
by <- c("cluster_id", "sample_id")
ms <- CATALYST:::.agg(sce, by, "median", "exprsQuantNorm")

df <- melt(ms, varnames = c("antigen", by[length(by)]))
names(df)[ncol(df)] <- "cluster_id"
i <- match(df$sample_id, sce$sample_id)
j <- setdiff(names(colData(sce)), c(names(df), "cluster_id"))
df <- cbind(df, colData(sce)[i, j, drop = FALSE])
cat("\n\n")
openxlsx::write.xlsx(as.data.frame(df), file = file.path(excelDir, "Median_Markers_Per_Cluster_Expression.xlsx"))
individualClusterMarkerBoxplots <- setNames(lapply(levels(markerBoxplotsMeltDF[[paste0(mergeBy, knn)]]), function(i) {
  cat("##### Cluster", i, "\n\n")
  df1 <- df[df$cluster_id == as.character(i),]
  g1 <- ggplot(df1, aes(x = antigen, y = value, fill = condition)) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0.25, 0.5, 0.75, 1)) +
    geom_hline(yintercept = c(0.25, 0.5, 0.75), linetype = "dashed", alpha = 0.3) + 
    geom_vline(xintercept = (seq_along(names(markers_to_use_and_their_type))+0.5)[-length(names(markers_to_use_and_their_type))], linetype = "dotted", alpha = 0.3) +
    geom_boxplot(alpha = 0.2, show.legend = F, outlier.shape = NA) +
    geom_point(position=position_jitterdodge(), pch = 21, size = 2, stroke = 0.2) +
    scale_fill_manual(values = coloursList$condition) +
    theme_prism(axis_text_angle = 45) +
    ylab("Median Quantile Normalised Expression") + xlab("Condition") + ggtitle(paste("Cluster", i))
  print(g1)
  cat("\n\n")
  return("g1" = g1)
}), levels(markerBoxplotsMeltDF[[paste0(mergeBy, knn)]]))
```

```{r, label = "save markers per cluster boxplots", include = F}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, "Markers_per_cluster_boxplots"))
  lapply(names(individualClusterMarkerBoxplots), function(x) {
    pdf(file.path(pdfDir, "Markers_per_cluster_boxplots", paste0(x, ".pdf")), width = 12, height = 5)
    print(individualClusterMarkerBoxplots[[x]])
    dev.off()
  })
}
```

```{r, label = "markers_boxplots2", fig.width=18, fig.height=5, results = "asis"}
for (i in seq_along(selectedClustersList)) {
  if (length(selectedClustersList[[i]]) > 0) {
    cat("####", names(selectedClustersList)[[i]], "\n\n")
    for (j in (selectedClustersList[[i]])) {
      df1 <- df[df$cluster_id == as.character(j),]
      g1 <- ggplot(df1, aes(x = antigen, y = value, fill = condition)) +
        geom_boxplot(alpha = 0.2, show.legend = F, outlier.shape = NA) +
        geom_point(position=position_jitterdodge(), pch = 21, size = 2, stroke = 0.2) +
        geom_hline(yintercept = c(0.25, 0.5, 0.75), linetype = "dashed", alpha = 0.3) + 
        geom_vline(xintercept = (seq_along(names(markers_to_use_and_their_type))+0.5)[-length(names(markers_to_use_and_their_type))], linetype = "dotted", alpha = 0.3) +
        scale_fill_manual(values = coloursList$condition) +
        theme_prism(axis_text_angle = 45) +
        ylab("Median Quantile Normalised Expression") + xlab("Condition") + ggtitle(paste("Cluster", j))
      print(g1)
    }
  } else {
    print("No DA clusters to see here...")
  }
  cat("\n\n")
}
```

### DA {.tabset}

Differential abundance of clusters between conditions. Each cluster was tested individually between the two groups specified in each test, e.g., cluster 1 in KO vs WT. Each test gives a log2 fold change, and p-values. Below is the results for each test specified in the metadata. The heatmap shows the expression of each cluster per sample, in order of p-value (most statistically differentially abundant at the top). If you have entered more than one test, then the heatmaps will show the same data but the clusters will be in different orders.

To the right of the heatmap are two annotations. The first is red or blue, and denotes the log2 fold change of that cluster. The second is the p-value, and is either green (adjusted p-value \< 0.05) or grey. The results of this heatmap are the same as under the `Heatmap -> Sample Per Cluster` but with the rows re-ordered and with the log2fc and p-values.

```{r, label = "htmltags1", include = F}
htmltools::tagList(DT::datatable(cars))
```

```{r, label = "da_plot", fig.height=6, fig.width=13, results='asis'}
daResultsList <- setNames(lapply(names(daList), function(cont) {
  cat("\n\n")
  cat("####", cont, "Top 20 {.tabset} \n\n")
  pDH <- plotDiffHeatmap(sce, daList[[cont]], normalize = TRUE, all = TRUE, top_n = 20)
  for (con in conditions) {
    pDH@top_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  }
  print(pDH)
  cat("\n\n")
  cx <- cont %>% gsub(" ", "_", .) %>% gsub("-", "_", .) %>% gsub("\\+", ".", .) %>% gsub("\\/", ".", .)
  ## daDF <- as.data.frame(daList[[cont]][which(daList[[cont]][[daPValToUse]] <= 0.05), c("cluster_id", "logFC", "p_val", "p_adj")])
  daDF <- as.data.frame(daList[[cont]][, c("cluster_id", "logFC", "p_val", "p_adj")])
  daDF <- daDF[order(daDF[[daPValToUse]]),]
  print(htmltools::tagList(
    datatable(daDF, rownames = F, width = "600px") %>% 
      formatSignif(c("p_val", "p_adj", "logFC"), digits = 3) %>% 
      formatStyle(columns = c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")), fontWeight = "bold") %>%
      formatStyle(columns = "logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold")
  ))
  cat("\n\n")
  openxlsx::write.xlsx(x = as.data.frame(daList[[cont]]), file = file.path(excelDir, paste0(cont, "_DA_Results.xlsx")))
  return(list("pDH" = pDH, "cx" = cx, "daDF" = daDF))
}), names(daList))
```

```{r, label = "save DA results", include = F}
if (gimmePDFs) {
  lapply(names(daResultsList), function(x) {
    pdf(file = file.path(pdfDir, paste0(x, "_DA_heatmap.pdf")), height = 6, width = 13)
    print(daResultsList[[x]][["pDH"]])
    dev.off()
  })
}
```

### DS {.tabset}

```{r, label = "htmltags2", include = F}
htmltools::tagList(DT::datatable(cars))
```

```{r, label = "ds_res", results = 'asis', fig.height=12, fig.width=10, fig.align='left'}
for (cont in names(dsList)[grep("Merged", names(dsList), invert = TRUE)]) {
  cat("\n\n")
  cat("####", cont, "{.tabset} \n\n")
  
  dsPThreshToUse <- 0.05
  
  openxlsx::write.xlsx(x = as.data.frame(dsList[[cont]]$tbl_DS), file = file.path(excelDir, paste0(cont, "_DS_Full_Results.xlsx")))
  
  dsUp <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC >= 0)]), decreasing = T) %>% .[. >= 1] %>% data.frame
  dsDown <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC <= 0)]), decreasing = T) %>% .[. >= 1] %>% data.frame
  
  if (nrow(dsUp) == 1) {
    dsUp <- data.frame("Marker" = rownames(dsUp), "Number of clusters" = dsUp[1,])
  }
  if (nrow(dsDown) == 1) {
    dsDown <- data.frame("Marker" = rownames(dsDown), "Number of clusters" = dsDown[1,])
  }
  
  cat("##### Markers up in", strsplit(cont, split = " over ")[[1]][1], "\n\n")
  if (nrow(dsUp) >= 1) {
    print(
      kable(
        x = dsUp, 
        col.names = c("Marker", "Number of clusters")) %>% 
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")
    )
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][1], "\n\n")
  }
  
  cat("##### Markers up in", strsplit(cont, split = " over ")[[1]][2], "\n\n")
  if (nrow(dsDown) >= 1) {
  print(
    kable(
      x = dsDown, 
      col.names = c("Marker", "Number of clusters")) %>% 
      kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")
  )
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][2], "\n\n")
  }
  
  cat("##### Heatmap \n\n")
  pDH <- plotDiffHeatmap(sce, dsList[[cont]]$tbl_DS, normalize = TRUE, all = TRUE, fdr = dsPThreshToUse)
  for (con in conditions) {
    pDH@top_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  }
  print(pDH)
  cat("\n\n")
  cat("##### Full Results \n\n")
  cat("This table shows the log2FC and p-values of each marker per cluster in", cont, "\n\n")
  cat("Positive log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][1], "in a given cluster. \n\n")
  cat("Negative log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][3], "in a given cluster. \n\n")

  dsTab1 <- as.data.frame(dsList[[cont]]$tbl_DS[, c("cluster_id", "marker_id", "logFC", "p_val", "p_adj")])
  dsTab1 <- dsTab1[order(dsTab1$p_val), ]
  print(
    htmltools::tagList(DT::datatable(
      data = dsTab1,
      rownames = FALSE,
      width = "700px"
      ) %>%
        DT::formatSignif(c("logFC", "p_val", "p_adj"), digits = 3) %>%
        DT::formatStyle("logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold") %>%
        DT::formatStyle(c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")),fontWeight = "bold")
    )
  )
}
```

### DS - Merged {.tabset}

Same as the other differential state results, but instead we have merged all clusters prior to DS, so now we see what could be considered differential abundance of each marker between conditions.

```{r, label = "htmltags3", include = F}
htmltools::tagList(DT::datatable(cars))
```

```{r, label = "ds_res merged", results = 'asis', fig.height=12, fig.width=10, fig.align='left'}
for (cont in names(dsList)[grep("Merged", names(dsList), invert = FALSE)]) {
  cat("\n\n")
  cat("####", cont, "{.tabset} \n\n")
  
  openxlsx::write.xlsx(x = as.data.frame(dsList[[cont]]$tbl_DS), file = file.path(excelDir, paste0(cont, "_DS_Merged_Results.xlsx")))
  dsUp <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC >= 0)]), decreasing = T) %>% .[. >= 1]
  dsDown <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC <= 0)]), decreasing = T) %>% .[. >= 1]
  
  cat("##### Markers up in", strsplit(cont, split = " over ")[[1]][1], "\n\n")
  if (length(dsUp >= 1)) {
    print(dsUp)
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][1], "\n\n")
  }
  
  cat("##### Markers up in", strsplit(cont, split = " over ")[[1]][2], "\n\n")
  if (length(dsDown >= 1)) {
  print(dsDown)
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][2], "\n\n")
  }
  cat("##### Heatmap \n\n")
  pDH <- plotDiffHeatmap(x = sceDS, y = dsList[[cont]]$tbl_DS, normalize = TRUE, all = TRUE)
  for (con in conditions) {
    pDH@top_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  }
  print(pDH)
  cat("\n\n")
  cat("##### Full Results \n\n")
  cat("Positive log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][1], "in a given cluster. \n\n")
  cat("Negative log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][3], "in a given cluster. \n\n")
  
  dsTab1 <- as.data.frame(dsList[[cont]]$tbl_DS[, c("cluster_id", "marker_id", "logFC", "p_val", "p_adj")])
  dsTab1 <- dsTab1[order(dsTab1$p_val), ]
  print(
    htmltools::tagList(DT::datatable(
      data = dsTab1,
      rownames = FALSE,
      width = "700px"
      ) %>%
        DT::formatSignif(c("logFC", "p_val", "p_adj"), digits = 3) %>%
        DT::formatStyle("logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold") %>%
        DT::formatStyle(c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")),fontWeight = "bold")
    )
  )
}
```

## Session Info

```{r, label = "file info"}
if (is.null(RDataFolder)) {
  cat("The files generated in this report have been saved in:", qsDir)
} else {
  cat("The results in this report were generated using the files in", RDataFolder)
}
```

```{r, label = save_sesh, include = T}
sessionInfo()
```

