---
title: "{{PIPELINE_NAME}}"

format:
  html:
    # ──────────────────────────────────────────────
    # Layout — no TOC; navigation via top-level tabset
    page-layout: full
    smooth-scroll: true
    embed-resources: true

    # ──────────────────────────────────────────────
    # Code — single toggle in header, not per-chunk
    code-tools: true

    # ──────────────────────────────────────────────
    # Light / dark mode
    theme:
      light: cosmo
      dark: cyborg
    color-mode:
      toggle: true
      respect-prefers-color-scheme: true

    # ──────────────────────────────────────────────
    # Grid sizing
    grid:
      body-width: 1600px
      sidebar-width: 250px
      margin-width: 100px

    # ──────────────────────────────────────────────
    # Header (banner with title overlay, tab cleanup)
    include-in-header:
      - marmot_header.html
      - text: |
          <style>
          .cell-output-display img,
          .quarto-figure img,
          .figure img {
            max-width: 100% !important;
            height: auto !important;
          }
          </style>

    # ──────────────────────────────────────────────
    # Figure defaults
    fig-dpi: 120

# Execution
execute:
  echo: false
  warning: false
  message: false
  cache: false

# Table options
df-print: paged
tbl-cap-location: top

# Cross-references
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  title-delim: ":"

lightbox: true

# ──────────────────────────────────────────────
# Fix htmlwidgets / plots in non-active tabs
include-after-body:
  text: |
    <script>
    document.addEventListener("DOMContentLoaded", function () {
      document.querySelectorAll('button[data-bs-toggle="tab"]').forEach(function (btn) {
        btn.addEventListener("shown.bs.tab", function () {
          if (window.HTMLWidgets && HTMLWidgets.staticRender) {
            HTMLWidgets.staticRender();
          }
          window.dispatchEvent(new Event("resize"));
        });
      });
    });
    </script>
---

```{r knitr_settings, include = FALSE}
knitr::opts_chunk$set(
  comment = "",
  fig.width = 6,
  fig.height = 5,
  fig.retina = 2,
  out.width = "60%",
  dev = if (requireNamespace("ragg", quietly = TRUE)) "ragg_png" else "png"
)
timeRun <- format(Sys.time(), "%Y-%m-%d_%H.%M.%S")
options(bitmapType = "cairo", kableExtra_view_html = FALSE)
```

```{r input_settings}

## 🔧 File paths
fp <- "~/Desktop/FGCZ/MARMOT/Data/Paper/"      # Folder with all FCS files
md_fp <- "~/Desktop/FGCZ/MARMOT/Data/Paper/MARMOT_Metadata.xlsx"      # Full path to metadata Excel file
## 🧪 Preprocessing
downsampleTo <- NULL  # e.g. 1000 to downsample each FCS file, NULL = no downsampling

## 🔍 Clustering
clusteringMethodToUse <- "FlowSOM"   # One of: "MfastPG", "Mphenograph", "Mparc", "Rphenograph", "PARC", "FlowSOM"
markersToClusterBy <- "all"       # One of: "all", "type", "state"
kValuesIWant <- c(20, 40)
knn <- 20          # Must be in kValuesIWant

## 📉 Dimension Reduction
dimRedMethodToUse <- "UMAP"   # One of: "TSNE", "UMAP", "Mpacmap", "pacmap"
markersToDimRedBy <- "all"   # One of: "all", "type", "state"

## 🧼 QC
runQC <- "PeacoQC"      # One of: "FlowAI", "PeacoQC", or "None" to skip
useQC <- FALSE            # Whether to keep only QC-passed cells

## 💾 Output
RDataFolder <- NULL # Use existing results (skip re-analysis)
excludeTheseSamples <- NULL           # e.g. c("Sample_001", "Sample_026"), or NULL
gimmePDFs <- FALSE
greyscalePlots <- FALSE
quantileNormaliseAll <- FALSE

## 🧠 Resources
nCores <- 4
ramPerCore <- 6

## 🎨 Aesthetics
themeToUse <- "prism"    # ggplot2 theme
viridisColour <- "lisbon"   # Viridis, Scico, RColorBrewer Diverging, or Fun (Marmot, Swiss, Lucky)

## 🌉 scGate (optional)
runScGate <- FALSE    # Set TRUE to run scGate marker-pair gating (requires scGate, UCell, Seurat)

## 📁 Results directory (set by marmot(); leave NULL for manual runs)
resultsDir <- NULL

## Done editing? Knit away!
```

```{r, label = "settings_summary", results = 'asis'}
cat("::: {.callout-note collapse='true' title='Pipeline Settings Summary'}\n\n")
cat("- **Clustering method:** ", clusteringMethodToUse, "\n")
cat("- **k values:** ", paste(kValuesIWant, collapse = ", "), " (knn = ", knn, ")\n")
cat("- **Dim. reduction:** ", dimRedMethodToUse, "\n")
cat("- **QC:** ", runQC, if (useQC) " (applied)" else " (not applied)", "\n")
cat("- **Downsample:** ", if (!is.null(downsampleTo)) paste0(downsampleTo, " cells") else "No", "\n")
cat("- **Cores:** ", nCores, "\n")
cat("- **scGate:** ", if (runScGate) "Enabled" else "Disabled", "\n")
cat(":::\n\n")
```

```{r check_variables, include = FALSE}

required_vars <- c(
  "fp", "md_fp", "downsampleTo",
  "clusteringMethodToUse", "markersToClusterBy", "kValuesIWant", "knn",
  "dimRedMethodToUse", "markersToDimRedBy",
  "runQC", "useQC", "RDataFolder", "excludeTheseSamples",
  "greyscalePlots", "gimmePDFs", "quantileNormaliseAll",
  "nCores", "themeToUse", "viridisColour"
)

missing_vars <- required_vars[!vapply(required_vars, exists, logical(1))]

if (length(missing_vars)) {
  stop(
    "Missing required variables: ",
    paste(missing_vars, collapse = ", "),
    ".\nPlease define them in the 'input settings' chunk.",
    call. = FALSE
  )
}
```

```{r palette_helper, include = FALSE}
viridis_palettes <- c("magma", "inferno", "plasma", "viridis", "cividis", "rocket", "mako", "turbo")
scico_palettes   <- c("bam", "berlin", "brocO", "corkO", "lapaz", "lisbon", "romaO", "vikO")
fun_palettes <- list(
  Marmot = c("#4a3728", "#6b5344", "#8c7a6b", "#a89888", "#bfb5a8", "#d4ccc4", "#9e9e9e", "#787878"),
  Swiss  = c("#ffffff", "#ffe5e5", "#ffb3b3", "#ff6666", "#ff1a1a", "#cc0000", "#990000")
)

get_palette_colours <- function(palette, n = 64) {
  if (palette %in% viridis_palettes) {
    viridis::viridis(n, option = palette)
  } else if (palette %in% scico_palettes) {
    scico::scico(n, palette = palette)
  } else if (palette == "Lucky") {
    colorRampPalette(sample(grDevices::colors(distinct = TRUE), 3))(n)
  } else if (palette %in% names(fun_palettes)) {
    colorRampPalette(fun_palettes[[palette]])(n)
  } else {
    # Diverging (RColorBrewer)
    colorRampPalette(RColorBrewer::brewer.pal(11, palette))(n)
  }
}
```

```{r load_packages, include = FALSE}

# Use pacman for convenience
suppressWarnings(suppressPackageStartupMessages(
  pacman::p_load(
    BiocGenerics, S4Vectors, SummarizedExperiment, SingleCellExperiment,
    flowCore, FlowSOM, ggplot2, dplyr, tidyr, purrr, tibble, readr,
    CATALYST, readxl, cowplot, diffcyt, ggpubr,
    ggprism, gridExtra, rstatix, plotly, grDevices,
    RColorBrewer, reshape2, grid, ComplexHeatmap, circlize, clustree, kableExtra,
    DT, flowAI, parallel, qs2, ggrepel, gtools, reticulate, ggbeeswarm, matrixStats,
    PeacoQC, MARMOT
  )
))

set.seed(2)

# Define class union for ExpData (if not already present)
if (!methods::isClass("ExpData")) {
  try({
    methods::setClassUnion("ExpData", c("matrix", "SummarizedExperiment"), where = .GlobalEnv)
  }, silent = TRUE)
}
```

```{r pipeline_helpers, include = FALSE}
# Parse "CellType: Marker1 Marker2" entries from the Marker Pairs column
parse_marker_pairs <- function(smd) {
  if (!"Marker Pairs" %in% colnames(smd)) return(NULL)
  pairs <- smd[["Marker Pairs"]][!is.na(smd[["Marker Pairs"]])]
  if (length(pairs) == 0) return(NULL)
  types <- unique(gsub("\\:.*", "", pairs))
  names(pairs) <- gsub("\\:.*", "", pairs)
  models <- lapply(types, function(tp) {
    entry <- pairs[names(pairs) == tp][[1]]
    markers <- strsplit(entry, "\\:\\s*")[[1]][[2]]
    strsplit(markers, "\\s+")[[1]]
  })
  names(models) <- types
  list(pairs = unname(pairs), types = types, models = models)
}

# Build coloursList entries for scGate results
setup_scgate_colours <- function(gated_cells, model_names, palette) {
  cols <- list()
  gate_pal <- palette[seq_along(gated_cells)]
  names(gate_pal) <- gated_cells
  cols[["Gated_Cells"]] <- gate_pal
  for (x in model_names) {
    y <- gsub(" |-", "", x)
    cols[[paste0("is_", y)]] <- setNames("steelblue", x)
  }
  cols
}

# Rewrite ComplexHeatmap top-annotation colours from coloursList. Four slots
# need to be poked per condition because ComplexHeatmap caches color_mapping
# both on the anno_list slot and inside the draw closure's var_env.
apply_anno_colours <- function(pDH, conditions, coloursList) {
  for (con in conditions) {
    cm <- pDH@top_annotation@anno_list[[con]]@color_mapping
    new_cols <- coloursList[[con]][match(names(cm@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@color_mapping@colors <- new_cols
    pDH@top_annotation@anno_list[[con]]@color_mapping@full_col <- new_cols
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- new_cols
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- new_cols
  }
  pDH
}

# Build inputs for the `Heatmaps of DA Clusters` chunk. This replaces the
# old `function for custom heatmap` chunk that leaked x / z / lgd_aes into
# the global env via CATALYST internals.
build_da_heatmap_inputs <- function(sce, k, assay = "exprs",
                                    fun = "median", q = 0.01) {
  x <- sce[unique(CATALYST:::.get_features(sce, NULL)), ]
  CATALYST:::.check_k(x, k)
  x$cluster_id <- cluster_ids(x, k)

  # "first" scale path: scale the assay, then aggregate.
  z_assay <- CATALYST:::.scale_exprs(assay(x, assay), 1, q)
  assay(x, assay, FALSE) <- z_assay
  z <- CATALYST:::.agg(x, "cluster_id", fun, assay)
  z <- t(z)

  qs <- round(quantile(z, c(0.01, 0.99)) * 5) / 5
  lgd_aes <- list(
    at = seq(qs[1], qs[2], 0.2),
    title_gp = grid::gpar(fontsize = 10, fontface = "bold", lineheight = 0.8)
  )

  list(x = x, z = z, lgd_aes = lgd_aes)
}
```

```{r load_rdata, include = FALSE, results = 'asis'}

if (!is.null(RDataFolder)) {
  # Auto-resolve: if user pointed to the top-level Results_Files dir, descend into R_files/
  h5ad_path <- file.path(RDataFolder, "marmot_results.h5ad")
  if (!file.exists(h5ad_path) &&
       file.exists(file.path(RDataFolder, "R_files", "marmot_results.h5ad"))) {
    RDataFolder <- file.path(RDataFolder, "R_files")
    h5ad_path <- file.path(RDataFolder, "marmot_results.h5ad")
    message("ℹ️ Auto-resolved RDataFolder to: ", RDataFolder)
  }
  if (!file.exists(h5ad_path)) {
    stop("No h5ad data found in ", RDataFolder,
         ". Please re-run the pipeline to generate h5ad output.")
  }
  message(glue::glue("🔁 Loading results from h5ad: {h5ad_path}"))
  MARMOT::load_h5ad_to_env(h5ad_path, envir = .GlobalEnv)
  # Also load FCS frame objects (stored as qs2 — not decomposable to h5ad)
  qs_path <- file.path(RDataFolder, "framesList.qs2")
  if (file.exists(qs_path)) {
    assign("framesList", qs2::qs_read(qs_path, nthreads = nCores), envir = .GlobalEnv)
  }

  # Load DA/DS result objects (S4 structures that can't survive h5ad)
  da_ds_path <- file.path(RDataFolder, "da_ds_results.qs2")
  if (file.exists(da_ds_path)) {
    da_ds <- qs2::qs_read(da_ds_path, nthreads = nCores)
    for (nm in names(da_ds)) {
      if (!is.null(da_ds[[nm]])) assign(nm, da_ds[[nm]], envir = .GlobalEnv)
    }
    message("Loaded (qs2): da_ds_results.qs2")
  }

  # Reconstruct panel from rowData(sce) — create_panel chunk is skipped on reload
  rd <- as.data.frame(SummarizedExperiment::rowData(sce))
  panel <- data.frame(
    fcs_colname  = rownames(rd),
    marker_name  = if ("marker_name" %in% names(rd)) rd$marker_name else rownames(rd),
    antigen      = if ("marker_name" %in% names(rd)) rd$marker_name else rownames(rd),
    marker_class = if ("marker_class" %in% names(rd)) rd$marker_class else "none",
    use_channel  = TRUE,
    stringsAsFactors = FALSE
  )

  # Inject md columns into colData — prepData did this on fresh run via
  # md_cols = list(factors = conditions), but h5ad doesn't preserve the merge.
  cd <- as.data.frame(SummarizedExperiment::colData(sce))
  md_reload <- as.data.frame(md)
  md_extra_cols <- setdiff(names(md_reload), c("file_name", names(cd)))
  if (length(md_extra_cols) > 0) {
    lookup <- md_reload[, c("sample_id", md_extra_cols), drop = FALSE]
    matched <- lookup[match(cd$sample_id, lookup$sample_id), md_extra_cols, drop = FALSE]
    rownames(matched) <- rownames(cd)
    for (col in md_extra_cols) {
      SummarizedExperiment::colData(sce)[[col]] <- matched[[col]]
    }
    message("ℹ️  Injected md columns into colData: ", paste(md_extra_cols, collapse = ", "))
  }
}

# Pick the merge method depending on clustering backend
mergeBy <- dplyr::case_when(
  clusteringMethodToUse %in% c("Rphenograph", "Mphenograph", "MfastPG") ~ "k",
  clusteringMethodToUse %in% c("PARC", "Mparc") ~ "p",
  clusteringMethodToUse == "FlowSOM" ~ "meta",
  TRUE ~ NA_character_  # temp placeholder
)
if (is.na(mergeBy)) stop("Unsupported clustering method: ", clusteringMethodToUse)

# Heatmap colour palette
hmPalToUse <- if (greyscalePlots) {
  RColorBrewer::brewer.pal(9, "Greys")
} else {
  rev(RColorBrewer::brewer.pal(11, "RdBu"))
}
```

```{r import_metadata, results = "hide"}

# Load metadata
md  <- readxl::read_xlsx(md_fp, sheet = "File Data")
smd <- readxl::read_xlsx(md_fp, sheet = "Study Data")

# Paths
fcsDir <- fp

# Order sample IDs naturally
md <- md[gtools::mixedorder(md$sample_id), ]
md$sample_id <- factor(md$sample_id, levels = md$sample_id)

# Apply condition ordering if available
conditionOrder <- smd$`Conditions Order` |> na.omit()
if (length(conditionOrder) > 0) {
  md$condition <- factor(md$condition, levels = conditionOrder)
} else {
  md$condition <- as.factor(md$condition)
}

# Extract marker names
useMarkers <- smd[[colnames(smd)[grep("include|cluster", colnames(smd))]]]
useMarkers <- useMarkers[!is.na(useMarkers)]
names(useMarkers) <- gsub("-|\\ |\\/", "_", useMarkers)

# Extract marker type vector (used downstream for marker_class and conditional sections)
# Named vector: names = marker names, values = "type"/"state"
# Named lookup prevents ifelse recycling when param_df has extra scatter/time channels
mt_mask <- !is.na(smd$`Marker Type`)
include_col <- colnames(smd)[grep("include|cluster", colnames(smd))]
markerType <- setNames(smd$`Marker Type`[mt_mask], smd[[include_col]][mt_mask])

# Define feature type from marker type
featureType <- if ("Marker Type" %in% names(smd)) {
  types <- na.omit(smd$`Marker Type`)
  if ("type" %in% types) "type" else NULL
} else {
  stop("Metadata file must contain 'Marker Type' column.")
}

# Markers to exclude
removeMarkers <- na.omit(smd$`Markers to exclude completely`)

# Cofactors
cfToUse <- as.numeric(na.omit(smd$`Cofactors for markers to use`))

# Contrasts
# Filter out empty or malformed contrasts (e.g. bare " over " from empty dropdowns)
contrastsToTest <- na.omit(smd$`Conditions To Test`)
contrastsToTest <- trimws(contrastsToTest)
contrastsToTest <- gsub("\\s+", " ", contrastsToTest)
contrastsToTest <- contrastsToTest[grepl(".+ over .+", contrastsToTest)]
contrastsToTestS <- paste(contrastsToTest, collapse = ",")
cTTForPlots <- contrastsToTest

# Subsampling settings
raw_subsamp <- na.omit(smd$`Cells per condition in UMAPs etc.`)
drCellPerCondition <- setNames(
  vapply(raw_subsamp, function(expr) as.numeric(eval(parse(text = expr))), numeric(1)),
  conditionOrder
)

# Identify all condition columns
conditions <- c("condition", setdiff(names(md), c("file_name", "sample_id", "condition")))
conditions <- gsub("-", ".", conditions)

# Exclude samples if specified
if (!is.null(excludeTheseSamples)) {
  md <- md[!md$sample_id %in% excludeTheseSamples, ]
}

# Sort again by condition
md <- md[order(md$condition), ]
md$patient_id <- factor(md$sample_id, levels = md$sample_id)
md$sample_id  <- factor(md$sample_id, levels = md$sample_id)
```

```{r cluster_colours, results = "hide"}

# Define base palette used in CATALYST-inspired plots
catalystCols <- c(
  "#DC050C", "#FB8072", "#1965B0", "#7BAFDE", "#882E72", "#B17BA6",
  "#FF7F00", "#FDB462", "#E7298A", "#E78AC3", "#33A02C", "#B2DF8A",
  "#55A1B1", "#8DD3C7", "#A6761D", "#E6AB02", "#7570B3", "#BEAED4",
  "#666666", "#999999", "#AA8282", "#D4B7B7", "#8600BF", "#BA5CE3",
  "#808000", "#AEAE5C", "#1E90FF", "#00BFFF", "#56FF0D", "#CFBF11"
)

# Darken the palette and append (to allow >30 colours with contrast)
darkenedCols <- colorspace::darken(catalystCols, 0.4)
catalystCols <- paste0(c(catalystCols, darkenedCols), "FF")  # Add full opacity
```

```{r sample_colours, results = "hide", dependson = -1, eval = is.null(RDataFolder)}

# Collect all unique factor levels from the specified metadata columns
allConditionLevels <- unlist(lapply(conditions, \(x) levels(factor(md[[x]]))), use.names = FALSE)

# Assign colours to each unique level using a palette combo (expand as needed)
palettePool <- c(
  RColorBrewer::brewer.pal(9, "Set1"),
  RColorBrewer::brewer.pal(8, "Dark2"),
  RColorBrewer::brewer.pal(12, "Paired"),
  RColorBrewer::brewer.pal(8, "Set3")
)
if (length(allConditionLevels) > length(palettePool)) {
  palettePool <- colorRampPalette(palettePool)(length(allConditionLevels))
}

allConditionLevelColours <- setNames(palettePool[seq_along(allConditionLevels)], allConditionLevels)

# Create named list of colours per condition
coloursList <- setNames(
  lapply(conditions, \(x) {
    levels <- levels(factor(md[[x]]))
    allConditionLevelColours[levels]
  }),
  conditions
)

# Assign sample-level colours
sampleCols <- setNames(
  colorRampPalette(brewer.pal(8, "Dark2"))(length(levels(md$sample_id))),
  levels(md$sample_id)
)

# Assign condition-level colours
conditionCols <- setNames(
  colorRampPalette(brewer.pal(8, "Set1"))(length(levels(md$condition))),
  levels(md$condition)
)
```

```{r import_fcs_scan, include = FALSE, results = "asis", eval = is.null(RDataFolder)}

message("📥 Scanning FCS files...")

# List all .fcs files (case-insensitive — some cytometers export as .FCS)
fcs_files <- list.files(fcsDir, pattern = "\\.fcs$", full.names = TRUE, ignore.case = TRUE)
fcs_files <- fcs_files[grep(paste(md$file_name, collapse = "|"), fcs_files, ignore.case = TRUE)]
fcs_names <- tools::file_path_sans_ext(basename(fcs_files))

# ── Header scan — get cell counts without loading data ────────────────────────
fcs_headers <- sapply(fcs_files, function(f) {
  hdr <- flowCore::read.FCSheader(f)
  c(tot = as.numeric(hdr[[1]]["$TOT"]),
    par = as.numeric(hdr[[1]]["$PAR"]))
})
originalCellCounts <- setNames(fcs_headers["tot", ], fcs_names)
total_cells_raw    <- sum(originalCellCounts)
n_params           <- fcs_headers["par", 1]
n_samples          <- length(fcs_files)

# ── Compute RAM budget and per-sample cap ─────────────────────────────────────
sys_ram_bytes <- tryCatch({
  os <- Sys.info()[["sysname"]]
  if (os == "Darwin") {
    as.numeric(system("sysctl -n hw.memsize", intern = TRUE))
  } else if (os == "Linux") {
    as.numeric(system("awk '/MemTotal/{print $2 * 1024}' /proc/meminfo", intern = TRUE))
  } else {
    as.numeric(trimws(system("wmic OS get TotalVisibleMemorySize /value 2>nul", intern = TRUE) |>
      grep("=", value = TRUE) |> sub(".*=", "", x = _))) * 1024
  }
}, error = function(e) NA_real_)
sys_ram_gb <- if (!is.na(sys_ram_bytes)) sys_ram_bytes / 1024^3 else Inf

# Available RAM: lower of system RAM and user-declared budget
user_budget_gb     <- ramPerCore * nCores
available_gb       <- min(sys_ram_gb, user_budget_gb)
pipeline_budget_gb <- available_gb * 0.6   # 60% for pipeline, 40% for OS/R/Quarto
peak_multiplier    <- 2.5                  # measured 2.2× on real data; 2.5× for safety
bytes_per_cell     <- n_params * 8

max_total_cells <- floor((pipeline_budget_gb * 1024^3) / (bytes_per_cell * peak_multiplier))
per_sample_cap  <- floor(max_total_cells / n_samples)
needs_cap       <- total_cells_raw > max_total_cells

message(sprintf("📊 Dataset: %s cells × %d params across %d samples",
                format(total_cells_raw, big.mark = ","), n_params, n_samples))
message(sprintf("📊 RAM: %.0f GB system, %.0f GB user budget, %.1f GB pipeline budget (60%%)",
                sys_ram_gb, user_budget_gb, pipeline_budget_gb))

if (needs_cap) {
  message(sprintf(
    paste0("⚠️  Dataset has %s cells but this machine can safely hold %s. ",
           "Capping to %s cells per sample (%s total)."),
    format(total_cells_raw, big.mark = ","),
    format(max_total_cells, big.mark = ","),
    format(per_sample_cap, big.mark = ","),
    format(per_sample_cap * n_samples, big.mark = ",")))
} else {
  message(sprintf("📊 Dataset fits in RAM budget (%s cells < %s max)",
                  format(total_cells_raw, big.mark = ","),
                  format(max_total_cells, big.mark = ",")))
}
```

```{r import_ram_cap_notice, results = "asis", eval = is.null(RDataFolder)}
if (needs_cap) {
  cat(sprintf(
    paste0("::: {.callout-warning title='RAM Cap Applied'}\n",
           "Your dataset has **%s cells** across %d samples, but this machine ",
           "can safely hold **%s cells** in memory (%.0f GB system RAM, ",
           "%.1f GB pipeline budget at 60%%).\n\n",
           "Each sample has been capped to **%s cells**. QC ran on the full dataset. ",
           "Clustering, dimensionality reduction, and all downstream analysis use the capped data. ",
           "To use more cells, run on a machine with more RAM or increase `ramPerCore`.\n",
           ":::\n"),
    format(total_cells_raw, big.mark = ","), n_samples,
    format(max_total_cells, big.mark = ","),
    sys_ram_gb, pipeline_budget_gb,
    format(per_sample_cap, big.mark = ",")))
}
```

```{r create_panel, results = "hide", eval = is.null(RDataFolder)}

# Extract parameter metadata from one frame (just need pData, not expression data)
.tmp_ff <- flowCore::read.FCS(fcs_files[1], transformation = FALSE,
                               truncate_max_range = FALSE, which.lines = 1)
param_df <- pData(parameters(.tmp_ff)) |>
  as.data.frame() |>
  dplyr::mutate(desc = ifelse(is.na(desc), name, desc))
rm(.tmp_ff)

# Build panel
panel <- param_df |>
  dplyr::select(fcs_colname = name, marker_name = desc) |>
  dplyr::mutate(
    antigen       = marker_name,
    marker_class  = ifelse(marker_name %in% names(markerType), markerType[marker_name], "none"),
    use_channel   = marker_name %in% useMarkers
  ) |>
  dplyr::filter(!marker_name %in% removeMarkers) |>
  dplyr::filter(marker_name %in% useMarkers) |>
  dplyr::arrange(match(marker_name, useMarkers))

# Sanity check
missing <- setdiff(useMarkers, panel$marker_name)
if (length(missing)) {
  stop("❌ Markers in metadata not found in FCS: ", paste(missing, collapse = ", "))
}

message("⚙️ Preparing cofactors for arcsinh transformation...")

# Base cofactor vector, matched to FCS column names
cf <- setNames(cfToUse, panel$fcs_colname[match(useMarkers, panel$marker_name)])

# Fill in default (5) for other channels
all_channels <- panel$fcs_colname
extra_channels <- setdiff(all_channels, names(cf))
cf[extra_channels] <- 2500 # Default cofactor for unused channels

# Ensure correct order
cf <- cf[match(all_channels, names(cf))]

# Alternative version cf2, named by marker name
cf2 <- setNames(cf[match(panel$fcs_colname, names(cf))], panel$marker_name)
cf2 <- cf2[panel$use_channel]

asinh_transforms <- transformList(
  names(cf),
  lapply(cf, function(cof) arcsinhTransform(a = 0, b = 1 / cof, c = 0))
)
```

```{r check_qc_viability, include = FALSE, eval = is.null(RDataFolder)}
if (runQC %in% c("FlowAI", "PeacoQC")) {
  sample_sizes <- originalCellCounts
  min_events <- min(sample_sizes)
  small_sample <- names(which.min(sample_sizes))

  if (min_events < 2000) {
    warning("Sample '", small_sample, "' has only ", min_events,
            " events. Skipping QC.")
    qc_skip_reason <- paste0(
      "Sample **", small_sample, "** has only **", min_events,
      "** events — too few for any automated QC tool (minimum ~2,000). ",
      "QC has been **skipped** for this run. ",
      "With so few events, automated QC is tenuous at best — ",
      "consider acquiring more events or excluding very small samples.")
    runQC <- "None"
    useQC <- FALSE

  } else if (runQC == "PeacoQC" && min_events < 5000) {
    warning("Sample '", small_sample, "' has only ", min_events,
            " events. Skipping PeacoQC.")
    qc_skip_reason <- paste0(
      "Sample **", small_sample, "** has only **", min_events,
      "** events — PeacoQC requires ~5,000+ events per sample for its ",
      "MAD outlier detection. QC has been **skipped** for this run. ",
      "With so few events, automated QC is tenuous at best. ",
      "To run QC on this dataset, try **FlowAI** (works with ~2,000+ events), ",
      "or set `runQC` to `None` to skip QC entirely.")
    runQC <- "None"
    useQC <- FALSE
  }
}
```

```{r run_flow_auto_qc, results = "asis", include = FALSE, dependson = -1, eval = is.null(RDataFolder)}

if (runQC == "FlowAI") {

  message("🔍 Starting automated quality control...")

  metricsQC <- "FR_FS"

  qc_dir <- file.path(fp, "resultsQC_flowAI")
  expected_qc_files <- gsub("\\.fcs$", "_QC.fcs", md$file_name)
  qc_present <- dir.exists(qc_dir) &&
    all(expected_qc_files %in% list.files(qc_dir, pattern = "_QC\\.fcs", ignore.case = TRUE))

  if (qc_present) {
    message("📁 QC files found on disk — skipping recomputation.")
  } else {
    message("⚙️  Running FlowAI per-file at full resolution...")
    dir.create(qc_dir, showWarnings = FALSE, recursive = TRUE)

    for (i in seq_along(fcs_files)) {
      message(sprintf("  [%d/%d] %s (%s cells)",
                      i, length(fcs_files), basename(fcs_files[i]),
                      format(originalCellCounts[i], big.mark = ",")))
      ff <- flowCore::read.FCS(fcs_files[i], transformation = FALSE,
                                truncate_max_range = FALSE)
      flow_auto_qc(
        fcsfiles       = ff,
        folder_results = qc_dir,
        remove_from    = metricsQC,
        fcs_highQ      = "_QC_highQ",
        fcs_lowQ       = "_QC_lowQ"
      )
      rm(ff); invisible(gc())
    }
  }

  # Load and annotate QC summary table
  fcsFilesQC <- list.files(qc_dir, full.names = TRUE, pattern = "_QC_highQ\\.fcs", ignore.case = TRUE)
  fcsFilesQC <- fcsFilesQC[grep(paste(gsub("\\.fcs", "", md$file_name), collapse = "|"), fcsFilesQC, ignore.case = TRUE)]
  fcsNamesQC <- sub("_QC_highQ\\.fcs$", "", basename(fcsFilesQC))

  QCmini <- readr::read_tsv(file.path(qc_dir, "QCmini.txt"), show_col_types = FALSE) |>
    dplyr::distinct(`Name file`, .keep_all = TRUE)
  QCmini <- QCmini[grep(paste0(fcsNamesQC, collapse = "|"), QCmini$`Name file`), ]

  qc_names_clean <- sub("_QC_highQ|\\.fcs", "", QCmini$`Name file`)
  md_names_clean <- sub("\\.fcs$", "", md$file_name)
  QCmini$`Sample Name` <- md$sample_id[match(qc_names_clean, md_names_clean)]
  QCmini <- QCmini[gtools::mixedorder(QCmini$`Name file`), ]

  message("🎉 FlowAI QC complete.")
}
```

```{r run_peacoqc, results = "asis", include = FALSE, dependson = -1, eval = is.null(RDataFolder)}
if (runQC == "PeacoQC") {

  message("🔍 Starting automated quality control with PeacoQC...🦚")

  qc_dir <- file.path(fp, "resultsQC_peacoQC")
  qc_present <- dir.exists(qc_dir) && {
    expected_qc_files <- gsub("\\.fcs$", "_QC.fcs", md$file_name)
    actual_qc_paths <- list.files(qc_dir, pattern = "_QC\\.fcs$", recursive = TRUE, full.names = TRUE, ignore.case = TRUE)
    all(expected_qc_files %in% basename(actual_qc_paths))
  }

  if (qc_present) {
    message("📁 QC files found on disk — skipping recomputation.")
  } else {
    message("⚙️  Running PeacoQC per-file at full resolution...")

    dir.create(qc_dir, showWarnings = FALSE, recursive = TRUE)

    # Determine fluorescence channels from one frame's metadata
    all_channels <- param_df$name
    fluor_channels <- all_channels[!grepl("^(FSC|SSC|Time)", all_channels, ignore.case = TRUE)]
    fluor_channels <- fluor_channels[!grepl("(Width|Height|Trigger)", fluor_channels, ignore.case = TRUE)]
    message("🔬 Using channels for QC: ", paste(fluor_channels, collapse = ", "))
    message("🚫 Excluded channels: ", paste(setdiff(all_channels, fluor_channels), collapse = ", "))

    for (i in seq_along(fcs_files)) {
      fcs_name <- fcs_names[i]
      message(sprintf("  [%d/%d] %s (%s cells)",
                      i, length(fcs_files), fcs_name,
                      format(originalCellCounts[i], big.mark = ",")))
      ff <- flowCore::read.FCS(fcs_files[i], transformation = FALSE,
                                truncate_max_range = FALSE)
      ff_t <- flowCore::transform(ff, asinh_transforms)
      rm(ff)

      PeacoQC::PeacoQC(
        ff = ff_t, channels = fluor_channels,
        plot = TRUE, save_fcs = TRUE,
        output_directory = qc_dir, name_directory = fcs_name,
        report = TRUE, determine_good_cells = "all"
      )
      rm(ff_t); invisible(gc())
    }

    # Generate combined QC summary table
    qc_files <- list.files(path = qc_dir, pattern = "report\\.txt", recursive = TRUE, full.names = TRUE)
    QCmini <- readr::read_tsv(qc_files[[1]], show_col_types = FALSE)
    if (length(qc_files) > 1) {
      for (j in 2:length(qc_files)) {
        xx <- readr::read_tsv(qc_files[[j]], show_col_types = FALSE)
        QCmini[j, ] <- xx[1, ]
      }
    }
    readr::write_tsv(QCmini, file.path(qc_dir, "QCmini.txt"))
    message("✅ PeacoQC processing complete.")
  }

  # Load QC summary table
  fcsFilesQC <- list.files(qc_dir, full.names = TRUE, pattern = "_QC\\.fcs$", recursive = TRUE, ignore.case = TRUE)
  fcsFilesQC <- fcsFilesQC[grep(paste(md$file_name, collapse = "|"), fcsFilesQC, ignore.case = TRUE)]
  fcsNamesQC <- sub("_QC\\.fcs$", "", basename(fcsFilesQC))

  QCmini <- readr::read_tsv(file.path(qc_dir, "QCmini.txt"), show_col_types = FALSE)
  QCmini <- QCmini[grep(paste0(fcsNamesQC, collapse = "|"), QCmini$Filename), ]

  message("🎉 PeacoQC QC complete.")
}
```

```{r import_fcs_load, include = FALSE, results = "asis", eval = is.null(RDataFolder)}

message("📥 Loading FCS data into memory...")

# Helper: read one FCS file and apply RAM cap if needed
read_and_cap <- function(f) {
  ff <- flowCore::read.FCS(f, transformation = FALSE, truncate_max_range = FALSE)
  if (needs_cap && nrow(ff) > per_sample_cap) {
    set.seed(2)
    ff <- ff[sort(sample(nrow(ff), per_sample_cap)), ]
  }
  ff
}

# Load only the data that will actually be used for analysis
if (useQC && runQC == "FlowAI" && exists("qc_dir") && dir.exists(qc_dir)) {
  # useQC=TRUE, FlowAI: load only QC'd FCS files
  fcsFilesQC <- list.files(qc_dir, full.names = TRUE, pattern = "_QC_highQ\\.fcs", ignore.case = TRUE)
  fcsFilesQC <- fcsFilesQC[grep(paste(gsub("\\.fcs", "", md$file_name), collapse = "|"), fcsFilesQC, ignore.case = TRUE)]
  fcsNamesQC <- sub("_QC_highQ\\.fcs$", "", basename(fcsFilesQC))

  framesList <- list()
  framesList[["FlowAI"]] <- setNames(
    lapply(fcsFilesQC, read_and_cap),
    paste0(fcsNamesQC, ".fcs")
  )
  invisible(lapply(framesList[["FlowAI"]], range, type = "data"))

} else if (useQC && runQC == "PeacoQC" && exists("qc_dir") && dir.exists(qc_dir)) {
  # useQC=TRUE, PeacoQC: load only QC'd FCS files
  fcsFilesQC <- list.files(qc_dir, full.names = TRUE, pattern = "_QC\\.fcs$", recursive = TRUE, ignore.case = TRUE)
  fcsFilesQC <- fcsFilesQC[grep(paste(md$file_name, collapse = "|"), fcsFilesQC, ignore.case = TRUE)]
  fcsNamesQC <- sub("_QC\\.fcs$", "", basename(fcsFilesQC))

  framesList <- list()
  framesList[["PeacoQC"]] <- setNames(
    lapply(fcsFilesQC, read_and_cap),
    paste0(fcsNamesQC, ".fcs")
  )
  invisible(lapply(framesList[["PeacoQC"]], range, type = "data"))

} else {
  # useQC=FALSE or no QC: load raw files → build Transformed
  frames <- setNames(lapply(fcs_files, read_and_cap), fcs_names)
  if (!is.null(excludeTheseSamples)) {
    keep_samples <- md$file_name[!md$sample_id %in% excludeTheseSamples]
    frames <- frames[keep_samples]
  }
  framesList <- list("Untransformed" = frames)
  rm(frames)
}

invisible(gc())

effective_cells <- sum(sapply(framesList[[1]], nrow))
message(sprintf("📊 Loaded %s cells (%.1f GB estimated)",
                format(effective_cells, big.mark = ","),
                effective_cells * bytes_per_cell / 1024^3))
```

```{r arcsinh_transform_apply, results = "hide", eval = is.null(RDataFolder)}
# Build Transformed from Untransformed (useQC=FALSE / no QC path)
if ("Untransformed" %in% names(framesList)) {
  framesList$Transformed <- lapply(seq_along(framesList$Untransformed), function(i) {
    flowCore::transform(framesList$Untransformed[[i]], asinh_transforms)
  }) |> setNames(names(framesList$Untransformed))
}

# Transform FlowAI QC'd data — FlowAI runs on raw data, so QC'd files are untransformed.
# PeacoQC doesn't need this: it runs on already-transformed data.
if ("FlowAI" %in% names(framesList)) {
  framesList[["FlowAI"]] <- lapply(seq_along(framesList[["FlowAI"]]), function(i) {
    flowCore::transform(framesList[["FlowAI"]][[i]], asinh_transforms)
  }) |> setNames(names(framesList[["FlowAI"]]))
}
```

```{r arrange_frames_list, include = FALSE, results = "asis", eval = is.null(RDataFolder)}

message("📦 Converting frame lists into flowSets...")

# Convert to flowSet and set $CYT keyword to "FACS"
# Also enforce sampleNames from the list names: read.FCS() embeds the on-disk
# filename (e.g. "Sample_001_QC.fcs") as the frame identifier, which would
# cause CATALYST::prepData() to fail matching against md$file_name.
fsList <- purrr::imap(framesList, function(sample_list, name) {
  fs <- as(sample_list, "flowSet")
  sampleNames(fs) <- names(sample_list)
  for (j in seq_along(fs)) {
    keyword(fs[[j]])[["$CYT"]] <- "FACS"
  }
  fs
})
```

```{r plot_cofactor_histograms, results = "hide", eval = is.null(RDataFolder)}

message("📉 Generating cofactor histograms for visualising transformations...")

cofactorsToTry <- c(5, 50, 500, 1000, 2500, 5000, 10000, 15000)

# Aggressively downsample (use whatever data was loaded — raw or QC'd)
framesList[["Marker Histos"]] <- lapply(framesList[[1]], function(fcs) {
  if (nrow(fcs) > 1000) {
    set.seed(2)
    fcs[sample(nrow(fcs), 1000), ]
  } else {
    fcs
  }
})

# Function to strip data from ggplot object
minify_plot_data <- function(p, colname, group_col = "condition") {
  # Keep only the columns needed for plotting (x and fill/colour)
  vars_needed <- unique(c(colname, group_col))
  p$data <- p$data[, vars_needed, drop = FALSE]
  p
}

# Plot generation without persistent cofactorDFList
cofactorPlotList <- lapply(useMarkers, function(marker) {
  colname <- panel$fcs_colname[panel$marker_name == marker]
  marker_expr <- purrr::map(framesList[["Marker Histos"]], ~ exprs(.x)[, colname])
  
  # Flat vector of expression values
  expr_vec <- unlist(marker_expr)
  sample_vec <- rep(names(marker_expr), times = purrr::map_int(marker_expr, length))
  cond_vec <- md$condition[match(gsub("\\.fcs$", "", sample_vec), gsub("\\.fcs$", "", md$file_name))]

  # Build only what's needed: wide format DF with just transformed values + condition
  df <- tibble::tibble(
    condition = cond_vec,
    pretransform = expr_vec
  )

  # Transform in-place
  transformed_cols <- purrr::map(cofactorsToTry, function(cof) {
    tibble::tibble(!!paste0("Cofactor_", cof) := asinh(expr_vec / cof))
  }) |> purrr::list_cbind()

  df <- dplyr::bind_cols(df, transformed_cols)

  # Generate plots and strip ggplot data
  plot_names <- c("pretransform", paste0("Cofactor_", cofactorsToTry))
  plot_set <- purrr::map(plot_names, function(colname) {
    p <- ggplot(df, aes(x = .data[[colname]], fill = condition, colour = condition)) +
      geom_density(alpha = 0.1) +
      ggtitle(colname) +
      theme_prism() +
      theme(text = element_text(size = 15), legend.position = "none") +
      scale_colour_manual(values = coloursList$condition) +
      scale_fill_manual(values = coloursList$condition) +
      geom_vline(xintercept = mean(df[[colname]], na.rm = TRUE), linetype = "dashed", colour = "red")
    minify_plot_data(p, colname)
  })

  setNames(plot_set, plot_names)
})

names(cofactorPlotList) <- useMarkers
```

```{r create_sce, results = "hide", eval = is.null(RDataFolder)}
if (useQC) {
  message("✅ Using QC-filtered data for downstream analysis.")
  fsToUse <- fsList[[runQC]]
} else {
  fsToUse <- fsList[["Transformed"]]
}

if (!is.null(downsampleTo)) {
  message("🔽 Downsampling to ", downsampleTo, " cells per sample.")
  fsToUse <- fsApply(fsToUse, function(x) {
    set.seed(2)
    n_events <- nrow(x)
    sample_size <- min(downsampleTo, n_events)
    sampled_indices <- sample(n_events, sample_size)
    x[sampled_indices, ]
  })
}

message("📦 Creating SingleCellExperiment objects with CATALYST::prepData...")

sce <- CATALYST::prepData(
  x = fsToUse,
  panel = panel,
  md = md,
  features = panel$fcs_colname,
  transform = FALSE,
  md_cols = list(factors = conditions)
)
# exprs and exprsTransformed are identical to counts at this stage (transformation
# was applied at the flowFrame level, before prepData).  Share the same matrix
# object to avoid tripling memory — R's copy-on-modify means no duplication until
# one is actually written to (which the pipeline does not do).
.shared_mat <- assay(sce, "counts")
assay(sce, "exprs") <- .shared_mat
assay(sce, "exprsTransformed") <- .shared_mat
rm(.shared_mat)

# 99% quantile normalisation
message("📐 Applying quantile normalisation...")

exprs_mat <- assay(sce, "exprs")
qs <- matrixStats::rowQuantiles(exprs_mat, probs = c(0.01, 0.99))
norm_exprs <- sweep(exprs_mat, 1, qs[, 1], "-")
norm_exprs <- sweep(norm_exprs, 1, qs[, 2] - qs[, 1], "/")
norm_exprs[norm_exprs < 0 | is.na(norm_exprs)] <- 0
norm_exprs[norm_exprs > 1] <- 1

assay(sce, "exprsQuantNorm") <- norm_exprs

# Also include a scaled version
scale.coef <- matrixStats::rowQuantiles(exprs_mat, probs = c(0.01, 0.99))
assay(sce, "norm") <- t(scale(t(exprs_mat), center = scale.coef[, 1], scale = scale.coef[, 2] - scale.coef[, 1]))

# Free intermediates and bulky framesList entries no longer needed for analysis.
# framesList.qs (for FCS export) is saved separately; fsList/fsToUse are done.
rm(exprs_mat, norm_exprs, qs, scale.coef)
if (exists("fsList")) rm(fsList)
if (exists("fsToUse")) rm(fsToUse)
invisible(gc())
```

```{r select_assay_and_markers, results = "hide"}

# Define which markers to use for clustering
if (!is.null(featureType) && markersToClusterBy == "type") {
  featuresToCluster <- panel$marker_name[panel$marker_class == "type"]
} else if (!is.null(featureType) && markersToClusterBy == "state") {
  featuresToCluster <- panel$marker_name[panel$marker_class == "state"]
} else if (markersToClusterBy %in% c("type", "state", "all")) {
  featuresToCluster <- panel$marker_name
} else {
  warning("⚠️  Unknown value for `markersToClusterBy`, falling back to all markers.")
  featuresToCluster <- panel$marker_name
}

if(quantileNormaliseAll) {
  assay(sce, "exprs") <- assay(sce, "exprsQuantNorm")
}
set.seed(2)
if (quantileNormaliseAll) {
  exprsToUse = "exprsQuantNorm"
} else {
  exprsToUse = "exprsTransformed"
}

# Define condition levels early (needed for fig.height calculations in later chunks)
conds <- as.factor(sce@metadata$experiment_info$condition)
condLevels <- levels(conds)
```

```{r clustering_graph, results = "hide", eval = is.null(RDataFolder)}

if (clusteringMethodToUse %in% c("MfastPG", "Mphenograph", "Rphenograph")) {
  message("🧠 Running clustering using ", clusteringMethodToUse, "...")

  # Expression matrix for clustering
  es <- t(assay(sce, exprsToUse)[featuresToCluster, ])

  cluster_fn <- function(k) {
    set.seed(2)
    if (clusteringMethodToUse == "MfastPG") {
      clusters <- MARMOT::MfastPG(data = es, k = k, seed = 2, verbose = TRUE)
    } else if (clusteringMethodToUse == "Mphenograph") {
      clusters <- MARMOT::Mphenograph(data = es, k = k, seed = 2, verbose = TRUE)
    } else if (clusteringMethodToUse == "Rphenograph") {
      clusters <- Rphenograph::Rphenograph(data = es, k = k)
    }
    ids <- factor(as.character(igraph::membership(clusters[[2]])))
    ids <- factor(ids, levels = gtools::mixedsort(levels(ids)))
    list(kids = ids, clusters = clusters)
  }

  kList <- setNames(lapply(kValuesIWant, cluster_fn), paste0("k", kValuesIWant))

  # Store all clusterings in SCE object
  for (k in names(kList)) {
    sce[[k]] <- kList[[k]]$kids
  }

  # Select default cluster to use
  selected_k <- paste0("k", knn)
  sce$cluster_id <- sce[[selected_k]]
  
  metadata(sce)$cluster_codes <- eval(parse(text = paste0(
    "data.frame(", clusteringMethodToUse , " = factor(levels(sce[[selected_k]])))"
  )))
  kids <- sce[[paste0("k", knn)]]
  numbersOfClusters <- gtools::mixedsort(levels(kids))
}
```

```{r parc_clustering, results = "hide", dependson = "input settings", eval = is.null(RDataFolder)}

if (clusteringMethodToUse == "PARC") {
  if (!MARMOT:::use_marmot_python()) {
    message("Python PARC unavailable -- falling back to Mparc (R).")
    clusteringMethodToUse <- "Mparc"
  } else {
  message("🐍 Running PARC clustering...")

  set.seed(2)
  esPARC <- t(assay(sce, exprsToUse)[featuresToCluster, ])

  # Run PARC in an isolated Python subprocess. Same process as R would collide
  # libomp runtimes (R's data.table libomp vs conda's annoy/hnswlib libomp on
  # macOS arm64) and segfault. The subprocess writes/reads raw float64 via
  # temp files so reticulate is bypassed entirely.
  parcClustList <- setNames(lapply(kValuesIWant, function(k) {
    labels <- MARMOT:::run_parc_subprocess(
      esPARC, knn = k, num_threads = nCores
    )
    as.integer(labels)
  }), paste0("p", kValuesIWant))

  for (k in kValuesIWant) {
    cluster_labels <- factor(as.character(parcClustList[[paste0("p", k)]] + 1))
    cluster_labels <- factor(cluster_labels, levels = gtools::mixedsort(levels(cluster_labels)))
    sce[[paste0("p", k)]] <- cluster_labels
  }

  chosen_k <- paste0("p", knn)
  sce$cluster_id <- sce[[chosen_k]]
  metadata(sce)$cluster_codes <- eval(parse(text = paste0(
    "data.frame(", clusteringMethodToUse, " = factor(levels(sce[[chosen_k]])))"
  )))
  kids <- sce[[paste0("p", knn)]]
  numbersOfClusters <- gtools::mixedsort(levels(kids))
  }
}
```

```{r mparc_clustering, results = "hide", dependson = "input settings", eval = is.null(RDataFolder)}

if (clusteringMethodToUse == "Mparc") {
  message("Running Mparc clustering (R)...")

  set.seed(2)
  esPARC <- t(assay(sce, exprsToUse)[featuresToCluster, ])

  parcClustList <- setNames(lapply(kValuesIWant, function(k) {
    Mparc(esPARC, k = as.integer(k), num_threads = as.integer(nCores))
  }), paste0("p", kValuesIWant))

  for (k in kValuesIWant) {
    cluster_labels <- factor(as.character(parcClustList[[paste0("p", k)]]))
    cluster_labels <- factor(cluster_labels, levels = gtools::mixedsort(levels(cluster_labels)))
    sce[[paste0("p", k)]] <- cluster_labels
  }

  chosen_k <- paste0("p", knn)
  sce$cluster_id <- sce[[chosen_k]]
  metadata(sce)$cluster_codes <- eval(parse(text = paste0(
    "data.frame(", clusteringMethodToUse, " = factor(levels(sce[[chosen_k]])))"
  )))
  kids <- sce[[paste0("p", knn)]]
  numbersOfClusters <- gtools::mixedsort(levels(kids))
}
```

```{r FlowSOM, results = "hide", dependson = "input settings", eval = is.null(RDataFolder)}

if (clusteringMethodToUse == "FlowSOM") {
  message("🌾 Running FlowSOM clustering...")

  sceSOM <- cluster(
    sce,
    features = featuresToCluster,
    xdim = 10,
    ydim = 10,
    maxK = max(kValuesIWant),
    seed = 42
  )

  # Store all clusterings with "k" prefix
  for (k in kValuesIWant) {
    raw_ids <- cluster_ids(sceSOM, paste0("meta", k))
    prefixed <- factor(as.character(raw_ids))
    prefixed <- factor(prefixed, levels = gtools::mixedsort(levels(prefixed)))
    sce[[paste0("meta", k)]] <- prefixed
  }

  # Assign selected clustering
  selected_k <- paste0("meta", knn)
  sce$cluster_id <- sce[[selected_k]]

  metadata(sce)$cluster_codes <- data.frame(FlowSOM = factor(levels(sce[[selected_k]])))
  kids <- sce[[paste0("meta", knn)]]
  numbersOfClusters <- gtools::mixedsort(levels(kids))
}
```

```{r derive_cluster_info, include = FALSE}
# Derive kids / numbersOfClusters from SCE — works for both fresh run and reload.
# Do NOT use the kids variable set inside each clustering chunk: the PARC path
# stores raw integers (not a factor), and on reload all clustering chunks are skipped.
kids <- sce$cluster_id
numbersOfClusters <- gtools::mixedsort(levels(kids))

# Derive featuresToUmap from panel — works for both fresh run and reload.
# On fresh run this is redundant (dim_reduction chunk sets it too), but on reload
# the dim_reduction chunk is skipped and featuresToUmap would be missing.
if (!is.null(featureType) && markersToDimRedBy == "type") {
  featuresToUmap <- panel$marker_name[panel$marker_class == "type"]
} else if (!is.null(featureType) && markersToDimRedBy == "state") {
  featuresToUmap <- panel$marker_name[panel$marker_class == "state"]
} else {
  featuresToUmap <- panel$marker_name
}
```

```{r dim_reduction, results = "hide", dependson = "input settings", eval = is.null(RDataFolder)}

message("📉 Running dimensionality reduction: ", dimRedMethodToUse)

# Fix seed so DR is identical regardless of which clustering method ran before this
set.seed(2)

# Select features for DR
if (!is.null(featureType) && markersToDimRedBy == "type") {
  featuresToUmap <- panel$marker_name[panel$marker_class == "type"]
} else if (!is.null(featureType) && markersToDimRedBy == "state") {
  featuresToUmap <- panel$marker_name[panel$marker_class == "state"]
} else {
  featuresToUmap <- panel$marker_name
}

n_cells_dr <- ncol(sce)

# Always run PCA first (needed for pacmap, helpful for others)
message("  ▸ PCA on ", n_cells_dr, " cells × ", length(featuresToUmap), " features...")
t0 <- Sys.time()
sce <- runDR(x = sce, dr = "PCA", features = featuresToUmap, assay = exprsToUse)
message("    done (", format(round(Sys.time() - t0, 1)), ")")

# PaCMAP — runs in an isolated Python subprocess (see R/run_python_dr.R).
# In-process reticulate collides libomp runtimes with R's data.table on
# macOS arm64 and segfaults during ANNOY's NN search. Falls back to Mpacmap
# (pure R) if the p4r conda env is missing or the subprocess errors.
if (dimRedMethodToUse == "pacmap") {
  es <- as.matrix(t(assay(sce, exprsToUse)[featuresToUmap, ]))
  embedding <- MARMOT:::run_pacmap_subprocess(es, apply_pca = FALSE, verbose = TRUE)
  if (is.null(embedding)) {
    message("PaCMAP failed -- falling back to Mpacmap (R).")
    dimRedMethodToUse <- "Mpacmap"
  } else {
    colnames(embedding) <- c("x", "y")
    reducedDim(sce, dimRedMethodToUse) <- embedding
  }
}

# Mpacmap (pure R) — also used as fallback when Python PaCMAP is unavailable
if (dimRedMethodToUse == "Mpacmap") {
  message("Running Mpacmap (R)...")
  es <- as.data.frame(t(assay(sce, exprsToUse)[featuresToUmap, ]))
  embedding <- Mpacmap(as.matrix(es), apply_pca = FALSE, verbose = TRUE)
  colnames(embedding) <- c("x", "y")
  reducedDim(sce, "Mpacmap") <- embedding
}

# Native DRs
sce <- runDR(x = sce, dr = "UMAP", features = featuresToUmap, assay = exprsToUse)
sce <- runDR(x = sce, dr = "TSNE", features = featuresToUmap, assay = exprsToUse)
```

```{r, label = "make_seurat_object", include = FALSE}
if (is.null(RDataFolder) && runScGate) {
  if (requireNamespace("Seurat", quietly = TRUE)) {
    message("🖼️ Generating Seurat object for scGate...")
    scData <- suppressWarnings(Seurat::as.Seurat(sce, counts = "exprsTransformed", data = "exprsTransformed"))
    scData <- Seurat::ScaleData(scData, assay = "originalexp", verbose = FALSE)
  } else {
    warning("Seurat is not installed — skipping scGate.")
    runScGate <- FALSE
  }
}
```

```{r, label = "run_scGate", include = FALSE}
if (is.null(RDataFolder) && runScGate) {
  if (requireNamespace("scGate", quietly = TRUE) && requireNamespace("UCell", quietly = TRUE)) {
    message("Running scGate on marker pairs...")
    mp <- parse_marker_pairs(smd)

    if (is.null(mp) || length(mp$models) == 0) {
      warning("No marker pairs defined in metadata — skipping scGate.")
      runScGate <- FALSE
    } else {
      models <- mp$models

    for (x in names(models)) {
      if (!models[[x]][[1]] %in% rownames(scData)) {
        hit <- agrep(models[[x]][[1]], rownames(scData), value = TRUE)
        if (length(hit) == 0) {
          warning("No marker match for '", models[[x]][[1]], "' in model '", x, "' — skipping model.")
          next
        }
        models[[x]][[1]] <- hit[[1]]
      }
      if (!models[[x]][[2]] %in% rownames(scData)) {
        hit <- agrep(models[[x]][[2]], rownames(scData), value = TRUE)
        if (length(hit) == 0) {
          warning("No marker match for '", models[[x]][[2]], "' in model '", x, "' — skipping model.")
          next
        }
        models[[x]][[2]] <- hit[[1]]
      }
      model <- scGate::gating_model(name = x, signature = c(models[[x]][[1]], models[[x]][[2]]))
      scData <- scGate::scGate(data = scData, model = model, pos.thr = 0.65)
      y <- gsub("[ -]", "", x)
      # Only mark cells that actually passed the gate (Positive), not all evaluated cells
      scData[[paste0("is_", y)]] <- NA_character_
      scData[[paste0("is_", y)]][which(scData$scGate_multi == "Positive")] <- x
    }

    # Determine Gated_Cells: highest UCell score, but only for cells that passed at least one gate
    ucell_cols <- grep("UCell", colnames(scData@meta.data), value = TRUE)
    is_cols <- paste0("is_", gsub("[ -]", "", names(models)))
    is_cols <- is_cols[is_cols %in% colnames(scData@meta.data)]
    any_positive <- apply(scData@meta.data[, is_cols, drop = FALSE], 1,
                          function(row) any(!is.na(row)))
    if (length(ucell_cols) == 0) {
      warning("No UCell score columns found — assigning Gated_Cells from is_* columns only.")
      scData$Gated_Cells <- ifelse(
        any_positive,
        apply(scData@meta.data[, is_cols, drop = FALSE], 1,
              function(row) { v <- row[!is.na(row)]; if (length(v)) v[[1]] else "Negative" }),
        "Negative"
      )
    } else {
      scData$Gated_Cells <- ifelse(
        any_positive,
        sapply(seq_len(nrow(scData@meta.data)), function(i) {
          gsub("_UCell", "", ucell_cols[which.max(scData@meta.data[i, ucell_cols])])
        }),
        "Negative"
      )
    }

    # Transfer scGate results back to SCE
    sce$Gated_Cells <- scData$Gated_Cells
    scgate_cols <- grep("^is_|UCell", colnames(scData@meta.data), value = TRUE)
    for (col in scgate_cols) {
      sce[[col]] <- scData@meta.data[[col]]
    }

    # Add gate colours
    gate_colours <- setup_scgate_colours(unique(scData$Gated_Cells), names(models), catalystCols)
    for (nm in names(gate_colours)) coloursList[[nm]] <- gate_colours[[nm]]
    } # end: mp non-NULL branch
  } else {
    warning("scGate and/or UCell are not installed — skipping scGate analysis.")
    runScGate <- FALSE
  }
}
```

```{r prepare_dimreds_all, results = "hide", dependson = c("input settings", "make sce"), eval = is.null(RDataFolder)}

message("🧭 Preparing dimensionality reduction dataframes for all methods...")

# Get available DR methods with 2D coords
dr_methods <- reducedDimNames(sce)
valid_methods <- dr_methods[sapply(dr_methods, function(m) ncol(reducedDim(sce, m)) >= 2)]

# Function to build one DR dataframe
build_dr_df <- function(method) {
  coords <- reducedDim(sce, method)[, 1:2]
  colnames(coords) <- c("x", "y")
  df <- data.frame(
    colData(sce),
    sce_idx = seq_len(ncol(sce)),
    coords,
    DRMethod = method,
    check.names = FALSE
  )
  
  # Add cluster annotations
  if (!is.null(metadata(sce)$cluster_codes)) {
    df <- dplyr::left_join(df, cluster_codes(sce), by = c("cluster_id" = clusteringMethodToUse))
  }
  df[[clusteringMethodToUse]] <- df$cluster_id

  # Add quantile-normalised expression
  qn <- t(assay(sce, "exprsQuantNorm"))
  df <- dplyr::bind_cols(df, as.data.frame(qn))
  
  # Drop only rows with missing DR coordinates (not all NAs — scGate is_* columns
  # have NA for negative cells and must be preserved)
  df <- df[!is.na(df$x) & !is.na(df$y), , drop = FALSE]
  
  return(df)
}

# Build full DR list
umapDFList <- lapply(valid_methods, build_dr_df)
names(umapDFList) <- valid_methods

# Optionally downsample per condition
if (!is.null(drCellPerCondition)) {
  set.seed(2)
  for (method in valid_methods) {
    df <- umapDFList[[method]]
    down_df <- df |>
      dplyr::group_by(sample_id) |>
      dplyr::group_modify(~ {
        cond <- unique(md$condition[md$sample_id == .y$sample_id])
        target_n <- as.numeric(drCellPerCondition[as.character(cond)])
        if (nrow(.x) > target_n) .x[sample(nrow(.x), target_n), ] else .x
      }) |>
      dplyr::ungroup()
    down_df <- droplevels(down_df)
    umapDFList[[paste0("Downsampled.", method)]] <- down_df
  }
}

# Create label vectors for each DR method
labsList <- lapply(valid_methods, function(method) {
  source_group <- if (!is.null(drCellPerCondition)) paste0("Downsampled.", method) else method
  df <- umapDFList[[source_group]]
  labs <- df |>
    dplyr::group_by(condition) |>
    dplyr::summarise(n = dplyr::n(), .groups = "drop") |>
    dplyr::mutate(label = paste0(condition, "\n n = ", n)) |>
    tibble::deframe()
  names(labs) <- levels(factor(df$condition))
  return(labs)
})
names(labsList) <- valid_methods

umapDFList$All <- umapDFList[[dimRedMethodToUse]]
umapDFList$Downsampled <- umapDFList[[paste0("Downsampled.", dimRedMethodToUse)]]
```

```{r cluster_colours2, results = "hide", dependson = c("input settings", "generate UMAP df"), eval = is.null(RDataFolder)}

message("🎨 Assigning colours to clusters...")

# Base cluster_id colours
cluster_levels <- levels(factor(umapDFList$All$cluster_id))
coloursList[["cluster_id"]] <- setNames(catalystCols[seq_along(cluster_levels)], cluster_levels)

# Add colours for each k-based clustering
for (k in kValuesIWant) {
  colname <- paste0(mergeBy, k)
  if (colname %in% names(umapDFList$All)) {
    cluster_levels_k <- levels(factor(umapDFList$All[[colname]]))
    coloursList[[colname]] <- setNames(catalystCols[seq_along(cluster_levels_k)], cluster_levels_k)
  }
}
```

```{r da_ds_analysis, results = "hide", dependson = c("input settings", "make sce", "import md file"), eval = is.null(RDataFolder)}
message("🚀 Starting differential abundance (DA) and differential state (DS) analysis...")

daPValToUse <- "p_adj"

# ───────────────────────────────────────────────────────
# 🧹 Tidy condition names and prepare design matrix
ei <- metadata(sce)$experiment_info
ei$condition <- gsub(" ", ".", ei$condition)

ei$conditionToTest <- ei$condition |>
  gsub(pattern = " ", replacement = ".", x = _) |>
  gsub(pattern = "-", replacement = "0", x = _) |>
  gsub(pattern = "\\+", replacement = "1", x = _) |>
  gsub(pattern = "/", replacement = "", x = _) |>
  as.factor()

cTTforEdgeR <- contrastsToTest |>
  gsub(pattern = " ", replacement = ".", x = _) |>
  gsub(pattern = "-", replacement = "0", x = _) |>
  gsub(pattern = "\\+", replacement = "1", x = _) |>
  gsub(pattern = "\\.over\\.", replacement = "-", x = _) |>
  gsub(pattern = "\\.vs\\.", replacement = "-", x = _) |>
  gsub(pattern = "/", replacement = "", x = _)

cTTForPlots <- contrastsToTest

design <- model.matrix(~0 + conditionToTest, data = ei)
colnames(design) <- gsub("conditionToTest", "", colnames(design))

# ───────────────────────────────────────────────────────
# ✍️ Create contrast matrix
message("🧮 Constructing contrasts for edgeR...")
contrasts <- setNames(lapply(cTTforEdgeR, \(ctr) {
  eval(parse(text = paste0("limma::makeContrasts(", ctr, ", levels = design)")))
}), cTTforEdgeR)

# ───────────────────────────────────────────────────────
# 📊 Differential abundance analysis (DA)
message("📈 Running DA analysis using diffcyt...")
daList <- setNames(lapply(seq_along(contrasts), function(i) {
  res <- diffcyt(
    d_input            = sce,
    design             = design,
    contrast           = contrasts[[i]],
    analysis_type      = "DA",
    method_DA          = "diffcyt-DA-edgeR",
    clustering_to_use  = clusteringMethodToUse,
    min_cells          = 5,
    transform          = FALSE,
    normalize          = TRUE
  )
  as.data.frame(rowData(res$res))
}), cTTForPlots)

# ───────────────────────────────────────────────────────
# 🧬 Differential state analysis (DS)
message("🧪 Running DS analysis using diffcyt...")
sceDS <- sce
if (!any(rowData(sceDS)$marker_class == "state")) {
  message("\u26a0\ufe0f No state markers defined \u2014 treating all markers as state for DS analysis.")
  rowData(sceDS)$marker_class <- "state"
}
ds_formula1 <- createFormula(ei, cols_fixed = "conditionToTest", cols_random = "sample_id")

dsList <- setNames(vector("list", length(contrasts)), cTTForPlots)

for (i in seq_along(contrasts)) {
  res_DS <- diffcyt(
    d_input           = sceDS,
    clustering_to_use = clusteringMethodToUse,
    analysis_type     = "DS",
    method_DS         = "diffcyt-DS-limma",
    design            = design,
    contrast          = contrasts[[i]],
    transform         = FALSE,
    normalize         = TRUE,
    verbose           = FALSE
  )
  dsList[[i]] <- list(res_DS = res_DS, tbl_DS = rowData(res_DS$res))
}

# ───────────────────────────────────────────────────────
# 🔗 Merge clusters and rerun DS
message("🔁 Merging clusters and rerunning DS on merged data...")
mergedID <- paste0(clusteringMethodToUse, "_merged")
mergedTable <- data.frame(
  old_cluster = levels(as.factor(sce[[paste0(mergeBy, knn)]])),
  new_cluster = "all"
)

sceDS <- mergeClusters(
  x     = sceDS,
  k     = clusteringMethodToUse,
  id    = mergedID,
  table = mergedTable
)

ds_formula1 <- createFormula(ei, cols_fixed = "conditionToTest")

for (i in seq_along(contrasts)) {
  res_DS <- diffcyt(
    d_input           = sceDS,
    formula           = ds_formula1,
    clustering_to_use = mergedID,
    analysis_type     = "DS",
    method_DS         = "diffcyt-DS-limma",
    design            = design,
    contrast          = contrasts[[i]],
    transform         = FALSE,
    normalize         = TRUE,
    verbose           = FALSE
  )
  dsList[[paste(cTTForPlots[i], "Merged")]] <- list(
    res_DS = res_DS,
    tbl_DS = rowData(res_DS$res)
  )
}

# ───────────────────────────────────────────────────────
# 🔍 Identify significant DA clusters
message("🔬 Extracting significant DA clusters...")

# Pre-allocate list for efficiency
selectedClustersList <- vector("list", length(contrasts) * 2)
names_list <- character(length(contrasts) * 2)

for (i in seq_along(contrasts)) {
  contrast_conditions <- strsplit(contrastsToTest[i], " over ")[[1]]
  da_table <- daList[[i]]
  
  # Filter for complete cases and significant results
  significant_rows <- complete.cases(da_table[c("cluster_id", "logFC", daPValToUse)]) &
                     da_table[[daPValToUse]] <= 0.05
  
  significant_data <- da_table[significant_rows, ]
  
  # Extract clusters by direction
  up_clusters <- as.character(unique(significant_data$cluster_id[significant_data$logFC >= 0]))
  down_clusters <- as.character(unique(significant_data$cluster_id[significant_data$logFC <= 0]))
  
  # Store results
  idx_up <- 2 * i - 1
  idx_down <- 2 * i
  
  selectedClustersList[[idx_up]] <- up_clusters
  selectedClustersList[[idx_down]] <- down_clusters
  
  names_list[idx_up] <- paste0("Contrast ", i, ": ", contrast_conditions[1], " Up")
  names_list[idx_down] <- paste0("Contrast ", i, ": ", contrast_conditions[2], " Down")
}

names(selectedClustersList) <- names_list

# ───────────────────────────────────────────────────────
# 🎨 Assign colours to selected clusters
message("🎨 Assigning colours to selected clusters...")
for (i in seq_along(selectedClustersList)) {
  label <- names(selectedClustersList)[i]
  clusterIDs <- selectedClustersList[[i]]
  sc <- coloursList[[paste0(mergeBy, knn)]][clusterIDs]
  coloursList[[label]] <- sc
}

# 🗺️ Map clusters to colours
message("🗺️ Generating cluster-to-colour mapping list...")
clustersToMapList <- lapply(names(selectedClustersList), function(label) {
  ids <- selectedClustersList[[label]]
  clrs <- coloursList[[label]]
  names(clrs) <- ids
  clrs
})
names(clustersToMapList) <- names(selectedClustersList)

message("✅ DA and DS analysis complete.")

# Save DA/DS results to h5ad (runs even during reload since DA/DS always re-runs)
h5ad_save_dir <- if (!is.null(RDataFolder)) {
  RDataFolder
} else if (exists("rFilesDir")) {
  rFilesDir
} else NULL

if (!is.null(h5ad_save_dir)) {
  h5ad_path <- file.path(h5ad_save_dir, "marmot_results.h5ad")
  if (file.exists(h5ad_path)) {
    message("💾 Saving updated DA/DS results to h5ad...")
    MARMOT::save_da_ds_h5ad(h5ad_path, daList, dsList, selectedClustersList, daPValToUse)
  }
}
```

```{r output_directories, results = "hide"}

# Create Results root directory if not resuming from saved RData
if (is.null(RDataFolder)) {
  if (is.null(resultsDir)) {
    # Manual run (RStudio) — create results dir with timestamp
    resultsDir <- file.path(fp, paste0("Results_Files_", timeRun))
    dir.create(resultsDir, showWarnings = FALSE, recursive = TRUE)
  }

  rFilesDir <- file.path(resultsDir, "R_files")
  dir.create(rFilesDir, showWarnings = FALSE)
}

# When reloading, RDataFolder points to R_files/ — go up one level for sibling dirs
reloadResultsDir <- if (!is.null(RDataFolder)) dirname(RDataFolder) else NULL

# PDFs
if (gimmePDFs) {
  pdfDir <- if (is.null(RDataFolder)) {
    file.path(resultsDir, "PDF_figures")
  } else {
    file.path(reloadResultsDir, "PDF_figures")
  }
  dir.create(pdfDir, showWarnings = FALSE)
}

# Excel outputs
excelDir <- if (is.null(RDataFolder)) {
  file.path(resultsDir, "Excel_Files")
} else {
  file.path(reloadResultsDir, "Excel_Files")
}
dir.create(excelDir, showWarnings = FALSE)
```

```{r write_results_to_file, eval = is.null(RDataFolder), include = FALSE}

message("💾 Saving outputs...")

# Save FCS frame objects as qs2 — flowFrame/flowSet cannot be stored in Parquet
if (exists("framesList", envir = .GlobalEnv)) {
  out_path <- file.path(rFilesDir, "framesList.qs2")
  qs2::qs_save(get("framesList", envir = .GlobalEnv), file = out_path,
               nthreads = nCores)
  message("Saved (qs2): framesList")
}

# Save DA/DS result objects as qs2 — S4 structures can't round-trip through h5ad
da_ds_save <- list(
  daList = if (exists("daList")) daList else NULL,
  dsList = if (exists("dsList")) dsList else NULL,
  selectedClustersList = if (exists("selectedClustersList")) selectedClustersList else NULL,
  clustersToMapList = if (exists("clustersToMapList")) clustersToMapList else NULL,
  sceDS = if (exists("sceDS")) sceDS else NULL,
  daPValToUse = if (exists("daPValToUse")) daPValToUse else "p_adj"
)
qs2::qs_save(da_ds_save, file.path(rFilesDir, "da_ds_results.qs2"), nthreads = nCores)
message("Saved (qs2): da_ds_results.qs2")

# Save h5ad — main data store for Shiny app and reproducibility
message("💾 Saving h5ad format...")
MARMOT::save_h5ad_data(rFilesDir, envir = .GlobalEnv)

# Write pipeline_settings.json sidecar so Tauri GUI can reload & lock fields
settings_list <- list(
  clusteringMethodToUse = as.character(clusteringMethodToUse),
  markersToClusterBy    = as.character(markersToClusterBy),
  kValuesIWant          = paste(kValuesIWant, collapse = ", "),
  knn                   = as.character(knn),
  dimRedMethodToUse     = as.character(dimRedMethodToUse),
  markersToDimRedBy     = as.character(markersToDimRedBy),
  runQC                 = as.character(runQC),
  useQC                 = toupper(as.character(useQC)),
  runScGate             = toupper(as.character(runScGate)),
  downsampleTo          = if (is.null(downsampleTo)) "" else as.character(downsampleTo),
  quantileNormaliseAll  = toupper(as.character(quantileNormaliseAll)),
  gimmePDFs             = toupper(as.character(gimmePDFs)),
  greyscalePlots        = toupper(as.character(greyscalePlots)),
  themeToUse            = as.character(themeToUse),
  viridisColour         = as.character(viridisColour),
  nCores                = as.character(nCores)
)
jsonlite::write_json(settings_list, file.path(rFilesDir, "pipeline_settings.json"),
                     auto_unbox = TRUE, pretty = TRUE)
message("Saved: pipeline_settings.json")

# Copy metadata Excel into results folder for self-contained archive
file.copy(md_fp, resultsDir, overwrite = TRUE)
```

```{r nice_umap_function}

# Adaptive point size: 25/sqrt(n) clamped to [0.05, 0.5]
# ~1000 cells → 0.5, ~10k → 0.25, ~50k → 0.11, ~100k → 0.08, ~500k+ → 0.05
n_dr_cells <- nrow(umapDFList[[ifelse(is.null(drCellPerCondition), "All", "Downsampled")]])
dr_point_size <- max(0.05, min(0.5, 25 / sqrt(n_dr_cells)))

plotUMAPNice <- function(
  df = umapDF,
  colour_by = "condition",
  show_density = FALSE,
  size = dr_point_size,
  alpha = 1,
  text_size = 12,
  facet_by = NULL,
  show_grids = FALSE,
  name = "Normalised\nExpression\n",
  show_cluster_labels = FALSE,
  cluster_labels = "cluster_id",
  border = FALSE,
  border_size = 3,
  border_density = 1,
  border_colour = "black"
) {
  gp <- ggplot(df, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[[colour_by]]))

  # Density overlay
  if (show_density) {
    gp <- gp +
      geom_density_2d(
        data = df[, c("x", "y")],
        aes(x = x, y = y),
        colour = "lightgrey",
        size = 0.5,
        bins = 20,
        alpha = 0.8
      ) +
      coord_cartesian(
        xlim = range(df$x, na.rm = TRUE) + c(-1, 1),
        ylim = range(df$y, na.rm = TRUE) + c(-1, 1)
      )
  }

  if (border && nrow(df) > 10) {
    kde <- MASS::kde2d(df[["x"]], df[["y"]], n = 100L)
    ix  <- pmax(1L, pmin(findInterval(df[["x"]], kde$x), length(kde$x)))
    iy  <- pmax(1L, pmin(findInterval(df[["y"]], kde$y), length(kde$y)))
    cell_density <- kde$z[cbind(ix, iy)]
    border_df <- df[cell_density < quantile(cell_density, border_density), ]
    gp <- gp +
      geom_point(data = border_df, size = size * border_size,
                 colour = border_colour, show.legend = FALSE) +
      geom_point(colour = "grey75", size = size, show.legend = FALSE) +
      geom_point(size = size, alpha = alpha)
  } else {
    gp <- gp + geom_point(size = size, alpha = alpha, shape = 16)
  }

  gp <- gp +
    labs(x = paste(dimRedMethodToUse, "dim 1"),
         y = paste(dimRedMethodToUse, "dim 2"),
         title = colour_by) +
    coord_fixed() +
    theme_prism() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      legend.text = element_text(size = 10),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5)
    )

  # Guides
  if (colour_by == "condition") {
    gp <- gp + guides(colour = guide_legend(override.aes = list(size = 2)))
  }

  # Cluster label overlays
  if (show_cluster_labels) {
    label_coords <- data.frame(
      df |>
        dplyr::group_by(.data[[cluster_labels]]) |>
        dplyr::summarise(x = median(x), y = median(y)))
    if (colour_by %in% useMarkers) {
      label_coords[[colour_by]] <- lapply(label_coords[[cluster_labels]], function(level) {
        mean(df[[colour_by]][df[[cluster_labels]] == level])
      }) |> unlist()
      gp <- gp + geom_label_repel(data = label_coords, aes(x = x, y = y, label = .data[[cluster_labels]], fill = .data[[colour_by]]), colour = "black", show.legend = FALSE)
    } else {
      gp <- gp + geom_label_repel(data = label_coords, aes(x = x, y = y, label = .data[[cluster_labels]], fill = .data[[cluster_labels]]), colour = "black", show.legend = FALSE)
    }
  }

  if (!show_grids) {
    gp <- gp +
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  }

  # Gradient or manual colour scale
  if (colour_by %in% useMarkers) {
    gp <- gp + scale_colour_gradientn(name, colours = get_palette_colours(viridisColour))
    gp <- gp + scale_fill_gradientn(name, colours = get_palette_colours(viridisColour))
  } else {
    gp <- gp + scale_colour_manual(values = coloursList[[colour_by]])
    gp <- gp + scale_fill_manual(values = coloursList[[colour_by]])
  }

  # Faceting
  if (!is.null(facet_by)) {
    gp <- gp + facet_wrap(facets = as.formula(paste("~", facet_by)))
  }

  return(gp)
}
```

Started on `r paste(timeRun)`

::::::::::::: panel-tabset
## QC

:::::::::: panel-tabset

```{r, label = "print QC", results = 'asis'}
if (exists("qc_skip_reason")) {
  cat("### QC Cell Filtering \n\n")
  cat("::: {.callout-warning title='QC Skipped'}\n\n")
  cat(qc_skip_reason)
  cat("\n\n:::\n\n")
} else if (runQC %in% c("FlowAI", "PeacoQC") && exists("QCmini")) {
  cat("### QC Cell Filtering \n\n")
  if(runQC %in% c("FlowAI", "PeacoQC") && !useQC) {
    cat("\n\n")
    cat("::: {.callout-warning title='QC Not Applied'}\n\n")
    cat("You ran QC using", runQC, ", but chose **not** to apply it.",
        "The data here is presented purely for your information and does not reflect any of the results in this report.",
        "If you wish to also use only the cells that passed QC, please change the `useQC` variable to TRUE in the metadata file.")
    cat("\n:::\n\n")
  }
  print(kable(QCmini, format = "html", escape = FALSE, col.names = gsub("\\.", "", names(QCmini))) |> kable_styling())
  cat("\n\n")
  openxlsx::write.xlsx(as.data.frame(QCmini), file = file.path(excelDir, "QCtable.xlsx"))
} else if (!is.null(RDataFolder)) {
  cat("### QC Cell Filtering \n\n")
  cat("QC data not available on reload.\n\n")
}
```

### Cofactor Histograms

:::::::: panel-tabset

```{r, label = "print cofactor plots", results = 'asis', fig.height=12, fig.width=12, eval = exists("cofactorPlotList")}
for (marker in useMarkers) {
  cat("####", marker, "\n")
  cat(paste(marker, "Cofactor you chose:", cf2[[marker]]))
  cat("\n\n")
  grid.arrange(grobs = cofactorPlotList[[marker]], ncol = 3, top = marker)
  cat("\n\n")
}
```

```{r, label = "cofactor_dims", include = FALSE}
nrowsCofactors <- ceiling(length(useMarkers)/4)
```

::::::::

### Marker Histograms

These are the histograms of all markers using the selected cofactors. Downsampled to 2000 cells per sample.

```{r, label = "plot marker histograms 2", fig.height = (nrowsCofactors*2)+1, fig.width = 15}
y <- assay(sce, "exprs")
df <- data.frame(t(y), colData(sce), check.names = FALSE)
gg_df <- melt(df, value.name = "value", variable.name = "antigen", 
  id.vars = names(colData(sce)))
mh <- ggplot(gg_df, aes(x = value, y = after_stat(ndensity), fill = condition, color = condition)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~antigen, scales = "free") + 
  ylab("Normalised Density") + 
  xlab("Transformed Value") +
  theme_prism(axis_text_angle = 90) + theme(panel.grid = element_blank(),
    strip.background = element_blank(), strip.text = element_text(face = "bold"),
    axis.text = element_text(color = "black"), axis.title = element_text(color = "black"),
    text = element_text(size = 15)) +
  scale_colour_manual(values = coloursList$condition) +
  scale_fill_manual(values = coloursList$condition)
mh
```

```{r, label = "save marker histogram", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "marker_histograms.pdf"), width = 15, height = (nrowsCofactors*2), bg = "transparent")
  print(mh)
  dev.off()
}
```

### Cell Counts

::: {.callout-note collapse="true" title="Sample and Cell Counts"}

The number of samples per condition is:

```{r, label = "print cell counts1"}
table(md$condition)
```

```{r, label = "warn qc counts"}
if (useQC && !is.null(downsampleTo)) {
  cat("\n\n")
  print("You chose to downsample AND use the QC-filtered counts. Therefore, cells were QC'd and then downsampled. The numbers here reflect this two-step approach.")
  cat("\n\n")
}
```

The number of cells per condition is:

```{r, label = "print cell counts2"}
table(sce$condition)
```

The number of cells per sample is:

```{r, label = "print cell counts3"}
n_cells(sce)
```
:::

```{r, label = "plot cell counts 1", fig.height=5, fig.width=max(6, nrow(md) * 0.6)}
plotCounts(sce, color_by = "condition") + scale_fill_manual(values = coloursList$condition) + theme_prism(axis_text_angle = 45)
```

```{r, label = "plot cell counts 2", fig.height=5, fig.width=max(6, nrow(md) * 0.6)}
plotCounts(sce, color_by = "condition", group_by = "sample_id") + scale_fill_manual(values = coloursList$condition) + theme_prism(axis_text_angle = 45)
```

```{r, label = "plot full counts", results = 'asis', fig.height=5, fig.width=max(6, nrow(md) * 0.6)}
if (!is.null(downsampleTo) && exists("originalCellCounts")) {
  cat("\n\nHowever, because you downsampled, here are those numbers for the full dataset. The black line is what you downsampled to.\n\n")
  cat("\n\n")
  originalCellNumbers <- data.frame(
    file_name = paste0(names(originalCellCounts), ".fcs"),
    Cells = as.numeric(originalCellCounts)
  )
  originalCellNumbers <- dplyr::left_join(originalCellNumbers, md, by = "file_name")
  barplotAllCells <- ggplot(originalCellNumbers, aes(x = sample_id, y = Cells, fill = condition)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = coloursList$condition) +
    labs(x = "Sample ID", y = "Number of Cells", title = "All Cells (before downsampling)") +
    theme_prism(axis_text_angle = 45) +
    geom_hline(yintercept = downsampleTo, linetype = "dashed", colour = "black")
  print(barplotAllCells)
  cat("\n\n")
}
cat("\n\n")
```

### Pseudo-bulk MDS Plots

```{r, label = "build mds plots", include = FALSE}
mdsList <- setNames(lapply(conditions, function(x) {
  mds1 <- CATALYST::pbMDS(x = sce, color_by = x, size_by = TRUE, features = featureType)
  mds1 <- mds1 + scale_colour_manual(values = coloursList[[x]]) + theme_prism(base_size = 16) + theme(panel.grid = element_blank())
  mds1
}), conditions)
```

:::::::: {.panel-tabset}

```{r, label = "plot mds by condition", fig.height=5, fig.width=7, results = 'asis'}
invisible(lapply(names(mdsList), function(x) {
  cat("####", x, "\n")
  print(mdsList[[x]])
  cat("\n\n")
}))
```

::::::::

```{r, label = "save mds plots", include = FALSE}
if (gimmePDFs) {
  lapply(names(mdsList), function(x) {
    pdf(file = file.path(pdfDir, paste0(x, "_MDS.pdf")), width = 7, height = 5, bg = "transparent")
    print(mdsList[[x]])
    dev.off()
  })
}
```

::::::::::

## `r paste(clusteringMethodToUse)` Cluster Analysis

:::::::::: panel-tabset
### K Tree

```{r, label = "clustering_summary", results = 'asis'}
cat("::: {.callout-tip title='Clustering Summary'}\n\n")
cat("- **Method:** ", clusteringMethodToUse, "\n")
cat("- **k values tested:** ", paste(kValuesIWant, collapse = ", "), "\n")
cat("- **Selected k:** ", knn, " → ", length(unique(cluster_ids(sce, clusteringMethodToUse))), " clusters\n")
cat("- **Clustering markers:** ", paste(featuresToCluster, collapse = ", "), "\n")
cat(":::\n\n")
```

```{r, label = "get cluster numbers", include = FALSE}
numbersOfClustersPerKnn <- setNames(lapply(paste0(mergeBy, kValuesIWant), function(x) {
  gtools::mixedsort(levels(sce[[x]]))
}), paste0(mergeBy, kValuesIWant))
```

```{r, label = "clustree", fig.height = ceiling(length(kValuesIWant)*3), fig.width = ceiling(max(lengths(numbersOfClustersPerKnn))/2.5)}
if(length(kValuesIWant) >= 2) {
  # Subset colData to only clustering columns + sample metadata to avoid
  # prefix collisions (e.g. patient_id collides with PARC prefix "p")
  k_cols <- paste0(mergeBy, kValuesIWant)
  keep_cols <- c(k_cols, "cluster_id", "sample_id")
  keep_cols <- intersect(keep_cols, names(colData(sce)))
  sce_ct <- sce
  colData(sce_ct) <- colData(sce_ct)[, keep_cols, drop = FALSE]
  clustree::clustree(sce_ct, prefix = mergeBy, exprs = "exprsTransformed")
}
```

```{r, label = "marker pair scatterplots", results = 'asis', fig.height=5, fig.width=5}
runMarkerPairPlots <- ("type" %in% markerType) && ("state" %in% markerType) && ("Marker Pairs" %in% colnames(smd))
if (runMarkerPairPlots) {
  cat("\n\n")
  cat("### Marker-Pair Scatterplots \n\n")
  es <- as.data.frame(t(assay(sce, "exprsTransformed")))
  es[[mergeBy]] <- sce[[paste0(mergeBy, knn)]]
  set.seed(2)
  if (!is.null(downsampleTo)) {
    if (nrow(es) > downsampleTo) {
      es <- es[sample(rownames(es), downsampleTo),]
    }
  }
  esQN <- as.data.frame(t(assay(sce, "exprsQuantNorm")))
  esQN[[mergeBy]] <- sce[[paste0(mergeBy, knn)]]
  if (!is.null(downsampleTo)) {
    if (nrow(esQN) > downsampleTo) {
      esQN <- esQN[sample(rownames(esQN), downsampleTo),]
    }
  }
  
  mp <- parse_marker_pairs(smd)
  markerPairs <- mp$pairs
  markerPairTypes <- mp$types
  markerPairsList <- setNames(lapply(markerPairTypes, function(x) {
    toPlot <- markerPairs[grep(paste0("^", x), markerPairs)]
    plots <- setNames(lapply(toPlot, function(y) {
      markersToPlot <- strsplit(strsplit(y, "\\: ")[[1]][[2]], " ")[[1]]
      clr_vals <- coloursList[[paste0(mergeBy, knn)]]
      if (!quantileNormaliseAll) {
        median_es <- do.call(rbind, lapply(unique(es[[mergeBy]]), function(cl) {
          data.frame(V1 = cl, x = median(es[[markersToPlot[1]]][es[[mergeBy]] == cl]),
                     y = median(es[[markersToPlot[2]]][es[[mergeBy]] == cl]))
        }))
        colnames(median_es)[1] <- mergeBy
        q1 <- ggplot(es, aes(x = .data[[markersToPlot[1]]], y = .data[[markersToPlot[2]]], color = .data[[mergeBy]])) +
          geom_point(size = 0.001) +
          geom_label_repel(data = median_es, aes(x = x, y = y, label = .data[[mergeBy]], fill = .data[[mergeBy]]),
                           colour = "black", show.legend = FALSE, size = 4) +
          scale_color_manual(values = clr_vals) +
          scale_fill_manual(values = clr_vals) +
          theme_prism() +
          theme(legend.position = "hide") +
          ggtitle(paste0(x, "\nTransformed"))
      }
      median_esQN <- do.call(rbind, lapply(unique(esQN[[mergeBy]]), function(cl) {
        data.frame(V1 = cl, x = median(esQN[[markersToPlot[1]]][esQN[[mergeBy]] == cl]),
                   y = median(esQN[[markersToPlot[2]]][esQN[[mergeBy]] == cl]))
      }))
      colnames(median_esQN)[1] <- mergeBy
      p1 <- ggplot(esQN, aes(x = .data[[markersToPlot[1]]], y = .data[[markersToPlot[2]]], color = .data[[mergeBy]])) +
        geom_point(size = 0.001) +
        geom_label_repel(data = median_esQN, aes(x = x, y = y, label = .data[[mergeBy]], fill = .data[[mergeBy]]),
                         colour = "black", show.legend = FALSE, size = 4) +
        scale_color_manual(values = clr_vals) +
        scale_fill_manual(values = clr_vals) +
        theme_prism() +
        theme(legend.position = "hide") +
        ggtitle(paste0(x, "\nTransformed & QN"))
      if (!quantileNormaliseAll) {
        return(list(q1, p1))
      } else {
        return(list(p1))
      }
    }), toPlot)
  }), markerPairTypes)
  
  cat(":::::::: {.panel-tabset}\n\n")
  lapply(names(markerPairsList), function(x) {
    cat("####", x, " \n\n")
    cat(":::::: {.panel-tabset}\n\n")
    lapply(names(markerPairsList[[x]]), function(y) {
      cat("#####", y, " \n\n")
      print(markerPairsList[[x]][[y]][[1]])
      if (length(markerPairsList[[x]][[y]]) == 2)
        print(markerPairsList[[x]][[y]][[2]])
      cat("\n\n")
    })
    cat("::::::\n\n")
    cat("\n\n")
  })
  cat("::::::::\n\n")
  cat("\n\n")
}
```

```{r, label = "save marker pair scatterplots", results = "hide", include = FALSE}
if (gimmePDFs && exists("markerPairsList")) {
  lapply(names(markerPairsList), function(x) {
    lapply(names(markerPairsList[[x]]), function(y) {
      pdf(file = file.path(pdfDir, paste0(y, "_scatter.pdf")), width = 5, height = 5, bg = "transparent")
      print(markerPairsList[[x]][[y]][[1]])
      if (length(markerPairsList[[x]][[y]]) == 2)
        print(markerPairsList[[x]][[y]][[2]])
      dev.off()
    })
  })
}
```

### Heatmap

:::::::: panel-tabset
#### Marker Per Cluster

:::::: panel-tabset

```{r, label = "type marker heatmap", results = 'asis', fig.height=max(6, ceiling(length(numbersOfClusters)/2.5)), fig.width=max(8, ceiling(length(useMarkers)/2.5))}
heatmapsList <- list()
if("type" %in% markerType && "state" %in% markerType) {
  cat("##### Type Markers Only \n\n")
  pET <- plotExprHeatmap(
    x = sce,
    assay = exprsToUse,
    features = "type",
    by = "cluster_id",
    k = clusteringMethodToUse,
    bars = TRUE,
    perc = TRUE,
    col_clust = FALSE,
    hm_pal = hmPalToUse,
    k_pal = coloursList[[paste0(mergeBy, knn)]]
    )
  heatmapsList[["Type"]] <- pET
  ComplexHeatmap::draw(pET, merge_legend = TRUE)
  pETdf <- pET@matrix[row_order(pET),]
  topLineageTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(pETdf))
  )
  topLineageTable$Top_Lineage_Markers <- NA
  for (i in seq_len(nrow(topLineageTable))) {
    row_vals <- pETdf[i,]; topLineageTable$Top_Lineage_Markers[i] <- paste(names(head(sort(row_vals[row_vals > 0.4], decreasing = TRUE), n = 2)), collapse = " ")
  }
  topLineageTable <- topLineageTable[mixedorder(topLineageTable$Cluster), ]
  print(
    htmltools::tagList(
      DT::datatable(
        data = topLineageTable, 
        rownames = FALSE, 
        colnames = c("Cluster", "Top 2 Type Markers"), 
        filter = "top", 
        width = 800, 
        caption = "Top Lineage Markers have to have mean expression > 0.4 to be included, so you may only see 1 marker")
      )
    )
  cat("\n\n")
  
  cat("##### State Markers Only \n\n")
  pES <- plotExprHeatmap(
    x = sce, 
    features = "state", 
    by = "cluster_id", 
    k = clusteringMethodToUse,
    bars = TRUE, 
    perc = TRUE, 
    col_clust = FALSE, 
    hm_pal = hmPalToUse,
    k_pal = coloursList[[paste0(mergeBy, knn)]]
    )
  heatmapsList[["State"]] <- pES
  ComplexHeatmap::draw(pES, merge_legend = TRUE)
  pESdf <- pES@matrix[row_order(pES),]
  topStateTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(pESdf))
  )
  topStateTable$Top_State_Markers <- NA
  for (i in seq_len(nrow(topStateTable))) {
    topStateTable$Top_State_Markers[i] <- paste(names(head(sort(pESdf[i,], decreasing = TRUE), n = 3)), collapse = " ")
    topStateTable$Bottom_State_Markers[i] <- paste(names(tail(sort(pESdf[i,], decreasing = TRUE), n = 3)), collapse = " ")
  }
  topStateTable <- topStateTable[mixedorder(topStateTable$Cluster), ]
  print(
    htmltools::tagList(
      DT::datatable(
        data = topStateTable, 
        rownames = FALSE, 
        colnames = c("Cluster", "Top 3 State Markers", "Bottom 3 State Markers"), 
        filter = "top", 
        width = 800
        )
      )
    )
  cat("\n\n")
}
```

```{r, label = "save type/state heatmaps", results = "hide", include = FALSE}
if (gimmePDFs) {
  if (length(heatmapsList) == 2) {
    for (n in names(heatmapsList)) {
      pdf(file = file.path(pdfDir, paste0(n, "_Marker_Heatmap.pdf")), height = max(6, ceiling(length(numbersOfClusters)/2.5)), width = max(8, ceiling(length(useMarkers)/3)))
      ComplexHeatmap::draw(heatmapsList[[n]], merge_legend = TRUE)
      dev.off()
    }
  }
}
```

##### All Markers

```{r, label = "plot marker expression cluster", fig.height=max(6, ceiling(length(numbersOfClusters)/2.5)), fig.width=max(10, ceiling(length(useMarkers)/2))}
exH <- plotExprHeatmap(x = sce, features = NULL, by = "cluster_id", k = clusteringMethodToUse, bars = TRUE, perc = TRUE, col_clust = FALSE, hm_pal = hmPalToUse, k_pal = coloursList[[paste0(mergeBy, knn)]])
ComplexHeatmap::draw(exH, merge_legend = TRUE)
```

```{r, label = "top marker table", include = TRUE}
if("type" %in% markerType && "state" %in% markerType) {
  topMarkerTable <- dplyr::left_join(topLineageTable, topStateTable)
  openxlsx::write.xlsx(as.data.frame(topMarkerTable), file.path(excelDir, "topMarkerTable.xlsx"))
  DT::datatable(
    data = topMarkerTable, 
    rownames = FALSE, 
    filter = "top", 
    width = 800, 
    colnames = c("Cluster ID", "Top 2 Lineage Markers", "Top 3 State Markers", "Bottom 3 State Markers"),
    caption = "Top Lineage Markers have to have mean expression > 0.4 to be included, so you may only see 1 marker")
} else { 
  f <- NULL
  exT <- plotExprHeatmap(x = sce, features = f, by = "cluster_id", k = clusteringMethodToUse, bars = TRUE, perc = TRUE, col_clust = FALSE)
  exTDF <- exT@matrix[row_order(exT),]
  topMarkerTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(exTDF))
  )
  topMarkerTable$Top_Markers <- NA
  for (i in seq_len(nrow(topMarkerTable))) {
    topMarkerTable$Top_Markers[i] <- paste(names(head(sort(exTDF[i,], decreasing = TRUE), n = 3)), collapse = " ")
  }
  DT::datatable(topMarkerTable, rownames = FALSE, filter = "top", width = 800)
}
```

::::::

```{r, label = "save expression cluster", include = FALSE}
if (gimmePDFs) {
  pdf(
    file = file.path(pdfDir, "marker_per_cluster_heatmap.pdf"),
    width = max(10, ceiling(length(useMarkers)/2)),
    height = max(6, ceiling(length(numbersOfClusters)/2.5)), bg = "transparent"
    )
  ComplexHeatmap::draw(plotExprHeatmap(x = sce, features = NULL, by = "cluster_id", k = clusteringMethodToUse, bars = TRUE, perc = TRUE, col_clust = FALSE, hm_pal = hmPalToUse, k_pal = coloursList[[paste0(mergeBy, knn)]]), merge_legend = TRUE)
  dev.off()
}
```

#### Sample Per Cluster

:::::: panel-tabset

##### All Clusters

```{r, label = "sample per cluster heatmap", fig.height=ceiling(length(numbersOfClusters)/3.5), fig.width=(ceiling(length(md$sample_id)/1.2))}
pFH <- plotFreqHeatmap(
  x = sce,
  k = clusteringMethodToUse,
  normalize = TRUE,
  col_clust = FALSE,
  bars = TRUE,
  perc = TRUE,
  hm_pal = hmPalToUse,
  k_pal = coloursList[[paste0(mergeBy, knn)]]
  )
pFH <- apply_anno_colours(pFH, conditions, coloursList)
pFH
```

```{r, label = "save cluster per sample", include = FALSE}
if (gimmePDFs) {
  pdf(
    file = file.path(pdfDir, "sample_per_cluster_heatmap.pdf"), 
    height = ceiling(length(numbersOfClusters)/4),
    width = (ceiling(length(md$sample_id)/1.6)),
    bg = "transparent"
  )
  print(pFH)
  dev.off()
}
```

::::::
::::::::

### Ridge Plot

```{r, label = "ridge plot", fig.height=ceiling(length(useMarkers)/3)*2.5, fig.width=16}
pCE <- plotClusterExprs(sce, k = clusteringMethodToUse, features = NULL)
pCE$facet$params$ncol <- 12
pCE$facet$params$nrow <- ceiling(length(useMarkers)/12)
pCE + theme_prism()
```

```{r, label = "save ridge plot", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Ridge_Plot.pdf"), width = 16, height = ceiling(length(useMarkers)/3)*2.5)
  print(pCE + theme_prism())
  dev.off()
}
```

### Barplots

:::::::: panel-tabset

#### Cells per Cluster

```{r, label = "cells per cluster", fig.height=4, fig.width=8}
ns <- table(cluster_id = cluster_ids(sce, clusteringMethodToUse), sample_id = sample_ids(sce))
df <- as.data.frame(ns)
df$cluster_id <- factor(df$cluster_id, levels = gtools::mixedsort(unique(as.character(df$cluster_id))))
df$condition <- md$condition[match(df$sample_id, md$sample_id)]
cellsPerClusterBarplot <- ggplot(df, aes(x = cluster_id, y = Freq, fill = cluster_id)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = coloursList[[paste0(mergeBy, knn)]]) + 
  theme_bw() +
  ylab("Number of cells") + 
  xlab("Cluster") + 
  theme_prism(axis_text_angle = 90) +
  theme(legend.position = "hide", panel.grid.major.x = element_blank())
cellsPerClusterBarplot
```

```{r, label = "cells per cluster facet", fig.height = ceiling(length(condLevels)/2)*4, fig.width = 16}
cellsPerClusterBarplot + facet_wrap(~condition, scales = "free") + theme_prism(axis_text_angle = 90) + theme(legend.position = "hide")
```

```{r, label = "cells per cluster facet sample", fig.height=(4*(ceiling(length(unique(df$sample_id)))/3)), fig.width = 18}
cellsPerClusterBarplot + facet_wrap(~sample_id, ncol = 3, scales = "free")
```

```{r, label = "marker frequency per cluster", fig.height=6, include = FALSE}
x1 <- as.data.frame(t(assay(sce, "exprsQuantNorm")))
x1$cluster_id <- cluster_ids(sce, clusteringMethodToUse)

x1mean <- data.frame("cluster_id" = gtools::mixedsort(levels(as.factor(x1$cluster_id))))
for (m in useMarkers) {
  x1mean[[m]] <- NA
  for (cl in levels(as.factor(x1$cluster_id))) {
    x1mean[cl,m] <- mean(x1[[m]][which(x1$cluster_id == cl)])
  }
}
x1rf <- x1mean
for (m in useMarkers) {
  for (cl in levels(as.factor(x1$cluster_id))) {
    x1rf[cl,m] <- x1mean[cl,m]/sum(x1mean[cl,-1])
  }
}
x1meanmelt <- melt(x1mean)
x1meanmelt$cluster_id <- factor(x1meanmelt$cluster_id, levels = unique(x1meanmelt$cluster_id))
x1rfmelt <- melt(x1rf)
x1rfmelt$cluster_id <- factor(x1rfmelt$cluster_id, levels = unique(x1rfmelt$cluster_id))
ggplot(x1rfmelt, aes(x = cluster_id, y = value, fill = variable)) + 
  geom_bar(stat = "identity") + 
  theme_prism() + 
  scale_fill_manual(values = coloursList[[paste0(mergeBy, knn)]]) +
  xlab("Cluster ID") +
  ylab("Proportion of the Quantile Scaled Mean")
```

```{r, label = "cells per cluster per sample table"}
.cells_wide <- tidyr::pivot_wider(df, names_from = cluster_id, values_from = Freq) |> dplyr::select(-condition) |> tibble::column_to_rownames("sample_id") |> t() |> data.frame() |> tibble::rownames_to_column("cluster_id")
DT::datatable(.cells_wide, rownames = FALSE)
openxlsx::write.xlsx(
  x = .cells_wide,
  file = file.path(excelDir, "Cells_Per_Cluster_Per_Sample.xlsx")
)
```

#### Cluster Frequency

```{r, label = "Cluster barplot 1", fig.width = ceiling(length(condLevels)/2)*10, fig.height=5}
{ sorted_ids <- gtools::mixedsort(names(coloursList[[paste0(mergeBy, knn)]]))
  sorted_pal <- coloursList[[paste0(mergeBy, knn)]][sorted_ids]
  plotAbundances(sce, k = clusteringMethodToUse, by = "sample_id", k_pal = coloursList[[paste0(mergeBy, knn)]]) + facet_grid(~condition, space = "free", scales="free") + theme_prism(axis_text_angle = 45) + scale_fill_manual(values = sorted_pal, breaks = sorted_ids) }
```

```{r, label = "save cluster barplot", include = FALSE}
if (gimmePDFs) {
  sorted_ids <- gtools::mixedsort(names(coloursList[[paste0(mergeBy, knn)]]))
  sorted_pal <- coloursList[[paste0(mergeBy, knn)]][sorted_ids]
  pdf(file = file.path(pdfDir, "cluster_barplot.pdf"), width = ceiling(length(condLevels)/2)*10, height = 5, bg = "transparent")
  print(plotAbundances(sce, k = clusteringMethodToUse, by = "sample_id", k_pal = coloursList[[paste0(mergeBy, knn)]]) + facet_grid(~condition, space = "free", scales="free") + theme_prism() + scale_fill_manual(values = sorted_pal, breaks = sorted_ids))
  dev.off()
}
```

```{r, label = "Cluster barplot 2"}
FDR_cutoff <- 0.05

ns <- table(cluster_id = cluster_ids(sce, clusteringMethodToUse), sample_id = sample_ids(sce))
fq <- prop.table(ns, 2) * 100
df <- as.data.frame(fq)
df$cluster_id <- factor(df$cluster_id, levels = gtools::mixedsort(unique(as.character(df$cluster_id))))

m <- match(df$sample_id, sce$sample_id)
for (i in c(NULL, "condition")) df[[i]] <- sce[[i]][m]

dfList <- list()
for (i in seq_along(clustersToMapList)) {
  dfC1 <- df[df$cluster_id %in% names(clustersToMapList[[i]]), ]
  dfList[[names(selectedClustersList)[[i]]]] <- dfC1
}
```

```{r, label = "Cluster barplot 3", fig.width=10, fig.height=5, results = 'asis'}
selectedBarplotList <- setNames(lapply(seq_along(dfList), function(i) {
  cat("\n\n#### DA:", names(dfList)[i], "\n\n")
  if (nrow(dfList[[i]]) >= 1) {
    g <- ggplot(dfList[[i]], aes(x = .data[["sample_id"]], y = .data[["Freq"]], fill = .data[["cluster_id"]]))
    g <- g + geom_bar(stat = "identity")
    g <- g + facet_grid(~condition, scales = "free", space = "free")
    g <- g + scale_fill_manual(values = coloursList[[names(selectedClustersList)[[i]]]])
    g <- g + ggtitle(names(dfList)[i])
    g <- g + scale_x_discrete(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0), labels = seq(0, 100, 25), limits = c(0, 100))
    g <- g + theme_prism(axis_text_angle = 90)
    g <- g + theme(panel.border = element_blank(), panel.spacing.x = unit(1, "lines"))
    g <- g + ylab("Proportion [%]") + xlab("Sample ID")
    print(g)
    cx <- names(dfList)[[i]] |> gsub(pattern = ": .*", replacement = "", x = _) |> gsub(pattern = "Contrast ", replacement = "Contrast_", x = _) |> gsub(pattern = "\\/", replacement = "", x = _)
    sx <- names(dfList)[[i]] |> gsub(pattern = ".*\\: ", replacement = "", x = _) |> gsub(pattern = " ", replacement = "_", x = _) |> gsub(pattern = "\\/", replacement = "", x = _)
    return(list("g" = g, "cx" = cx, "sx" = sx))
  } else {
    cat("There are no DA clusters in this group.\n")
  }
  cat("\n\n")
}), names(dfList))
```

::::::::

```{r, label = "Save Cluster Barplots 3", include = FALSE}
if (gimmePDFs) {
  lapply(names(selectedBarplotList), function(x) {
    if (is.null(selectedBarplotList[[x]])) return(NULL)
    pdf(file = paste0(pdfDir, "/", selectedBarplotList[[x]][["cx"]], "_", selectedBarplotList[[x]][["sx"]], "_barplot.pdf"), width = 10, height = 5)
    print(selectedBarplotList[[x]][["g"]])
    dev.off()
  })
}
```

### Cluster Boxplots

```{r, label = "cluster boxplot dimensions", include = FALSE}
nColsToPlot <- 6
nGroupsToPlot <- nlevels(sce$condition)
nClustersToPlot <- nlevels(sce$cluster_id)
nRowsToPlot <- ceiling(nClustersToPlot/nColsToPlot)
heightToPlot <- ceiling(nRowsToPlot*5)
widthToPlot <- ceiling((5*1.2)*nColsToPlot)
```

```{r, label = "cluster boxplot", fig.width = widthToPlot, fig.height = heightToPlot}
## From CATALYST plotFreqHeatmap function
ns <- table(sce$cluster_id, sce$sample_id)
fq <- prop.table(ns, 2)
fq.z <- as.matrix(unclass(fq))
clusterBoxPlotsDF <- CATALYST:::.z_normalize(asin(sqrt(fq.z)))
clusterBoxPlotsDF <- as.data.frame(t(clusterBoxPlotsDF))
clusterBoxPlotsDF <- tibble::rownames_to_column(clusterBoxPlotsDF, "sample_id") |> dplyr::left_join(md)
clusterBoxPlotsDFMelt <- melt(clusterBoxPlotsDF)

ctsList <- lapply(names(daList), function(da) { strsplit(da, " over ")[[1]] }) |> setNames(names(daList))

clusterPvalues <- lapply(levels(kids), function(k) {
  pvals <- lapply(names(daList), function(da) { daList[[da]]$p_adj[daList[[da]]$cluster_id == k] }) |> setNames(names(daList)) |> unlist()
  if(any(is.na(pvals))) {pvals[is.na(pvals)] <- 1}
  y_max <- max(clusterBoxPlotsDFMelt$value)
  spacing <- (y_max - min(clusterBoxPlotsDFMelt$value)) * 0.2
  y_positions <- y_max + (1:length(ctsList)) * spacing
  pval_df <- data.frame(
    group1 = sapply(ctsList, `[`, 1),
    group2 = sapply(ctsList, `[`, 2),
    p = pvals,
    y.position = y_positions
  )
  pval_df <- add_significance(pval_df, p.col = "p", output.col = "p.signif")
  
  pval_df
}) |> setNames(levels(kids))
maxY <- lapply(seq_along(clusterPvalues), function(i) {clusterPvalues[[i]]$y.position}) |> unlist() |> max()

clusterBoxplotList <- lapply(levels(kids), function(k) {
  ctp <- as.character(k)
  ggplot(clusterBoxPlotsDF, aes(x = .data[["condition"]], y = .data[[ctp]])) + 
    ## geom_vline(xintercept = (seq_along(condLevels)+0.5)[-length(condLevels)], linetype = "dashed", alpha = 0.6) +
    scale_y_continuous(limits = c(min(clusterBoxPlotsDFMelt$value), maxY)) +
    geom_boxplot(aes(fill = condition), alpha = 0.2, show.legend = FALSE, outlier.shape = NA) +
    geom_beeswarm(aes(fill = condition), pch = 21, size = 3, stroke = 0.3, cex = 2, corral = "random", corral.width = 0.9) +
    scale_fill_manual(values = coloursList$condition) +
    stat_pvalue_manual(clusterPvalues[[ctp]], label = "p.signif", tip.length = 0, hide.ns = TRUE, label.size = 9, bracket.size = 0.8, bracket.shorten = 0.1, bracket.nudge.y = -0.1) +
    theme_prism(axis_text_angle = 45) +
    ggtitle(paste("Cluster", ctp)) +
    theme(legend.position = "hide", axis.title = element_blank())
})
ylab <- textGrob("Normalised Frequency", rot = 90, gp = gpar(fontface = "bold", fontsize = 16))
xlab <- textGrob("Condition", gp = gpar(fontface = "bold", fontsize = 16))
grid.arrange(grobs = clusterBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
```

```{r, label = "save cluster boxplot", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Cluster_Boxplot.pdf"), width = widthToPlot, height = heightToPlot)
  grid.arrange(grobs = clusterBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
  dev.off()
}
```

### Markers per Condition

```{r, label = "generate marker tables"}
medianScaledDF <- CATALYST:::.agg(sce, "sample_id", "median", exprsToUse)
markerBoxplotsMeltDF <- melt(medianScaledDF, varnames = c("antigen", "sample_id"[length("sample_id")]))
i <- match(markerBoxplotsMeltDF$sample_id, sce$sample_id)
j <- setdiff(names(colData(sce)), c(names(markerBoxplotsMeltDF), "cluster_id"))
markerBoxplotsMeltDF <- cbind(markerBoxplotsMeltDF, colData(sce)[i, j, drop = FALSE])
medianScaledDF <- medianScaledDF |>
  t() |>
  data.frame(check.names = FALSE) |>
  tibble::rownames_to_column("sample_id") |>
  dplyr::left_join(md, by = "sample_id")
openxlsx::write.xlsx(as.data.frame(medianScaledDF), file = file.path(excelDir, "Median_Markers_Per_Sample_Expression.xlsx"))
```

```{r, label = "draw marker boxplots", fig.height = heightToPlot*0.7, fig.width = widthToPlot}
markerPvalues <- lapply(useMarkers, function(marker) {
  pvals <- lapply(names(dsList)[grep("Merged", names(dsList))], function(ds) { dsList[[ds]]$tbl_DS$p_adj[dsList[[ds]]$tbl_DS$marker_id == marker] }) |> setNames(names(dsList)[grep("Merged", names(dsList))]) |> unlist()
  if(length(pvals) == 0) {pvals <- 1}
  y_max <- max(markerBoxplotsMeltDF$value)
  spacing <- (y_max - min(markerBoxplotsMeltDF$value)) * 0.1
  y_positions <- y_max + (1:length(ctsList)) * spacing
  pval_df <- data.frame(
    group1 = sapply(ctsList, `[`, 1),
    group2 = sapply(ctsList, `[`, 2),
    p = pvals,
    y.position = y_positions
  )
  pval_df <- add_significance(pval_df, p.col = "p", output.col = "p.signif")
  pval_df
}) |> setNames(useMarkers)
maxY <- lapply(seq_along(markerPvalues), function(i) {markerPvalues[[i]]$y.position}) |> unlist() |> max()

markerBoxplotList <- lapply(useMarkers, function(marker) {
  ggplot(medianScaledDF, aes(x = .data[["condition"]], y = .data[[marker]])) + 
    ## geom_vline(xintercept = (seq_along(condLevels)+0.5)[-length(condLevels)], linetype = "dashed", alpha = 0.6) +
    scale_y_continuous(limits = c(min(markerBoxplotsMeltDF$value), maxY)) +
    geom_boxplot(aes(fill = condition), alpha = 0.2, show.legend = FALSE, outlier.shape = NA) +
    geom_beeswarm(aes(fill = condition), pch = 21, size = 3, stroke = 0.3, cex = 2, corral = "random", corral.width = 0.9) +
    scale_fill_manual(values = coloursList$condition) +
    stat_pvalue_manual(markerPvalues[[marker]], label = "p.signif", tip.length = 0, hide.ns = TRUE, label.size = 9, bracket.size = 0.8, bracket.shorten = 0.1, bracket.nudge.y = -0.1) +
    theme_prism(axis_text_angle = 45) +
    ggtitle(marker) +
    theme(legend.position = "hide", axis.title = element_blank())
})
ylabtext <- switch(exprsToUse, "exprsTransformed" = "Transformed", "exprsQuantNorm" = "Quantile Normalised")
ylab <- textGrob(paste0("Median ", ylabtext, " Expression"), rot = 90, gp = gpar(fontface = "bold", fontsize = 16))
xlab <- textGrob("Condition", gp = gpar(fontface = "bold", fontsize = 16))
grid.arrange(grobs = markerBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
```

### Markers within Clusters

:::::::: panel-tabset

```{r, label = "markers per cluster boxplot", fig.width=18, fig.height=5, results = 'asis'}
by <- c("cluster_id", "sample_id")
ms <- CATALYST:::.agg(sce, by, "median", "exprsQuantNorm")

df <- melt(ms, varnames = c("antigen", by[length(by)]))
names(df)[ncol(df)] <- "cluster_id"
i <- match(df$sample_id, sce$sample_id)
j <- setdiff(names(colData(sce)), c(names(df), "cluster_id"))
df <- cbind(df, colData(sce)[i, j, drop = FALSE])
cat("\n\n")
openxlsx::write.xlsx(as.data.frame(df), file = file.path(excelDir, "Median_Markers_Per_Cluster_Expression.xlsx"))
individualClusterMarkerBoxplots <- setNames(lapply(levels(markerBoxplotsMeltDF[[paste0(mergeBy, knn)]]), function(i) {
  cat("#### Cluster", i, "\n\n")
  df1 <- df[df$cluster_id == as.character(i),]
  g1 <- ggplot(df1, aes(x = antigen, y = value, fill = condition)) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0.25, 0.5, 0.75, 1)) +
    geom_hline(yintercept = c(0.25, 0.5, 0.75), linetype = "dashed", alpha = 0.3) +
    geom_vline(xintercept = (seq_along(useMarkers)+0.5)[-length(useMarkers)], linetype = "dotted", alpha = 0.3) +
    geom_boxplot(alpha = 0.2, show.legend = FALSE, outlier.shape = NA) +
    geom_point(position=position_jitterdodge(), pch = 21, size = 2, stroke = 0.2) +
    scale_fill_manual(values = coloursList$condition) +
    theme_prism(axis_text_angle = 45) +
    ylab("Median Quantile Normalised Expression") + xlab("Condition") + ggtitle(paste("Cluster", i))
  print(g1)
  cat("\n\n")
  return("g1" = g1)
}), levels(markerBoxplotsMeltDF[[paste0(mergeBy, knn)]]))
```

```{r, label = "save markers per cluster boxplots", include = FALSE}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, "Markers_per_cluster_boxplots"))
  lapply(names(individualClusterMarkerBoxplots), function(x) {
    pdf(file.path(pdfDir, "Markers_per_cluster_boxplots", paste0(x, ".pdf")), width = 12, height = 5)
    print(individualClusterMarkerBoxplots[[x]])
    dev.off()
  })
}
```

```{r, label = "markers_boxplots2", fig.width=18, fig.height=5, results = "asis"}
for (i in seq_along(selectedClustersList)) {
  if (length(selectedClustersList[[i]]) > 0) {
    cat("####", names(selectedClustersList)[[i]], "\n\n")
    for (j in (selectedClustersList[[i]])) {
      df1 <- df[df$cluster_id == as.character(j),]
      g1 <- ggplot(df1, aes(x = antigen, y = value, fill = condition)) +
        geom_boxplot(alpha = 0.2, show.legend = FALSE, outlier.shape = NA) +
        geom_point(position=position_jitterdodge(), pch = 21, size = 2, stroke = 0.2) +
        geom_hline(yintercept = c(0.25, 0.5, 0.75), linetype = "dashed", alpha = 0.3) +
        geom_vline(xintercept = (seq_along(useMarkers)+0.5)[-length(useMarkers)], linetype = "dotted", alpha = 0.3) +
        scale_fill_manual(values = coloursList$condition) +
        theme_prism(axis_text_angle = 45) +
        ylab("Median Quantile Normalised Expression") + xlab("Condition") + ggtitle(paste("Cluster", j))
      print(g1)
    }
  } else {
    cat("No DA clusters to see here...\n")
  }
  cat("\n\n")
}
```

::::::::

::::::::::

```{r, label = "save marker boxplots", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Marker_per_Sample_boxplot.pdf"), width = widthToPlot, height = heightToPlot)
  grid.arrange(grobs = markerBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
  dev.off()
}
```

## `r paste(dimRedMethodToUse)`

::: {.callout-tip collapse="true" title="Markers used for DR generation"}

```{r, label = "show features to DR"}
paste(featuresToUmap)
```
:::

:::::::::: panel-tabset

### Full `r paste(dimRedMethodToUse)`

:::::::: panel-tabset

```{r, label = "DR all cells", fig.height=6, fig.width=7, results = 'asis'}
if (!is.null(drCellPerCondition)) {
  cat("#### Clusters: All Cells \n\n")
  cat("Since you clustered on all available cells (after initial downsampling), here is a DR plot of all available cells. All subsequent DR plots are subset from this main DR plot.\n\n")
  dr0 <- plotUMAPNice(df = umapDFList$All, colour_by = "cluster_id", show_density = FALSE, alpha = 0.9, show_cluster_labels = TRUE, show_grids = FALSE, text_size = 14) +
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
  dr0
}
```

```{r, label = "save DF all cells", include = FALSE}
if (!is.null(drCellPerCondition)) {
  if (gimmePDFs) {
    pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_All_Cells.pdf")), width = (3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, height = 5)
    print(dr0)
    dev.off()
  }
}
```

```{r, label = "DR all cells unlabelled", fig.height=6, fig.width=(3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, results = 'asis'}
if (!is.null(drCellPerCondition)) {
  cat("\n\n")
  dr0a <- plotUMAPNice(df = umapDFList$All, colour_by = "cluster_id", show_density = FALSE, alpha = 0.9, show_cluster_labels = FALSE, show_grids = FALSE, text_size = 14) +
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
  dr0a
  cat("\n\n")
}
```

```{r, label = "save DF all cells unlabelled", include = FALSE}
if (!is.null(drCellPerCondition)) {
  if (gimmePDFs) {
    pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_All_Cells_unlabelled.pdf")), width = (3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, height = 5)
    print(dr0a)
    dev.off()
  }
}
```

#### Clusters: Selected Cells 

```{r, label = "main cluster DR 1", fig.height=6, fig.width=7}
if (is.null(drCellPerCondition)) {
  umapsDFToPlot <- "All"
} else {
  umapsDFToPlot <- "Downsampled"
}
openxlsx::write.xlsx(x = as.data.frame(umapDFList[[umapsDFToPlot]]), file = file.path(excelDir, paste0(dimRedMethodToUse, "_Coordinates.xlsx")))
dr1a <- plotUMAPNice(
  df = as.data.frame(umapDFList[[umapsDFToPlot]]),
  colour_by = "cluster_id",
  show_density = FALSE,
  alpha = 0.9,
  show_cluster_labels = TRUE,
  show_grids = FALSE,
  text_size = 14
  ) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr1a
```

```{r, label = "main cluster DR 2", fig.height=6, fig.width=7}
dr1b <- plotUMAPNice(df = as.data.frame(umapDFList[[umapsDFToPlot]]), colour_by = "cluster_id", show_density = FALSE, alpha = 0.8, show_cluster_labels = FALSE, show_grids = FALSE, text_size = 14) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr1b
```

```{r, label =  "save full DR", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_full_labelled.pdf")), width = 7, height = 6, bg = "transparent")
  print(dr1a)
  dev.off()
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_full_unlabelled.pdf")), width = 7, height = 6, bg = "transparent")
  print(dr1b)
  dev.off()
}
```

#### All Samples: Other k values

```{r, label = "plot k DRs", fig.height=ceiling(6*length(kValuesIWant)), fig.width=9}
pDRList <- setNames(lapply(kValuesIWant, function(i) {
  plotUMAPNice(df = umapDFList[[umapsDFToPlot]], colour_by = paste0(mergeBy, i), show_grids = FALSE) +
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
}), paste0(mergeBy, kValuesIWant))
plot_grid(plotlist = pDRList, ncol = 1, align = "hv")
```

::::::::

### Condition `r paste(dimRedMethodToUse)`

```{r, label = "make DRs split by condition"}
## Get the number of cells per condition for the UMAP
conds <- as.factor(sce@metadata$experiment_info$condition)
cellN <- n_cells(sce)
names(cellN) <- conds
condLevels <- levels(conds)
condLabels <- NA
for (i in seq_along(condLevels)) {
  condLabels[i] <- paste(condLevels[i], "\n n =", as.numeric(nrow(umapDFList[[umapsDFToPlot]][which(umapDFList[[umapsDFToPlot]]$condition == condLevels[i]),])))
}
names(condLabels) <- condLevels
dr2a <- plotUMAPNice(df = as.data.frame(umapDFList[[umapsDFToPlot]]), colour_by = "cluster_id", show_density = FALSE, alpha = 0.85, text_size = 14, show_cluster_labels = TRUE, show_grids = FALSE) +
  facet_wrap("condition", labeller = labeller("condition" = condLabels), ncol = 2) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
```

```{r, label = "plot DRs split by condition", fig.height=ceiling(length(condLevels)/2)*6, fig.width=12}
dr2a
```

```{r, label = "plot DRs split by condition 2", fig.height=ceiling(length(condLevels)/2)*6, fig.width=12}
dr2b <- plotUMAPNice(df = as.data.frame(umapDFList[[umapsDFToPlot]]), colour_by = "cluster_id", show_density = FALSE, alpha = 0.85, text_size = 14, show_cluster_labels = FALSE, show_grids = FALSE) +
  facet_wrap("condition", labeller = labeller("condition" = condLabels), ncol = 2) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr2b
```

```{r, label = "save DRs split by condition", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_condition_labelled.pdf")), width = 12, height = ceiling(length(condLevels)/2)*6, bg = "transparent")
  print(dr2a)
  dev.off()
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_condition_unlabelled.pdf")), width = 12, height = ceiling(length(condLevels)/2)*6, bg = "transparent")
  print(dr2b)
  dev.off()
}
```

```{r, label = "full DR by condition", fig.height=(6+length(conditionOrder)*2.5), fig.width=7}
u1_by_cond <- plotUMAPNice(umapDFList[[umapsDFToPlot]], "condition")
u2_by_cond <- plotUMAPNice(df = umapDFList[[umapsDFToPlot]], colour_by = "condition", show_density = FALSE) +
  facet_wrap("condition", labeller = labeller("condition" = condLabels))
plot_grid(u1_by_cond, u2_by_cond, ncol = 1, align = "hv", rel_heights = c(1.4, 1))
```

### `r paste(dimRedMethodToUse)` Per Marker

All values have been normalised to 0 - 1.

```{r, label = "DR per marker 1", results = 'asis', include = FALSE}
UMAPList <- setNames(lapply(c("With", "Without"), function(yn) {
  var <- switch(
    yn,
    "With" = TRUE,
    "Without" = FALSE
  )
  l1 <- setNames(lapply(useMarkers, function(marker) {
    m1 <- plotUMAPNice(
      df = as.data.frame(umapDFList[[umapsDFToPlot]]),
      colour_by = marker,
      name = "Normalised\nExpression\n",
      show_cluster_labels = var
      )
    m1 <- m1 + ggtitle(marker) + theme(plot.title = element_text(hjust = 0.5, face = "bold"), axis.title = element_blank())
    m1
  }), useMarkers)
  l1[["legend"]] <- cowplot::get_legend(l1[[1]])
  for (marker in useMarkers) {
    l1[[marker]] <- l1[[marker]] + theme(legend.position = "hide")
  }
  return(l1)
}), c("With Labels", "Without Labels"))
```

:::::::: panel-tabset

```{r, label = "DR per per marker 2", fig.width=16, fig.height=ceiling(length(useMarkers)/3)*3, dependson="umaps_per_marker1", results = 'asis'}
invisible(lapply(names(UMAPList), function(yn){
  cat("####", yn, "\n\n")
  print(plot_grid(plotlist = UMAPList[[yn]], ncol = 4, align = "hv"))
  cat("\n\n")
}))
```

::::::::

```{r, label = "save DR per marker", include = FALSE}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, paste0("marker_", dimRedMethodToUse)))
  lapply(names(UMAPList), function(m) {
    dir.create(file.path(pdfDir, paste0("marker_", dimRedMethodToUse), m))
    pdf(file = file.path(pdfDir, paste0(m, "_all_markers_", dimRedMethodToUse, ".pdf")), width = 16, height = ceiling(length(useMarkers)/3)*3)
    print(plot_grid(plotlist = UMAPList[[m]], ncol = 4))
    dev.off()
    lapply(names(UMAPList[[m]]), function(n) {
      pdf(file = file.path(pdfDir, paste0("marker_", dimRedMethodToUse), m, paste0(n, "_", dimRedMethodToUse, ".pdf")), width = 4, height = 4, bg = "transparent")
      if (n != "legend") {
        print(UMAPList[[m]][[n]])
      } else if (n == "legend") {
        plot(UMAPList[[m]][[n]])
      }
      dev.off()
    })
  })
}
```

### `r paste(dimRedMethodToUse)` Per Marker Per Condition

```{r, label = "DR per marker per cond 1", results = 'asis', include = FALSE}
DRperCondList <- setNames(lapply(c("With", "Without"), function(yn) {
  var <- switch(
    yn,
    "With" = TRUE,
    "Without" = FALSE
  )
  l1 <- setNames(lapply(useMarkers, function(marker) {
    m1 <- plotUMAPNice(
      df = as.data.frame(umapDFList[[umapsDFToPlot]]),
      colour_by = marker,
      name = "Normalised\nExpression\n",
      show_cluster_labels = var
      ) + 
      facet_wrap(~condition, ncol = 2) +
      ggtitle(marker)
    m1 <- m1 + theme(plot.title = element_text(hjust = 0.5, face = "bold"), axis.title = element_blank(), legend.position = "hide")
    m1
  }), useMarkers)
  return(l1)
}), c("With Labels", "Without Labels"))
```

:::::::: panel-tabset

```{r, label = "DR per marker per cond 2", fig.width=10, fig.height=5*(ceiling(length(condLevels)/2)), results = 'asis'}
invisible(lapply(names(DRperCondList), function(yn){
  cat("####", yn, "\n\n")
  cat(":::::: {.panel-tabset}\n\n")
  invisible(lapply(names(DRperCondList[[yn]]), function(marker) {
    cat("#####", marker, "\n\n")
    print(DRperCondList[[yn]][[marker]])
    cat("\n\n")
  }))
  cat("::::::\n\n")
}))
```

::::::::

```{r, label = "save DR per marker per cond", include = FALSE}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse)))
  lapply(names(DRperCondList), function(m) {
    dir.create(file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse), m))
    lapply(names(DRperCondList[[m]]), function(n) {
      pdf(file = file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse), m, paste0(n, "_", dimRedMethodToUse, ".pdf")), width = 10, height = 5*(ceiling(length(condLevels)/2)), bg = "transparent")
      if (n != "legend") {
        print(DRperCondList[[m]][[n]])
      } else if (n == "legend") {
        plot(DRperCondList[[m]][[n]])
      }
      dev.off()
    })
  })
}
```

```{r, label = "scgate_setup", include = FALSE}
show_scgate <- runScGate && "Gated_Cells" %in% colnames(colData(sce))
if (show_scgate) {
  mp <- parse_marker_pairs(smd)
  scgate_models <- if (!is.null(mp)) mp$models else list()
}
```

```{r, label = "scgate_header", results = 'asis', eval = show_scgate}
cat("### scGate Gating \n\n")
```

:::::::: {.panel-tabset}

#### All

```{r, label = "scgate_all_plot", fig.height=6, fig.width=18, eval = show_scgate}
print(
  plotUMAPNice(
    df = umapDFList$Downsampled,
    colour_by = "Gated_Cells",
    show_density = FALSE,
    alpha = 1,
    facet_by = "condition",
    show_cluster_labels = TRUE,
    cluster_labels = "Gated_Cells"
  )
)
```

```{r, label = "scgate_model_plots", results = 'asis', fig.height=6, fig.width=6, eval = show_scgate}
invisible(lapply(names(scgate_models), function(mpt) {
  cat("####", mpt, "\n\n")
  y <- gsub("[ -]", "", mpt)
  m1 <- scgate_models[[mpt]][[1]]
  m2 <- scgate_models[[mpt]][[2]]
  if (!m1 %in% rownames(sce)) {
    hit <- agrep(m1, rownames(sce), value = TRUE)
    if (length(hit) == 0) return(invisible(NULL))
    m1 <- hit[[1]]
  }
  if (!m2 %in% rownames(sce)) {
    hit <- agrep(m2, rownames(sce), value = TRUE)
    if (length(hit) == 0) return(invisible(NULL))
    m2 <- hit[[1]]
  }

  dr_theme <- theme_prism() + theme(
    axis.text = element_blank(), axis.ticks = element_blank(),
    axis.line = element_blank(), legend.position = "none",
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5)
  )

  # Plot 1: DimPlot — is this cell type?
  p1 <- plotUMAPNice(
    df = umapDFList$Downsampled,
    colour_by = paste0("is_", y),
    show_density = FALSE,
    alpha = 1
  ) + ggtitle(paste0("Are ", mpt, "?")) +
    theme(legend.position = "none")
  print(p1)
  cat("\n\n")

  # Plot 2: FeaturePlot — UCell score
  ucell_col <- paste0(mpt, "_UCell")
  p2 <- ggplot(umapDFList$Downsampled, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[[ucell_col]])) +
    geom_point(size = dr_point_size, shape = 16) +
    scale_colour_gradientn(colours = hcl.colors(64, "mako")) +
    labs(
      x = paste(dimRedMethodToUse, "dim 1"),
      y = paste(dimRedMethodToUse, "dim 2"),
      title = paste0(mpt, " UCell Score")
    ) + dr_theme
  print(p2)
  cat("\n\n")

  # Plot 3: Joint density — ks::kde weighted scatter
  df_joint <- umapDFList$Downsampled
  emb_mat <- as.matrix(df_joint[, c("x", "y")])
  w_joint <- rep(0, nrow(df_joint))
  for (mk in c(m1, m2)) {
    if (mk %in% colnames(df_joint)) {
      v <- as.numeric(df_joint[[mk]])
      v[is.na(v)] <- 0
      w_joint <- w_joint + v
    }
  }
  if (sum(w_joint) > 0) {
    w_norm <- w_joint / sum(w_joint) * length(w_joint)
    dens <- ks::kde(emb_mat, w = w_norm, eval.points = emb_mat)
    df_joint$density <- dens$estimate
  } else {
    df_joint$density <- 0
  }
  df_joint <- df_joint[order(df_joint$density), ]

  p3 <- ggplot(df_joint, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[["density"]])) +
    geom_point(size = dr_point_size, alpha = 0.7, shape = 16) +
    scale_colour_gradientn(colours = get_palette_colours(viridisColour)) +
    ggtitle(paste0(m1, " + ", m2, " density")) +
    coord_fixed() + dr_theme +
    labs(x = paste(dimRedMethodToUse, "dim 1"),
         y = paste(dimRedMethodToUse, "dim 2"),
         colour = "Density")
  print(p3)
  cat("\n\n")
}))
```

::::::::

::::::::::

## DA / DS

:::::::::: panel-tabset
### DA

::: {.callout-note collapse="true" title="About Differential Abundance"}

Differential abundance of clusters between conditions. Each cluster was tested individually between the two groups specified in each test, e.g., cluster 1 in KO vs WT. Each test gives a log2 fold change, and p-values. Below is the results for each test specified in the metadata. The heatmap shows the expression of each cluster per sample, in order of p-value (most statistically differentially abundant at the top). If you have entered more than one test, then the heatmaps will show the same data but the clusters will be in different orders.

To the right of the heatmap are two annotations. The first is red or blue, and denotes the log2 fold change of that cluster. The second is the p-value, and is either green (adjusted p-value \< 0.05) or grey. The results of this heatmap are the same as under the `Heatmap -> Sample Per Cluster` but with the rows re-ordered and with the log2fc and p-values.
:::

```{r, label = "htmltags1", include = FALSE}
htmltools::tagList(DT::datatable(data.frame(x = 1)))
```

:::::::: panel-tabset

```{r, label = "da_plot", fig.height=6, fig.width=13, results='asis'}
daResultsList <- setNames(lapply(names(daList), function(cont) {
  cat("\n\n")
  cat("####", cont, "Top 20 \n\n")
  pDH <- plotDiffHeatmap(sce, daList[[cont]], normalize = TRUE, all = TRUE, top_n = 20, col_anno = conditions)
  pDH <- apply_anno_colours(pDH, conditions, coloursList)
  print(pDH)
  cat("\n\n")
  cx <- cont |> gsub(pattern = " ", replacement = "_", x = _) |> gsub(pattern = "-", replacement = "_", x = _) |> gsub(pattern = "\\+", replacement = ".", x = _) |> gsub(pattern = "\\/", replacement = ".", x = _)
  ## daDF <- as.data.frame(daList[[cont]][which(daList[[cont]][[daPValToUse]] <= 0.05), c("cluster_id", "logFC", "p_val", "p_adj")])
  daDF <- as.data.frame(daList[[cont]][, c("cluster_id", "logFC", "p_val", "p_adj")])
  daDF <- daDF[order(daDF[[daPValToUse]]),]
  print(htmltools::tagList(
    datatable(daDF, rownames = FALSE, width = "600px") |>
      formatSignif(c("p_val", "p_adj", "logFC"), digits = 3) |>
      formatStyle(columns = c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")), fontWeight = "bold") |>
      formatStyle(columns = "logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold")
  ))
  cat("\n\n")
  openxlsx::write.xlsx(x = as.data.frame(daList[[cont]]), file = file.path(excelDir, paste0(cont, "_DA_Results.xlsx")))
  return(list("pDH" = pDH, "cx" = cx, "daDF" = daDF))
}), names(daList))
```

::::::::

```{r, label = "save DA results", include = FALSE}
if (gimmePDFs) {
  lapply(names(daResultsList), function(x) {
    pdf(file = file.path(pdfDir, paste0(x, "_DA_heatmap.pdf")), height = 6, width = 13)
    print(daResultsList[[x]][["pDH"]])
    dev.off()
  })
}
```

### DS

::: {.callout-note collapse="true" title="About Differential State"}

Differential state (DS) analysis tests whether the expression of individual markers within each cluster differs between conditions. Unlike DA (which tests cluster abundance), DS tests for changes in marker expression levels within each cluster.
:::

```{r, label = "htmltags2", include = FALSE}
htmltools::tagList(DT::datatable(data.frame(x = 1)))
```

:::::::: panel-tabset

```{r, label = "ds_res", results = 'asis', fig.height=12, fig.width=10}
dsPThreshToUse <- 0.05
dsResultsList <- setNames(lapply(names(dsList)[grep("Merged", names(dsList), invert = TRUE)], function(cont) {
  cat("\n\n")
  cat("####", cont, "\n\n")

  openxlsx::write.xlsx(x = as.data.frame(dsList[[cont]]$tbl_DS), file = file.path(excelDir, paste0(cont, "_DS_Full_Results.xlsx")))

  dsUp_tbl <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC >= 0)]), decreasing = TRUE); dsUp <- data.frame(dsUp_tbl[dsUp_tbl >= 1])
  dsDown_tbl <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC <= 0)]), decreasing = TRUE); dsDown <- data.frame(dsDown_tbl[dsDown_tbl >= 1])

  if (nrow(dsUp) == 1) {
    dsUp <- data.frame("Marker" = rownames(dsUp), "Number of clusters" = dsUp[1,])
  }
  if (nrow(dsDown) == 1) {
    dsDown <- data.frame("Marker" = rownames(dsDown), "Number of clusters" = dsDown[1,])
  }

  cat("**Markers up in", strsplit(cont, split = " over ")[[1]][1], "**\n\n")
  if (nrow(dsUp) >= 1) {
    print(
      kable(
        x = dsUp,
        col.names = c("Marker", "Number of clusters")) |>
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")
    )
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][1], "\n\n")
  }

  cat("**Markers up in", strsplit(cont, split = " over ")[[1]][2], "**\n\n")
  if (nrow(dsDown) >= 1) {
    print(
      kable(
        x = dsDown,
        col.names = c("Marker", "Number of clusters")) |>
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")
    )
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][2], "\n\n")
  }

  cat("**Heatmap**\n\n")
  pDH <- plotDiffHeatmap(sce, dsList[[cont]]$tbl_DS, normalize = TRUE, all = TRUE, fdr = dsPThreshToUse, col_anno = conditions, hm_pal = hmPalToUse)
  pDH <- apply_anno_colours(pDH, conditions, coloursList)
  print(pDH)
  cat("\n\n")
  cat("**Full Results**\n\n")
  cat("This table shows the log2FC and p-values of each marker per cluster in", cont, "\n\n")
  cat("Positive log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][1], "in a given cluster. \n\n")
  cat("Negative log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][3], "in a given cluster. \n\n")

  dsTab1 <- as.data.frame(dsList[[cont]]$tbl_DS[, c("cluster_id", "marker_id", "logFC", "p_val", "p_adj")])
  dsTab1 <- dsTab1[order(dsTab1$p_val), ]
  print(
    htmltools::tagList(DT::datatable(
      data = dsTab1,
      rownames = FALSE,
      width = "700px"
      ) |>
        DT::formatSignif(c("logFC", "p_val", "p_adj"), digits = 3) |>
        DT::formatStyle("logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold") |>
        DT::formatStyle(c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")),fontWeight = "bold")
    )
  )
  cat("\n\n")
  return(list("pDH" = pDH, "dsTab1" = dsTab1))
}), names(dsList)[grep("Merged", names(dsList), invert = TRUE)])
```

::::::::

### DS - Merged

Same as the other differential state results, but instead we have merged all clusters prior to DS, so now we see what could be considered differential abundance of each marker between conditions.

```{r, label = "htmltags3", include = FALSE}
htmltools::tagList(DT::datatable(data.frame(x = 1)))
```

:::::::: panel-tabset

```{r, label = "ds_res merged", results = 'asis', fig.height=12, fig.width=10}
dsMergedResultsList <- setNames(lapply(names(dsList)[grep("Merged", names(dsList), invert = FALSE)], function(cont) {
  cat("\n\n")
  cat("####", cont, "\n\n")

  openxlsx::write.xlsx(x = as.data.frame(dsList[[cont]]$tbl_DS), file = file.path(excelDir, paste0(cont, "_DS_Merged_Results.xlsx")))
  dsUp_tbl <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC >= 0)]), decreasing = TRUE); dsUp <- dsUp_tbl[dsUp_tbl >= 1]
  dsDown_tbl <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC <= 0)]), decreasing = TRUE); dsDown <- dsDown_tbl[dsDown_tbl >= 1]

  cat("**Markers up in", strsplit(cont, split = " over ")[[1]][1], "**\n\n")
  if (length(dsUp) >= 1) {
    print(dsUp)
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][1], "\n\n")
  }

  cat("**Markers up in", strsplit(cont, split = " over ")[[1]][2], "**\n\n")
  if (length(dsDown) >= 1) {
    print(dsDown)
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][2], "\n\n")
  }
  cat("**Heatmap**\n\n")
  pDH <- plotDiffHeatmap(x = sceDS, y = dsList[[cont]]$tbl_DS, normalize = TRUE, all = TRUE, col_anno = conditions, hm_pal = hmPalToUse)
  pDH <- apply_anno_colours(pDH, conditions, coloursList)
  print(pDH)
  cat("\n\n")
  cat("**Full Results**\n\n")
  cat("Positive log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][1], "in a given cluster. \n\n")
  cat("Negative log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][3], "in a given cluster. \n\n")

  dsTab1 <- as.data.frame(dsList[[cont]]$tbl_DS[, c("cluster_id", "marker_id", "logFC", "p_val", "p_adj")])
  dsTab1 <- dsTab1[order(dsTab1$p_val), ]
  print(
    htmltools::tagList(DT::datatable(
      data = dsTab1,
      rownames = FALSE,
      width = "700px"
      ) |>
        DT::formatSignif(c("logFC", "p_val", "p_adj"), digits = 3) |>
        DT::formatStyle("logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold") |>
        DT::formatStyle(c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")),fontWeight = "bold")
    )
  )
  cat("\n\n")
  return(list("pDH" = pDH, "dsTab1" = dsTab1))
}), names(dsList)[grep("Merged", names(dsList), invert = FALSE)])
```

::::::::

### Heatmaps of DA Clusters

```{r, label = "build da heatmap inputs", include = FALSE}
.da_hm <- build_da_heatmap_inputs(sce, clusteringMethodToUse)
```

:::::::: panel-tabset

```{r, label = "heatmap by clusters", fig.height = 4, fig.width = 8, results = 'asis', message = FALSE}
heatmapClustersList <- list()
for (i in seq_along(selectedClustersList)) {
  if (length(selectedClustersList[[i]]) >= 2) {

    cat("####", names(selectedClustersList)[i], "\n\n")
    cl <- selectedClustersList[[i]]
    x1 <- filterSCE(.da_hm$x, cluster_id %in% cl, k = clusteringMethodToUse)
    x1$cluster_id <- droplevels(x1$cluster_id)

    ns <- table(.da_hm$x[["cluster_id"]])
    ns2 <- ns[which(names(ns) %in% cl)]
    fq <- round(ns/sum(ns) * 100, 2)
    txt <- sprintf("%s%%(%s)", fq, names(fq))
    txt <- txt[which(gsub(".*\\(|\\).*", "", txt) %in% names(ns2))]
    foo <- row_anno_text(txt, just = "center", gp = gpar(fontsize = 8), location = unit(0.5, "npc"))
    right_anno <- rowAnnotation(n_cells = row_anno_barplot(x = as.matrix(ns2), width = unit(2, "cm"), gp = gpar(fill = "grey", col = "white"), border = FALSE, axis = TRUE, bar_width = 0.8), foo = foo)

    left_anno <- CATALYST:::.anno_clusters(x1, "cluster_id", NULL, coloursList[[names(selectedClustersList)[[i]]]], NULL)

    z1 <- .da_hm$z[selectedClustersList[[i]], ]
    h1 <- Heatmap(
      matrix = z1,
      name = paste0("median scaled\nexpression"),
      col = hmPalToUse,
      cluster_rows = TRUE,
      cluster_columns = FALSE,
      show_row_dend = TRUE,
      show_column_dend = TRUE,
      show_row_names = FALSE,
      row_names_side = "left",
      left_annotation = left_anno,
      right_annotation = right_anno,
      rect_gp = gpar(col = "white"),
      heatmap_legend_param = .da_hm$lgd_aes
    )
    print(h1)
    heatmapClustersList[[i]] <- h1
    cx <- names(selectedClustersList)[[i]] |> gsub(pattern = ": .*", replacement = "", x = _) |> gsub(pattern = "Contrast ", replacement = "Contrast_", x = _) |> gsub(pattern = "\\/", replacement = ".", x = _)
    sx <- names(selectedClustersList)[[i]] |> gsub(pattern = ".*\\: ", replacement = "", x = _) |> gsub(pattern = " ", replacement = "_", x = _) |> gsub(pattern = "\\/", replacement = ".", x = _)
    if (gimmePDFs) {
      pdf(file = paste0(pdfDir, "/", cx, "_", sx, "_heatmaps.pdf"), width = 10, height = ceiling(length(cl)*0.5)+1)
      print(h1)
      dev.off()
    }
    cat("\n\n")
  } else {
    cat("####", names(selectedClustersList)[i], "\n\n")
    cat("There is only one (or zero) cluster(s) in this group, and the heatmaps currently do not support a single cluster. Sorry!")
    cat("\n\n")
  }
}
```

::::::::

### `r paste(dimRedMethodToUse)` of DA Clusters

:::::::: panel-tabset

#### All Cells

```{r, label = "DR of selected clusters", fig.height=5, fig.width=5, results = 'asis'}
DRSelectedClusters <- setNames(lapply(seq_along(clustersToMapList), function(i) {
  clustersToMap <- names(clustersToMapList[[i]])
  cTM <- clustersToMapList[[i]]
  umapDF <- umapDFList[[umapsDFToPlot]]
  umapDF2 <- umapDF[umapDF[["cluster_id"]] %in% clustersToMap,]
  umapDF2[["cluster_id"]] <- droplevels(umapDF2[["cluster_id"]])
  umapDF3 <- umapDF[!umapDF[["cluster_id"]] %in% clustersToMap, ]

  ## Get the mean x/y coordinate for each cluster so we can add labels manually 
  cluster_centre <- umapDF2 |>
  dplyr::group_by(cluster_id) |>
  dplyr::summarize(center_x = mean(x), center_y = mean(y))

  gp <- ggplot(umapDF, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[["cluster_id"]]))

  gp <- gp + geom_point(data = umapDF3, size = dr_point_size, colour = "grey75", alpha = 0.3, shape = 16, show.legend = FALSE)
  gp <- gp + geom_point(data = umapDF2, size = dr_point_size * 1.5, alpha = 0.95)
  gp <- gp + geom_label_repel(data = cluster_centre, aes(x = center_x, y = center_y, label = cluster_id, fill = cluster_id), size = 4, colour = "black")
  gp <- gp + scale_color_manual(values = c(cTM, na.value = "grey33"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + scale_fill_manual(values = c(cTM, na.value = "grey33"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + labs(x = paste(dimRedMethodToUse, "dim 1"), y = paste(dimRedMethodToUse, "dim 2"))
  gp <- gp + ggtitle(names(clustersToMapList)[i])
  gp <- gp + coord_fixed() +
    theme_prism() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      legend.text = element_text(size = 10),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5))
  gp <- gp + guides(colour = guide_legend(override.aes = list(size = 3)))
  gp <- gp + theme(legend.position = "hide")
  print(gp)
  cx <- names(clustersToMapList)[[i]] |> gsub(pattern = ": .*", replacement = "", x = _) |> gsub(pattern = "Contrast ", replacement = "Contrast_", x = _) |> gsub(pattern = "\\/", replacement = "", x = _)
  sx <- names(clustersToMapList)[[i]] |> gsub(pattern = ".*\\: ", replacement = "", x = _) |> gsub(pattern = " ", replacement = "_", x = _) |> gsub(pattern = "\\/", replacement = "", x = _)
  return(list("gp" = gp, "cx" = cx, "sx" = sx))
}), names(clustersToMapList))
```

```{r, label = "save DR of selected clusters", include = FALSE}
if (gimmePDFs) {
  lapply(names(DRSelectedClusters), function(x) {
    pdf(file = paste0(pdfDir, "/", DRSelectedClusters[[x]][["cx"]], "_", DRSelectedClusters[[x]][["sx"]], "_", dimRedMethodToUse, "_full.pdf"), width = 5, height = 5)
    print(DRSelectedClusters[[x]][["gp"]])
    dev.off()
  })
}
```

```{r, label = "DR of selected clusters 2", fig.height= 6*(ceiling(length(condLevels)/2)), fig.width=10, results = 'asis'}
DRSelectedClusters2 <- setNames(lapply(seq_along(clustersToMapList), function(i) {
  cat("####" , names(clustersToMapList)[i], "\n\n")
  clustersToMap <- (names(clustersToMapList[[i]]))
  cTM <- clustersToMapList[[i]]
  umapDF <- umapDFList[[umapsDFToPlot]]
  umapDF2 <- umapDF[umapDF[["cluster_id"]] %in% clustersToMap,]
  umapDF2[["cluster_id"]] <- droplevels(umapDF2[["cluster_id"]])
  umapDF3 <- umapDF[!umapDF[["cluster_id"]] %in% clustersToMap, ]

  ## Get the mean x/y coordinate for each cluster so we can add labels manually 
  cluster_centre <- umapDF2 |>
  dplyr::group_by(cluster_id) |>
  dplyr::summarize(center_x = mean(x), center_y = mean(y))

  gp <- ggplot(umapDF, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[["cluster_id"]]))

  gp <- gp + geom_point(data = umapDF3, size = dr_point_size, colour = "grey75", alpha = 0.3, shape = 16, show.legend = FALSE)
  gp <- gp + geom_point(data = umapDF2, size = dr_point_size * 1.5, alpha = 0.95)
  gp <- gp + geom_label_repel(data = cluster_centre, aes(x = center_x, y = center_y, label = cluster_id, fill = cluster_id), size = 4, colour = "black", show.legend = FALSE)
  gp <- gp + scale_color_manual(
    values = c(cTM, na.value = "black"),
    breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + scale_fill_manual(values = c(cTM, na.value = "black"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + ylim(min(umapDF$y[!is.na(umapDF$y)])-1, max(umapDF$y[!is.na(umapDF$y)])+1)
  gp <- gp + xlim(min(umapDF$x[!is.na(umapDF$x)])-1, max(umapDF$x[!is.na(umapDF$x)])+1)
  gp <- gp + labs(x = paste(dimRedMethodToUse, "dim 1"), y = paste(dimRedMethodToUse, "dim 2"))
  gp <- gp + facet_wrap(~ condition, ncol = 2)
  gp <- gp + ggtitle(names(clustersToMapList)[i])
  gp <- gp + coord_fixed() +
    theme_prism() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      legend.text = element_text(size = 10),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5),
      legend.position = "hide")
  gp <- gp + guides(colour = guide_legend(override.aes = list(size = 3)))
  print(gp)
  cat("\n\n")
  return(gp)
}), names(clustersToMapList))
```

::::::::
::::::::::

```{r, label = "save DR of selected clusters 2", include = FALSE}
if (gimmePDFs) {
  lapply(names(DRSelectedClusters2), function(x) {
    pdf(file = paste0(pdfDir, "/", x, "_", dimRedMethodToUse, "_subset.pdf"), width = 10, height =  6*(ceiling(length(condLevels)/2)))
    print(DRSelectedClusters2[[x]])
    dev.off()
  })
}
```

## Session Info

```{r, label = "file info"}
if (is.null(RDataFolder)) {
  cat("The files generated in this report have been saved in:", rFilesDir)
} else {
  cat("The results in this report were generated using the files in", RDataFolder)
}
```

```{r, label = save_sesh, include = TRUE}
sessionInfo()
```

:::::::::::::
