---
title: "{{PIPELINE_NAME}}"
author: "MARMOT Pipeline"
date: today

format:
  html:
    # ──────────────────────────────────────────────
    # Layout — no TOC; navigation via top-level tabset
    page-layout: full
    smooth-scroll: true
    embed-resources: true

    # ──────────────────────────────────────────────
    # Code — single toggle in header, not per-chunk
    code-tools: true

    # ──────────────────────────────────────────────
    # Light / dark mode
    theme:
      light: cosmo
      dark: cyborg
    color-mode:
      toggle: true
      respect-prefers-color-scheme: true

    # ──────────────────────────────────────────────
    # Grid sizing
    grid:
      body-width: 1600px
      sidebar-width: 250px
      margin-width: 100px

    # ──────────────────────────────────────────────
    # Responsive figures
    include-in-header:
      - marmot_header.html
      - text: |
          <style>
          .cell-output-display img,
          .quarto-figure img,
          .figure img {
            max-width: 100% !important;
            height: auto !important;
          }
          </style>

    # ──────────────────────────────────────────────
    # Figure defaults
    fig-dpi: 150

# Execution
execute:
  echo: false
  warning: false
  message: false
  cache: false

# Table options
df-print: paged
tbl-cap-location: top

# Cross-references
crossref:
  fig-title: "Figure"
  tbl-title: "Table"
  title-delim: ":"

lightbox: true

# ──────────────────────────────────────────────
# Fix htmlwidgets / plots in non-active tabs
include-after-body:
  text: |
    <script>
    document.addEventListener("DOMContentLoaded", function () {
      document.querySelectorAll('button[data-bs-toggle="tab"]').forEach(function (btn) {
        btn.addEventListener("shown.bs.tab", function () {
          if (window.HTMLWidgets && HTMLWidgets.staticRender) {
            HTMLWidgets.staticRender();
          }
          window.dispatchEvent(new Event("resize"));
        });
      });
    });
    </script>
---

```{r knitr_settings, include = FALSE}
knitr::opts_chunk$set(
  comment = "",
  fig.width = 8,
  fig.height = 8,
  dev = "ragg_png"
)
timeRun <- format(Sys.time(), "%Y-%m-%d_%H.%M.%S")
options(bitmapType = "cairo", kableExtra_view_html = FALSE)
```

```{r input_settings}

## 🔧 File paths
fp <- "~/Desktop/FGCZ/MARMOT/Data/Paper/"      # Folder with all FCS files
md_fp <- "~/Desktop/FGCZ/MARMOT/Data/Paper/MARMOT_Metadata.xlsx"      # Full path to metadata Excel file
condaDir <- "/Users/peterleary/miniforge3/bin/mamba"      # Where conda/mamba lives (for PARC/PaCMAP)
parcScriptDir <- "~/Desktop/FGCZ/MARMOT/" # Folder with f_pacmap.py and f_parc.py

## 🧪 Preprocessing
downsampleTo <- NULL  # e.g. 1000 to downsample each FCS file, NULL = no downsampling

## 🔍 Clustering
clusteringMethodToUse <- "FlowSOM"   # One of: "MfastPG", "Mphenograph", "Rphenograph", "PARC", "FlowSOM"
markersToClusterBy <- "all"       # One of: "all", "type", "state"
kValuesIWant <- c(20, 40)
knn <- 20          # Must be in kValuesIWant

## 📉 Dimension Reduction
dimRedMethodToUse <- "UMAP"   # One of: "TSNE", "UMAP", "pacmap"
markersToDimRedBy <- "all"   # One of: "all", "type", "state"

## 🧼 QC
runQC <- "PeacoQC"      # One of: "FlowAI", "PeacoQC", or "None" to skip
useQC <- FALSE            # Whether to keep only QC-passed cells

## 💾 Output
RDataFolder <- NULL # Use existing results (skip re-analysis)
excludeTheseSamples <- NULL           # e.g. c("Sample_001", "Sample_026"), or NULL
gimmePDFs <- FALSE
greyscalePlots <- FALSE
quantileNormaliseAll <- FALSE

## 🧠 Parallelism
runInParallel <- TRUE
nCores <- 4
ramPerCore <- 6

## 🎨 Aesthetics
themeToUse <- "prism"    # ggplot2 theme
viridisColour <- "lisbon"   # Palette: "lisbon", "berlin", "vik", "cork", "batlow", "lapaz", "magma", "inferno", "plasma", "viridis", "mako", "rocket", "turbo"

## 🌉 scGate (optional)
runScGate <- FALSE    # Set TRUE to run scGate marker-pair gating (requires scGate, UCell, Seurat)

## Done editing? Knit away!
```

```{r, label = "settings_summary", results = 'asis'}
cat("::: {.callout-note collapse='true' title='Pipeline Settings Summary'}\n\n")
cat("- **Clustering method:** ", clusteringMethodToUse, "\n")
cat("- **k values:** ", paste(kValuesIWant, collapse = ", "), " (knn = ", knn, ")\n")
cat("- **Dim. reduction:** ", dimRedMethodToUse, "\n")
cat("- **QC:** ", runQC, if (useQC) " (applied)" else " (not applied)", "\n")
cat("- **Downsample:** ", if (!is.null(downsampleTo)) paste0(downsampleTo, " cells") else "No", "\n")
cat("- **Parallel:** ", if (runInParallel) paste0("Yes (", nCores, " cores)") else "No", "\n")
cat("- **scGate:** ", if (runScGate) "Enabled" else "Disabled", "\n")
cat(":::\n\n")
```

```{r check_variables, include = FALSE}

required_vars <- c(
  "fp", "md_fp", "downsampleTo",
  "clusteringMethodToUse", "markersToClusterBy", "kValuesIWant", "knn",
  "dimRedMethodToUse", "markersToDimRedBy",
  "runQC", "useQC", "RDataFolder", "excludeTheseSamples",
  "greyscalePlots", "gimmePDFs", "quantileNormaliseAll",
  "runInParallel", "nCores", "ramPerCore", "themeToUse", "viridisColour"
)

missing_vars <- required_vars[!vapply(required_vars, exists, logical(1))]

if (length(missing_vars)) {
  stop(
    "Missing required variables: ",
    paste(missing_vars, collapse = ", "),
    ".\nPlease define them in the 'input settings' chunk.",
    call. = FALSE
  )
}
```

```{r resolve_paths, include = FALSE}

# Auto-detect PARC/PaCMAP paths when not explicitly set in the metadata
if (is.null(parcScriptDir) || !nzchar(trimws(parcScriptDir))) {
  parcScriptDir <- system.file("python", package = "MARMOT")
  if (nzchar(parcScriptDir)) {
    message("🔧 Auto-detected PARC/PaCMAP scripts: ", parcScriptDir)
  }
}

if (is.null(condaDir) || !nzchar(trimws(condaDir))) {
  condaDir <- tryCatch(
    reticulate::conda_binary(),
    error = function(e) NULL
  )
  if (!is.null(condaDir)) {
    message("🔧 Auto-detected conda binary: ", condaDir)
  }
}
```

```{r load_packages, include = FALSE}

# Use pacman for convenience
suppressWarnings(suppressPackageStartupMessages(
  pacman::p_load(
    BiocGenerics, S4Vectors, SummarizedExperiment, SingleCellExperiment, future.apply,
    flowCore, FlowSOM, ggplot2, dplyr, tidyr, purrr, tibble, readr,
    CATALYST, readxl, cowplot, diffcyt, ggpubr,
    ggprism, gridExtra, rstatix, plotly, grDevices,
    RColorBrewer, reshape2, grid, ComplexHeatmap, circlize, clustree, kableExtra,
    DT, flowAI, parallel, qs2, ggrepel, gtools, reticulate, ggbeeswarm, matrixStats,
    PeacoQC, MARMOT
  )
))

# Parallel processing setup
plan(multisession, workers = nCores)
options(future.globals.maxSize = (ramPerCore * nCores) * 1024^3)
set.seed(2)

# Define class union for ExpData (if not already present)
if (!methods::isClass("ExpData")) {
  try({
    methods::setClassUnion("ExpData", c("matrix", "SummarizedExperiment"), where = .GlobalEnv)
  }, silent = TRUE)
}
```

```{r pipeline_helpers, include = FALSE}
# Parse "CellType: Marker1 Marker2" entries from the Marker Pairs column
parse_marker_pairs <- function(smd) {
  if (!"Marker Pairs" %in% colnames(smd)) return(NULL)
  pairs <- smd[["Marker Pairs"]][!is.na(smd[["Marker Pairs"]])]
  if (length(pairs) == 0) return(NULL)
  types <- unique(gsub("\\:.*", "", pairs))
  names(pairs) <- gsub("\\:.*", "", pairs)
  models <- lapply(types, function(tp) {
    entry <- pairs[names(pairs) == tp][[1]]
    markers <- strsplit(entry, "\\:\\s*")[[1]][[2]]
    strsplit(markers, "\\s+")[[1]]
  })
  names(models) <- types
  list(pairs = unname(pairs), types = types, models = models)
}

# Build coloursList entries for scGate results
setup_scgate_colours <- function(gated_cells, model_names, palette) {
  cols <- list()
  gate_pal <- palette[seq_along(gated_cells)]
  names(gate_pal) <- gated_cells
  cols[["Gated_Cells"]] <- gate_pal
  for (x in model_names) {
    y <- gsub(" |-", "", x)
    cols[[paste0("is_", y)]] <- setNames("steelblue", x)
  }
  cols
}
```

```{r load_rdata, include = FALSE, results = 'asis'}

if (!is.null(RDataFolder)) {
  # Auto-resolve: if user pointed to the top-level Results_Files dir, descend into R_files/
  if (!file.exists(file.path(RDataFolder, "parquet", "_manifest.json")) &&
       file.exists(file.path(RDataFolder, "R_files", "parquet", "_manifest.json"))) {
    RDataFolder <- file.path(RDataFolder, "R_files")
    message("ℹ️ Auto-resolved RDataFolder to: ", RDataFolder)
  }
  pq_manifest <- file.path(RDataFolder, "parquet", "_manifest.json")
  if (!file.exists(pq_manifest)) {
    stop("No Parquet data found in ", RDataFolder,
         "/parquet/. Please re-run the pipeline to generate Parquet output.")
  }
  message(glue::glue("🔁 Loading results from Parquet: {RDataFolder}/parquet/"))
  MARMOT::load_parquet_to_env(file.path(RDataFolder, "parquet"), envir = .GlobalEnv)
  # Also load FCS frame objects (stored as qs2 — not decomposable to Parquet)
  for (qs_obj in c("framesList", "frames")) {
    qs_path <- file.path(RDataFolder, paste0(qs_obj, ".qs2"))
    if (file.exists(qs_path)) {
      assign(qs_obj, qs2::qs_read(qs_path, nthreads = nCores), envir = .GlobalEnv)
    }
  }
}

# Pick the merge method depending on clustering backend
mergeBy <- dplyr::case_when(
  clusteringMethodToUse %in% c("Rphenograph", "Mphenograph", "MfastPG") ~ "k",
  clusteringMethodToUse == "PARC" ~ "p",
  clusteringMethodToUse == "FlowSOM" ~ "meta",
  TRUE ~ NA_character_  # temp placeholder
)
if (is.na(mergeBy)) stop("Unsupported clustering method: ", clusteringMethodToUse)

# Heatmap colour palette
hmPalToUse <- if (greyscalePlots) {
  RColorBrewer::brewer.pal(9, "Greys")
} else {
  rev(RColorBrewer::brewer.pal(11, "RdBu"))
}
```

```{r import_metadata, results = "hide"}

# Load metadata
md  <- readxl::read_xlsx(md_fp, sheet = "File Data")
smd <- readxl::read_xlsx(md_fp, sheet = "Study Data")

# Paths
fcsDir <- fp

# Order sample IDs naturally
md <- md[gtools::mixedorder(md$sample_id), ]
md$sample_id <- factor(md$sample_id, levels = md$sample_id)

# Apply condition ordering if available
conditionOrder <- smd$`Conditions Order` |> na.omit()
if (length(conditionOrder) > 0) {
  md$condition <- factor(md$condition, levels = conditionOrder)
} else {
  md$condition <- as.factor(md$condition)
}

# Extract marker names
useMarkers <- smd[[colnames(smd)[grep("include|cluster", colnames(smd))]]]
useMarkers <- useMarkers[!is.na(useMarkers)]
names(useMarkers) <- gsub("-|\\ |\\/", "_", useMarkers)

if ("Marker Type" %in% colnames(smd)) {
  markerType <- smd$`Marker Type`[!is.na(smd$`Marker Type`)]
  if ("type" %in% markerType) {
    featureType <- "type"
  } else {
    featureType <- NULL
  }
} else {
  knitr::knit_exit()
}

# Define feature type from marker type
featureType <- if ("Marker Type" %in% names(smd)) {
  types <- na.omit(smd$`Marker Type`)
  if ("type" %in% types) "type" else NULL
} else {
  stop("Metadata file must contain 'Marker Type' column.")
}

# Markers to exclude
removeMarkers <- na.omit(smd$`Markers to exclude completely`)

# Cofactors
cfToUse <- as.numeric(na.omit(smd$`Cofactors for markers to use`))

# Contrasts
contrastsToTest <- na.omit(smd$`Conditions To Test`)
contrastsToTestS <- paste(contrastsToTest, collapse = ",")

# Subsampling settings
raw_subsamp <- na.omit(smd$`Cells per condition in UMAPs etc.`)
drCellPerCondition <- setNames(
  lapply(raw_subsamp, function(expr) eval(parse(text = expr))),
  conditionOrder
)

# Identify all condition columns
conditions <- c("condition", setdiff(names(md), c("file_name", "sample_id", "condition")))
conditions <- gsub("-", ".", conditions)

# Exclude samples if specified
if (!is.null(excludeTheseSamples)) {
  md <- md[!md$sample_id %in% excludeTheseSamples, ]
}

# Sort again by condition
md <- md[order(md$condition), ]
md$patient_id <- factor(md$sample_id, levels = md$sample_id)
md$sample_id  <- factor(md$sample_id, levels = md$sample_id)
```

```{r cluster_colours, results = "hide"}

# Define base palette used in CATALYST-inspired plots
catalystCols <- c(
  "#DC050C", "#FB8072", "#1965B0", "#7BAFDE", "#882E72", "#B17BA6",
  "#FF7F00", "#FDB462", "#E7298A", "#E78AC3", "#33A02C", "#B2DF8A",
  "#55A1B1", "#8DD3C7", "#A6761D", "#E6AB02", "#7570B3", "#BEAED4",
  "#666666", "#999999", "#AA8282", "#D4B7B7", "#8600BF", "#BA5CE3",
  "#808000", "#AEAE5C", "#1E90FF", "#00BFFF", "#56FF0D", "#CFBF11"
)

# Darken the palette and append (to allow >30 colours with contrast)
darkenedCols <- colorspace::darken(catalystCols, 0.4)
catalystCols <- paste0(c(catalystCols, darkenedCols), "FF")  # Add full opacity
```

```{r sample_colours, results = "hide", dependson = -1, eval = is.null(RDataFolder)}

# Collect all unique factor levels from the specified metadata columns
allConditionLevels <- unlist(lapply(conditions, \(x) levels(factor(md[[x]]))), use.names = FALSE)

# Assign colours to each unique level using a palette combo (expand as needed)
palettePool <- c(
  RColorBrewer::brewer.pal(9, "Set1"),
  RColorBrewer::brewer.pal(8, "Dark2"),
  RColorBrewer::brewer.pal(12, "Paired")
)

allConditionLevelColours <- setNames(palettePool[seq_along(allConditionLevels)], allConditionLevels)

# Create named list of colours per condition
coloursList <- setNames(
  lapply(conditions, \(x) {
    levels <- levels(factor(md[[x]]))
    allConditionLevelColours[levels]
  }),
  conditions
)

# Assign sample-level colours
sampleCols <- setNames(
  colorRampPalette(brewer.pal(8, "Dark2"))(length(levels(md$sample_id))),
  levels(md$sample_id)
)

# Assign condition-level colours
conditionCols <- setNames(
  colorRampPalette(brewer.pal(8, "Set1"))(length(levels(md$condition))),
  levels(md$condition)
)
```

```{r import_fcs_files, include = FALSE, results = "asis", eval = is.null(RDataFolder)}

message("📥 Importing FCS files...")

# List all .fcs files
fcs_files <- list.files(fcsDir, pattern = "\\.fcs$", full.names = TRUE)
fcs_files <- fcs_files[grep(paste(md$file_name, collapse = "|"), fcs_files)]
fcs_names <- tools::file_path_sans_ext(basename(fcs_files))

# Read each FCS file into a named list
frames <- setNames(
  lapply(fcs_files, function(f) {
    flowCore::read.FCS(f, transformation = FALSE, truncate_max_range = FALSE)
  }),
  fcs_names
)

# Exclude samples, if specified
if (!is.null(excludeTheseSamples)) {
  keep_samples <- md$file_name[!md$sample_id %in% excludeTheseSamples]
  frames <- frames[keep_samples]
}

framesList <- list(
  "Untransformed" = frames
)
```

```{r create_panel, results = "hide", eval = is.null(RDataFolder)}

# Extract parameter metadata
param_df <- pData(parameters(framesList$Untransformed[[1]])) |>
  as.data.frame() |>
  dplyr::mutate(desc = ifelse(is.na(desc), name, desc))

# Build panel
panel <- param_df |>
  dplyr::select(fcs_colname = name, marker_name = desc) |>
  dplyr::mutate(
    antigen       = marker_name,
    marker_class  = ifelse(marker_name %in% useMarkers, markerType, "none"),
    use_channel   = marker_name %in% useMarkers
  ) |>
  dplyr::filter(!marker_name %in% removeMarkers) |>
  dplyr::filter(marker_name %in% useMarkers) |>
  dplyr::arrange(match(marker_name, useMarkers))

# Sanity check
missing <- setdiff(useMarkers, panel$marker_name)
if (length(missing)) {
  stop("❌ Markers in metadata not found in FCS: ", paste(missing, collapse = ", "))
}

message("⚙️ Preparing cofactors for arcsinh transformation...")

# Base cofactor vector, matched to FCS column names
cf <- setNames(cfToUse, panel$fcs_colname[match(useMarkers, panel$marker_name)])

# Fill in default (5) for other channels
all_channels <- panel$fcs_colname
extra_channels <- setdiff(all_channels, names(cf))
cf[extra_channels] <- 2500 # Default cofactor for unused channels

# Ensure correct order
cf <- cf[match(all_channels, names(cf))]

# Alternative version cf2, named by marker name
cf2 <- setNames(cf[match(panel$fcs_colname, names(cf))], panel$marker_name)
cf2 <- cf2[panel$use_channel]

# Assume `fs` is your flowSet and `cofactors` is a named numeric vector
asinh_transforms <- transformList(
  names(cf),
  lapply(cf, function(cof) arcsinhTransform(a = 0, b = 1 / cof, c = 0))
)
# Apply to the flowSet
framesList$Transformed <- future_lapply(seq_along(framesList$Untransformed), function(i) {
  flowCore::transform(framesList$Untransformed[[i]], asinh_transforms)
}) |> setNames(names(framesList$Untransformed))
```

```{r run_flow_auto_qc, results = "asis", include = FALSE, dependson = -1, eval = is.null(RDataFolder)}

if (runQC == "FlowAI") {

  message("🔍 Starting automated quality control...")

  metricsQC <- "FR_FS"

  qc_dir <- file.path(fp, "resultsQC_flowAI")
  expected_qc_files <- gsub(".fcs$", "_QC.fcs", md$file_name)
  qc_present <- dir.exists(qc_dir) &&
    all(expected_qc_files %in% list.files(qc_dir, pattern = "_QC.fcs"))

  if (qc_present) {
    message("📁 QC files found — loading from disk.")
  } else {
    message("⚙️  QC files not found — running `flow_auto_qc()`.")
    flow_auto_qc(
      fcsfiles      = as(framesList$Untransformed, "flowSet"),
      folder_results = qc_dir,
      remove_from   = metricsQC,
      fcs_highQ     = "_QC_highQ",
      fcs_lowQ      = "_QC_lowQ"
    )
  }

  # Load the QC FCS files
  fcsFilesQC <- list.files(qc_dir, full.names = TRUE, pattern = "_QC_highQ.fcs")
  fcsFilesQC <- fcsFilesQC[grep(paste(gsub("\\.fcs", "", md$file_name), collapse = "|"), fcsFilesQC)]
  fcsNames <- sub("_QC_highQ.fcs$", "", basename(fcsFilesQC))

  message("📥 Loading ", length(fcsFilesQC), " QC FCS files into memory...")
  framesList[["FlowAI"]] <- setNames(
    lapply(fcsFilesQC, \(f) read.FCS(f, truncate_max_range = FALSE, transformation = FALSE)),
    paste0(fcsNames, ".fcs")
  )

  # Force evaluation of ranges to avoid downstream surprises
  invisible(lapply(framesList[["FlowAI"]], range, type = "data"))

  # Load and annotate QC summary table
  QCmini <- readr::read_tsv(file.path(qc_dir, "QCmini.txt"), show_col_types = FALSE) |>
    dplyr::distinct(`Name file`, .keep_all = TRUE)
  QCmini <- QCmini[grep(paste0(fcsNames, collapse = "|"), QCmini$`Name file`), ]

  message("🧾 Processing QC summary metadata...")

  qc_names_clean <- sub("_QC_highQ|\\.fcs", "", QCmini$`Name file`)
  md_names_clean <- sub("\\.fcs$", "", md$file_name)

  QCmini$`Sample Name` <- md$sample_id[match(qc_names_clean, md_names_clean)]
  QCmini <- QCmini[gtools::mixedorder(QCmini$`Name file`), ]

  # Use QC data as default if requested
  if (useQC) {
    
    framesList <- framesList[c("FlowAI", setdiff(names(framesList), "FlowAI"))]
  } else {
    message("📦 QC-filtered data loaded but not used in primary workflow.")
  }

  message("🎉 QC step complete.")
}
```

```{r run_peacoqc, results = "asis", include = FALSE, dependson = -1, eval = is.null(RDataFolder)}
if (runQC == "PeacoQC") {

  message("🔍 Starting automated quality control with PeacoQC...🦚")

  qc_dir <- file.path(fp, "resultsQC_peacoQC")
  qc_present <- dir.exists(qc_dir) && {
    expected_qc_files <- gsub(".fcs$", "_QC.fcs", md$file_name)
    actual_qc_paths <- list.files(qc_dir, pattern = "_QC\\.fcs$", recursive = TRUE, full.names = TRUE)
    all(expected_qc_files %in% basename(actual_qc_paths))
  }

  if (qc_present) {
    message("📁 QC files found — loading from disk.")
  } else {
    message("⚙️  QC files not found — running PeacoQC...")

    dir.create(qc_dir, showWarnings = FALSE, recursive = TRUE)

    fs <- as(framesList$Transformed, "flowSet")
    all_channels <- colnames(fs[[1]])
    fluor_channels <- all_channels[!grepl("^(FSC|SSC|Time)", all_channels, ignore.case = TRUE)]
    # Also exclude common technical channels if present
    fluor_channels <- fluor_channels[!grepl("(Width|Height|Trigger)", fluor_channels, ignore.case = TRUE)]
    message("🔬 Using channels for QC: ", paste(fluor_channels, collapse = ", "))
    message("🚫 Excluded channels: ", paste(setdiff(all_channels, fluor_channels), collapse = ", "))
    
    fs_qc <- lapply(seq_along(fs), function(i) {
      fcs_name <- sampleNames(fs)[i]
      ff <- fs[[i]]
      qc_result <- PeacoQC::PeacoQC(
        ff = ff, 
        channels = fluor_channels, 
        plot = TRUE, 
        save_fcs = TRUE, 
        output_directory = qc_dir, 
        name_directory = fcs_name, 
        report = TRUE, 
        determine_good_cells = "all", 
        time_channel_parameter = "Time"
      )
      keep <- qc_result$GoodCells
      ff[keep, ]
    })
    fs_qc <- flowSet(fs_qc)
    
    # Generate a QC summary table
    qc_files <- list.files(path = qc_dir, pattern = "report.txt", recursive = TRUE, full.names = TRUE)
    QCmini <- readr::read_tsv(qc_files[[1]], show_col_types = FALSE)
    if (length(qc_files) > 1) {
      for (i in 2:length(qc_files)) {
        xx <- readr::read_tsv(qc_files[[i]], show_col_types = FALSE)
        QCmini[i,] <- xx[1,]
      }
    }
    readr::write_tsv(QCmini, file.path(qc_dir, "QCmini.txt"))
    message("✅ PeacoQC processing complete. QC files saved to: ", qc_dir)
  }
  
  # Load the QC FCS files
  fcsFilesQC <- list.files(qc_dir, full.names = TRUE, pattern = "_QC.fcs$", recursive = TRUE)
  fcsFilesQC <- fcsFilesQC[grep(paste(md$file_name, collapse = "|"), fcsFilesQC)]
  fcsNames <- sub("_QC.fcs$", "", basename(fcsFilesQC))

  message("📥 Loading ", length(fcsFilesQC), " QC FCS files into memory...")
  framesList[["PeacoQC"]] <- setNames(
    lapply(fcsFilesQC, \(f) read.FCS(f, truncate_max_range = FALSE, transformation = FALSE)),
    paste0(fcsNames, ".fcs")
  )

  # Force evaluation of ranges
  invisible(lapply(framesList[["PeacoQC"]], range, type = "data"))

  # Load QC summary table
  QCmini <- readr::read_tsv(file.path(qc_dir, "QCmini.txt"), show_col_types = FALSE)
  QCmini <- QCmini[grep(paste0(fcsNames, collapse = "|"), QCmini$Filename), ]
  message("🧾 Processing QC summary metadata...")

  message("🎉 QC step complete.")
}
```

```{r arrange_frames_list, include = FALSE, results = "asis", eval = is.null(RDataFolder)}

message("📦 Converting frame lists into flowSets...")

# Convert to flowSet and set $CYT keyword to "FACS"
# Also enforce sampleNames from the list names: read.FCS() embeds the on-disk
# filename (e.g. "Sample_001_QC.fcs") as the frame identifier, which would
# cause CATALYST::prepData() to fail matching against md$file_name.
fsList <- purrr::imap(framesList, function(sample_list, name) {
  fs <- as(sample_list, "flowSet")
  sampleNames(fs) <- names(sample_list)
  for (j in seq_along(fs)) {
    keyword(fs[[j]])[["$CYT"]] <- "FACS"
  }
  fs
})
```

```{r plot_cofactor_histograms, results = "hide", eval = is.null(RDataFolder)}

message("📉 Generating cofactor histograms for visualising transformations...")

cofactorsToTry <- c(5, 50, 500, 1000, 2500, 5000, 10000, 15000)

# Aggressively downsample
framesList[["Marker Histos"]] <- lapply(framesList$Untransformed, function(fcs) {
  if (nrow(fcs) > 1000) {
    set.seed(2)
    fcs[sample(nrow(fcs), 1000), ]
  } else {
    fcs
  }
})

# Function to strip data from ggplot object
minify_plot_data <- function(p, colname, group_col = "condition") {
  # Keep only the columns needed for plotting (x and fill/colour)
  vars_needed <- unique(c(colname, group_col))
  p$data <- p$data[, vars_needed, drop = FALSE]
  p
}

# Plot generation without persistent cofactorDFList
cofactorPlotList <- lapply(useMarkers, function(marker) {
  colname <- panel$fcs_colname[panel$marker_name == marker]
  marker_expr <- purrr::map(framesList[["Marker Histos"]], ~ exprs(.x)[, colname])
  
  # Flat vector of expression values
  expr_vec <- unlist(marker_expr)
  sample_vec <- rep(names(marker_expr), times = purrr::map_int(marker_expr, length))
  cond_vec <- md$condition[match(sample_vec, gsub("\\.fcs", "", md$file_name))]

  # Build only what's needed: wide format DF with just transformed values + condition
  df <- tibble::tibble(
    condition = cond_vec,
    pretransform = expr_vec
  )

  # Transform in-place
  transformed_cols <- purrr::map(cofactorsToTry, function(cof) {
    tibble::tibble(!!paste0("Cofactor_", cof) := asinh(expr_vec / cof))
  }) |> purrr::list_cbind()

  df <- dplyr::bind_cols(df, transformed_cols)

  # Generate plots and strip ggplot data
  plot_names <- c("pretransform", paste0("Cofactor_", cofactorsToTry))
  plot_set <- purrr::map(plot_names, function(colname) {
    p <- ggplot(df, aes(x = .data[[colname]], fill = condition, colour = condition)) +
      geom_density(alpha = 0.1) +
      ggtitle(colname) +
      theme_prism() +
      theme(text = element_text(size = 15), legend.position = "none") +
      scale_colour_manual(values = coloursList$condition) +
      scale_fill_manual(values = coloursList$condition) +
      geom_vline(xintercept = mean(df[[colname]], na.rm = TRUE), linetype = "dashed", colour = "red")
    minify_plot_data(p, colname)
  })

  setNames(plot_set, plot_names)
})

names(cofactorPlotList) <- useMarkers
```

```{r create_sce, results = "hide", eval = is.null(RDataFolder)}
if (useQC) {
  message("✅ Using QC-filtered data for downstream analysis.")
  fsToUse <- fsList[[runQC]]
} else {
  fsToUse <- fsList[["Transformed"]]
}

if (!is.null(downsampleTo)) {
  message("🔽 Downsampling to ", downsampleTo, " cells per sample.")
  fsToUse <- fsApply(fsToUse, function(x) {
    set.seed(2)
    n_events <- nrow(x)
    sample_size <- min(downsampleTo, n_events)
    sampled_indices <- sample(n_events, sample_size)
    x[sampled_indices, ]
  })
}

message("📦 Creating SingleCellExperiment objects with CATALYST::prepData...")

sce <- CATALYST::prepData(
  x = fsToUse,
  panel = panel,
  md = md,
  features = panel$fcs_colname,
  transform = FALSE,
  md_cols = list(factors = conditions)
)
assay(sce, "exprs") <- assay(sce, "counts")
assay(sce, "exprsTransformed") <- assay(sce, "exprs")

# 99% quantile normalisation
message("📐 Applying quantile normalisation...")

exprs_mat <- assay(sce, "exprs")
qs <- matrixStats::rowQuantiles(exprs_mat, probs = c(0.01, 0.99))
norm_exprs <- sweep(exprs_mat, 1, qs[, 1], "-")
norm_exprs <- sweep(norm_exprs, 1, qs[, 2] - qs[, 1], "/")
norm_exprs[norm_exprs < 0 | is.na(norm_exprs)] <- 0
norm_exprs[norm_exprs > 1] <- 1

assay(sce, "exprsQuantNorm") <- norm_exprs

# Also include a scaled version
scale.coef <- matrixStats::rowQuantiles(exprs_mat, probs = c(0.01, 0.99))
assay(sce, "norm") <- t(scale(t(exprs_mat), center = scale.coef[, 1], scale = scale.coef[, 2] - scale.coef[, 1]))
```

```{r select_assay_and_markers, results = "hide"}

# Define which markers to use for clustering
if (!is.null(featureType) && markersToClusterBy == "type") {
  featuresToCluster <- panel$marker_name[panel$marker_class == "type"]
} else if (!is.null(featureType) && markersToClusterBy == "state") {
  featuresToCluster <- panel$marker_name[panel$marker_class == "state"]
} else if (markersToClusterBy %in% c("type", "state", "all")) {
  featuresToCluster <- panel$marker_name
} else {
  warning("⚠️  Unknown value for `markersToClusterBy`, falling back to all markers.")
  featuresToCluster <- panel$marker_name
}

if(quantileNormaliseAll) {
  assay(sce, "exprs") <- assay(sce, "exprsQuantNorm")
}
set.seed(2)
if (quantileNormaliseAll) {
  exprsToUse = "exprsQuantNorm"
} else {
  exprsToUse = "exprsTransformed"
}

# Define condition levels early (needed for fig.height calculations in later chunks)
conds <- as.factor(sce@metadata$experiment_info$condition)
condLevels <- levels(conds)
```

```{r clustering_graph, results = "hide", eval = is.null(RDataFolder)}

if (clusteringMethodToUse %in% c("MfastPG", "Mphenograph", "Rphenograph")) {
  message("🧠 Running clustering using ", clusteringMethodToUse, "...")

  # Expression matrix for clustering
  es <- t(assay(sce, exprsToUse)[featuresToCluster, ])

  cluster_fn <- function(k) {
    set.seed(2)
    if (clusteringMethodToUse == "MfastPG") {
      clusters <- MARMOT::MfastPG(data = es, k = k, seed = 2, verbose = TRUE)
    } else if (clusteringMethodToUse == "Mphenograph") {
      clusters <- MARMOT::Mphenograph(data = es, k = k, seed = 2, verbose = TRUE)
    } else if (clusteringMethodToUse == "Rphenograph") {
      clusters <- Rphenograph::Rphenograph(data = es, k = k)
    }
    ids <- factor(as.character(igraph::membership(clusters[[2]])))
    ids <- factor(ids, levels = gtools::mixedsort(levels(ids)))
    list(kids = ids, clusters = clusters)
  }

  # Run clustering (parallel if requested)
  run_fn <- if (runInParallel) future_lapply else lapply
  kList <- setNames(run_fn(kValuesIWant, cluster_fn), paste0("k", kValuesIWant))

  # Store all clusterings in SCE object
  for (k in names(kList)) {
    sce[[k]] <- kList[[k]]$kids
  }

  # Select default cluster to use
  selected_k <- paste0("k", knn)
  sce$cluster_id <- sce[[selected_k]]
  
  metadata(sce)$cluster_codes <- eval(parse(text = paste0(
    "data.frame(", clusteringMethodToUse , " = factor(levels(sce[[selected_k]])))"
  )))
  kids <- sce[[paste0("k", knn)]]
  numbersOfClusters <- gtools::mixedsort(levels(kids))
}
```

```{r parc_clustering, results = "hide", dependson = "input settings", eval = is.null(RDataFolder)}

if (clusteringMethodToUse == "PARC") {
  message("🐍 Running PARC clustering via basilisk...")

  set.seed(2)
  esPARC <- t(assay(sce, exprsToUse)[featuresToCluster, ])

  parcClustList <- basilisk::basiliskRun(env = MARMOT:::p4r_env,
    fun = function(data, k_values, n_threads) {
      reticulate::source_python(system.file("python", "f_parc.py", package = "MARMOT"))
      setNames(lapply(k_values, function(k) {
        as.integer(unlist(parc_clust(data, knn = as.integer(k),
                                     num_threads = as.integer(n_threads))))
      }), paste0("p", k_values))
    },
    data = esPARC, k_values = kValuesIWant, n_threads = nCores
  )

  for (k in kValuesIWant) {
    cluster_labels <- factor(as.character(parcClustList[[paste0("p", k)]] + 1))
    cluster_labels <- factor(cluster_labels, levels = gtools::mixedsort(levels(cluster_labels)))
    sce[[paste0("p", k)]] <- cluster_labels
  }

  chosen_k <- paste0("p", knn)
  sce$cluster_id <- sce[[chosen_k]]
  metadata(sce)$cluster_codes <- eval(parse(text = paste0(
    "data.frame(", clusteringMethodToUse, " = factor(levels(sce[[chosen_k]])))"
  )))
  kids <- sce[[paste0("p", knn)]]
  numbersOfClusters <- gtools::mixedsort(levels(kids))
}
```

```{r FlowSOM, results = "hide", dependson = "input settings", eval = is.null(RDataFolder)}

if (clusteringMethodToUse == "FlowSOM") {
  message("🌾 Running FlowSOM clustering...")

  sceSOM <- cluster(
    sce,
    features = featuresToCluster,
    xdim = 10,
    ydim = 10,
    maxK = max(kValuesIWant),
    seed = 42
  )

  # Store all clusterings with "k" prefix
  for (k in kValuesIWant) {
    raw_ids <- cluster_ids(sceSOM, paste0("meta", k))
    prefixed <- factor(as.character(raw_ids))
    prefixed <- factor(prefixed, levels = gtools::mixedsort(levels(prefixed)))
    sce[[paste0("meta", k)]] <- prefixed
  }

  # Assign selected clustering
  selected_k <- paste0("meta", knn)
  sce$cluster_id <- sce[[selected_k]]

  metadata(sce)$cluster_codes <- data.frame(FlowSOM = factor(levels(sce[[selected_k]])))
  kids <- sce[[paste0("meta", knn)]]
  numbersOfClusters <- gtools::mixedsort(levels(kids))
}
```

```{r derive_cluster_info, include = FALSE}
# Derive kids / numbersOfClusters from SCE — works for both fresh run and reload.
# Do NOT use the kids variable set inside each clustering chunk: the PARC path
# stores raw integers (not a factor), and on reload all clustering chunks are skipped.
kids <- sce$cluster_id
numbersOfClusters <- gtools::mixedsort(levels(kids))

# Derive featuresToUmap from panel — works for both fresh run and reload.
# On fresh run this is redundant (dim_reduction chunk sets it too), but on reload
# the dim_reduction chunk is skipped and featuresToUmap would be missing.
if (!is.null(featureType) && markersToDimRedBy == "type") {
  featuresToUmap <- panel$marker_name[panel$marker_class == "type"]
} else if (!is.null(featureType) && markersToDimRedBy == "state") {
  featuresToUmap <- panel$marker_name[panel$marker_class == "state"]
} else {
  featuresToUmap <- panel$marker_name
}
```

```{r dim_reduction, results = "hide", dependson = "input settings", eval = is.null(RDataFolder)}

message("📉 Running dimensionality reduction: ", dimRedMethodToUse)

# Select features for DR
if (!is.null(featureType) && markersToDimRedBy == "type") {
  featuresToUmap <- panel$marker_name[panel$marker_class == "type"]
} else if (!is.null(featureType) && markersToDimRedBy == "state") {
  featuresToUmap <- panel$marker_name[panel$marker_class == "state"]
} else {
  featuresToUmap <- panel$marker_name
}

# Always run PCA first (needed for pacmap, helpful for others)
sce <- runDR(x = sce, dr = "PCA", features = featuresToUmap, assay = exprsToUse)

# PaCMAP (via basilisk)
if (dimRedMethodToUse == "pacmap") {
  es <- as.data.frame(t(assay(sce, exprsToUse)[featuresToUmap, ]))
  embedding <- tryCatch(
    basilisk::basiliskRun(env = MARMOT:::p4r_env, fun = function(data) {
      reticulate::source_python(system.file("python", "f_pacmap.py", package = "MARMOT"))
      pacmap_fit(data, verbose = TRUE, apply_pca = FALSE)
    }, data = es),
    error = function(e) { message("PaCMAP failed: ", e$message); NULL }
  )
  if (is.null(embedding)) {
    message("⚠️ PaCMAP unavailable → skipping.")
    dimRedMethodToUse <- "UMAP"
  } else {
    colnames(embedding) <- c("x", "y")
    reducedDim(sce, dimRedMethodToUse) <- embedding
  }
}

# Native DRs
sce <- runDR(x = sce, dr = "UMAP", features = featuresToUmap, assay = exprsToUse)
sce <- runDR(x = sce, dr = "TSNE", features = featuresToUmap, assay = exprsToUse)
```

```{r, label = "make_seurat_object", include = FALSE}
if (is.null(RDataFolder) && runScGate) {
  if (requireNamespace("Seurat", quietly = TRUE)) {
    message("🖼️ Generating Seurat object for scGate...")
    scData <- suppressWarnings(Seurat::as.Seurat(sce, counts = exprsToUse, data = "exprsQuantNorm"))
    scData <- Seurat::ScaleData(scData, assay = "originalexp", verbose = FALSE)
  } else {
    warning("Seurat is not installed — skipping scGate.")
    runScGate <- FALSE
  }
}
```

```{r, label = "run_scGate", include = FALSE}
if (is.null(RDataFolder) && runScGate) {
  if (requireNamespace("scGate", quietly = TRUE) && requireNamespace("UCell", quietly = TRUE)) {
    message("Running scGate on marker pairs...")
    mp <- parse_marker_pairs(smd)

    if (is.null(mp) || length(mp$models) == 0) {
      warning("No marker pairs defined in metadata — skipping scGate.")
      runScGate <- FALSE
    } else {
      models <- mp$models

    for (x in names(models)) {
      if (!models[[x]][[1]] %in% rownames(scData)) {
        hit <- agrep(models[[x]][[1]], rownames(scData), value = TRUE)
        if (length(hit) == 0) {
          warning("No marker match for '", models[[x]][[1]], "' in model '", x, "' — skipping model.")
          next
        }
        models[[x]][[1]] <- hit[[1]]
      }
      if (!models[[x]][[2]] %in% rownames(scData)) {
        hit <- agrep(models[[x]][[2]], rownames(scData), value = TRUE)
        if (length(hit) == 0) {
          warning("No marker match for '", models[[x]][[2]], "' in model '", x, "' — skipping model.")
          next
        }
        models[[x]][[2]] <- hit[[1]]
      }
      model <- scGate::gating_model(name = x, signature = c(models[[x]][[1]], models[[x]][[2]]))
      scData <- scGate::scGate(data = scData, model = model, pos.thr = 0.65)
      y <- gsub("[ -]", "", x)
      # Only mark cells that actually passed the gate (Positive), not all evaluated cells
      scData[[paste0("is_", y)]] <- NA_character_
      scData[[paste0("is_", y)]][which(scData$scGate_multi == "Positive")] <- x
    }

    # Determine Gated_Cells: highest UCell score, but only for cells that passed at least one gate
    ucell_cols <- grep("UCell", colnames(scData@meta.data), value = TRUE)
    is_cols <- paste0("is_", gsub("[ -]", "", names(models)))
    is_cols <- is_cols[is_cols %in% colnames(scData@meta.data)]
    any_positive <- apply(scData@meta.data[, is_cols, drop = FALSE], 1,
                          function(row) any(!is.na(row)))
    if (length(ucell_cols) == 0) {
      warning("No UCell score columns found — assigning Gated_Cells from is_* columns only.")
      scData$Gated_Cells <- ifelse(
        any_positive,
        apply(scData@meta.data[, is_cols, drop = FALSE], 1,
              function(row) { v <- row[!is.na(row)]; if (length(v)) v[[1]] else "Negative" }),
        "Negative"
      )
    } else {
      scData$Gated_Cells <- ifelse(
        any_positive,
        sapply(seq_len(nrow(scData@meta.data)), function(i) {
          gsub("_UCell", "", ucell_cols[which.max(scData@meta.data[i, ucell_cols])])
        }),
        "Negative"
      )
    }

    # Transfer scGate results back to SCE
    sce$Gated_Cells <- scData$Gated_Cells
    scgate_cols <- grep("^is_|UCell", colnames(scData@meta.data), value = TRUE)
    for (col in scgate_cols) {
      sce[[col]] <- scData@meta.data[[col]]
    }

    # Add gate colours
    gate_colours <- setup_scgate_colours(unique(scData$Gated_Cells), names(models), catalystCols)
    for (nm in names(gate_colours)) coloursList[[nm]] <- gate_colours[[nm]]
    } # end: mp non-NULL branch
  } else {
    warning("scGate and/or UCell are not installed — skipping scGate analysis.")
    runScGate <- FALSE
  }
}
```

```{r prepare_dimreds_all, results = "hide", dependson = c("input settings", "make sce"), eval = is.null(RDataFolder)}

message("🧭 Preparing dimensionality reduction dataframes for all methods...")

# Get available DR methods with 2D coords
dr_methods <- reducedDimNames(sce)
valid_methods <- dr_methods[sapply(dr_methods, function(m) ncol(reducedDim(sce, m)) >= 2)]

# Function to build one DR dataframe
build_dr_df <- function(method) {
  coords <- reducedDim(sce, method)[, 1:2]
  colnames(coords) <- c("x", "y")
  df <- data.frame(
    colData(sce),
    sce_idx = seq_len(ncol(sce)),
    coords,
    DRMethod = method,
    check.names = FALSE
  )
  
  # Add cluster annotations
  if (!is.null(metadata(sce)$cluster_codes)) {
    df <- dplyr::left_join(df, cluster_codes(sce), by = c("cluster_id" = clusteringMethodToUse))
  }
  df[[clusteringMethodToUse]] <- df$cluster_id

  # Add quantile-normalised expression
  qn <- t(assay(sce, "exprsQuantNorm"))
  df <- dplyr::bind_cols(df, as.data.frame(qn))
  
  # Drop only rows with missing DR coordinates (not all NAs — scGate is_* columns
  # have NA for negative cells and must be preserved)
  df <- df[!is.na(df$x) & !is.na(df$y), , drop = FALSE]
  
  return(df)
}

# Build full DR list
umapDFList <- lapply(valid_methods, build_dr_df)
names(umapDFList) <- valid_methods

# Optionally downsample per condition
if (!is.null(drCellPerCondition)) {
  for (method in valid_methods) {
    df <- umapDFList[[method]]
    down_df <- df |>
      dplyr::group_by(sample_id) |>
      dplyr::group_modify(~ {
        cond <- unique(md$condition[md$sample_id == .y$sample_id])
        target_n <- as.numeric(drCellPerCondition[as.character(cond)])
        if (nrow(.x) > target_n) .x[sample(nrow(.x), target_n), ] else .x
      }) |>
      dplyr::ungroup()
    umapDFList[[paste0("Downsampled.", method)]] <- down_df
  }
}

# Create label vectors for each DR method
labsList <- lapply(valid_methods, function(method) {
  source_group <- if (!is.null(drCellPerCondition)) paste0("Downsampled.", method) else method
  df <- umapDFList[[source_group]]
  labs <- df |>
    dplyr::group_by(condition) |>
    dplyr::summarise(n = dplyr::n(), .groups = "drop") |>
    dplyr::mutate(label = paste0(condition, "\n n = ", n)) |>
    tibble::deframe()
  names(labs) <- levels(factor(df$condition))
  return(labs)
})
names(labsList) <- valid_methods

umapDFList$All <- umapDFList[[dimRedMethodToUse]]
umapDFList$Downsampled <- umapDFList[[paste0("Downsampled.", dimRedMethodToUse)]]
```

```{r cluster_colours2, results = "hide", dependson = c("input settings", "generate UMAP df"), eval = is.null(RDataFolder)}

message("🎨 Assigning colours to clusters...")

# Base cluster_id colours
cluster_levels <- levels(factor(umapDFList$All$cluster_id))
coloursList[["cluster_id"]] <- setNames(catalystCols[seq_along(cluster_levels)], cluster_levels)

# Add colours for each k-based clustering
for (k in kValuesIWant) {
  colname <- paste0(mergeBy, k)
  if (colname %in% names(umapDFList$All)) {
    cluster_levels_k <- levels(factor(umapDFList$All[[colname]]))
    coloursList[[colname]] <- setNames(catalystCols[seq_along(cluster_levels_k)], cluster_levels_k)
  }
}
```

```{r da_ds_analysis, results = "hide", dependson = c("input settings", "make sce", "import md file")}
message("🚀 Starting differential abundance (DA) and differential state (DS) analysis...")

daPValToUse <- "p_adj"

# ───────────────────────────────────────────────────────
# 🧹 Tidy condition names and prepare design matrix
ei <- metadata(sce)$experiment_info
ei$condition <- gsub(" ", ".", ei$condition)

ei$conditionToTest <- ei$condition |>
  gsub(pattern = " ", replacement = ".", x = _) |>
  gsub(pattern = "-", replacement = "0", x = _) |>
  gsub(pattern = "\\+", replacement = "1", x = _) |>
  gsub(pattern = "/", replacement = "", x = _) |>
  as.factor()

cTTforEdgeR <- contrastsToTest |>
  gsub(pattern = " ", replacement = ".", x = _) |>
  gsub(pattern = "-", replacement = "0", x = _) |>
  gsub(pattern = "\\+", replacement = "1", x = _) |>
  gsub(pattern = "\\.over\\.", replacement = "-", x = _) |>
  gsub(pattern = "\\.vs\\.", replacement = "-", x = _) |>
  gsub(pattern = "/", replacement = "", x = _)

cTTForPlots <- contrastsToTest

design <- model.matrix(~0 + conditionToTest, data = ei)
colnames(design) <- gsub("conditionToTest", "", colnames(design))

# ───────────────────────────────────────────────────────
# ✍️ Create contrast matrix
message("🧮 Constructing contrasts for edgeR...")
contrasts <- setNames(lapply(cTTforEdgeR, \(ctr) {
  eval(parse(text = paste0("limma::makeContrasts(", ctr, ", levels = design)")))
}), cTTforEdgeR)

# ───────────────────────────────────────────────────────
# 📊 Differential abundance analysis (DA)
message("📈 Running DA analysis using diffcyt...")
daList <- setNames(lapply(seq_along(contrasts), function(i) {
  res <- diffcyt(
    d_input            = sce,
    design             = design,
    contrast           = contrasts[[i]],
    analysis_type      = "DA",
    method_DA          = "diffcyt-DA-edgeR",
    clustering_to_use  = clusteringMethodToUse,
    min_cells          = 5,
    transform          = FALSE,
    normalize          = TRUE
  )
  as.data.frame(rowData(res$res))
}), cTTForPlots)

# ───────────────────────────────────────────────────────
# 🧬 Differential state analysis (DS)
message("🧪 Running DS analysis using diffcyt...")
if (is.null(featureType)) {
  rowData(sce)$marker_class <- "state"
}
sceDS <- sce
ds_formula1 <- createFormula(ei, cols_fixed = "conditionToTest", cols_random = "sample_id")

dsList <- setNames(vector("list", length(contrasts)), cTTForPlots)

for (i in seq_along(contrasts)) {
  res_DS <- diffcyt(
    d_input           = sceDS,
    clustering_to_use = clusteringMethodToUse,
    analysis_type     = "DS",
    method_DS         = "diffcyt-DS-limma",
    design            = design,
    contrast          = contrasts[[i]],
    transform         = FALSE,
    normalize         = TRUE,
    verbose           = FALSE
  )
  dsList[[i]] <- list(res_DS = res_DS, tbl_DS = rowData(res_DS$res))
}

# ───────────────────────────────────────────────────────
# 🔗 Merge clusters and rerun DS
message("🔁 Merging clusters and rerunning DS on merged data...")
mergedID <- paste0(clusteringMethodToUse, "_merged")
mergedTable <- data.frame(
  old_cluster = levels(as.factor(sce[[paste0(mergeBy, knn)]])),
  new_cluster = "all"
)

sceDS <- mergeClusters(
  x     = sceDS,
  k     = clusteringMethodToUse,
  id    = mergedID,
  table = mergedTable
)

ds_formula1 <- createFormula(ei, cols_fixed = "conditionToTest")

for (i in seq_along(contrasts)) {
  res_DS <- diffcyt(
    d_input           = sceDS,
    formula           = ds_formula1,
    clustering_to_use = mergedID,
    analysis_type     = "DS",
    method_DS         = "diffcyt-DS-limma",
    design            = design,
    contrast          = contrasts[[i]],
    transform         = FALSE,
    normalize         = TRUE,
    verbose           = FALSE
  )
  dsList[[paste(cTTForPlots[i], "Merged")]] <- list(
    res_DS = res_DS,
    tbl_DS = rowData(res_DS$res)
  )
}

# ───────────────────────────────────────────────────────
# 🔍 Identify significant DA clusters
message("🔬 Extracting significant DA clusters...")

# Pre-allocate list for efficiency
selectedClustersList <- vector("list", length(contrasts) * 2)
names_list <- character(length(contrasts) * 2)

for (i in seq_along(contrasts)) {
  contrast_conditions <- strsplit(contrastsToTest[i], " over ")[[1]]
  da_table <- daList[[i]]
  
  # Filter for complete cases and significant results
  significant_rows <- complete.cases(da_table[c("cluster_id", "logFC", daPValToUse)]) &
                     da_table[[daPValToUse]] <= 0.05
  
  significant_data <- da_table[significant_rows, ]
  
  # Extract clusters by direction
  up_clusters <- as.character(unique(significant_data$cluster_id[significant_data$logFC >= 0]))
  down_clusters <- as.character(unique(significant_data$cluster_id[significant_data$logFC <= 0]))
  
  # Store results
  idx_up <- 2 * i - 1
  idx_down <- 2 * i
  
  selectedClustersList[[idx_up]] <- up_clusters
  selectedClustersList[[idx_down]] <- down_clusters
  
  names_list[idx_up] <- paste0("Contrast ", i, ": ", contrast_conditions[1], " Up")
  names_list[idx_down] <- paste0("Contrast ", i, ": ", contrast_conditions[2], " Up")
}

names(selectedClustersList) <- names_list

# ───────────────────────────────────────────────────────
# 🎨 Assign colours to selected clusters
message("🎨 Assigning colours to selected clusters...")
for (i in seq_along(selectedClustersList)) {
  label <- names(selectedClustersList)[i]
  clusterIDs <- selectedClustersList[[i]]
  sc <- coloursList[[paste0(mergeBy, knn)]][clusterIDs]
  coloursList[[label]] <- sc
}

# 🗺️ Map clusters to colours
message("🗺️ Generating cluster-to-colour mapping list...")
clustersToMapList <- lapply(names(selectedClustersList), function(label) {
  ids <- selectedClustersList[[label]]
  clrs <- coloursList[[label]]
  names(clrs) <- ids
  clrs
})
names(clustersToMapList) <- names(selectedClustersList)

message("✅ DA and DS analysis complete.")

# Save DA/DS results to Parquet (runs even during reload since DA/DS always re-runs)
pq_dir <- if (!is.null(RDataFolder)) {
  file.path(RDataFolder, "parquet")
} else if (exists("rFilesDir")) {
  file.path(rFilesDir, "parquet")
} else NULL

if (!is.null(pq_dir) && dir.exists(pq_dir)) {
  message("💾 Saving updated DA/DS results to Parquet...")
  MARMOT::save_da_ds_parquet(pq_dir, daList, dsList, selectedClustersList, daPValToUse)
}
```

```{r output_directories, results = "hide"}

# Create Results root directory if not resuming from saved RData
if (is.null(RDataFolder)) {
  resultsDir <- file.path(fp, paste0("Results_Files_", timeRun))
  dir.create(resultsDir, showWarnings = FALSE, recursive = TRUE)

  rFilesDir <- file.path(resultsDir, "R_files")
  dir.create(rFilesDir, showWarnings = FALSE)
}

# When reloading, RDataFolder points to R_files/ — go up one level for sibling dirs
reloadResultsDir <- if (!is.null(RDataFolder)) dirname(RDataFolder) else NULL

# PDFs
if (gimmePDFs) {
  pdfDir <- if (is.null(RDataFolder)) {
    file.path(resultsDir, "PDF_figures")
  } else {
    file.path(reloadResultsDir, "PDF_figures")
  }
  dir.create(pdfDir, showWarnings = FALSE)
}

# Excel outputs
excelDir <- if (is.null(RDataFolder)) {
  file.path(resultsDir, "Excel_Files")
} else {
  file.path(reloadResultsDir, "Excel_Files")
}
dir.create(excelDir, showWarnings = FALSE)
```

```{r write_results_to_file, eval = is.null(RDataFolder), include = FALSE}

message("💾 Saving outputs...")

# Save FCS frame objects as qs2 — flowFrame/flowSet cannot be stored in Parquet
for (qs_obj in c("framesList", "frames")) {
  if (exists(qs_obj, envir = .GlobalEnv)) {
    out_path <- file.path(rFilesDir, paste0(qs_obj, ".qs2"))
    qs2::qs_save(get(qs_obj, envir = .GlobalEnv), file = out_path,
                 nthreads = nCores)
    message("Saved (qs2): ", qs_obj)
  }
}

# Save Parquet — main data store for Shiny app and reproducibility
message("💾 Saving Parquet format...")
MARMOT::save_parquet_data(rFilesDir, envir = .GlobalEnv)

# Write pipeline_settings.json sidecar so Tauri GUI can reload & lock fields
settings_list <- list(
  clusteringMethodToUse = as.character(clusteringMethodToUse),
  markersToClusterBy    = as.character(markersToClusterBy),
  kValuesIWant          = paste(kValuesIWant, collapse = ", "),
  knn                   = as.character(knn),
  dimRedMethodToUse     = as.character(dimRedMethodToUse),
  markersToDimRedBy     = as.character(markersToDimRedBy),
  runQC                 = as.character(runQC),
  useQC                 = toupper(as.character(useQC)),
  runScGate             = toupper(as.character(runScGate)),
  downsampleTo          = if (is.null(downsampleTo)) "" else as.character(downsampleTo),
  quantileNormaliseAll  = toupper(as.character(quantileNormaliseAll)),
  gimmePDFs             = toupper(as.character(gimmePDFs)),
  greyscalePlots        = toupper(as.character(greyscalePlots)),
  themeToUse            = as.character(themeToUse),
  viridisColour         = as.character(viridisColour),
  runInParallel         = toupper(as.character(runInParallel)),
  nCores                = as.character(nCores),
  ramPerCore            = as.character(ramPerCore)
)
jsonlite::write_json(settings_list, file.path(rFilesDir, "pipeline_settings.json"),
                     auto_unbox = TRUE, pretty = TRUE)
message("Saved: pipeline_settings.json")

# Copy metadata Excel into results folder for self-contained archive
file.copy(file.path(fp, md_fp), resultsDir, overwrite = TRUE)

# Marker file so marmot.R can copy QMD/HTML into results folder after render
writeLines(resultsDir, file.path(fp, ".marmot_last_results"))
```

```{r nice_umap_function}

# Adaptive point size: 25/sqrt(n) clamped to [0.05, 0.5]
# ~1000 cells → 0.5, ~10k → 0.25, ~50k → 0.11, ~100k → 0.08, ~500k+ → 0.05
n_dr_cells <- nrow(umapDFList[[ifelse(is.null(drCellPerCondition), "All", "Downsampled")]])
dr_point_size <- max(0.05, min(0.5, 25 / sqrt(n_dr_cells)))

plotUMAPNice <- function(
  df = umapDF,
  colour_by = "condition",
  show_density = FALSE,
  size = dr_point_size,
  alpha = 1,
  text_size = 12,
  facet_by = NULL,
  show_grids = FALSE,
  name = "Normalised\nExpression\n",
  show_cluster_labels = FALSE,
  cluster_labels = "cluster_id",
  border = FALSE,
  border_size = 3,
  border_density = 1,
  border_colour = "black"
) {
  gp <- ggplot(df, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[[colour_by]]))

  # Density overlay
  if (show_density) {
    gp <- gp +
      geom_density_2d(
        data = df[, c("x", "y")],
        aes(x = x, y = y),
        colour = "lightgrey",
        size = 0.5,
        bins = 20,
        alpha = 0.8
      ) +
      coord_cartesian(
        xlim = range(df$x, na.rm = TRUE) + c(-1, 1),
        ylim = range(df$y, na.rm = TRUE) + c(-1, 1)
      )
  }

  if (border && nrow(df) > 10) {
    kde <- MASS::kde2d(df[["x"]], df[["y"]], n = 100L)
    ix  <- pmax(1L, pmin(findInterval(df[["x"]], kde$x), length(kde$x)))
    iy  <- pmax(1L, pmin(findInterval(df[["y"]], kde$y), length(kde$y)))
    cell_density <- kde$z[cbind(ix, iy)]
    border_df <- df[cell_density < quantile(cell_density, border_density), ]
    gp <- gp +
      geom_point(data = border_df, size = size * border_size,
                 colour = border_colour, show.legend = FALSE) +
      geom_point(colour = "grey75", size = size, show.legend = FALSE) +
      geom_point(size = size, alpha = alpha)
  } else {
    gp <- gp + geom_point(size = size, alpha = alpha, shape = 16)
  }

  gp <- gp +
    labs(x = paste(dimRedMethodToUse, "dim 1"),
         y = paste(dimRedMethodToUse, "dim 2"),
         title = colour_by) +
    coord_fixed() +
    theme_prism() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      legend.text = element_text(size = 10),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5)
    )

  # Guides
  if (colour_by == "condition") {
    gp <- gp + guides(colour = guide_legend(override.aes = list(size = 2)))
  }

  # Cluster label overlays
  if (show_cluster_labels) {
    label_coords <- data.frame(
      df |>
        dplyr::group_by(.data[[cluster_labels]]) |>
        dplyr::summarise(x = median(x), y = median(y)))
    if (colour_by %in% useMarkers) {
      label_coords[[colour_by]] <- lapply(levels(df[[cluster_labels]]), function(level) {
        mean(df[[colour_by]][df[[cluster_labels]] == level])
      }) |> unlist()
      gp <- gp + geom_label_repel(data = label_coords, aes(x = x, y = y, label = .data[[cluster_labels]], fill = .data[[colour_by]]), colour = "black", show.legend = FALSE)
    } else {
      gp <- gp + geom_label_repel(data = label_coords, aes(x = x, y = y, label = .data[[cluster_labels]], fill = .data[[cluster_labels]]), colour = "black", show.legend = FALSE)
    }
  }

  if (!show_grids) {
    gp <- gp +
      theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
  }

  # Gradient or manual colour scale
  if (colour_by %in% useMarkers) {
    gp <- gp + scale_colour_gradientn(name, colours = hcl.colors(64, viridisColour))
    gp <- gp + scale_fill_gradientn(name, colours = hcl.colors(64, viridisColour))
  } else {
    gp <- gp + scale_colour_manual(values = coloursList[[colour_by]])
    gp <- gp + scale_fill_manual(values = coloursList[[colour_by]])
  }

  # Faceting
  if (!is.null(facet_by)) {
    gp <- gp + facet_wrap(facets = as.formula(paste("~", facet_by)))
  }

  return(gp)
}
```

Started on `r paste(timeRun)`

::: {.callout-tip appearance="simple" collapse="true" title="Colour Key"}

- **Green / tip** — results summaries, key findings, cluster descriptions
- **Blue / note** — methodological details, parameter explanations
- **Amber / warning** — QC notes, filtering statistics, caveats
- **Red / important** — critical results, significant DA/DS findings
:::


::::::::::::: panel-tabset
## QC

:::::::::: panel-tabset

```{r, label = "print QC", results = 'asis'}
if (runQC %in% c("FlowAI", "PeacoQC") && exists("QCmini")) {
  cat("### QC Cell Filtering \n\n")
  if(runQC %in% c("FlowAI", "PeacoQC") && !useQC) {
    cat("\n\n")
    cat("::: {.callout-warning title='QC Not Applied'}\n\n")
    cat("You ran QC using", runQC, ", but chose **not** to apply it.",
        "The data here is presented purely for your information and does not reflect any of the results in this report.",
        "If you wish to also use only the cells that passed QC, please change the `useQC` variable to TRUE in the metadata file.")
    cat("\n:::\n\n")
  }
  print(kable(QCmini, format = "html", escape = FALSE, col.names = gsub("\\.", "", names(QCmini))) |> kable_styling())
  cat("\n\n")
  openxlsx::write.xlsx(as.data.frame(QCmini), file = file.path(excelDir, "QCtable.xlsx"))
} else if (!is.null(RDataFolder)) {
  cat("### QC Cell Filtering \n\n")
  cat("QC data not available on reload.\n\n")
}
```

### Cofactor Histograms

:::::::: panel-tabset

```{r, label = "print cofactor plots", results = 'asis', fig.height=12, fig.width=12, eval = exists("cofactorPlotList")}
for (marker in useMarkers) {
  cat("####", marker, "\n")
  cat(paste(marker, "Cofactor you chose:", cf2[[marker]]))
  cat("\n\n")
  grid.arrange(grobs = cofactorPlotList[[marker]], ncol = 3, top = marker)
  cat("\n\n")
}
```

```{r, label = "cofactor_dims", include = FALSE}
nrowsCofactors <- ceiling(length(useMarkers)/4)
```

::::::::

### Marker Histograms

These are the histograms of all markers using the selected cofactors. Downsampled to 2000 cells per sample.

```{r, label = "plot marker histograms 2", fig.height = (nrowsCofactors*2)+1, fig.width = 15}
y <- assay(sce, "exprs")
df <- data.frame(t(y), colData(sce), check.names = FALSE)
gg_df <- melt(df, value.name = "value", variable.name = "antigen", 
  id.vars = names(colData(sce)))
mh <- ggplot(gg_df, aes(x = value, y = after_stat(ndensity), fill = condition, color = condition)) +
  geom_density(alpha = 0.1) +
  facet_wrap(~antigen, scales = "free") + 
  ylab("Normalised Density") + 
  xlab("Transformed Value") +
  theme_prism(axis_text_angle = 90) + theme(panel.grid = element_blank(),
    strip.background = element_blank(), strip.text = element_text(face = "bold"),
    axis.text = element_text(color = "black"), axis.title = element_text(color = "black"),
    text = element_text(size = 15)) +
  scale_colour_manual(values = coloursList$condition) +
  scale_fill_manual(values = coloursList$condition)
mh
```

```{r, label = "save marker histogram", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "marker_histograms.pdf"), width = 15, height = (nrowsCofactors*2), bg = "transparent")
  print(mh)
  dev.off()
}
```

### Cell Counts

::: {.callout-note collapse="true" title="Sample and Cell Counts"}

The number of samples per condition is:

```{r, label = "print cell counts1"}
table(md$condition)
```

```{r, label = "warn qc counts"}
if (useQC && !is.null(downsampleTo)) {
  cat("\n\n")
  print("You chose to downsample AND use the QC-filtered counts. Therefore, cells were QC'd and then downsampled. The numbers here reflect this two-step approach.")
  cat("\n\n")
}
```

The number of cells per condition is:

```{r, label = "print cell counts2"}
table(sce$condition)
```

The number of cells per sample is:

```{r, label = "print cell counts3"}
n_cells(sce)
```
:::

```{r, label = "plot cell counts 1", fig.height=4, fig.width=ceiling(length(conditions)*2)}
plotCounts(sce, color_by = "condition") + scale_fill_manual(values = coloursList$condition) + theme_prism(axis_text_angle = 45)
```

```{r, label = "plot cell counts 2", fig.height=4, fig.width=ceiling(length(conditions)*3)}
plotCounts(sce, color_by = "condition", group_by = "sample_id") + scale_fill_manual(values = coloursList$condition) + theme_prism(axis_text_angle = 45)
```

```{r, label = "plot full counts", results = 'asis', fig.height=4, fig.width=ceiling(length(conditions)*3)}
if (!is.null(downsampleTo) && exists("framesList") && "Untransformed" %in% names(framesList)) {
  cat("\n\nHowever, because you downsampled, here are those numbers for the full dataset. The black line is what you downsampled to.\n\n")
  cat("\n\n")
  originalCellNumbers <- lapply(names(framesList$Untransformed), function(x) {
    nrow(framesList$Untransformed[[x]])
  }) |> setNames(names(framesList$Untransformed))
  originalCellNumbers <- originalCellNumbers |> data.frame(check.names = FALSE) |> t() |> data.frame()
  colnames(originalCellNumbers)[[1]] <- "Cells"
  originalCellNumbers <- tibble::rownames_to_column(.data = originalCellNumbers, "file_name")
  originalCellNumbers$file_name <- paste0(originalCellNumbers$file_name, ".fcs")
  originalCellNumbers <- dplyr::left_join(originalCellNumbers, md, by = "file_name")
  barplotAllCells <- ggplot(originalCellNumbers, aes(x = sample_id, y = Cells, fill = condition)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = coloursList$condition) +
    labs(x = "Sample ID", y = "Number of Cells", title = "All Cells (before downsampling)") +
    theme_prism(axis_text_angle = 45) +
    geom_hline(yintercept = downsampleTo, linetype = "dashed", colour = "black")
  print(barplotAllCells)
  cat("\n\n")
}
```
### Pseudo-bulk MDS Plots

:::::::: panel-tabset

```{r, label = "plot mds by condition", fig.height=5, fig.width=7, results = 'asis'}
mdsList <- setNames(lapply(conditions, function(x) {
  mds1 <- CATALYST::pbMDS(x = sce, color_by = x, size_by = TRUE, features = featureType)
  mds1 <- mds1 + scale_colour_manual(values = coloursList[[x]]) + theme_prism(base_size = 16) + theme(panel.grid = element_blank())
  mds1
}), conditions)
invisible(lapply(names(mdsList), function(x) {
  cat("####", x, "\n")
  print(mdsList[[x]])
  cat("\n\n")
}))
```

::::::::

```{r, label = "save mds plots", include = FALSE}
if (gimmePDFs) {
  lapply(names(mdsList), function(x) {
    pdf(file = file.path(pdfDir, paste0(x, "_MDS.pdf")), width = 7, height = 5, bg = "transparent")
    print(mdsList[[x]])
    dev.off()
  })
}
```

### Non-redundancy score plot

```{r, label = "plot nrs", fig.height=4, fig.width=12}
plotNRS(sce, color_by = "condition", features = featureType) + scale_colour_manual(values = coloursList$condition)  + theme_prism(axis_text_angle = 90)
```

### Marker Expression Heatmap

```{r, label = "plot marker expr heatmap", fig.height=8, fig.width=12}
pEH <- plotExprHeatmap(sce, bin_anno = TRUE, row_anno = TRUE)
for (con in conditions) {
  pEH@left_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pEH@left_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pEH@left_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pEH@left_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pEH@left_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pEH@left_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pEH@left_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pEH@left_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
}
pEH
```

::::::::::

```{r, label = "save expression marker", include = FALSE, eval = gimmePDFs}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "marker_heatmap.pdf"), width = 12, height = (nrowsCofactors*2), bg = "transparent")
  print(pEH)
  dev.off()
}
```

## `r paste(clusteringMethodToUse)` Cluster Analysis

:::::::::: panel-tabset
### K Tree

```{r, label = "clustering_summary", results = 'asis'}
cat("::: {.callout-tip title='Clustering Summary'}\n\n")
cat("- **Method:** ", clusteringMethodToUse, "\n")
cat("- **k values tested:** ", paste(kValuesIWant, collapse = ", "), "\n")
cat("- **Selected k:** ", knn, " → ", length(unique(cluster_ids(sce, clusteringMethodToUse))), " clusters\n")
cat("- **Clustering markers:** ", paste(featuresToCluster, collapse = ", "), "\n")
cat(":::\n\n")
```

```{r, label = "get cluster numbers", include = FALSE}
numbersOfClustersPerKnn <- setNames(lapply(paste0(mergeBy, kValuesIWant), function(x) {
  gtools::mixedsort(levels(sce[[x]]))
}), paste0(mergeBy, kValuesIWant))
```

```{r, label = "clustree", fig.height = ceiling(length(kValuesIWant)*3), fig.width = ceiling(max(lengths(numbersOfClustersPerKnn))/2.5)}
if(length(kValuesIWant) >= 2) {
  clustree::clustree(sce, prefix = mergeBy, exprs = "exprsTransformed")
}
```

```{r, label = "marker pair scatterplots", results = 'asis', fig.height=5, fig.width=5}
runMarkerPairPlots <- ("type" %in% markerType) && ("state" %in% markerType) && ("Marker Pairs" %in% colnames(smd))
if (runMarkerPairPlots) {
  cat("\n\n")
  cat("### Marker-Pair Scatterplots \n\n")
  es <- as.data.frame(t(assay(sce, "exprsTransformed")))
  es[[mergeBy]] <- sce[[paste0(mergeBy, knn)]]
  if (!is.null(downsampleTo)) {
    if (nrow(es) > downsampleTo) {
      es <- es[sample(rownames(es), downsampleTo),]
    }
  }
  esQN <- as.data.frame(t(assay(sce, "exprsQuantNorm")))
  esQN[[mergeBy]] <- sce[[paste0(mergeBy, knn)]]
  if (!is.null(downsampleTo)) {
    if (nrow(esQN) > downsampleTo) {
      esQN <- esQN[sample(rownames(esQN), downsampleTo),]
    }
  }
  
  mp <- parse_marker_pairs(smd)
  markerPairs <- mp$pairs
  markerPairTypes <- mp$types
  markerPairsList <- setNames(lapply(markerPairTypes, function(x) {
    toPlot <- markerPairs[grep(paste0("^", x), markerPairs)]
    plots <- setNames(lapply(toPlot, function(y) {
      markersToPlot <- strsplit(strsplit(y, "\\: ")[[1]][[2]], " ")[[1]]
      clr_vals <- coloursList[[paste0(mergeBy, knn)]]
      if (!quantileNormaliseAll) {
        median_es <- do.call(rbind, lapply(unique(es[[mergeBy]]), function(cl) {
          data.frame(V1 = cl, x = median(es[[markersToPlot[1]]][es[[mergeBy]] == cl]),
                     y = median(es[[markersToPlot[2]]][es[[mergeBy]] == cl]))
        }))
        colnames(median_es)[1] <- mergeBy
        q1 <- ggplot(es, aes(x = .data[[markersToPlot[1]]], y = .data[[markersToPlot[2]]], color = .data[[mergeBy]])) +
          geom_point(size = 0.001) +
          geom_label_repel(data = median_es, aes(x = x, y = y, label = .data[[mergeBy]], fill = .data[[mergeBy]]),
                           colour = "black", show.legend = FALSE, size = 4) +
          scale_color_manual(values = clr_vals) +
          scale_fill_manual(values = clr_vals) +
          theme_prism() +
          theme(legend.position = "hide") +
          ggtitle(paste0(x, "\nTransformed"))
      }
      median_esQN <- do.call(rbind, lapply(unique(esQN[[mergeBy]]), function(cl) {
        data.frame(V1 = cl, x = median(esQN[[markersToPlot[1]]][esQN[[mergeBy]] == cl]),
                   y = median(esQN[[markersToPlot[2]]][esQN[[mergeBy]] == cl]))
      }))
      colnames(median_esQN)[1] <- mergeBy
      p1 <- ggplot(esQN, aes(x = .data[[markersToPlot[1]]], y = .data[[markersToPlot[2]]], color = .data[[mergeBy]])) +
        geom_point(size = 0.001) +
        geom_label_repel(data = median_esQN, aes(x = x, y = y, label = .data[[mergeBy]], fill = .data[[mergeBy]]),
                         colour = "black", show.legend = FALSE, size = 4) +
        scale_color_manual(values = clr_vals) +
        scale_fill_manual(values = clr_vals) +
        theme_prism() +
        theme(legend.position = "hide") +
        ggtitle(paste0(x, "\nTransformed & QN"))
      if (!quantileNormaliseAll) {
        return(list(q1, p1))
      } else {
        return(list(p1))
      }
    }), toPlot)
  }), markerPairTypes)
  
  cat(":::::::: {.panel-tabset}\n\n")
  lapply(names(markerPairsList), function(x) {
    cat("####", x, " \n\n")
    cat(":::::: {.panel-tabset}\n\n")
    lapply(names(markerPairsList[[x]]), function(y) {
      cat("#####", y, " \n\n")
      print(markerPairsList[[x]][[y]][[1]])
      if (length(markerPairsList[[x]][[y]]) == 2)
        print(markerPairsList[[x]][[y]][[2]])
      cat("\n\n")
    })
    cat("::::::\n\n")
    cat("\n\n")
  })
  cat("::::::::\n\n")
  cat("\n\n")
}
```

```{r, label = "save marker pair scatterplots", results = "hide", include = FALSE}
if (gimmePDFs && exists("markerPairsList")) {
  lapply(names(markerPairsList), function(x) {
    lapply(names(markerPairsList[[x]]), function(y) {
      pdf(file = file.path(pdfDir, paste0(y, "_scatter.pdf")), width = 5, height = 5, bg = "transparent")
      print(markerPairsList[[x]][[y]][[1]])
      if (length(markerPairsList[[x]][[y]]) == 2)
        print(markerPairsList[[x]][[y]][[2]])
      dev.off()
    })
  })
}
```

### Heatmap

:::::::: panel-tabset
#### Marker Per Cluster

:::::: panel-tabset

```{r, label = "type marker heatmap", results = 'asis', fig.height=ceiling(length(numbersOfClusters)/4), fig.width=ceiling(length(useMarkers)/3.5)}
heatmapsList <- list()
if("type" %in% markerType && "state" %in% markerType) {
  cat("##### Type Markers Only \n\n")
  pET <- plotExprHeatmap(
    x = sce, 
    assay = exprsToUse, 
    features = "type", 
    by = "cluster_id",
    k = clusteringMethodToUse, 
    bars = TRUE,
    perc = TRUE, 
    col_clust = FALSE, 
    hm_pal = hmPalToUse,
    k_pal = coloursList[[paste0(mergeBy, knn)]]
    )
  heatmapsList[["Type"]] <- pET
  print(pET)
  pETdf <- pET@matrix[row_order(pET),]
  topLineageTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(pETdf))
  )
  topLineageTable$Top_Lineage_Markers <- NA
  for (i in seq_along(1:nrow(topLineageTable))) {
    row_vals <- pETdf[i,]; topLineageTable$Top_Lineage_Markers[i] <- paste(names(head(sort(row_vals[row_vals > 0.4], decreasing = TRUE), n = 2)), collapse = " ")
  }
  topLineageTable <- topLineageTable[mixedorder(topLineageTable$Cluster), ]
  print(
    htmltools::tagList(
      DT::datatable(
        data = topLineageTable, 
        rownames = FALSE, 
        colnames = c("Cluster", "Top 2 Type Markers"), 
        filter = "top", 
        width = 800, 
        caption = "Top Lineage Markers have to have mean expression > 0.4 to be included, so you may only see 1 marker")
      )
    )
  cat("\n\n")
  
  cat("##### State Markers Only \n\n")
  pES <- plotExprHeatmap(
    x = sce, 
    features = "state", 
    by = "cluster_id", 
    k = clusteringMethodToUse,
    bars = TRUE, 
    perc = TRUE, 
    col_clust = FALSE, 
    hm_pal = hmPalToUse,
    k_pal = coloursList[[paste0(mergeBy, knn)]]
    )
  heatmapsList[["State"]] <- pES
  print(pES)
  pESdf <- pES@matrix[row_order(pES),]
  topStateTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(pESdf))
  )
  topStateTable$Top_State_Markers <- NA
  for (i in seq_along(1:nrow(topStateTable))) {
    topStateTable$Top_State_Markers[i] <- paste(names(head(sort(pESdf[i,], decreasing = TRUE), n = 3)), collapse = " ")
    topStateTable$Bottom_State_Markers[i] <- paste(names(tail(sort(pESdf[i,], decreasing = TRUE), n = 3)), collapse = " ")
  }
  topStateTable <- topStateTable[mixedorder(topStateTable$Cluster), ]
  print(
    htmltools::tagList(
      DT::datatable(
        data = topStateTable, 
        rownames = FALSE, 
        colnames = c("Cluster", "Top 3 State Markers", "Bottom 3 State Markers"), 
        filter = "top", 
        width = 800
        )
      )
    )
  cat("\n\n")
}
```

```{r, label = "save type/state heatmaps", results = "hide", include = FALSE}
if (gimmePDFs) {
  if (length(heatmapsList) == 2) {
    for (n in names(heatmapsList)) {
      pdf(file = file.path(pdfDir, paste0(n, "_Marker_Heatmap.pdf")), height = ceiling(length(numbersOfClusters)/4.5), width = ceiling(length(useMarkers)/4))
      print(heatmapsList[[n]])
      dev.off()
    }
  }
}
```

##### All Markers

```{r, label = "plot marker expression cluster", fig.height=ceiling(length(numbersOfClusters)/4), fig.width=if((length(useMarkers)/2) < 10){10}else{(ceiling(length(useMarkers)/2.5))}}
exH <- plotExprHeatmap(x = sce, features = NULL, by = "cluster_id", k = clusteringMethodToUse, bars = TRUE, perc = TRUE, col_clust = FALSE, hm_pal = hmPalToUse, k_pal = coloursList[[paste0(mergeBy, knn)]])
exH
```

```{r, label = "top marker table", include = TRUE}
if("type" %in% markerType && "state" %in% markerType) {
  topMarkerTable <- dplyr::left_join(topLineageTable, topStateTable)
  openxlsx::write.xlsx(as.data.frame(topMarkerTable), file.path(excelDir, "topMarkerTable.xlsx"))
  DT::datatable(
    data = topMarkerTable, 
    rownames = FALSE, 
    filter = "top", 
    width = 800, 
    colnames = c("Cluster ID", "Top 2 Lineage Markers", "Top 3 State Markers", "Bottom 3 State Markers"),
    caption = "Top Lineage Markers have to have mean expression > 0.4 to be included, so you may only see 1 marker")
} else { 
  f <- NULL
  exT <- plotExprHeatmap(x = sce, features = f, by = "cluster_id", k = clusteringMethodToUse, bars = TRUE, perc = TRUE, col_clust = FALSE)
  exTDF <- exT@matrix[row_order(exT),]
  topMarkerTable <- data.frame(
    "Cluster" = paste("Cluster", rownames(exTDF))
  )
  topMarkerTable$Top_Markers <- NA
  for (i in seq_along(1:nrow(topMarkerTable))) {
    topMarkerTable$Top_Markers[i] <- paste(names(head(sort(exTDF[i,], decreasing = TRUE), n = 3)), collapse = " ")
  }
  DT::datatable(topMarkerTable, rownames = FALSE, filter = "top", width = 800)
}
```

::::::

```{r, label = "save expression cluster", include = FALSE}
if (gimmePDFs) {
  pdf(
    file = file.path(pdfDir, "marker_per_cluster_heatmap.pdf"),
    width = if((length(useMarkers)/2) < 10){10}else{(ceiling(length(useMarkers)/3))},
    height = ceiling(length(numbersOfClusters)/4.5), bg = "transparent"
    )
  print(plotExprHeatmap(x = sce, features = NULL, by = "cluster_id", k = clusteringMethodToUse, bars = TRUE, perc = TRUE, col_clust = FALSE, hm_pal = hmPalToUse, k_pal = coloursList[[paste0(mergeBy, knn)]]))
  dev.off()
}
```

#### Sample Per Cluster

:::::: panel-tabset

##### All Clusters

```{r, label = "sample per cluster heatmap", fig.height=ceiling(length(numbersOfClusters)/3.5), fig.width=(ceiling(length(md$sample_id)/1.2))}
pFH <- plotFreqHeatmap(
  x = sce,
  k = clusteringMethodToUse, 
  normalize = TRUE, 
  col_clust = FALSE, 
  bars = TRUE, 
  perc = TRUE, 
  hm_pal = hmPalToUse,
  k_pal = coloursList[[paste0(mergeBy, knn)]]
  )
for (con in conditions) {
  pFH@top_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pFH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pFH@top_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pFH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pFH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pFH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  pFH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pFH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
}
pFH
```

```{r, label = "save cluster per sample", include = FALSE}
if (gimmePDFs) {
  pdf(
    file = file.path(pdfDir, "sample_per_cluster_heatmap.pdf"), 
    height = ceiling(length(numbersOfClusters)/4),
    width = (ceiling(length(md$sample_id)/1.6)),
    bg = "transparent"
  )
  print(pFH)
  dev.off()
}
```

```{r, label = "function for custom heatmap"}
x <- sce
by <- "cluster_id"
k <- clusteringMethodToUse
assay = "exprs"
fun = "median"
scale = "first"
q = 0.01
features = NULL
hm_pal = rev(brewer.pal(11, "RdYlBu"))
left_anno = TRUE
row_clust = TRUE
row_anno = TRUE
col_anno = TRUE
perc = TRUE
row_clust = TRUE 
col_clust = TRUE
row_dend = TRUE
col_dend = TRUE
linkage = c("average")

x <- x[unique(CATALYST:::.get_features(x, features)), ]
if (by != "sample_id") {
  CATALYST:::.check_k(x, k)
  x$cluster_id <- cluster_ids(x, k)
}
if (by == "both") {
  by <- c("cluster_id", "sample_id")
}
.do_agg <- function() {
  z <- CATALYST:::.agg(x, by, fun, assay)
  if (length(by) == 1) {
    return(z)
  }
  set_rownames(do.call("rbind", z), levels(x$cluster_id))
}
.do_scale <- function() {
  if (scale == "first") {
    z <- assay(x, assay)
    z <- CATALYST:::.scale_exprs(z, 1, q)
    assay(x, assay, FALSE) <- z
    return(x)
  } else {
    CATALYST:::.scale_exprs(z, 1, q)
  }
}
z <- switch(scale,
            first = {
              x <- .do_scale()
              .do_agg()
            },
            last = {
              z <- .do_agg()
              .do_scale()
            },
            never = {
              .do_agg()
            }
)
if (length(by) == 1) {
  z <- t(z)
}
if (scale != "never" && !(assay == "counts" && fun == "sum")) {
  qs <- round(quantile(z, c(0.01, 0.99)) * 5) / 5
  lgd_aes <- list(at = seq(qs[1], qs[2], 0.2))
} else {
  lgd_aes <- list()
}
lgd_aes$title_gp <- gpar(
  fontsize = 10, fontface = "bold",
  lineheight = 0.8
)
```

##### Selected Clusters

:::: panel-tabset

```{r, label = "heatmap by clusters", fig.height = 4, fig.width = 8, results = 'asis', message = FALSE}
heatmapClustersList <- list()
for (i in seq_along(selectedClustersList)) {
  if (length(selectedClustersList[[i]]) >= 2) {
    
    cat("######", names(selectedClustersList)[i], "\n\n")
    cl <- selectedClustersList[[i]]
    x1 <- filterSCE(x, cluster_id %in% cl, k = clusteringMethodToUse)
    x1$cluster_id <- droplevels(x1$cluster_id)
    
    ns <- table(x[[by[1]]])
    ns2 <- ns[which(names(ns) %in% cl)]
    fq <- round(ns/sum(ns) * 100, 2)
    txt <- sprintf("%s%%(%s)", fq, names(fq))
    txt <- txt[which(gsub(".*\\(|\\).*", "", txt) %in% names(ns2))]
    foo <- row_anno_text(txt, just = "center", gp = gpar(fontsize = 8), location = unit(0.5, "npc"))
    right_anno <- rowAnnotation(n_cells = row_anno_barplot(x = as.matrix(ns2), width = unit(2, "cm"), gp = gpar(fill = "grey", col = "white"), border = FALSE, axis = TRUE, bar_width = 0.8), foo = foo)
    
    left_anno <- CATALYST:::.anno_clusters(x1, "cluster_id", NULL, coloursList[[names(selectedClustersList)[[i]]]], NULL)
    
    z1 <- z[selectedClustersList[[i]], ]
    h1 <- Heatmap(
      matrix = z1, 
      name = paste0("median scaled\nexpression"),
      col = hmPalToUse,
      cluster_rows = TRUE,
      cluster_columns = FALSE,
      show_row_dend = TRUE,
      show_column_dend = TRUE,
      show_row_names = (is.null(left_anno) || isTRUE(by == "sample_id")) && !perc, row_names_side = ifelse(by[1] == "cluster_id" || isFALSE(row_anno) && !row_dend || isFALSE(row_clust), "left", "right"),
      left_annotation = left_anno,
      right_annotation = right_anno,
      rect_gp = gpar(col = "white"),
      heatmap_legend_param = lgd_aes
    )
    print(h1)
    heatmapClustersList[[i]] <- h1
    cx <- names(selectedClustersList)[[i]] |> gsub(pattern = ": .*", replacement = "", x = _) |> gsub(pattern = "Contrast ", replacement = "Contrast_", x = _) |> gsub(pattern = "\\/", replacement = ".", x = _)
    sx <- names(selectedClustersList)[[i]] |> gsub(pattern = ".*\\: ", replacement = "", x = _) |> gsub(pattern = " ", replacement = "_", x = _) |> gsub(pattern = "\\/", replacement = ".", x = _)
    if (gimmePDFs) {
      pdf(file = paste0(pdfDir, "/", cx, "_", sx, "_heatmaps.pdf"), width = 10, height = ceiling(length(cl)*0.5)+1)
      print(h1)
      dev.off()
    }
    cat("\n\n")
  } else {
    cat("######", names(selectedClustersList)[i], "\n\n")
    cat("There is only one (or zero) cluster(s) in this group, and the heatmaps currently do not support a single cluster. Sorry!")
    cat("\n\n")
  }
}
```

```{r, label = "Prepare Seurat object for heatmap", eval = FALSE}
#### Per-Cell Heatmap
scData <- suppressWarnings(Seurat::as.Seurat(x = sce, counts = "exprsTransformed", data = "exprsQuantNorm"))
scData <- Seurat::ScaleData(scData, assay = "originalexp", verbose = FALSE)
Seurat::Idents(scData) <- "cluster_id"
Seurat::DoHeatmap(
  object = scData,
  assay = "originalexp",
  slot = "scale.data",
  features = gsub("_", "-", useMarkers),
  group.by = "cluster_id",
  group.colors = coloursList[[paste0(mergeBy, knn)]],
  size = 4,
  vjust = 0.1,
  group.bar = TRUE
)
```
::::
::::::
::::::::

### Ridge Plot

```{r, label = "ridge plot", fig.height=ceiling(length(useMarkers)/3)*2.5, fig.width=16}
pCE <- plotClusterExprs(sce, k = clusteringMethodToUse, features = NULL)
pCE$facet$params$ncol <- 12
pCE$facet$params$nrow <- ceiling(length(useMarkers)/12)
pCE + theme_prism()
```

```{r, label = "save ridge plot", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Ridge_Plot.pdf"), width = 16, height = ceiling(length(useMarkers)/3)*2.5)
  print(pCE + theme_prism())
  dev.off()
}
```

### Barplots

:::::::: panel-tabset

#### Cells per Cluster

```{r, label = "cells per cluster", fig.height=4, fig.width=8}
ns <- table(cluster_id = cluster_ids(sce, clusteringMethodToUse), sample_id = sample_ids(sce))
df <- as.data.frame(ns)
df$cluster_id <- factor(df$cluster_id, levels = gtools::mixedsort(unique(as.character(df$cluster_id))))
df$condition <- md$condition[match(df$sample_id, md$sample_id)]
cellsPerClusterBarplot <- ggplot(df, aes(x = cluster_id, y = Freq, fill = cluster_id)) +
  geom_bar(stat = "identity") + 
  scale_fill_manual(values = coloursList[[paste0(mergeBy, knn)]]) + 
  theme_bw() +
  ylab("Number of cells") + 
  xlab("Cluster") + 
  theme_prism(axis_text_angle = 90) +
  theme(legend.position = "hide", panel.grid.major.x = element_blank())
cellsPerClusterBarplot
```

```{r, label = "cells per cluster facet", fig.height = ceiling(length(condLevels)/2)*4, fig.width = 16}
cellsPerClusterBarplot + facet_wrap(~condition, scales = "free") + theme_prism(axis_text_angle = 90) + theme(legend.position = "hide")
```

```{r, label = "cells per cluster facet sample", fig.height=(4*(ceiling(length(unique(df$sample_id)))/3)), fig.width = 18}
cellsPerClusterBarplot + facet_wrap(~sample_id, ncol = 3, scales = "free")
```

```{r, label = "marker frequency per cluster", fig.height=6, include = FALSE}
x1 <- as.data.frame(t(assay(sce, "exprsQuantNorm")))
x1$cluster_id <- cluster_ids(sce, clusteringMethodToUse)

x1mean <- data.frame("cluster_id" = gtools::mixedsort(levels(as.factor(x1$cluster_id))))
for (m in useMarkers) {
  x1mean[[m]] <- NA
  for (cl in levels(as.factor(x1$cluster_id))) {
    x1mean[cl,m] <- mean(x1[[m]][which(x1$cluster_id == cl)])
  }
}
x1rf <- x1mean
for (m in useMarkers) {
  for (cl in levels(as.factor(x1$cluster_id))) {
    x1rf[cl,m] <- x1mean[cl,m]/sum(x1mean[cl,-1])
  }
}
x1meanmelt <- melt(x1mean)
x1meanmelt$cluster_id <- factor(x1meanmelt$cluster_id, levels = unique(x1meanmelt$cluster_id))
x1rfmelt <- melt(x1rf)
x1rfmelt$cluster_id <- factor(x1rfmelt$cluster_id, levels = unique(x1rfmelt$cluster_id))
ggplot(x1rfmelt, aes(x = cluster_id, y = value, fill = variable)) + 
  geom_bar(stat = "identity") + 
  theme_prism() + 
  scale_fill_manual(values = coloursList[[paste0(mergeBy, knn)]]) +
  xlab("Cluster ID") +
  ylab("Proportion of the Quantile Scaled Mean")
```

```{r, label = "cells per cluster per sample table"}
.cells_wide <- tidyr::pivot_wider(df, names_from = cluster_id, values_from = Freq) |> dplyr::select(-condition) |> tibble::column_to_rownames("sample_id") |> t() |> data.frame() |> tibble::rownames_to_column("cluster_id")
DT::datatable(.cells_wide, rownames = FALSE)
openxlsx::write.xlsx(
  x = .cells_wide,
  file = file.path(excelDir, "Cells_Per_Cluster_Per_Sample.xlsx")
)
```

#### Cluster Frequency

```{r, label = "Cluster barplot 1", fig.width = ceiling(length(condLevels)/2)*10, fig.height=5}
{ sorted_ids <- gtools::mixedsort(names(coloursList[[paste0(mergeBy, knn)]]))
  sorted_pal <- coloursList[[paste0(mergeBy, knn)]][sorted_ids]
  plotAbundances(sce, k = clusteringMethodToUse, by = "sample_id", k_pal = coloursList[[paste0(mergeBy, knn)]]) + facet_grid(~condition, space = "free", scales="free") + theme_prism(axis_text_angle = 45) + scale_fill_manual(values = sorted_pal, breaks = sorted_ids) }
```

```{r, label = "save cluster barplot", include = FALSE}
if (gimmePDFs) {
  sorted_ids <- gtools::mixedsort(names(coloursList[[paste0(mergeBy, knn)]]))
  sorted_pal <- coloursList[[paste0(mergeBy, knn)]][sorted_ids]
  pdf(file = file.path(pdfDir, "cluster_barplot.pdf"), width = ceiling(length(condLevels)/2)*10, height = 5, bg = "transparent")
  print(plotAbundances(sce, k = clusteringMethodToUse, by = "sample_id", k_pal = coloursList[[paste0(mergeBy, knn)]]) + facet_grid(~condition, space = "free", scales="free") + theme_prism() + scale_fill_manual(values = sorted_pal, breaks = sorted_ids))
  dev.off()
}
```

```{r, label = "Cluster barplot 2"}
FDR_cutoff <- 0.05

ns <- table(cluster_id = cluster_ids(sce, clusteringMethodToUse), sample_id = sample_ids(sce))
fq <- prop.table(ns, 2) * 100
df <- as.data.frame(fq)
df$cluster_id <- factor(df$cluster_id, levels = gtools::mixedsort(unique(as.character(df$cluster_id))))

m <- match(df$sample_id, sce$sample_id)
for (i in c(NULL, "condition")) df[[i]] <- sce[[i]][m]

dfList <- list()
for (i in seq_along(clustersToMapList)) {
  dfC1 <- df[df$cluster_id %in% names(clustersToMapList[[i]]), ]
  dfList[[names(selectedClustersList)[[i]]]] <- dfC1
}
```

```{r, label = "Cluster barplot 3", fig.width=10, fig.height=5, results = 'asis'}
selectedBarplotList <- setNames(lapply(seq_along(dfList), function(i) {
  cat("\n\n#### DA:", names(dfList)[i], "\n\n")
  if (nrow(dfList[[i]]) >= 1) {
    g <- ggplot(dfList[[i]], aes(x = .data[["sample_id"]], y = .data[["Freq"]], fill = .data[["cluster_id"]]))
    g <- g + geom_bar(stat = "identity")
    g <- g + facet_grid(~condition, scales = "free", space = "free")
    g <- g + scale_fill_manual(values = coloursList[[names(selectedClustersList)[[i]]]])
    g <- g + ggtitle(names(dfList)[i])
    g <- g + scale_x_discrete(expand = c(0, 0)) + scale_y_continuous(expand = c(0, 0), labels = seq(0, 100, 25), limits = c(0, 100))
    g <- g + theme_prism()
    g <- g + theme(panel.border = element_blank(), panel.spacing.x = unit(1, "lines"))
    g <- g + ylab("Proportion [%]") + xlab("Sample ID")
    print(g)
    cx <- names(dfList)[[i]] |> gsub(pattern = ": .*", replacement = "", x = _) |> gsub(pattern = "Contrast ", replacement = "Contrast_", x = _) |> gsub(pattern = "\\/", replacement = "", x = _)
    sx <- names(dfList)[[i]] |> gsub(pattern = ".*\\: ", replacement = "", x = _) |> gsub(pattern = " ", replacement = "_", x = _) |> gsub(pattern = "\\/", replacement = "", x = _)
    return(list("g" = g, "cx" = cx, "sx" = sx))
  } else {
    cat("There are no DA clusters in this group.\n")
  }
  cat("\n\n")
}), names(dfList))
```

::::::::

```{r, label = "Save Cluster Barplots 3", include = FALSE}
if (gimmePDFs) {
  lapply(names(selectedBarplotList), function(x) {
    if (is.null(selectedBarplotList[[x]])) return(NULL)
    pdf(file = paste0(pdfDir, "/", selectedBarplotList[[x]][["cx"]], "_", selectedBarplotList[[x]][["sx"]], "_barplot.pdf"), width = 10, height = 5)
    print(selectedBarplotList[[x]][["g"]])
    dev.off()
  })
}
```

### Cluster Boxplots

```{r, label = "cluster boxplot dimensions", include = FALSE}
nColsToPlot <- 6
nGroupsToPlot <- nlevels(sce$condition)
nClustersToPlot <- nlevels(sce$cluster_id)
nRowsToPlot <- ceiling(nClustersToPlot/nColsToPlot)
heightToPlot <- ceiling(nRowsToPlot*5)
widthToPlot <- ceiling((5*1.2)*nColsToPlot)
```

```{r, label = "cluster boxplot", fig.width = widthToPlot, fig.height = heightToPlot}
## From CATALYST plotFreqHeatmap function
ns <- table(sce$cluster_id, sce$sample_id)
fq <- prop.table(ns, 2)
fq.z <- as.matrix(unclass(fq))
clusterBoxPlotsDF <- CATALYST:::.z_normalize(asin(sqrt(fq.z)))
clusterBoxPlotsDF <- as.data.frame(t(clusterBoxPlotsDF))
clusterBoxPlotsDF <- tibble::rownames_to_column(clusterBoxPlotsDF, "sample_id") |> dplyr::left_join(md)
clusterBoxPlotsDFMelt <- melt(clusterBoxPlotsDF)

ctsList <- lapply(names(daList), function(da) { strsplit(da, " over ")[[1]] }) |> setNames(names(daList))

clusterPvalues <- lapply(levels(kids), function(k) {
  pvals <- lapply(names(daList), function(da) { daList[[da]]$p_adj[daList[[da]]$cluster_id == k] }) |> setNames(names(daList)) |> unlist()
  if(any(is.na(pvals))) {pvals[is.na(pvals)] <- 1}
  y_max <- max(clusterBoxPlotsDFMelt$value)
  spacing <- (y_max - min(clusterBoxPlotsDFMelt$value)) * 0.2
  y_positions <- y_max + (1:length(ctsList)) * spacing
  pval_df <- data.frame(
    group1 = sapply(ctsList, `[`, 1),
    group2 = sapply(ctsList, `[`, 2),
    p = pvals,
    y.position = y_positions
  )
  pval_df <- add_significance(pval_df, p.col = "p", output.col = "p.signif")
  
  pval_df
}) |> setNames(levels(kids))
maxY <- lapply(seq_along(clusterPvalues), function(i) {clusterPvalues[[i]]$y.position}) |> unlist() |> max()

clusterBoxplotList <- lapply(levels(kids), function(k) {
  ctp <- as.character(k)
  ggplot(clusterBoxPlotsDF, aes(x = .data[["condition"]], y = .data[[ctp]])) + 
    ## geom_vline(xintercept = (seq_along(condLevels)+0.5)[-length(condLevels)], linetype = "dashed", alpha = 0.6) +
    scale_y_continuous(limits = c(min(clusterBoxPlotsDFMelt$value), maxY)) +
    geom_boxplot(aes(fill = condition), alpha = 0.2, show.legend = FALSE, outlier.shape = NA) +
    geom_beeswarm(aes(fill = condition), pch = 21, size = 3, stroke = 0.3, cex = 2, corral = "random", corral.width = 0.9) +
    scale_fill_manual(values = coloursList$condition) +
    stat_pvalue_manual(clusterPvalues[[ctp]], label = "p.signif", tip.length = 0, hide.ns = TRUE, label.size = 9, bracket.size = 0.8, bracket.shorten = 0.1, bracket.nudge.y = -0.1) +
    theme_prism(axis_text_angle = 45) +
    ggtitle(paste("Cluster", ctp)) +
    theme(legend.position = "hide", axis.title = element_blank())
})
ylab <- textGrob("Normalised Frequency", rot = 90, gp = gpar(fontface = "bold", fontsize = 16))
xlab <- textGrob("Condition", gp = gpar(fontface = "bold", fontsize = 16))
grid.arrange(grobs = clusterBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
```

```{r, label = "save cluster boxplot", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Cluster_Boxplot.pdf"), width = widthToPlot, height = heightToPlot)
  grid.arrange(grobs = clusterBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
  dev.off()
}
```

### Markers per Condition

```{r, label = "generate marker tables"}
medianScaledDF <- CATALYST:::.agg(sce, "sample_id", "median", exprsToUse)
markerBoxplotsMeltDF <- melt(medianScaledDF, varnames = c("antigen", "sample_id"[length("sample_id")]))
i <- match(markerBoxplotsMeltDF$sample_id, sce$sample_id)
j <- setdiff(names(colData(sce)), c(names(markerBoxplotsMeltDF), "cluster_id"))
markerBoxplotsMeltDF <- cbind(markerBoxplotsMeltDF, colData(sce)[i, j, drop = FALSE])
medianScaledDF <- medianScaledDF |>
  t() |>
  data.frame(check.names = FALSE) |>
  tibble::rownames_to_column("sample_id") |>
  dplyr::left_join(md, by = "sample_id")
openxlsx::write.xlsx(as.data.frame(medianScaledDF), file = file.path(excelDir, "Median_Markers_Per_Sample_Expression.xlsx"))
```

```{r, label = "draw marker boxplots", fig.height = heightToPlot*0.7, fig.width = widthToPlot}
markerPvalues <- lapply(useMarkers, function(marker) {
  pvals <- lapply(names(dsList)[grep("Merged", names(dsList))], function(ds) { dsList[[ds]]$tbl_DS$p_adj[dsList[[ds]]$tbl_DS$marker_id == marker] }) |> setNames(names(dsList)[grep("Merged", names(dsList))]) |> unlist()
  if(length(pvals) == 0) {pvals <- 1}
  y_max <- max(markerBoxplotsMeltDF$value)
  spacing <- (y_max - min(markerBoxplotsMeltDF$value)) * 0.1
  y_positions <- y_max + (1:length(ctsList)) * spacing
  pval_df <- data.frame(
    group1 = sapply(ctsList, `[`, 1),
    group2 = sapply(ctsList, `[`, 2),
    p = pvals,
    y.position = y_positions
  )
  pval_df <- add_significance(pval_df, p.col = "p", output.col = "p.signif")
  pval_df
}) |> setNames(useMarkers)
maxY <- lapply(seq_along(markerPvalues), function(i) {markerPvalues[[i]]$y.position}) |> unlist() |> max()

markerBoxplotList <- lapply(useMarkers, function(marker) {
  ggplot(medianScaledDF, aes(x = .data[["condition"]], y = .data[[marker]])) + 
    ## geom_vline(xintercept = (seq_along(condLevels)+0.5)[-length(condLevels)], linetype = "dashed", alpha = 0.6) +
    scale_y_continuous(limits = c(min(markerBoxplotsMeltDF$value), maxY)) +
    geom_boxplot(aes(fill = condition), alpha = 0.2, show.legend = FALSE, outlier.shape = NA) +
    geom_beeswarm(aes(fill = condition), pch = 21, size = 3, stroke = 0.3, cex = 2, corral = "random", corral.width = 0.9) +
    scale_fill_manual(values = coloursList$condition) +
    stat_pvalue_manual(markerPvalues[[marker]], label = "p.signif", tip.length = 0, hide.ns = TRUE, label.size = 9, bracket.size = 0.8, bracket.shorten = 0.1, bracket.nudge.y = -0.1) +
    theme_prism(axis_text_angle = 45) +
    ggtitle(marker) +
    theme(legend.position = "hide", axis.title = element_blank())
})
ylabtext <- switch(exprsToUse, "exprsTransformed" = "Transformed", "exprsQuantNorm" = "Quantile Normalised")
ylab <- textGrob(paste0("Median ", ylabtext, " Expression"), rot = 90, gp = gpar(fontface = "bold", fontsize = 16))
xlab <- textGrob("Condition", gp = gpar(fontface = "bold", fontsize = 16))
grid.arrange(grobs = markerBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
```

::::::::::

```{r, label = "save marker boxplots", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, "Marker_per_Sample_boxplot.pdf"), width = widthToPlot, height = heightToPlot)
  grid.arrange(grobs = markerBoxplotList, ncol = nColsToPlot, left = ylab, bottom = xlab)
  dev.off()
}
```

## `r paste(dimRedMethodToUse)`

::: {.callout-tip collapse="true" title="Markers used for DR generation"}

```{r, label = "show features to DR"}
paste(featuresToUmap)
```
:::

:::::::::: panel-tabset

### Full `r paste(dimRedMethodToUse)`

:::::::: panel-tabset

```{r, label = "DR all cells", fig.height=6, fig.width=7, results = 'asis'}
if (!is.null(drCellPerCondition)) {
  cat("#### All Cells", dimRedMethodToUse, "\n\n")
  cat("Since you clustered on all available cells (after initial downsampling), here is a DR plot of all available cells. All subsequent DR plots are subset from this main DR plot.\n\n")
  dr0 <- plotUMAPNice(df = umapDFList$All, colour_by = "cluster_id", show_density = FALSE, alpha = 0.9, show_cluster_labels = TRUE, show_grids = FALSE, text_size = 14) +
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
  dr0
}
```

```{r, label = "save DF all cells", include = FALSE}
if (!is.null(drCellPerCondition)) {
  if (gimmePDFs) {
    pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_All_Cells.pdf")), width = (3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, height = 5)
    print(dr0)
    dev.off()
  }
}
```

```{r, label = "DR all cells unlabelled", fig.height=6, fig.width=(3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, results = 'asis'}
if (!is.null(drCellPerCondition)) {
  cat("\n\n")
  dr0a <- plotUMAPNice(df = umapDFList$All, colour_by = "cluster_id", show_density = FALSE, alpha = 0.9, show_cluster_labels = FALSE, show_grids = FALSE, text_size = 14) +
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
  dr0a
  cat("\n\n")
}
```

```{r, label = "save DF all cells unlabelled", include = FALSE}
if (!is.null(drCellPerCondition)) {
  if (gimmePDFs) {
    pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_All_Cells_unlabelled.pdf")), width = (3.5*ceiling(length(levels(umapDFList$All$cluster_id))/18))+0.5, height = 5)
    print(dr0a)
    dev.off()
  }
}
```

#### All Samples: Selected Clusters

```{r, label = "main cluster DR 1", fig.height=6, fig.width=7}
if (is.null(drCellPerCondition)) {
  umapsDFToPlot <- "All"
} else {
  umapsDFToPlot <- "Downsampled"
}
openxlsx::write.xlsx(x = as.data.frame(umapDFList[[umapsDFToPlot]]), file = file.path(excelDir, paste0(dimRedMethodToUse, "_Coordinates.xlsx")))
dr1a <- plotUMAPNice(
  df = as.data.frame(umapDFList[[umapsDFToPlot]]),
  colour_by = "cluster_id",
  show_density = FALSE,
  alpha = 0.9,
  show_cluster_labels = TRUE,
  show_grids = FALSE,
  text_size = 14
  ) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr1a
```

```{r, label = "main cluster DR 2", fig.height=6, fig.width=7}
dr1b <- plotUMAPNice(df = as.data.frame(umapDFList[[umapsDFToPlot]]), colour_by = "cluster_id", show_density = FALSE, alpha = 0.8, show_cluster_labels = FALSE, show_grids = FALSE, text_size = 14) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr1b
```

```{r, label =  "save full DR", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_full_labelled.pdf")), width = 7, height = 6, bg = "transparent")
  print(dr1a)
  dev.off()
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_full_unlabelled.pdf")), width = 7, height = 6, bg = "transparent")
  print(dr1b)
  dev.off()
}
```

#### All Samples: Other k values

```{r, label = "plot k DRs", fig.height=ceiling(6*length(kValuesIWant)), fig.width=9}
pDRList <- setNames(lapply(kValuesIWant, function(i) {
  plotUMAPNice(df = umapDFList[[umapsDFToPlot]], colour_by = paste0(mergeBy, i), show_grids = FALSE) +
    guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
}), paste0(mergeBy, kValuesIWant))
plot_grid(plotlist = pDRList, ncol = 1, align = "hv")
```

::::::::

### Condition `r paste(dimRedMethodToUse)`

```{r, label = "make DRs split by condition"}
## Get the number of cells per condition for the UMAP
conds <- as.factor(sce@metadata$experiment_info$condition)
cellN <- n_cells(sce)
names(cellN) <- conds
condLevels <- levels(conds)
condLabels <- NA
for (i in seq_along(condLevels)) {
  condLabels[i] <- paste(condLevels[i], "\n n =", as.numeric(nrow(umapDFList[[umapsDFToPlot]][which(umapDFList[[umapsDFToPlot]]$condition == condLevels[i]),])))
}
names(condLabels) <- condLevels
dr2a <- plotUMAPNice(df = as.data.frame(umapDFList[[umapsDFToPlot]]), colour_by = "cluster_id", show_density = FALSE, alpha = 0.85, text_size = 14, show_cluster_labels = TRUE, show_grids = FALSE) +
  facet_wrap("condition", labeller = labeller("condition" = condLabels), ncol = 2) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
```

```{r, label = "plot DRs split by condition", fig.height=ceiling(length(condLevels)/2)*6, fig.width=12}
dr2a
```

```{r, label = "plot DRs split by condition 2", fig.height=ceiling(length(condLevels)/2)*6, fig.width=12}
dr2b <- plotUMAPNice(df = as.data.frame(umapDFList[[umapsDFToPlot]]), colour_by = "cluster_id", show_density = FALSE, alpha = 0.85, text_size = 14, show_cluster_labels = FALSE, show_grids = FALSE) +
  facet_wrap("condition", labeller = labeller("condition" = condLabels), ncol = 2) +
  guides(colour = guide_legend(override.aes = list(size = 2), keyheight = 0.5))
dr2b
```

```{r, label = "save DRs split by condition", include = FALSE}
if (gimmePDFs) {
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_condition_labelled.pdf")), width = 12, height = ceiling(length(condLevels)/2)*6, bg = "transparent")
  print(dr2a)
  dev.off()
  pdf(file = file.path(pdfDir, paste0(dimRedMethodToUse, "_condition_unlabelled.pdf")), width = 12, height = ceiling(length(condLevels)/2)*6, bg = "transparent")
  print(dr2b)
  dev.off()
}
```

```{r, label = "full DR by condition", fig.height=(6+length(conditionOrder)*2.5), fig.width=7}
u1_by_cond <- plotUMAPNice(umapDFList[[umapsDFToPlot]], "condition")
u2_by_cond <- plotUMAPNice(df = umapDFList[[umapsDFToPlot]], colour_by = "condition", show_density = FALSE) +
  facet_wrap("condition", labeller = labeller("condition" = condLabels))
plot_grid(u1_by_cond, u2_by_cond, ncol = 1, align = "hv", rel_heights = c(1.4, 1))
```

### `r paste(dimRedMethodToUse)` Per Marker

All values have been normalised to 0 - 1.

```{r, label = "DR per marker 1", results = 'asis', include = FALSE}
UMAPList <- setNames(lapply(c("With", "Without"), function(yn) {
  var <- switch(
    yn,
    "With" = TRUE,
    "Without" = FALSE
  )
  l1 <- setNames(lapply(useMarkers, function(marker) {
    m1 <- plotUMAPNice(
      df = as.data.frame(umapDFList[[umapsDFToPlot]]),
      colour_by = marker,
      name = "Normalised\nExpression\n",
      show_cluster_labels = var
      )
    m1 <- m1 + ggtitle(marker) + theme(plot.title = element_text(hjust = 0.5, face = "bold"), axis.title = element_blank())
    med_expr <- median(m1$data[[marker]], na.rm = TRUE)
    pal <- hcl.colors(64, viridisColour)
    fill_col <- pal[max(1, round(med_expr * 63) + 1)]
    text_col <- ifelse(med_expr > 0.5, "white", "black")
    m1 <- m1 + annotate(geom = "label", x = max(m1$data$x)*0.8, y = max(m1$data$y)*0.95, label = marker, fill = fill_col, color = text_col, size = 6, fontface = "bold", label.padding = unit(0.3, "lines"))
    m1
  }), useMarkers)
  l1[["legend"]] <- cowplot::get_legend(l1[[1]])
  for (marker in useMarkers) {
    l1[[marker]] <- l1[[marker]] + theme(legend.position = "hide")
  }
  return(l1)
}), c("With Labels", "Without Labels"))
```

:::::::: panel-tabset

```{r, label = "DR per per marker 2", fig.width=16, fig.height=ceiling(length(useMarkers)/3)*3, dependson="umaps_per_marker1", results = 'asis'}
invisible(lapply(names(UMAPList), function(yn){
  cat("####", yn, "\n\n")
  print(plot_grid(plotlist = UMAPList[[yn]], ncol = 4, align = "hv"))
  cat("\n\n")
}))
```

::::::::

```{r, label = "save DR per marker", include = FALSE}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, paste0("marker_", dimRedMethodToUse)))
  lapply(names(UMAPList), function(m) {
    dir.create(file.path(pdfDir, paste0("marker_", dimRedMethodToUse), m))
    pdf(file = file.path(pdfDir, paste0(m, "_all_markers_", dimRedMethodToUse, ".pdf")), width = 16, height = ceiling(length(useMarkers)/3)*3)
    print(plot_grid(plotlist = UMAPList[[m]], ncol = 4))
    dev.off()
    lapply(names(UMAPList[[m]]), function(n) {
      pdf(file = file.path(pdfDir, paste0("marker_", dimRedMethodToUse), m, paste0(n, "_", dimRedMethodToUse, ".pdf")), width = 4, height = 4, bg = "transparent")
      if (n != "legend") {
        print(UMAPList[[m]][[n]])
      } else if (n == "legend") {
        plot(UMAPList[[m]][[n]])
      }
      dev.off()
    })
  })
}
```

### `r paste(dimRedMethodToUse)` Per Marker Per Condition

```{r, label = "DR per marker per cond 1", results = 'asis', include = FALSE}
DRperCondList <- setNames(lapply(c("With", "Without"), function(yn) {
  var <- switch(
    yn,
    "With" = TRUE,
    "Without" = FALSE
  )
  l1 <- setNames(lapply(useMarkers, function(marker) {
    m1 <- plotUMAPNice(
      df = as.data.frame(umapDFList[[umapsDFToPlot]]),
      colour_by = marker,
      name = "Normalised\nExpression\n",
      show_cluster_labels = var
      ) + 
      facet_wrap(~condition, ncol = 2) +
      ggtitle(marker)
    m1 <- m1 + theme(plot.title = element_text(hjust = 0.5, face = "bold"), axis.title = element_blank(), legend.position = "hide")
    m1
  }), useMarkers)
  return(l1)
}), c("With Labels", "Without Labels"))
```

:::::::: panel-tabset

```{r, label = "DR per marker per cond 2", fig.width=10, fig.height=5*(ceiling(length(condLevels)/2)), results = 'asis'}
invisible(lapply(names(DRperCondList), function(yn){
  cat("####", yn, "\n\n")
  cat(":::::: {.panel-tabset}\n\n")
  invisible(lapply(names(DRperCondList[[yn]]), function(marker) {
    cat("#####", marker, "\n\n")
    print(DRperCondList[[yn]][[marker]])
    cat("\n\n")
  }))
  cat("::::::\n\n")
}))
```

::::::::

```{r, label = "save DR per marker per cond", include = FALSE}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse)))
  lapply(names(DRperCondList), function(m) {
    dir.create(file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse), m))
    lapply(names(DRperCondList[[m]]), function(n) {
      pdf(file = file.path(pdfDir, paste0("marker_per_cond_", dimRedMethodToUse), m, paste0(n, "_", dimRedMethodToUse, ".pdf")), width = 10, height = 5*(ceiling(length(condLevels)/2)), bg = "transparent")
      if (n != "legend") {
        print(DRperCondList[[m]][[n]])
      } else if (n == "legend") {
        plot(DRperCondList[[m]][[n]])
      }
      dev.off()
    })
  })
}
```

```{r, label = "gated_dr_plots", results = 'asis', fig.height=6, fig.width=18}
if (runScGate && "Gated_Cells" %in% colnames(colData(sce))) {
  cat("### scGate Gating \n\n")
  cat(":::::::: {.panel-tabset}\n\n")
  cat("\n\n#### All \n\n")

  mp <- parse_marker_pairs(smd)
  models <- if (!is.null(mp)) mp$models else list()

  print(
    plotUMAPNice(
      df = umapDFList$Downsampled,
      colour_by = "Gated_Cells",
      show_density = FALSE,
      alpha = 1,
      facet_by = "condition",
      show_cluster_labels = TRUE,
      cluster_labels = "Gated_Cells"
    )
  )

  cat("\n\n")
  lapply(names(models), function(mpt) {
    cat("####", mpt, "\n\n")
    y <- gsub("[ -]", "", mpt)
    m1 <- models[[mpt]][[1]]
    m2 <- models[[mpt]][[2]]
    if (!m1 %in% rownames(sce)) {
      hit <- agrep(m1, rownames(sce), value = TRUE)
      if (length(hit) == 0) return(invisible(NULL))
      m1 <- hit[[1]]
    }
    if (!m2 %in% rownames(sce)) {
      hit <- agrep(m2, rownames(sce), value = TRUE)
      if (length(hit) == 0) return(invisible(NULL))
      m2 <- hit[[1]]
    }

    # Panel 1: DimPlot — is this cell type?
    p1 <- plotUMAPNice(
      df = umapDFList$Downsampled,
      colour_by = paste0("is_", y),
      show_density = FALSE,
      alpha = 1
    ) + ggtitle(paste0("Are ", mpt, "?")) +
      theme(legend.position = "none")

    # Panel 2: FeaturePlot — UCell score
    ucell_col <- paste0(mpt, "_UCell")
    p2 <- ggplot(umapDFList$Downsampled, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[[ucell_col]])) +
      geom_point(size = dr_point_size, shape = 16) +
      scale_colour_gradientn(colours = hcl.colors(64, "mako")) +
      labs(
        x = paste(dimRedMethodToUse, "dim 1"),
        y = paste(dimRedMethodToUse, "dim 2"),
        title = paste0(mpt, " UCell Score")
      ) +
      theme_prism() +
      theme(
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        legend.position = "none",
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5)
      )

    # Panel 3: Joint density — ks::kde weighted scatter (same as Shiny app)
    dr_theme <- theme_prism() + theme(
      axis.text = element_blank(), axis.ticks = element_blank(),
      axis.line = element_blank(), legend.position = "none",
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5)
    )
    df_joint <- umapDFList$Downsampled
    emb_mat <- as.matrix(df_joint[, c("x", "y")])

    # Sum expression weights across both markers
    w_joint <- rep(0, nrow(df_joint))
    for (mk in c(m1, m2)) {
      if (mk %in% colnames(df_joint)) {
        v <- as.numeric(df_joint[[mk]])
        v[is.na(v)] <- 0
        w_joint <- w_joint + v
      }
    }

    if (sum(w_joint) > 0) {
      w_norm <- w_joint / sum(w_joint) * length(w_joint)
      dens <- ks::kde(emb_mat, w = w_norm, eval.points = emb_mat)
      df_joint$density <- dens$estimate
    } else {
      df_joint$density <- 0
    }
    df_joint <- df_joint[order(df_joint$density), ]

    p3 <- ggplot(df_joint, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[["density"]])) +
      geom_point(size = dr_point_size, alpha = 0.7, shape = 16) +
      scale_colour_gradientn(colours = hcl.colors(64, viridisColour)) +
      ggtitle(paste0(m1, " + ", m2, " density")) +
      coord_fixed() + dr_theme +
      labs(x = paste(dimRedMethodToUse, "dim 1"),
           y = paste(dimRedMethodToUse, "dim 2"),
           colour = "Density")

    print(cowplot::plot_grid(p1, p2, p3, align = "hv", ncol = 3))
    cat("\n\n")
  })
  cat("::::::::\n\n")
}
```

::::::::::

## DA / DS

:::::::::: panel-tabset
### DA

::: {.callout-note collapse="true" title="About Differential Abundance"}

Differential abundance of clusters between conditions. Each cluster was tested individually between the two groups specified in each test, e.g., cluster 1 in KO vs WT. Each test gives a log2 fold change, and p-values. Below is the results for each test specified in the metadata. The heatmap shows the expression of each cluster per sample, in order of p-value (most statistically differentially abundant at the top). If you have entered more than one test, then the heatmaps will show the same data but the clusters will be in different orders.

To the right of the heatmap are two annotations. The first is red or blue, and denotes the log2 fold change of that cluster. The second is the p-value, and is either green (adjusted p-value \< 0.05) or grey. The results of this heatmap are the same as under the `Heatmap -> Sample Per Cluster` but with the rows re-ordered and with the log2fc and p-values.
:::

```{r, label = "htmltags1", include = FALSE}
htmltools::tagList(DT::datatable(data.frame(x = 1)))
```

:::::::: panel-tabset

```{r, label = "da_plot", fig.height=6, fig.width=13, results='asis'}
daResultsList <- setNames(lapply(names(daList), function(cont) {
  cat("\n\n")
  cat("####", cont, "Top 20 \n\n")
  pDH <- plotDiffHeatmap(sce, daList[[cont]], normalize = TRUE, all = TRUE, top_n = 20)
  for (con in conditions) {
    pDH@top_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  }
  print(pDH)
  cat("\n\n")
  cx <- cont |> gsub(pattern = " ", replacement = "_", x = _) |> gsub(pattern = "-", replacement = "_", x = _) |> gsub(pattern = "\\+", replacement = ".", x = _) |> gsub(pattern = "\\/", replacement = ".", x = _)
  ## daDF <- as.data.frame(daList[[cont]][which(daList[[cont]][[daPValToUse]] <= 0.05), c("cluster_id", "logFC", "p_val", "p_adj")])
  daDF <- as.data.frame(daList[[cont]][, c("cluster_id", "logFC", "p_val", "p_adj")])
  daDF <- daDF[order(daDF[[daPValToUse]]),]
  print(htmltools::tagList(
    datatable(daDF, rownames = FALSE, width = "600px") |>
      formatSignif(c("p_val", "p_adj", "logFC"), digits = 3) |>
      formatStyle(columns = c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")), fontWeight = "bold") |>
      formatStyle(columns = "logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold")
  ))
  cat("\n\n")
  openxlsx::write.xlsx(x = as.data.frame(daList[[cont]]), file = file.path(excelDir, paste0(cont, "_DA_Results.xlsx")))
  return(list("pDH" = pDH, "cx" = cx, "daDF" = daDF))
}), names(daList))
```

::::::::

```{r, label = "save DA results", include = FALSE}
if (gimmePDFs) {
  lapply(names(daResultsList), function(x) {
    pdf(file = file.path(pdfDir, paste0(x, "_DA_heatmap.pdf")), height = 6, width = 13)
    print(daResultsList[[x]][["pDH"]])
    dev.off()
  })
}
```

### DS

::: {.callout-note collapse="true" title="About Differential State"}

Differential state (DS) analysis tests whether the expression of individual markers within each cluster differs between conditions. Unlike DA (which tests cluster abundance), DS tests for changes in marker expression levels within each cluster.
:::

```{r, label = "htmltags2", include = FALSE}
htmltools::tagList(DT::datatable(data.frame(x = 1)))
```

:::::::: panel-tabset

```{r, label = "ds_res", results = 'asis', fig.height=12, fig.width=10}
dsPThreshToUse <- 0.05
dsResultsList <- setNames(lapply(names(dsList)[grep("Merged", names(dsList), invert = TRUE)], function(cont) {
  cat("\n\n")
  cat("####", cont, "\n\n")

  openxlsx::write.xlsx(x = as.data.frame(dsList[[cont]]$tbl_DS), file = file.path(excelDir, paste0(cont, "_DS_Full_Results.xlsx")))

  dsUp_tbl <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC >= 0)]), decreasing = TRUE); dsUp <- data.frame(dsUp_tbl[dsUp_tbl >= 1])
  dsDown_tbl <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC <= 0)]), decreasing = TRUE); dsDown <- data.frame(dsDown_tbl[dsDown_tbl >= 1])

  if (nrow(dsUp) == 1) {
    dsUp <- data.frame("Marker" = rownames(dsUp), "Number of clusters" = dsUp[1,])
  }
  if (nrow(dsDown) == 1) {
    dsDown <- data.frame("Marker" = rownames(dsDown), "Number of clusters" = dsDown[1,])
  }

  cat("**Markers up in", strsplit(cont, split = " over ")[[1]][1], "**\n\n")
  if (nrow(dsUp) >= 1) {
    print(
      kable(
        x = dsUp,
        col.names = c("Marker", "Number of clusters")) |>
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")
    )
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][1], "\n\n")
  }

  cat("**Markers up in", strsplit(cont, split = " over ")[[1]][2], "**\n\n")
  if (nrow(dsDown) >= 1) {
    print(
      kable(
        x = dsDown,
        col.names = c("Marker", "Number of clusters")) |>
        kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "left")
    )
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][2], "\n\n")
  }

  cat("**Heatmap**\n\n")
  pDH <- plotDiffHeatmap(sce, dsList[[cont]]$tbl_DS, normalize = TRUE, all = TRUE, fdr = dsPThreshToUse)
  for (con in conditions) {
    pDH@top_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  }
  print(pDH)
  cat("\n\n")
  cat("**Full Results**\n\n")
  cat("This table shows the log2FC and p-values of each marker per cluster in", cont, "\n\n")
  cat("Positive log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][1], "in a given cluster. \n\n")
  cat("Negative log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][3], "in a given cluster. \n\n")

  dsTab1 <- as.data.frame(dsList[[cont]]$tbl_DS[, c("cluster_id", "marker_id", "logFC", "p_val", "p_adj")])
  dsTab1 <- dsTab1[order(dsTab1$p_val), ]
  print(
    htmltools::tagList(DT::datatable(
      data = dsTab1,
      rownames = FALSE,
      width = "700px"
      ) |>
        DT::formatSignif(c("logFC", "p_val", "p_adj"), digits = 3) |>
        DT::formatStyle("logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold") |>
        DT::formatStyle(c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")),fontWeight = "bold")
    )
  )
  cat("\n\n")
  return(list("pDH" = pDH, "dsTab1" = dsTab1))
}), names(dsList)[grep("Merged", names(dsList), invert = TRUE)])
```

::::::::

### DS - Merged

Same as the other differential state results, but instead we have merged all clusters prior to DS, so now we see what could be considered differential abundance of each marker between conditions.

```{r, label = "htmltags3", include = FALSE}
htmltools::tagList(DT::datatable(data.frame(x = 1)))
```

:::::::: panel-tabset

```{r, label = "ds_res merged", results = 'asis', fig.height=12, fig.width=10}
dsMergedResultsList <- setNames(lapply(names(dsList)[grep("Merged", names(dsList), invert = FALSE)], function(cont) {
  cat("\n\n")
  cat("####", cont, "\n\n")

  openxlsx::write.xlsx(x = as.data.frame(dsList[[cont]]$tbl_DS), file = file.path(excelDir, paste0(cont, "_DS_Merged_Results.xlsx")))
  dsUp_tbl <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC >= 0)]), decreasing = TRUE); dsUp <- dsUp_tbl[dsUp_tbl >= 1]
  dsDown_tbl <- sort(table(dsList[[cont]]$tbl_DS$marker_id[which(dsList[[cont]]$tbl_DS[[daPValToUse]] <= dsPThreshToUse & dsList[[cont]]$tbl_DS$logFC <= 0)]), decreasing = TRUE); dsDown <- dsDown_tbl[dsDown_tbl >= 1]

  cat("**Markers up in", strsplit(cont, split = " over ")[[1]][1], "**\n\n")
  if (length(dsUp) >= 1) {
    print(dsUp)
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][1], "\n\n")
  }

  cat("**Markers up in", strsplit(cont, split = " over ")[[1]][2], "**\n\n")
  if (length(dsDown) >= 1) {
    print(dsDown)
    cat("\n\n")
  } else {
    cat("There are no up-regulated markers in", strsplit(cont, split = " over ")[[1]][2], "\n\n")
  }
  cat("**Heatmap**\n\n")
  pDH <- plotDiffHeatmap(x = sceDS, y = dsList[[cont]]$tbl_DS, normalize = TRUE, all = TRUE)
  for (con in conditions) {
    pDH@top_annotation@anno_list[[con]]@color_mapping@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@color_mapping@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@colors <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
    pDH@top_annotation@anno_list[[con]]@fun@var_env[["color_mapping"]]@full_col <- coloursList[[con]][match(names(pDH@top_annotation@anno_list[[con]]@color_mapping@colors), names(coloursList[[con]]))]
  }
  print(pDH)
  cat("\n\n")
  cat("**Full Results**\n\n")
  cat("Positive log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][1], "in a given cluster. \n\n")
  cat("Negative log2fc means the marker was higher in", strsplit(cont, split = "-| ")[[1]][3], "in a given cluster. \n\n")

  dsTab1 <- as.data.frame(dsList[[cont]]$tbl_DS[, c("cluster_id", "marker_id", "logFC", "p_val", "p_adj")])
  dsTab1 <- dsTab1[order(dsTab1$p_val), ]
  print(
    htmltools::tagList(DT::datatable(
      data = dsTab1,
      rownames = FALSE,
      width = "700px"
      ) |>
        DT::formatSignif(c("logFC", "p_val", "p_adj"), digits = 3) |>
        DT::formatStyle("logFC", color = styleInterval(cuts = 0, values = c("blue", "darkorange")), fontWeight = "bold") |>
        DT::formatStyle(c("p_val", "p_adj"), color = styleInterval(cuts = 0.05, values = c("green", "black")),fontWeight = "bold")
    )
  )
  cat("\n\n")
  return(list("pDH" = pDH, "dsTab1" = dsTab1))
}), names(dsList)[grep("Merged", names(dsList), invert = FALSE)])
```

::::::::

### Markers within Clusters

:::::::: panel-tabset

```{r, label = "markers per cluster boxplot", fig.width=18, fig.height=5, results = 'asis'}
by <- c("cluster_id", "sample_id")
ms <- CATALYST:::.agg(sce, by, "median", "exprsQuantNorm")

df <- melt(ms, varnames = c("antigen", by[length(by)]))
names(df)[ncol(df)] <- "cluster_id"
i <- match(df$sample_id, sce$sample_id)
j <- setdiff(names(colData(sce)), c(names(df), "cluster_id"))
df <- cbind(df, colData(sce)[i, j, drop = FALSE])
cat("\n\n")
openxlsx::write.xlsx(as.data.frame(df), file = file.path(excelDir, "Median_Markers_Per_Cluster_Expression.xlsx"))
individualClusterMarkerBoxplots <- setNames(lapply(levels(markerBoxplotsMeltDF[[paste0(mergeBy, knn)]]), function(i) {
  cat("#### Cluster", i, "\n\n")
  df1 <- df[df$cluster_id == as.character(i),]
  g1 <- ggplot(df1, aes(x = antigen, y = value, fill = condition)) +
    scale_y_continuous(limits = c(0, 1), breaks = c(0.25, 0.5, 0.75, 1)) +
    geom_hline(yintercept = c(0.25, 0.5, 0.75), linetype = "dashed", alpha = 0.3) + 
    geom_vline(xintercept = (seq_along(useMarkers)+0.5)[-length(useMarkers)], linetype = "dotted", alpha = 0.3) +
    geom_boxplot(alpha = 0.2, show.legend = FALSE, outlier.shape = NA) +
    geom_point(position=position_jitterdodge(), pch = 21, size = 2, stroke = 0.2) +
    scale_fill_manual(values = coloursList$condition) +
    theme_prism(axis_text_angle = 45) +
    ylab("Median Quantile Normalised Expression") + xlab("Condition") + ggtitle(paste("Cluster", i))
  print(g1)
  cat("\n\n")
  return("g1" = g1)
}), levels(markerBoxplotsMeltDF[[paste0(mergeBy, knn)]]))
```

```{r, label = "save markers per cluster boxplots", include = FALSE}
if (gimmePDFs) {
  dir.create(file.path(pdfDir, "Markers_per_cluster_boxplots"))
  lapply(names(individualClusterMarkerBoxplots), function(x) {
    pdf(file.path(pdfDir, "Markers_per_cluster_boxplots", paste0(x, ".pdf")), width = 12, height = 5)
    print(individualClusterMarkerBoxplots[[x]])
    dev.off()
  })
}
```

```{r, label = "markers_boxplots2", fig.width=18, fig.height=5, results = "asis"}
for (i in seq_along(selectedClustersList)) {
  if (length(selectedClustersList[[i]]) > 0) {
    cat("####", names(selectedClustersList)[[i]], "\n\n")
    for (j in (selectedClustersList[[i]])) {
      df1 <- df[df$cluster_id == as.character(j),]
      g1 <- ggplot(df1, aes(x = antigen, y = value, fill = condition)) +
        geom_boxplot(alpha = 0.2, show.legend = FALSE, outlier.shape = NA) +
        geom_point(position=position_jitterdodge(), pch = 21, size = 2, stroke = 0.2) +
        geom_hline(yintercept = c(0.25, 0.5, 0.75), linetype = "dashed", alpha = 0.3) + 
        geom_vline(xintercept = (seq_along(useMarkers)+0.5)[-length(useMarkers)], linetype = "dotted", alpha = 0.3) +
        scale_fill_manual(values = coloursList$condition) +
        theme_prism(axis_text_angle = 45) +
        ylab("Median Quantile Normalised Expression") + xlab("Condition") + ggtitle(paste("Cluster", j))
      print(g1)
    }
  } else {
    cat("No DA clusters to see here...\n")
  }
  cat("\n\n")
}
```

::::::::

### DR of Selected Clusters

:::::::: panel-tabset

#### All Cells

```{r, label = "DR of selected clusters", fig.height=5, fig.width=5, results = 'asis'}
DRSelectedClusters <- setNames(lapply(seq_along(clustersToMapList), function(i) {
  clustersToMap <- names(clustersToMapList[[i]])
  cTM <- clustersToMapList[[i]]
  umapDF <- umapDFList[[umapsDFToPlot]]
  umapDF2 <- umapDF[umapDF[["cluster_id"]] %in% clustersToMap,]
  umapDF2[["cluster_id"]] <- droplevels(umapDF2[["cluster_id"]])
  umapDF3 <- umapDF[!umapDF[["cluster_id"]] %in% clustersToMap, ]

  ## Get the mean x/y coordinate for each cluster so we can add labels manually 
  cluster_centre <- umapDF2 |>
  dplyr::group_by(cluster_id) |>
  dplyr::summarize(center_x = mean(x), center_y = mean(y))

  gp <- ggplot(umapDF, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[["cluster_id"]]))

  gp <- gp + geom_point(data = umapDF3, size = dr_point_size, colour = "grey75", alpha = 0.3, shape = 16, show.legend = FALSE)
  gp <- gp + geom_point(data = umapDF2, size = dr_point_size * 1.5, alpha = 0.95)
  gp <- gp + geom_label_repel(data = cluster_centre, aes(x = center_x, y = center_y, label = cluster_id, fill = cluster_id), size = 4, colour = "black")
  gp <- gp + scale_color_manual(values = c(cTM, na.value = "grey33"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + scale_fill_manual(values = c(cTM, na.value = "grey33"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + labs(x = paste(dimRedMethodToUse, "dim 1"), y = paste(dimRedMethodToUse, "dim 2"))
  gp <- gp + ggtitle(names(clustersToMapList)[i])
  gp <- gp + coord_fixed() +
    theme_prism() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      legend.text = element_text(size = 10),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5))
  gp <- gp + guides(colour = guide_legend(override.aes = list(size = 3)))
  gp <- gp + theme(legend.position = "hide")
  print(gp)
  cx <- names(clustersToMapList)[[i]] |> gsub(pattern = ": .*", replacement = "", x = _) |> gsub(pattern = "Contrast ", replacement = "Contrast_", x = _) |> gsub(pattern = "\\/", replacement = "", x = _)
  sx <- names(clustersToMapList)[[i]] |> gsub(pattern = ".*\\: ", replacement = "", x = _) |> gsub(pattern = " ", replacement = "_", x = _) |> gsub(pattern = "\\/", replacement = "", x = _)
  return(list("gp" = gp, "cx" = cx, "sx" = sx))
}), names(clustersToMapList))
```

```{r, label = "save DR of selected clusters", include = FALSE}
if (gimmePDFs) {
  lapply(names(DRSelectedClusters), function(x) {
    pdf(file = paste0(pdfDir, "/", DRSelectedClusters[[x]][["cx"]], "_", DRSelectedClusters[[x]][["sx"]], "_", dimRedMethodToUse, "_full.pdf"), width = 5, height = 5)
    print(DRSelectedClusters[[x]][["gp"]])
    dev.off()
  })
}
```

```{r, label = "DR of selected clusters 2", fig.height= 6*(ceiling(length(condLevels)/2)), fig.width=10, results = 'asis'}
DRSelectedClusters2 <- setNames(lapply(seq_along(clustersToMapList), function(i) {
  cat("####" , names(clustersToMapList)[i], "\n\n")
  clustersToMap <- (names(clustersToMapList[[i]]))
  cTM <- clustersToMapList[[i]]
  umapDF <- umapDFList[[umapsDFToPlot]]
  umapDF2 <- umapDF[umapDF[["cluster_id"]] %in% clustersToMap,]
  umapDF2[["cluster_id"]] <- droplevels(umapDF2[["cluster_id"]])
  umapDF3 <- umapDF[!umapDF[["cluster_id"]] %in% clustersToMap, ]

  ## Get the mean x/y coordinate for each cluster so we can add labels manually 
  cluster_centre <- umapDF2 |>
  dplyr::group_by(cluster_id) |>
  dplyr::summarize(center_x = mean(x), center_y = mean(y))

  gp <- ggplot(umapDF, aes(x = .data[["x"]], y = .data[["y"]], colour = .data[["cluster_id"]]))

  gp <- gp + geom_point(data = umapDF3, size = dr_point_size, colour = "grey75", alpha = 0.3, shape = 16, show.legend = FALSE)
  gp <- gp + geom_point(data = umapDF2, size = dr_point_size * 1.5, alpha = 0.95)
  gp <- gp + geom_label_repel(data = cluster_centre, aes(x = center_x, y = center_y, label = cluster_id, fill = cluster_id), size = 4, colour = "black", show.legend = FALSE)
  gp <- gp + scale_color_manual(
    values = c(cTM, na.value = "black"),
    breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + scale_fill_manual(values = c(cTM, na.value = "black"), breaks = levels(as.factor(umapDF2[["cluster_id"]])))
  gp <- gp + ylim(min(umapDF$y[!is.na(umapDF$y)])-1, max(umapDF$y[!is.na(umapDF$y)])+1)
  gp <- gp + xlim(min(umapDF$x[!is.na(umapDF$x)])-1, max(umapDF$x[!is.na(umapDF$x)])+1)
  gp <- gp + labs(x = paste(dimRedMethodToUse, "dim 1"), y = paste(dimRedMethodToUse, "dim 2"))
  gp <- gp + facet_wrap(~ condition, ncol = 2)
  gp <- gp + ggtitle(names(clustersToMapList)[i])
  gp <- gp + coord_fixed() +
    theme_prism() +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      legend.text = element_text(size = 10),
      panel.border = element_rect(colour = "black", fill = NA, linewidth = 1.5),
      legend.position = "hide")
  gp <- gp + guides(colour = guide_legend(override.aes = list(size = 3)))
  print(gp)
  cat("\n\n")
  return(gp)
}), names(clustersToMapList))
```

::::::::
::::::::::

```{r, label = "save DR of selected clusters 2", include = FALSE}
if (gimmePDFs) {
  lapply(names(DRSelectedClusters2), function(x) {
    pdf(file = paste0(pdfDir, "/", x, "_", dimRedMethodToUse, "_subset.pdf"), width = 10, height =  6*(ceiling(length(condLevels)/2)))
    print(DRSelectedClusters2[[x]])
    dev.off()
  })
}
```

## Session Info

```{r, label = "file info"}
if (is.null(RDataFolder)) {
  cat("The files generated in this report have been saved in:", rFilesDir)
} else {
  cat("The results in this report were generated using the files in", RDataFolder)
}
```

```{r, label = save_sesh, include = TRUE}
sessionInfo()
```

:::::::::::::
